### 2022-05-31 ms.Rnw
setting up the repo

%\input{stat-sinica-header.tex}

\documentclass[12pt]{article}

\setlength{\abovecaptionskip}{-2mm}

%%%%%%% START OF STAT SINICA SETTINGS %%%%%%%%%%

\textwidth=31.9pc
\textheight=46.5pc
\oddsidemargin=1pc
\evensidemargin=1pc
\headsep=15pt
\topmargin=.6cm
\parindent=1.7pc
\parskip=0pt

\usepackage[sectionbib]{natbib}
\usepackage{array,epsfig,fancyheadings,rotating}
%\usepackage[]{hyperref}  %<----modified by Ivan
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{sectsty, secdot}
\sectionfont{\fontsize{12}{15}\selectfont}
\sectionfont{\fontsize{12}{14pt plus.8pt minus .6pt}\selectfont}
\renewcommand{\theequation}{\thesection\arabic{equation}}
\subsectionfont{\fontsize{12}{14pt plus.8pt minus .6pt}\selectfont}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\pagestyle{fancy}
\def\n{\noindent}
\lhead[\fancyplain{} \leftmark]{}
\chead[]{}
\rhead[]{\fancyplain{}\rightmark}
\cfoot{}

%%%%%% END OF STAT SINICA SETTINGS %%%%%%%%5

% sec:intro.
% sec:alg. The \code{ibpf} algorithm for likelihood maximization
% sec:sim. Testing ibpf on a measles transmission model: simulation results
% sec:model. (subsection) The measles data and model
% sec:data. Data analysis results
% sec:discussion.

<<debug,echo=F>>=
run_level <- 1
if(is.numeric(run_level)){
  lik_run_level <- run_level
}
@

\input{header.tex}

\usepackage{amsmath,amssymb}
%\usepackage{graphicx,psfrag,epsf}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{soul}
\usepackage{xcolor}

\usepackage{url} % not crucial - just used below for the URL

%\newcommand\jpc[1]{{\textcolor{orange}{#1}}}
\newcommand\aak[1]{{\textcolor{blue}{#1}}}

%\usepackage{fullpage}
\bibliographystyle{apalike}

<<knitr_setup,include=FALSE,cache=FALSE,purl=FALSE,child="header.Rnw">>=
@

<<doParallel,include=F,cache=F>>=
cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()  
registerDoParallel(cores)

ggplot2::theme_set(ggplot2::theme_bw())
@

% \date{This manuscript was compiled on \today}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\renewcommand{\baselinestretch}{1.5}

\markright{ \hbox{\footnotesize\rm Statistica Sinica
%{\footnotesize\bf 24} (201?), 000-000
}\hfill\\[-13pt]
\hbox{\footnotesize\rm
%\href{http://dx.doi.org/10.5705/ss.20??.???}{doi:http://dx.doi.org/10.5705/ss.20??.???}
}\hfill }

\markboth{\hfill{\footnotesize\rm E. L. Ionides et al.} \hfill}
{\hfill {\footnotesize\rm An iterated block particle filter} \hfill}

\renewcommand{\thefootnote}{}
$\ $\par

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\fontsize{12}{14pt plus.8pt minus .6pt}\selectfont \vspace{0.8pc}
\centerline{\large\bf An iterated block particle filter for inference on coupled dynamic}
\vspace{2pt} 
\centerline{\large\bf systems with shared and unit-specific parameters}
\vspace{.4cm} 
\centerline{E. L. Ionides, J. Wheeler and N. Ning}
\vspace{.4cm} 
\centerline{University of Michigan, department of Statistics,}

 \vspace{.55cm} \fontsize{9}{11.5pt plus.8pt minus.6pt}\selectfont

\centerline{Submitted to {\it Statistica Sinica} special edition on}
\centerline{Recent developments of sequential Monte Carlo methods}
\centerline{and their applications}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{quotation}
\noindent {\it Abstract:}
We consider inference for a collection of partially observed, stochastic, interacting, nonlinear dynamic processes.
Each process is identified with a label called its unit, and our primary motivation arises in biological metapopulation systems where a unit corresponds to a spatially distinct sub-population.
Metapopulation systems are characterized by strong dependence through time within a single unit and relatively weak interactions between units, and these properties make block particle filters an effective tool for simulation-based likelihood evaluation. 
Iterated filtering algorithms can facilitate likelihood maximization for simulation-based filters.
We introduce a new iterated block particle filter algorithm applicable when parameters are unit-specific or shared between units.
We demonstrate this algorithm by performing inference on a coupled epidemiological model describing spatiotemporal measles case report data for twenty towns.

\vspace{9pt}
\noindent {\it Key words and phrases:}
sequential Monte Carlo; metapopulation; spatiotemporal; maximum likelihood estimation; partially observed Markov process.
\par
\end{quotation}\par

\def\thefigure{\arabic{figure}}
\def\thetable{\arabic{table}}

\renewcommand{\theequation}{\thesection.\arabic{equation}}

\fontsize{12}{14pt plus.8pt minus .6pt}\selectfont

%% \def\spacingset#1{\renewcommand{\baselinestretch}%
%% {#1}\small\normalsize} \spacingset{1}

\date{}

%% \spacingset{1.45}

\section{Introduction}
\label{sec:intro}

Statistical inference for high-dimensional partially-observed nonlinear dynamic systems arises in various scientific contexts.
Massive models and datasets are considered in the geophysical sciences, carried out under the name of data assimilation \citep{evensen09book}.
Population models in ecology and epidemiology can be characterized by high levels of stochasticity, nonlinearity, measurement error, and model uncertainty, leading to challenges of a somewhat different nature than geophysical models.
In addition, biological population systems may sometimes have low population due to local introductions or fade-outs of one or more constituent species.
Low population situations may require consideration of integer-valued counts rather than continuous population approximations.
Collections of biological populations measured at different spatial locations may have spatial interactions in addition to local population dynamics; such collections are called a metapopulation.
The study of spatiotemporal disease dynamics has motivated research into inference for metapopulation systems \citep{xia04,li20,park20,ionides21,cauchemez08,bjornstad19}.

Statistical inference for partially observed nonlinear biological systems was an open methodological challenge until recently even in the time series case \citep{bjornstad01}.
Advances in Monte Carlo methods based on the particle filter have made inference accessible on many low-dimensional problems, but the curse of dimensionality \citep{bengtsson08} prevents applicability of the basic particle filter on metapopulations with more than a few units.
Methods based on improving the proposal distribution for the particle filter may not fully resolve the curse of dimensionality \citep{snyder15}.
Situations where previous Monte Carlo methods can provably beat the curse of dimensionality may have limited applicability \citep{beskos17,park20,ionides21}.
Consequently, state-of-the-art analysis has depended on ensemble Kalman filter (EnKF) methods \citep{li20}.
EnKF algorithms scale well but are founded on an approximation that can be unsuitable for discrete populations with fade-out and re-introduction dynamics and other highly non-Gaussian features \citep{ionides21}.

In Sec.~\ref{sec:alg}, we propose an algorithm for inference on metapopulation dynamics which we call an iterated block particle filter (IBPF).
IBPF extends the block particle filter (BPF) of \citet{rebeschini15} to carry out maximum likelihood parameter estimation.
A previous IBPF algorithm was developed by \citet{ning21-ibpf} for the particular case that all estimated parameters are unit-specific, which is to say that the dynamics and measurement process for a unit $u$ are determined by a set of parameters $\psi_u$ specific to unit $u$.
The formal meaning of this assertion is given, together with the pseudocode for our algorithm, in Sec.~\ref{sec:alg}.
We propose an extension of their IBPF which additionally permits the estimation of a vector of shared parameters, $\phi$.
In this case, the full parameter vector is $\theta=(\phi,\psi_{1:U})$ where the $U$ units are labeled $\{1,\dots,\Unit\}$ which we denote by $\seq{1}{\Unit}$.
\citet{ning21-ibpf} developed theoretical justification for their algorithm, but our extension of IBPF currently carries only empirical support.

There may be scientific interest concerning which parameters in metapopulation systems are best understood as unit-specific, and which can reasonably be modeled as shared between units.
Another relevant possibility is that a parameter may differ between units as a shared function of unit-specific covariates, which formally is a special case of a shared parameter.
Addressing these issues is also prerequisite for studying questions about coupling of metapopulation systems via model-based inference from spatiotemporal data.
We will show empirically, in Sec.~\ref{sec:data}, that our {\ibpf} algorithm is applicable to an inference challenge in epidemiological metapopulation dynamics.
Our demonstration focuses on a dataset for weekly measles incidence in 20 towns in the United Kingdom (UK) during the pre-vaccination era \citep{he10} modeled using a previously studied metapopulation model \citep{park20,ionides21}.
Measles case reports have been a longstanding benchmark problem for inference on biological dynamics, motivating the development of time series methodology and, more recently, the progression from single populations to metapopulation systems.
Unlike previous attempts on sequential Monte Carlo inference for metapopulation models, we show that our algorithm can provide practical inference when parameters are either shared between units or differ between units.
Our data analysis results do not fully resolve open questions about what models for coupling between towns are supported by the data, and which parameters should be modeled as unit-specific.
Rather, we demonstrate steps toward this research goal.
We use a simulation study, in Sec.~\ref{sec:sim}, to show that our methodology can deliver a good approximation to the maximum likelihood estimate when fitting the model used to simulate the data.
This reassurance allows us to interpret our data analysis results as evidence of model misspecification, providing a guide for future investigations of these data as well as tools to carry out those investigations.

Optimization of high-dimensional, non-convex and potentially multi-modal functions, evaluated using stochastic methods, is not straightforward even when it is possible to evaluate the function within an acceptable level of error.
Therefore, we discuss approaches that assist the noisy likelihood searches, and suggest diagnostic plots to assess their success.

\section{The {\ibpf} algorithm for likelihood maximization}
\label{sec:alg}

A latent Markov process is denoted by $\{\myvec{X}_{\time},\time\in \seq{0}{\Time}\}$, with $\myvec{X}_{\time}=X_{1:\Unit,\time}$ taking values in a product space $\Xspace^\Unit$.
We define set-valued subscripts by $X_A=\{X_a,a\in A\}$ and $X_{A,B}=\{X_{a,b}, a\in A, b\in B\}$.
The discrete time process $\myvec{X}_{0:\Time}$ may arise from a continuous time Markov process $\{\myvec{X}(t), t_0\le t\le t_{\Time} \}$ observed at times $t_{1:\Time}$, and in this case we set $\myvec{X}_\time=\myvec{X}(t_\time)$.
The initial value $\myvec{X}_{0}$ may be stochastic or deterministic.
Observations are made on each unit, modeled by an observable process $\myvec{Y}_{1:\Time}=Y_{1:\Unit,1:\Time}$ which takes values  at each time $\time$ in a product space $\Yspace^\Unit$.
Observations are modeled as conditionally independent given the latent process.
The conditional independence of measurements applies over both time and the unit structure, so the collection $\big\{Y_{\unit\comma\time},\unit\in\seq{1}{\Unit},\time\in\seq{1}{\Time}\big\}$ is conditionally independent given
$\big\{X_{\unit\comma\time},\unit\in\seq{1}{\Unit},\time\in\seq{1}{\Time}\big\}$.
We suppose the existence of a joint density $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$  for $X_{1:\Unit,0:\Time}$ and $Y_{1:\Unit,1:\Time}$ with respect to some appropriate measure, following a notational convention that the subscripts of $f$ denote the joint or conditional density under consideration.
We suppose that $f$ depends on a parameter vector $\theta=(\phi,\psi_{1:U})$.
The data are $\data{y}_{\unit\comma\time}$ for unit $\unit$ at time $\time$.
This model is a special case of a  partially observed Markov process (POMP) also known as a state space model or hidden Markov model.
The additional unit structure, not generally required for a POMP, is appropriate for modeling interactions between units characterized by a spatial location, and so we call the model a SpatPOMP.
For metapopulation models, the units are not generally arranged on a spatial grid but comprise a collection of spatially distributed population centers.

A numerical challenge of fundamental statistical relevance is the maximization of the log-likelihood function of the data given the model, $\loglik(\theta)=\log f_{\myvec{Y}_{1:\Time}}(\data{\myvec{y}}_{1:\Time}\giventh \theta)$.
Numerical evaluation of the likelihood function is closely related to the filtering problem of evaluating $f_{\myvec{X}_n|\myvec{Y}_{1:n}}(\myvec{x}_n\given \data{\myvec{y}}_{1:n}\giventh \theta)$.
Including parameters into the latent space of a dynamic model can coerce a filter into calculating a Bayesian posterior distribution, though regularization is required to make the calculation numerically tractable using Monte Carlo methods \citep{kitagawa98,janeliu01}.
Iterating this Bayesian calculation recursively targets a maximum likelihood estimate (MLE), a strategy known as data cloning \citep{lele07,lele10}.
Adding noise to perturb the parameters at each time point can stabilize the numerics while retaining the capability to approximate the MLE \citep{ionides06-pnas,ionides11,ionides15}.
Many variations on this idea have been developed using different filter methods \citep{park20,li20,ionides21,manoli15} or employing different perturbation systems \citep{doucet13,nguyen17}.

Previous approaches to iterated perturbed particle filtering are not immediately applicable to BPF.
In the special case where each parameter is localized to an individual unit (i.e., all parameters are unit-specific) one can resample perturbed parameters at the block level and obtain a theoretical guarantee of approximating the MLE \citep{ning21-ibpf}.
Formally, we say that a parameter for a discrete-time SpatPOMP is specific to a unit $u$ if it is involved in specifying the measurement density $f_{Y_{u,n}|X_{u,n}}$ or transition density $f_{X_{u,n+1}|\myvec{X}_n}$ for some $n$, and it is not involved in $f_{Y_{v,m}|X_{v,m}}$ or $f_{X_{v,m+1}|\myvec{X}_m}$ for any $v\neq u$ and any $m$.
For a continuous-time SpatPOMP, we replace the requirement on $f_{X_{u,n+1}|\myvec{X}_n}$ with an equivalent requirement on a numerical solution over a small time increment $\delta$, as in equation \eqref{eq:extension} below.
A parameter which is not unit-specific is said to be shared between units.
There are intermediate possibilities, where a parameter shared for only a subset of all units, but such situations are not explored further here.
The special case where all parameters are unit-specific may arise sometimes, but models typically have some shared parameters which arise in transition densities and/or measurement densities for multiple units.

Our approach to iterated filtering for shared parameters involves constructing an extended model within which the a shared parameters is represented as a unit-specific parameter that happens to be constant across units.
We construct a spatiotemporally extended model by supposing a numerical solution for the transition from time $t$ to time $t+\delta$ for each unit $u$ with the form
\begin{equation}
\label{eq:extension}
  X_{\unit}(t+\delta) = X_\unit(t) + Q_u(\myvec{X}(t),\myvec{\eta}_t,\phi,\psi_{\unit},t,\delta),
\end{equation}
where the random vector $\myvec{\eta}_t=\eta_{1:\Unit,t}$ is shared for all $u\in \seq{1}{U}$ and does not depend on $\theta$.
If a representation \ref{eq:extension} exists, an extended model is defined by replacing $\phi$ with $\phi_{\unit}(t)$ and $\psi_u$ with $\psi_{u}(t)$.
Equation \ref{eq:extension} implicitly defines a continuous-time extended model by the limit as $\delta\to 0$, when that limit exists, but for simulation-based methods the numerical solution is of more immediate concern than this limit.
Admitting a minor abuse of notation, we subsequently use a density $f$ to denote both the original model and its extension for spatiotemporally varying parameters, with the context determining which one is intended.

In some situations, the extended model could have problematic properties.
For example, it could break conservation laws obeyed by the original system.
In other situations, the extended model may make scientific sense in its own right.
For example, in biological metapopulation systems it might be scientifically meaningful to consider a model where there is variation over space and time in the parameters describing the local dynamics.
Here, we are focusing on the hypothesis that there are some parameters that are fixed across space and time, but the specification in \eqref{eq:extension} requires this hypothesis to be nested within a more flexible alternative.

The IBPF algorithm described in Alg.~\ref{alg:ibpfilter} carries out a block particle filter on this extended model.
The extended parameters are given independent perturbations, but a spatial autoregressive step pulls the values of shared parameters toward their mean over the units.
Indeed, this autoregressive step is the only difference between  Alg.~\ref{alg:ibpfilter} and the algorithm for unit-specific parameters proposed by \citet{ning21-ibpf}.
This algorithm, in turn, is essentially the IF2 algorithm of \citet{ionides15} with the particle filter replaced by BPF.
Alg.~\ref{alg:ibpfilter} assumes implicit loops over $\np\ \text{in}\ \seq{1}{\Np}$ and $\unit\ \text{in}\ \seq{1}{\Unit}$, 
$\normal(\mu,\Sigma)$ denotes the normal distribution with mean $\mu$ and variance matrix $\Sigma$, and
$\sigma_\time$ is a $D\times D$ diagonal matrix with entries $\sigma_{d,\time}$.

The pseudocode in Alg.~\ref{alg:ibpfilter} represents our implementation of {\ibpf} as the R function \code{ibpf} in the open-source package \code{spatPomp} \citep{asfaw21github,asfaw21arxiv}.
We will contribute \code{ibpf} to \code{spatPomp}.
Aadditionally, the source code for all the results in this article is available at \url{https://github.com:ionides/ibpf_article}.
Various generalizations of this implementation are possible.
For example, iterated filtering theory does not rely on parameter perturbations following the normal distribution \citep{ionides15} though in practice we transform parameters to facilitate this convenient choice.
Also, the pseudocode does not include parameter-dependent random walk variance, or a different perturbation schedule for initial value parameters, both of which could be included.

%%%%%%% ibpf BBBBBBBBBBBBBBBBBBBBBBBBBBBBBBBBB %%%%%%%%%%%%%%%%%

\begin{algorithm}[H]	
   \caption{\textbf{IBPF}. \\
    {\bf Inputs}:
   simulator for the extended model, $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1}\giventh\myvec{\theta})$;
    evaluator for $f_{{Y}_{\unit,\time}|{X}_{\unit,\time}}({y}_{\unit,\time}\given {x}_{\unit,\time}\giventh\theta)$;
    simulator for extended initialization, $f_{\myvec{X}_0}(\myvec{x}_0\giventh\myvec{\theta})$;
    data, $\data{\myvec{y}}_{1:\Time}$;
    number of particles, $\Np$;    
    blocks, $\blocklist_{1:\Block}$;
    initial parameter swarm with decomposition into shared and unit-specific parameters, $\Theta^{0,\np}_{\unit}=(\Phi^{0,\np}_{\unit},\Psi^{0,\np}_{\unit})$;
    random walk perturbation, $\sigma_{d,\time}$.
%{\bf \code{spatPomp} implementation}:	
%     \code{ibpf(P,\,
%        params{\argequals}$\theta_0${\argequals}$(\phi_0,\psi_{1:\Unit,0})$,
%        Nbpf{\argequals}$\Nit$,
%        Np{\argequals}$\Np\!$,
%        rw.sd{\argequals}$\sigma$,
%        cooling.fraction.50{\argequals}$a$,
%        spat.regression{\argequals}$\spatReg$,
%        block\_list{\argequals}$\blocklist$\,)}	
%    where \code{P} is a	
%     \class{\spatPomp} object \citep{asfaw21arxiv}.
%
}\label{alg:ibpfilter}	
%\BlankLine	
\For{$\nit\ \mathrm{in}\ \seq{1}{\Nit}$}{	
  Perturb parameters:
    $\Theta^{F,\nit,\np}_{\unit,0}\sim \normal
    \left(
      \Theta^{\nit-1,\np}_{\unit}\param \sigma^2_{0} a^{2\nit/50}
    \right)$
  \;
  Initialization:  simulate $\myvec{X}_{0}^{F,\np}\sim {f}_{\myvec{X}_{0}}
    \left(\mydot\giventh{\Theta^{F,\nit,\np}_{1:\Unit,0}}\right)$\;
    \For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{ % For each time step
      Perturb parameters:
        $\Theta^{P,\nit,\np}_{\unit,\time}\sim \normal\left(
       \Theta^{F,\nit,\np}_{\unit,\time-1},\sigma^2_\time a^{2\nit/50} \right)$
     \;
      Prediction simulation: $\myvec{X}_{\time}^{P,\np}
        \sim {f}_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}\big(
        \mydot|\myvec{X}_{\time-1}^{F,\np};{\Theta^{P,\nit,\np}_{1:\Unit,\time}}\big)$	
      \;	
    \For{$\block\ \mathrm{in}\ \seq{1}{\Block}$}{
      Block prediction weights:
            $\displaystyle \blockweight^P_{\time,\np, \block}=
              \prod_{\unit \in \blocklist_{\block}}
              f_{Y_{\unit,\time}|X_{\unit,\time}}
              \big(
              \data{y}_{\unit,\time}\given X^{P,\np}_{\unit,\time}
                \giventh \Theta^{P,\nit,\np}_{\unit,\time} \big)$	
            \nllabel{alg:bpfilter:blockweights}
      \;
      Normalize weights:
        $\tilde{\blockweight}_{\time,\np, \block}= \blockweight^P_{\time,\np,\block}\Big/
        \sum_{i=1}^{\Np}\blockweight^P_{\time,i,\block}$\;
      Select resample indices:
        $\resampleIndex_{1:\Np,\block}$ with
        $\prob\left[\resampleIndex_{\np,\block}=\altNp\right] =
        \tilde{\blockweight}_{\time,\np,\block}$
      \;
         $X_{\blocklist_{\block},\time}^{F,\np}=X_{\blocklist_{\block},\time}^{P,\resampleIndex_{\np,\block}}$,
	$\Theta_{\blocklist_{\block},\time}^{F,\np}
	  =\Theta_{\blocklist_{\block},\time}^{P,\resampleIndex_{\np,\block}}
	  =\big(\Phi_{\blocklist_{\block},\time}^{P,\resampleIndex_{\np,\block}},
	    \Psi_{\blocklist_{\block},\time}^{P,\resampleIndex_{\np,\block}}\big)$
         \nllabel{alg:bpfilter:resample}
      \;
      block mean of shared parameters: $\mu_{\block,\time}^{} = \Np^{-1}\sum_{\np=1}^{\Np} \Phi^{F,\np}_{\blocklist_{\block},\time}$
    } %End of for loop over blocks
    overall mean of shared parameters: $\mu_{\time}^{} = \Block^{-1}\sum_{\block=1}^{\Block} \mu_{\block,\time}^{}$
    \;
    autoregressive correction:
    $\Phi^{F,\np}_{\blocklist_{\block},\time} = \Phi^{F,\np}_{\blocklist_{\block},\time} + \spatReg\big( \mu_{\time}^{} - \mu_{\block,\time}^{}  \big)$
  \;
  }%End timestep for Loop
  $\Theta_{\unit}^{\nit,\np}=\Theta^{F,\nit,\np}_{\unit,\Time}$
  \;  
} % end iterated filtering loop
\KwOut{	
    IBPF parameter swarm, $\Theta_{\unit}^{\Nit,\np}$
}	
\end{algorithm}

\subsection{Algorithmic parameters}

Model parameters are optimized on a transformed scale for which unit variation is scientifically meaningful.
In practice, this means working with positive parameters on a log scale, and $(0,1)$ interval-valued parameters on a logistic scale.
We follow standard iterated filtering practice by using independent random walks for each parameter on this transformed scale \citep{king16}.
We discovered that the large number of parameters following a random walk, in the presence of unit-specific parameters, can require considerably smaller random walk standard deviations than the values around $\sigma_{d,n}=0.02$ (i.e., 2\% perturbation per time point) that have been employed for iterated filtering of time series models.
After experimentation, we used $\sigma_{d,n}=0.005$ for the initial search and $\sigma_{d,n}=0.00125$ for subsequent refinements.
Occasionally, a parameter which can be estimated precisely from the data can benefit substantially from a smaller perturbation.
This is the case for one parameter in our measles analysis, an exponent $\alpha$ for which the scale of uncertainty is an order of magnitude smaller than other parameters; therefore, $\sigma_{d,n}$ for this parameter was scaled accordingly.
In principle, $\sigma_{d,n}$ can be a function of $n$ as well as the parameter, $d$.
The most common reason for using this flexibility is to avoid perturbing parameters during time intervals when there is no information about them: following an evolutionary analogy, evolution cannot operate effectively when there is mutation but no selection.
For an initial value parameter, meaning one which specifies only the latent process value at the initial time $t_0$, we use $\sigma_{d,n}=0$ for $n\ge 1$ and we doubled the value of $\sigma_{d,n}$ for $n=0$.

We set the cooling rate parameter to be $a=0.5$, corresponding to a 1\% reduction in the random walk standard deviation at each iteration.
The only additional algorithmic parameter over previous iterated filtering algorithms is the spatial autoregressive parameter, $\spatReg$.
Numerical experimentation suggests that the performance is not sensitive to the choice of $\spatReg>0$ (see Fig~\ref{fig:spatreg_boxplot} in Appendix A).
Subsequently, we used $\spatReg=0.1$.

Applying IBPF to real data forces us to address issues of model development, model misspecification, and performance in the presence of outliers.
Before tackling these issues, we use simulated data to demonstrate the capabilities of the methodology on a correctly specified model.

\section{Testing {\ibpf} on a measles transmission model: simulation results}
\label{sec:sim}

Measles transmission has provided a useful example of epidemiological dynamics (and therefore also ecological dynamics, for a host-pathogen ecosystem), with plentiful case report data and relatively simple biology.
Model-based analysis of measles time series data led to advances in understanding of the seasonality of infectious disease \citep{fine82}, critical community size \citep{bartlett57}, the recognition that relatively simple mechanistic models can provide a remarkably good description of the dynamics \citep{earn00}, and much other foundational research on disease dynamics.
Some progress has been on building and fitting spatiotemporal models for measles; e.g., \citet{xia04,eggo11,bjornstad19,becker20}.
However, the lack of suitable methodology to fit and assess a flexible class of coupled models is an obstacle to further progress \citep{becker20}.
Previous methodological work has used spatiotemporal measles models as a test problem \citep{park20,ionides21}.
However, these methods have fallen short as tools for data analysis due to numerical considerations.
Bearing all this in mind, measles provides a natural testing ground for our new methodologies.
We demonstrate that we now have the tools to carry out likelihood maximization and profile likelihood confidence interval construction on mechanistic statistical models that are appropriate for spatiotemporal metapopulation disease data.

We specifically set ourselves the task of fitting a spatiotemporal model to the case reports for 20 towns studied by \citet{he10}.
We require ability to handle discrete case counts which vary from zero to thousands of cases per week.
We consider a model with the same structure as \citet{he10}, namely a Markov chain with Gamma noise on the infection rate, but with the addition of a term for transmission between cities.
This requirement limits us to plug-and-play methodologies, which are those that require simulation from the latent process model but not the ability to  evaluate transition densities \citep{breto09,he10}.
The model equations are given in Sec.~\ref{sec:model}.

<<data-plot, echo=F, fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Weekly measles case reports for 20 UK towns.",fig.show = "hold">>=
library(spatPomp)
x <- he10 <- readRDS("sim/he10.rds")
df <- as.data.frame(x)
df[x@unit_obsnames] <- log10(df[x@unit_obsnames]+1)
unit_nm <- rlang::sym(x@unitname)
df[[unit_nm]] <- factor(df[[unit_nm]], levels = x@unit_names)
g <- ggplot2::ggplot(data = df,
  mapping = ggplot2::aes(x = !!rlang::sym(x@timename),
    y = !!rlang::sym(x@unit_obsnames))) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(unit_nm)+
  ggplot2::ylab("log10(reported cases)")+
  ggplot2::xlab("Year")
print(g)
@

<<sim-plot, echo=F, fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Simulated weekly measles case reports.",fig.show = "hold">>=
library(spatPomp)
x <- sim <- readRDS("sim/sim.rds")
df <- as.data.frame(x)
df[x@unit_obsnames] <- log10(df[x@unit_obsnames]+1)
unit_nm <- rlang::sym(x@unitname)
df[[unit_nm]] <- factor(df[[unit_nm]], levels = x@unit_names)
g <- ggplot2::ggplot(data = df,
  mapping = ggplot2::aes(x = !!rlang::sym(x@timename),
    y = !!rlang::sym(x@unit_obsnames))) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(unit_nm)+
  ggplot2::ylab("log10(simulated cases)")+
  ggplot2::xlab("Year")
print(g)
@

The data are shown in Fig~\ref{fig:data-plot} and simulations from the model are shown in Fig~\ref{fig:sim-plot}.
Parameters for the simulated model were based on the analysis of individual towns by \citet{he10} and adjusted to provide a visual match in the simulations.
A goal of the simulation study is to obtain appropriate algorithmic choices for the data analysis, and to develop methodology which can be successful when the truth is known before applying it to data.
We present the simulation study from this perspective, to describe the process of developing our data analysis, though we acknowledge one could revisit the simulation study based on the data analysis in Sec.~\ref{sec:data} using maximum likelihood estimates of the parameters.

A key data analysis question is to decide which parameters should be shared between units and which should be unit-specific, taking distinct values on each unit.
For a general SpatPOMP parameter, it is not necessarily well-defined how to extend the model so that each unit has a distinct parameter value, however our measles model falls into the structure defined in \eqref{eq:extension} and we suspect that similar considerations may arise in other situations.
Different numerical issues may arise as the number of unit-specific parameters increases.
Each unit-specific parameter contributes $U$ dimensions to the parameter space, potentially contributing difficulties associated with high-dimensional or weakly identified parameters.
On the other hand, information about a unit-specific parameters may be localized in neighboring units which could align conveniently with the structure of the BPF.
Estimation of shared parameters, by contrast, has the task of harvesting information from across all the units, which may have its own difficulties.

Estimated parameter values may have scientific interest, but we focus on the statistical task of likelihood maximization.
In the presence of weak identifiability, small differences in likelihood could lead to large differences in parameter estimates.
In such sitatuations, a scientist may choose to investigate what functions of the parameters can be inferred accurately without adding additional assumptions, or alternatively they may choose to investigate the consequences of placing constraints on some parameters to improve identifiability of the remainder.
Likelihood maximization permits such investigations, but they are beyond the scope of this article.

To facilitate data analysis, we wish to have methodology that operates across the full spectrum of decisions on shared versus unit-specific parameter designations.
We therefore test our method on two cases: (A) mostly shared parameters, choosing only the initial values and reporting rate to be unit-specific; (B) every parameter is unit-specific.
We also consider a third case; (C) every parameter is unit-specific and the dynamic coupling (in our context, movement of infected individuals between towns) is replaced by an external forcing of each unit (in our context, an immigration rate of infected individuals from outside the study population).
Case (C) provides a useful point of reference since it is a special case of a PanelPOMP model \citep{breto19} and also can be analyzed as a collection of separate POMP models.
We set up our model so that in case (C) we retrieve the analysis of \citet{he10}.
One may expect that methods which take advantage of the special structure of (C) should out-perform more general methods that permit coupling between units, so we expect the SpatPOMP methods to be less numerically efficient than applying POMP methods separately to each unit.
Questions at stake are: How much less efficient are the SpatPOMP methods? Are simulation-based SpatPOMP inference methods practical for situations such as the measles model of \citet{he10}?

The simulated model is drawn from model (A), which is nested within model (B) but not within (C).
For models B and C, with all parameters are unit-specific, we must estimate $20\times 13=260$ parameters.
This greatly exceeds the 7 shared parameters fitted by \citet{ionides21} for a measles metapopulation model using a large computational effort.
An iterated guided intermediate resampling filter (IGIRF) algorithm was used to fit 9 shared parameters and 3 unit-specific initial value parameters in a measles metapopulation model \citep{park20}.
That analysis used a customized treatment for the unit-specific initial value parameters and did not attempt to estimate other unit-specific parameters.
IGIRF is sensitive to the choice of the guide function, and the model-specific implementation used by \citet{park20} performed better than the generic implementation of IGIRF currently available in \code{spatPomp}.
To our knowledge, {\ibpf} is therefore advancing the current limits for scalability of likelihood-based inference for metapopulation dynamics.

<<sim-loglik-truth,echo=FALSE>>=
readRDS("w/w1_3/lik.rds") -> w
w6400 <- w[w[,"J"]==6400,]
loglik_truth <- logmeanexp(w6400[,"logLik"],se=TRUE)
@

Before engaging in likelihood maximization, one first wants to validate likelihood evaluation.
Likelihood evaluation for metapopulation models was the main topic of \citet{ionides21} and we do not repeat that analysis here.
In brief, the basic particle filter provides a consistent evaluation of the log-likelihood in a limit where the number of particles is large enough to make the Monte Carlo standard deviation small.
This is practical only when $U$ is small (say, $U\le 5$) but that can be used to calibrate the bias induced by BPF, which turns out to be small for our multi-town measles model when each town is its own block.
It may be surprising that resampling independently on each block (which is what the block particle filter does) is able to capture the dependence.
Heuristically, we note that the dynamic dependence between blocks is maintained by BPF, which updates particles according to the full coupled dynamics.
Whether this is sufficient to obtain a good approximation to the filter distribution is situation-dependent, but for the specific case of metapopulation models (for which the strongest coupling is within units rather than between units) the approximation can be empirically successful.
In principle, the approximation error can be reduced by increasing block size, but in practice the additional Monte Carlo variance acquired by doing this is not worthwhile in situations where the coupling is relatively weak.
The BPF log-likelihood evaluation at the true parameter value is $\loglik_{\mathrm{true}}=\Sexpr{myround(loglik_truth[1],1)}$ with a Monte Carlo standard error of $\Sexpr{myround(loglik_truth[2],1)}$.

In Fig.~\ref{fig:sim_search_boxplot} we investigate a sequence of successive searches for the MLE for the models A, B and C described above.
Each search was replicated $36$ times.
Search~1 was started with each parameter adjusted by a uniform $[-0.1,0.1]$ random perturbation on an appropriate dimensionless scale (log for non-negative parameters, logit for $[0,1]$ valued parameters).
This is a fairly small perturbation, but nevertheless sufficient to knock the likelihood of the parameter vector around 300 log units below the MLE (shown in Fig.~\ref{fig:search_diagnostics}).
The goal of this study is to show that the algorithm can succeed reliably on a relatively easy local optimization task.
Subsequent searches were started with four copies of each parameter vector with estimated likelihood in the top 25\% for the previous search.

The MLE is not known exactly in this case.
Wilks's theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+9)/2=44.5$ for A and $(13\times 20)/2=130$ for B.
We find log-likelihoods exceeding the truth by less than this, indicating some imperfection to the Monte Carlo maximization so far as Wilks's asymptotic result holds in this situation. 
Despite this limitation, searches that exceed the likelihood at the truth have found inferentially plausible sets of parameters which can be used to study the likelihood surface around the maximum.
For example, Monte Carlo profile confidence intervals can give proper coverage even in the presence of considerable Monte Carlo error \citep{ionides17,ning21}.

<<v6_start,echo=F>>=
readRDS("v4/v1a_3/initial_lik.rds") -> loglik_start
median_start <- median(loglik_start[,"start_loglik"])
@

<<sim_results,echo=F,results=F>>=
library(pomp)
truth_eval <- readRDS("w/w1_3/lik.rds")
truth_loglik<-logmeanexp(truth_eval[truth_eval[,"J"]==25600,"logLik"])
v1a <- read.csv(file="v4/v1a_3/global.csv")
v1b <- read.csv(file="v4/v1b_3/global.csv")
v1c <- read.csv(file="v4/v1c_3/global.csv")
v2a <- read.csv(file="v4/v2a_3/global.csv")
v2b <- read.csv(file="v4/v2b_3/global.csv")
v2c <- read.csv(file="v4/v2c_3/global.csv")
v3a <- read.csv(file="v4/v3a_3/global.csv")
v3b <- read.csv(file="v4/v3b_3/global.csv")
v3c <- read.csv(file="v4/v3c_3/global.csv")
v4a <- read.csv(file="v4/v4a_3/global.csv")
v4b <- read.csv(file="v4/v4b_3/global.csv")
v4c <- read.csv(file="v4/v4c_3/global.csv")
v5a <- read.csv(file="v4/v5a_3/global.csv")
v5b <- read.csv(file="v4/v5b_3/global.csv")
v5c <- read.csv(file="v4/v5c_3/global.csv")
@

<<z3_results,echo=F,results=F>>=
z1a <- read.csv(file="z3/z1a_3/global.csv")
z1b <- read.csv(file="z3/z1b_3/global.csv")
z1c <- read.csv(file="z3/z1c_3/global.csv")
z1d <- read.csv(file="z3/z1d_3/global.csv")
z2a <- read.csv(file="z3/z2a_3/global.csv")
z2b <- read.csv(file="z3/z2b_3/global.csv")
z2c <- read.csv(file="z3/z2c_3/global.csv")
z2d <- read.csv(file="z3/z2d_3/global.csv")
z3a <- read.csv(file="z3/z3a_3/global.csv")
z3b <- read.csv(file="z3/z3b_3/global.csv")
z3c <- read.csv(file="z3/z3c_3/global.csv")
z3d <- read.csv(file="z3/z3d_3/global.csv")
z4a <- read.csv(file="z3/z4a_3/global.csv")
z4b <- read.csv(file="z3/z4b_3/global.csv")
z4c <- read.csv(file="z3/z4c_3/global.csv")
z4d <- read.csv(file="z3/z4d_3/global.csv")
z5a <- read.csv(file="z3/z5a_3/global.csv")
z5b <- read.csv(file="z3/z5b_3/global.csv")
z5c <- read.csv(file="z3/z5c_3/global.csv")
z5d <- read.csv(file="z3/z5d_3/global.csv")
loglik_he <- sum(read.csv("data/he10mle.csv")$loglik)
# -40345.7
@

<<sim_search_boxplot, echo=F, fig.height=4, fig.width=5.5, out.width="5.5in", fig.cap = "Fitting different models to the simulated measles data, using an initial search and four further refinement searches. (A) $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B) $13\\times 20$ unit-specific parameters and no shared parameters. (C) Also all unit-specific, but with immigration rather than coupling, matching the model of He et al (2010). The horizontal dashed line denotes the likelihood obtained by He et al, using \\code{pomp}. Here, $J=4000$, and the random walk standard deviation (on log- and logit-transformed parameters) is 0.005 for the initial search and 0.00125 for subsequent refinements.",fig.show = "hold">>=
a <- cbind(
A1 = v1a$loglik,
A2 = v2a$loglik,
A3 = v3a$loglik,
A4 = v4a$loglik,
A5 = v5a$loglik
)

b <- cbind(
B1 = v1b$loglik,
B2 = v2b$loglik,
B3 = v3b$loglik,
B4 = v4b$loglik,
B5 = v5b$loglik
)

c <- cbind(
C1 = v1c$loglik,
C2 = v2c$loglik,
C3 = v3c$loglik,
C4 = v4c$loglik,
C5 = v5c$loglik
)

loglik_max <- max(cbind(a,b,c),truth_loglik)
boxplot(cbind(a,b,c),ylim=loglik_max-c(200,0),
ylab="Log-likelihood")
abline(h=truth_loglik,lty="dashed",col="blue")
@

<<search_diagnostics, echo=F, fig.height=6.5, fig.width=4.5, out.width="5in", fig.cap = "Three steps of a likelihood search on simulated data (left panels) and UK measles data (right panels). Final log-likelihood values are plotted against starting log-likelihood for an initial search (top row), a refinement of this search (second row) and a second refinement (third row). Starting values for each refinement were 4 copies of the top 25\\% of end points for the previous search. Unit-specific parameters were $\\rho$, $S_0$, $E_0$, $I_0$ (for the simulation study, model A) and all parameters (for the data, model B).">>=
par(mfcol=c(3,2))
par(mai=c(0.3,0.3,0.1,0.1))
par(omi=c(0.3,0.5,0.5,0.1))
##
## (1,1)
##
x <- v1a[,"start_loglik"]
y <- v1a$loglik
x_lim <- range(c(x,truth_loglik,max(y)))
y_lim <- range(c(y,truth_loglik,max(x)))
plot(x,y,xlim=x_lim,ylim=y_lim,
  xlab="",
  ylab="Initial search",
)
abline(a=0,b=1,lty="dashed",col="blue")
abline(h=truth_loglik,lty="dashed",col="blue")
abline(v=truth_loglik,lty="dashed",col="blue")
##
## (2,1)
##
x <- rep(v1a$loglik[1:18],each=2)
y <- v2a$loglik
x_lim <- range(c(x,truth_loglik,max(y)))
y_lim <- range(c(y,truth_loglik,max(x)))
plot(x,y,xlim=x_lim,ylim=y_lim,
  xlab="",
  ylab="1st refinement"
)
abline(a=0,b=1,lty="dashed",col="blue")
abline(h=truth_loglik,lty="dashed",col="blue")
abline(v=truth_loglik,lty="dashed",col="blue")
##
## (3,1)
##
x <- rep(v2a$loglik[1:18],each=2)
y <- v3a$loglik
x_lim <- range(c(x,truth_loglik,max(y)))
y_lim <- range(c(y,truth_loglik,max(x)))
plot(x,y,xlim=x_lim,ylim=y_lim,
    ylab="2nd refinement",
    xlab="Starting log-likelihood"
)
abline(a=0,b=1,lty="dashed",col="blue")
abline(h=truth_loglik,lty="dashed",col="blue")
abline(v=truth_loglik,lty="dashed",col="blue")
##
##
## (1,2)
##
  x <- z1b[,"start_loglik"]
  y <- z1b$loglik
  x_lim <- range(c(x,loglik_he,max(y)))
  y_lim <- range(c(y,loglik_he,max(x)))
  plot(x,y,xlim=x_lim,ylim=y_lim,
    xlab="",
    ylab="",
  )
  abline(a=0,b=1,lty="dashed",col="blue")
  abline(h=loglik_he,lty="dashed",col="blue")
  abline(v=loglik_he,lty="dashed",col="blue")
##
## (2,2)
##
  x <- rep(z1b$loglik[1:18],each=2)
  y <- z2b$loglik
  x_lim <- range(c(x,loglik_he,max(y)))
  y_lim <- range(c(y,loglik_he,max(x)))
  plot(x,y,xlim=x_lim,ylim=y_lim,
    xlab="",
    ylab=""
  )
  abline(a=0,b=1,lty="dashed",col="blue")
  abline(h=loglik_he,lty="dashed",col="blue")
  abline(v=loglik_he,lty="dashed",col="blue")
##
## (3,2)
##
  x <- rep(z2b$loglik[1:18],each=2)
  y <- z3b$loglik
  x_lim <- range(c(x,loglik_he,max(y)))
  y_lim <- range(c(y,loglik_he,max(x)))
  plot(x,y,xlim=x_lim,ylim=y_lim,
    xlab="Starting log-likelihood",
    ylab=""
  )
  abline(a=0,b=1,lty="dashed",col="blue")
  abline(h=loglik_he,lty="dashed",col="blue")
  abline(v=loglik_he,lty="dashed",col="blue")

LINE<-0.25
CEX <- 0.9
mtext("Initial search",side=2,line=LINE,outer=TRUE,at=0.833,cex=CEX)
mtext("First refinement",side=2,line=LINE,outer=TRUE,at=0.5,cex=CEX)
mtext("Second refinement",side=2,line=LINE,outer=TRUE,at=0.167,cex=CEX)
mtext("Simulated data",side=3,line=LINE,outer=TRUE,at=0.25,cex=CEX)
mtext("UK towns",side=3,line=LINE,outer=TRUE,at=0.75,cex=CEX)
mtext("Log-likelihood",side=1,line=LINE,outer=TRUE,at=0.25,cex=CEX)
mtext("Log-likelihood",side=1,line=LINE,outer=TRUE,at=0.75,cex=CEX)

@

In Fig.~\ref{fig:search_diagnostics} we take a closer look at convergence diagnostics corresponding to A for the simulated data (the first column) and B for the UK measles data (the second column).
For now, we focus on the first column.
The first row plots the log-likelihood obtained after the initial search against the log-likelihood of the randomly selected starting value.
The horizontal and vertical dashed lines denote the likelihood at the truth, and the diagonal dashed line represents equality, so that points above the diagonal show improvement after the search.
We find that {\ibpf} robustly and rapidly approaches a neighborhood of the MLE, as measured by likelihood.
The second row shows that further investigation of the more successful searches can reliably obtain likelihood values higher than those at the truth.
However, for a Monte Carlo search based on Monte Carlo likelihood evaluation, it may be anticipated that pinpointing the exact maximum in a high-dimensional space is problematic.
The third row shows that continued searching does not lead to substantially better outcomes.
When using these methods, we emphasize the need to make proper inferences despite imperfect maximization \citep{ionides17,ning21}.

The results in Fig.~\ref{fig:sim_search_boxplot} and the first column of  Fig.~\ref{fig:search_diagnostics} document that {\ibpf} can be effective on simulated data, but do not guarantee comparable performance for our data analysis.
Indeed, model misspecification that is inevitable for data analysis may be expected to add difficulties to filtering and therefore to numerical methods based on filtering.
Rather, we view the simulation study as a lower bound on the effort required to carry out effective inference on data, and a starting point for investigating effective ways to proceed with the data analysis. Before moving on to the data analysis, we briefly describe the details of the model.

\subsection{The measles data and model}
\label{sec:model}

The data we consider here are weekly case reports for twenty towns, identical to those studied by \citet{he10}.
Some previous analyses have used counts aggregated over two week intervals \citep{park20,ionides21} since these were available for more cities, but our goal here is to extend the analysis of \citet{he10}.
Apart from this, our model matches \citet{ionides21} and for completeness we repeat the description here.
We compartmentalize the population of each city into susceptible ($S$), exposed ($E$), infectious ($I$), and recovered/removed ($R$) categories.
The numbers of individuals in each compartment for city $\unit$ at time $t$ are denoted by integer-valued random variables $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$.
The population dynamics are written in terms of counting processes $N_{AB,\unit}(t)$ enumerating cumulative transitions in city $\unit$, up to time $t$, between compartments $A$ and $B$.
Here, $A,B\in \{S,E,I,R,B,D\}$ with $B$ denoting a source compartment for immigration or birth, and $D$ a sink compartment for emigration or death.
We enumerate the $U=20$ cities studied by \citet{he10} in decreasing size, so that $\unit=1$ corresponds to London.
Our model is described by the following system of stochastic differential equations, for $\unit=1,\dots, \Unit$,
\begin{equation}
\nonumber
%\label{eq:measles:system}
\begin{array}{lllllll}
\displaystyle dS_\unit(t) &=& dN_{BS,\unit}(t) &-& dN_{SE,\unit}(t) &-& dN_{SD,\unit}(t), \\
\displaystyle dE_\unit(t) &=& dN_{SE,\unit}(t) &-& dN_{EI,\unit}(t) &-& dN_{ED,\unit}(t), \\
\displaystyle dI_\unit(t) &=& dN_{EI,\unit}(t) &-& dN_{IR,\unit}(t) &-& dN_{ID,\unit}(t). 
\end{array}
\end{equation}
The total population $P_\unit(t)=S_\unit(t)+E_\unit(t)+I_\unit(t)+R_\unit(t)$ is calculated by smoothing census data and is treated as known.
The number of recovered individuals $R_\unit(t)$ in city $\unit$ is therefore defined implicitly.
$N_{SE,\unit}(t)$ is modeled as negative binomial death processes \citep{breto09,breto11}
with over-dispersion parameter $\sigma_{SE}$, and rate given by
\begin{eqnarray}
\nonumber
\mathbb{E} \big[ N_{SE,\unit}(t+dt) - N_{SE,\unit}(t) \big] 
&=& 
\beta(t) \, S_\unit(t) 
\Big[ 
  \left( \frac{I_\unit+\iota}{P_\unit} \right)^\alpha
\\
\label{eq:dEdt}
&& \hspace{-3.8cm}
 + \sum_{\altUnit \neq \unit} \frac{v_{\unit\altUnit}}{P_\unit} 
  \left\{ 
    \left(
      \frac{ I_{\altUnit} }{ P_{\altUnit} }
    \right)^\alpha - 
    \left(
      \frac{I_\unit}{P_\unit}  
    \right)^\alpha
  \right\}
\Big] dt + o(dt),\label{eqn:transmissionrate}
\end{eqnarray}
where $\beta(t)$ models seasonality driven by high contact rates between children at school, described by
\begin{equation}
\nonumber
  \beta(t)=\begin{cases}
\big(1+\amplitude(1-p)p^{-1} \big)\, \meanBeta & \mbox{ during school term},\\
\big( 1-\amplitude\big) \, \meanBeta& \mbox{ during vacation},
  \end{cases} \label{eq:term}
\end{equation}
with $p = 0.759$ being the proportion of the year taken up by the school terms, $\meanBeta$ is the mean transmission rate, and $\amplitude$ measures the reduction of transmission during school holidays.
In \eqref{eq:dEdt}, $\alpha$ is a mixing exponent modeling inhomogeneous contact rates within a city, and $\iota$ models immigration of infected individuals which is appropriate when analyzing a subset of cities that cannot be treated as a closed system.
The number of travelers from city $\unit$ to $\altUnit$ is denoted by $v_{\unit\altUnit}$. 
Here, $v_{\unit\altUnit}$ is constructed using the gravity model of \cite{xia04}, 
\[
v_{\unit\altUnit} = \gravity \cdot \frac{\;\overline{\dist}\;}{\bar{P}^2} \cdot \frac{P_\unit \cdot P_{\altUnit}}{\dist(\unit,\altUnit)},
\]
where $\dist(\unit,\altUnit)$ denotes the distance between city $\unit$ and city $\altUnit$, $P_\unit$ is the average population for city $\unit$ across time, $\bar{P}$ is the average population across cities, and $\overline{\dist}$ is the average distance between a randomly chosen pair of cities.
Here, we model $v_{\unit\altUnit}$ as fixed through time and symmetric between any two arbitrary cities, though a natural extension would allow for temporal variation and asymmetric movement between two cities.
The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{\bullet D,\unit}(t)$ are modeled as conditional Poisson process with per-capita rates $\mu_{EI}$, $\mu_{IR}$ and $\mu_{\bullet D}$ respectively, and we fix $\mu_{\bullet D}=(50 \mbox{ year})^{-1}$.
The birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson processes with rate $\mu_{BS,\unit}(t)$, given by interpolated census data.

To describe the measurement process, let $Z_{\unit\comma\time}=N_{IR\comma\unit}(t_\time)-N_{IR\comma\unit}(t_{\time-1})$ be the number of removed infected individuals in the $n$th reporting interval.
Suppose that cases are quarantined once they are identified, so that reported cases comprise a fraction $\rho$ of these removal events.
The case report $\data{y}_{\unit\comma\time}$ is modeled as a realization of a discretized conditionally Gaussian random variable $Y_{\unit\comma\time}$, defined for $y>0$ via
\begin{eqnarray}
\nonumber
\prob\big[Y_{\unit\comma\time}{=}y\mid Z_{\unit\comma\time}{=}z\big] &=& \Phi\big(y+0.5; \rho z,\rho(1-\rho)z+\psi^2\rho^2z^2\big)
\\
&&\hspace{-20mm}
- \Phi\big(y-0.5; \rho z,\rho(1-\rho)z+\psi^2\rho^2z^2\big)
\label{eq:obs}
\end{eqnarray}
where $\Phi(\cdot;\mu,\sigma^2)$ is the $\normal(\mu,\sigma^2)$ cumulative distribution function, and $\psi$ models overdispersion relative to the binomial distribution.
For $y=0$, we replace $y-0.5$ by $-\infty$ in \eqref{eq:obs}.

Three data points were treated as missing by \citet{he10} due to presumed recording errors, and we followed that decision.
The general SpatPOMP framework has flexibility that permits modeling of missing data in various ways. 
We chose to include a special value $\mathrm{NA}$ in $\Yspace$ and set $Y_{u,n}$ to be $\mathrm{NA}$ with probability one when $\data{y}_{u,n}$ is missing.

\section{Demonstration on data}
\label{sec:data}

<<data_search_boxplot, echo=F, fig.height=4, fig.width=5.5, out.width="5.5in", fig.cap = "Fitting different models to the UK measles data using the method tested on simulated data. (A) $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B) $13\\times 20$ unit-specific parameters and no shared parameters. (C) Also all unit-specific, but with immigration rather than coupling, matching the model of He et al (2010). The horizontal dashed line denotes the likelihood obtained by He et al, using \\code{pomp}.",fig.show = "hold">>=
a <- cbind(
A1 = z1a$loglik,
A2 = z2a$loglik,
A3 = z3a$loglik,
A4 = z4a$loglik,
A5 = z5a$loglik
)

b <- cbind(
B1 = z1b$loglik,
B2 = z2b$loglik,
B3 = z3b$loglik,
B4 = z4b$loglik,
B5 = z5b$loglik
)

c <- cbind(
C1 = z1c$loglik,
C2 = z2c$loglik,
C3 = z3c$loglik,
C4 = z4c$loglik,
C5 = z5c$loglik
)

d <- cbind(
D1 = z1d$loglik,
D2 = z2d$loglik,
D3 = z3d$loglik,
D4 = z4d$loglik,
D5 = z5d$loglik
)
loglik_max <- max(cbind(a,b,c),loglik_he)
boxplot(cbind(a,b,c),ylim=loglik_max-c(5500,0),
  ylab="Log-likelihood")
abline(h=loglik_he,lty="dashed",col="blue")
@

Fig.~\ref{fig:data_search_boxplot} shows results for fitting models A, B and C to UK measles data using successive rounds of {\ibpf}, applying the same method used for the simulation study shown in Fig.~\ref{fig:sim_search_boxplot}.
The dashed lines show the sum of the log-likelihoods obtained by \citet{he10}, $\ell_{\mathrm He}=\Sexpr{myround(loglik_he,1)}$.
The value $\ell_{\mathrm He}$ corresponds to model C.
BPF with each town forming a separate block, applied to an uncoupled model such as C, is equivalent to carrying out independent particle filters for each unit.
Indeed, if the published parameter values from \citet{he10} are inserted into the \code{spatPomp} object for model C, then the \code{bpfilter()} function can be used to carry out BPF, which retrieves $\ell_{\mathrm He}$, up to Monte Carlo error.
Each of models A, B and C in Fig.~\ref{fig:data_search_boxplot} shows shortfall relative to $\ell_{\mathrm He}$.

The largest shortfall is for A, and an explanation is that A has more shared parameters than the evidence in the data supports.
The results of \citet{he10} suggest that the data are explained better when various parameters are a function of the town population.
However, it remains an open problem to determine suitable functional forms for this relationship, and to establish regularities across towns that can be represented by shared parameters.
This may be investigated using panel methods such as PanelPOMP models \citep{breto19} in addition to consideration of SpatPOMP models.

Models B and C give rise to comparable likelihoods, which is in contrast to Fig.~\ref{fig:sim_search_boxplot}.
Fitting to simulated data from B (in the special case where all unit-specific parameters are equal across all units) the shortfall for C in Fig.~\ref{fig:sim_search_boxplot} indicates that we obtain a substantially worse fit by approximating coupling by an uncoupled reservoir of infection.
If the actual data were also explained substantially better by a coupled model, we would expect to see comparable results in the data analysis.
Since we do not, we conclude that this coupled model is not a substantially better explanation than the uncoupled model.
Various other candidate coupling mechanisms have been proposed \citep{bjornstad19} but they have not yet been fitted to the full data, suggesting a lack of appropriate methodology to do so.
Instead, \citet{bjornstad19} looked at summary statistics based on local fade-outs and re-introductions.
Further investigation is needed to determine whether other models for coupling can show improvement over this gravity model on these data.
This should be carried out in conjunction with research into the specification of shared and unit-specific parameters.

We notice from Fig.~\ref{fig:data_search_boxplot} that model B has a small but distinct shortfall compared to $\ell_{\mathrm He}$.
On the simulated data, we do not observe this shortfall, and we deduce that the real data provides a more challenging optimization environment.
When carrying out hard optimization problems, it may be possible to develop helpful strategies specific to the model and data in question.
For example, one could try merging unit-specific parameters from different searches, using the likelihood component for each unit to assess successful choices.
Optimization heuristics such as this do not have general theoretical support; if they obtain higher likelihoods that is sufficient justification. 
For analysis of panel time series data using PanelPOMP models, such methods have clearer theoretical support and have been found useful \citep{breto19}.

\section{Discussion}
\label{sec:discussion}

Theoretical interest in BPF was inspired by \citet{rebeschini15} who showed mathematically that BPF can, under particular conditions, enjoy linear scaling properties.
\citet{rebeschini15} expressed pessimism about the practical applicability of their algorithm, which may help to explain why the practical use of this algorithm has been limited.
The algorithm was independently invented under the name of a factored particle filter by \citet{ng02} who offered an empirical justification.
However, the proposal of \citet{ng02} also saw only limited use.
The numerical results in Fig.~3 of \citet{ionides21} suggest that BPF is particularly well suited to metapopulation models.
In a metapopulation model, one expects most of the population dynamics to occur at a local level, among the individuals at one spatial unit.
As a broad generalization, in ecological systems dispersion of individuals between spatial units is rare but dynamically important.
Edge effects between blocks, which may be a serious problem for BPF in a system with stronger spatial coupling \citep{ionides21} can therefore be a relatively minor concern for metapopulation models.
Blocks of size one unit are therefore a natural choice for block filtering of metapopulation models when the constituent populations are spatially distinct.

This article has investigated a likelihood maximization approach to inference.
Much research has been done on inference for high-dimensional partially observed stochastic dynamic systems, and we have cited above only the most directly relevant work.
A Bayesian inference approach based on the ensemble Kalman filter was developed by \citet{katzfuss19}.
An expectation-maximization approach based on a block particle filter was presented by \citet{finke17}.
Spatiotemporal models having a convenient factorization across units were studied by \citet{beskos17} and \citet{xu19}.

Computation time for Fig.~\ref{fig:sim_search_boxplot} and Fig.~\ref{fig:data_search_boxplot} was approximately 24hr on a single core of a compute node for 4000 particles iterated 100 times over the 730 time points (weekly data for 14 years) for 20 cities.
Each box in these figures involves 36 replications and thus took 24hr on all cores of one node on a computing cluster.
This level of effort is tolerable, but does suggest that the methods will not readily extend to tasks that are considerably larger. 
In particular, we do not expect applicability to situations where it is necessary to consider $10^3$ or more spatial units.
In such situations, it may be necessary to make additional approximations such as those inherent in the ensemble Kalman filter \citep{evensen09,katzfuss19} or other numerical filtering techniques \citep{whitehouse22}.
However, we anticipate that many practical metapopulation analyses can be addressed within the scope we have demonstrated.
The majority of the computational effort is spent on simulating from the dynamic model.
In the \code{spatPomp} R package, as in \code{pomp}, the user supplies a snippet of C code for simulating a single particle between a single arbitrary pair of times, and the package provides a vectorized form of this computation which is used by the inference methodologies within the package.
As a consequence, competitive computational performance is obtained even though the majority of the package is written in R \citep[Table~1]{FitzJohn20}.

\vspace{3mm}

%\newpage

\noindent {\bf Code and data availability}

\vspace{1mm}

\noindent The data we analyze are in the public domain, and are available together with source code at \url{https://github.com:ionides/ibpf_article}.

\vspace{3mm}

\noindent {\bf Acknowledgments}

\vspace{1mm}

\noindent This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108, and National Institutes of Health grant 1-R01-AI143852. We acknowledge an anonymous referee from a previous article who recommended that we should try implementing a block particle filter.

\bibliography{bib-ibpf}

% \vspace{8mm}
\newpage

\noindent{\bf \Large Appendix A}

\vspace{3mm}

\noindent{\bf \large Varying the IBPF spatial autoregression parameter, $\spatReg$}

\vspace{-5mm}

<<spatreg_boxplot, echo=F, fig.height=4, fig.width=5.5, out.width="5.5in", fig.pos="h", fig.cap = paste0("Varying the spatial autoregression algorithmic parameter for shared parameters with simulated measles data. The log-likelihood was obtained using $M=100$ iterations of IBPF starting at random parameters (median starting log-likelihood, ",signif(median_start,3), "). The horizontal dashed line denotes the log-likelihood at the true parameters.")>>=
loglik_truth <- -38882.7 # from w1.r
spat_regression <- c(0, 1/64, 1/32, 1/16, 1/8, 1/4, 1/2, 1) 
spat_regression_char <- c("0", "1/64", "1/32", "1/16", "1/8", "1/4", "1/2", "1")
Experiment <- length(spat_regression)
results <- foreach(experiment=1:Experiment,.combine=rbind) %do% {
  r <- readRDS(file=paste0("v/v6_3/global_e",experiment,".rds"))
  cbind(r,sp_reg=spat_regression[experiment])
}
loglik_max <- max(c(loglik_truth,results$loglik))
results <- as.data.frame(results)
boxplot(loglik~sp_reg,data=results,ylim=c(-40000,loglik_max),
  ylab="Log-likelihood", xlab="spatial autoregression parameter",
  names=spat_regression_char)
abline(h=loglik_truth,col="blue",lty="dotted")
@

\end{document}

### 2022-05-31 ms.tex
setting up the repo

--- 2022-05-31 ms.Rnw
@@
 
-\documentclass[12pt]{article}
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{xcolor}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 
-<<debug,echo=F>>=
-run_level <- 1
-if(is.numeric(run_level)){
-  lik_run_level <- run_level
-}
-@
-
 \input{header.tex}
@@
 
-<<knitr_setup,include=FALSE,cache=FALSE,purl=FALSE,child="header.Rnw">>=
-@
-
-<<doParallel,include=F,cache=F>>=
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
-if(is.na(cores)) cores <- detectCores()  
-registerDoParallel(cores)
-
-ggplot2::theme_set(ggplot2::theme_bw())
-@
-
 % \date{This manuscript was compiled on \today}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-<<data-plot, echo=F, fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Weekly measles case reports for 20 UK towns.",fig.show = "hold">>=
-library(spatPomp)
-x <- he10 <- readRDS("sim/he10.rds")
-df <- as.data.frame(x)
-df[x@unit_obsnames] <- log10(df[x@unit_obsnames]+1)
-unit_nm <- rlang::sym(x@unitname)
-df[[unit_nm]] <- factor(df[[unit_nm]], levels = x@unit_names)
-g <- ggplot2::ggplot(data = df,
-  mapping = ggplot2::aes(x = !!rlang::sym(x@timename),
-    y = !!rlang::sym(x@unit_obsnames))) +
-  ggplot2::geom_line() +
-  ggplot2::facet_wrap(unit_nm)+
-  ggplot2::ylab("log10(reported cases)")+
-  ggplot2::xlab("Year")
-print(g)
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}
 
-<<sim-plot, echo=F, fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Simulated weekly measles case reports.",fig.show = "hold">>=
-library(spatPomp)
-x <- sim <- readRDS("sim/sim.rds")
-df <- as.data.frame(x)
-df[x@unit_obsnames] <- log10(df[x@unit_obsnames]+1)
-unit_nm <- rlang::sym(x@unitname)
-df[[unit_nm]] <- factor(df[[unit_nm]], levels = x@unit_names)
-g <- ggplot2::ggplot(data = df,
-  mapping = ggplot2::aes(x = !!rlang::sym(x@timename),
-    y = !!rlang::sym(x@unit_obsnames))) +
-  ggplot2::geom_line() +
-  ggplot2::facet_wrap(unit_nm)+
-  ggplot2::ylab("log10(simulated cases)")+
-  ggplot2::xlab("Year")
-print(g)
-@
+\includegraphics[width=6.5in]{figure/data-plot-1} \hfill{}
+
+\caption[Weekly measles case reports for 20 UK towns]{Weekly measles case reports for 20 UK towns.}\label{fig:data-plot}
+\end{figure}
+
+\end{knitrout}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}
+
+\includegraphics[width=6.5in]{figure/sim-plot-1} \hfill{}
+
+\caption[Simulated weekly measles case reports]{Simulated weekly measles case reports.}\label{fig:sim-plot}
+\end{figure}
+
+\end{knitrout}
 
@@
 
-<<sim-loglik-truth,echo=FALSE>>=
-readRDS("w/w1_3/lik.rds") -> w
-w6400 <- w[w[,"J"]==6400,]
-loglik_truth <- logmeanexp(w6400[,"logLik"],se=TRUE)
-@
-
 Before engaging in likelihood maximization, one first wants to validate likelihood evaluation.
@@
 In principle, the approximation error can be reduced by increasing block size, but in practice the additional Monte Carlo variance acquired by doing this is not worthwhile in situations where the coupling is relatively weak.
-The BPF log-likelihood evaluation at the true parameter value is $\loglik_{\mathrm{true}}=\Sexpr{myround(loglik_truth[1],1)}$ with a Monte Carlo standard error of $\Sexpr{myround(loglik_truth[2],1)}$.
+The BPF log-likelihood evaluation at the true parameter value is $\loglik_{\mathrm{true}}=-40612.5$ with a Monte Carlo standard error of $0.6$.
 
@@
 
-<<v6_start,echo=F>>=
-readRDS("v4/v1a_3/initial_lik.rds") -> loglik_start
-median_start <- median(loglik_start[,"start_loglik"])
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}
 
-<<sim_results,echo=F,results=F>>=
-library(pomp)
-truth_eval <- readRDS("w/w1_3/lik.rds")
-truth_loglik<-logmeanexp(truth_eval[truth_eval[,"J"]==25600,"logLik"])
-v1a <- read.csv(file="v4/v1a_3/global.csv")
-v1b <- read.csv(file="v4/v1b_3/global.csv")
-v1c <- read.csv(file="v4/v1c_3/global.csv")
-v2a <- read.csv(file="v4/v2a_3/global.csv")
-v2b <- read.csv(file="v4/v2b_3/global.csv")
-v2c <- read.csv(file="v4/v2c_3/global.csv")
-v3a <- read.csv(file="v4/v3a_3/global.csv")
-v3b <- read.csv(file="v4/v3b_3/global.csv")
-v3c <- read.csv(file="v4/v3c_3/global.csv")
-v4a <- read.csv(file="v4/v4a_3/global.csv")
-v4b <- read.csv(file="v4/v4b_3/global.csv")
-v4c <- read.csv(file="v4/v4c_3/global.csv")
-v5a <- read.csv(file="v4/v5a_3/global.csv")
-v5b <- read.csv(file="v4/v5b_3/global.csv")
-v5c <- read.csv(file="v4/v5c_3/global.csv")
-@
+\includegraphics[width=5.5in]{figure/sim_search_boxplot-1} \hfill{}
 
-<<z3_results,echo=F,results=F>>=
-z1a <- read.csv(file="z3/z1a_3/global.csv")
-z1b <- read.csv(file="z3/z1b_3/global.csv")
-z1c <- read.csv(file="z3/z1c_3/global.csv")
-z1d <- read.csv(file="z3/z1d_3/global.csv")
-z2a <- read.csv(file="z3/z2a_3/global.csv")
-z2b <- read.csv(file="z3/z2b_3/global.csv")
-z2c <- read.csv(file="z3/z2c_3/global.csv")
-z2d <- read.csv(file="z3/z2d_3/global.csv")
-z3a <- read.csv(file="z3/z3a_3/global.csv")
-z3b <- read.csv(file="z3/z3b_3/global.csv")
-z3c <- read.csv(file="z3/z3c_3/global.csv")
-z3d <- read.csv(file="z3/z3d_3/global.csv")
-z4a <- read.csv(file="z3/z4a_3/global.csv")
-z4b <- read.csv(file="z3/z4b_3/global.csv")
-z4c <- read.csv(file="z3/z4c_3/global.csv")
-z4d <- read.csv(file="z3/z4d_3/global.csv")
-z5a <- read.csv(file="z3/z5a_3/global.csv")
-z5b <- read.csv(file="z3/z5b_3/global.csv")
-z5c <- read.csv(file="z3/z5c_3/global.csv")
-z5d <- read.csv(file="z3/z5d_3/global.csv")
-loglik_he <- sum(read.csv("data/he10mle.csv")$loglik)
-# -40345.7
-@
+\caption[Fitting different models to the simulated measles data, using an initial search and four further refinement searches]{Fitting different models to the simulated measles data, using an initial search and four further refinement searches.
+(A) $4\times 20$ unit-specific parameters and $9$ shared parameters; (B) $13\times 20$ unit-specific parameters and no shared parameters.
+(C) Also all unit-specific, but with immigration rather than coupling, matching the model of He et al (2010).
+The horizontal dashed line denotes the likelihood obtained by He et al, using \code{pomp}.
+Here, $J=4000$, and the random walk standard deviation (on log- and logit-transformed parameters) is 0.005 for the initial search and 0.00125 for subsequent refinements.}\label{fig:sim_search_boxplot}
+\end{figure}
 
-<<sim_search_boxplot, echo=F, fig.height=4, fig.width=5.5, out.width="5.5in", fig.cap = "Fitting different models to the simulated measles data, using an initial search and four further refinement searches.
-(A) $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B) $13\\times 20$ unit-specific parameters and no shared parameters.
-(C) Also all unit-specific, but with immigration rather than coupling, matching the model of He et al (2010).
-The horizontal dashed line denotes the likelihood obtained by He et al, using \\code{pomp}.
-Here, $J=4000$, and the random walk standard deviation (on log- and logit-transformed parameters) is 0.005 for the initial search and 0.00125 for subsequent refinements.",fig.show = "hold">>=
-a <- cbind(
-A1 = v1a$loglik,
-A2 = v2a$loglik,
-A3 = v3a$loglik,
-A4 = v4a$loglik,
-A5 = v5a$loglik
-)
+\end{knitrout}
 
-b <- cbind(
-B1 = v1b$loglik,
-B2 = v2b$loglik,
-B3 = v3b$loglik,
-B4 = v4b$loglik,
-B5 = v5b$loglik
-)
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}
 
-c <- cbind(
-C1 = v1c$loglik,
-C2 = v2c$loglik,
-C3 = v3c$loglik,
-C4 = v4c$loglik,
-C5 = v5c$loglik
-)
+\includegraphics[width=5in]{figure/search_diagnostics-1} \hfill{}
 
-loglik_max <- max(cbind(a,b,c),truth_loglik)
-boxplot(cbind(a,b,c),ylim=loglik_max-c(200,0),
-ylab="Log-likelihood")
-abline(h=truth_loglik,lty="dashed",col="blue")
-@
+\caption[Three steps of a likelihood search on simulated data (left panels) and UK measles data (right panels)]{Three steps of a likelihood search on simulated data (left panels) and UK measles data (right panels).
+Final log-likelihood values are plotted against starting log-likelihood for an initial search (top row), a refinement of this search (second row) and a second refinement (third row).
+Starting values for each refinement were 4 copies of the top 25\% of end points for the previous search.
+Unit-specific parameters were $\rho$, $S_0$, $E_0$, $I_0$ (for the simulation study, model A) and all parameters (for the data, model B).}\label{fig:search_diagnostics}
+\end{figure}
 
-<<search_diagnostics, echo=F, fig.height=6.5, fig.width=4.5, out.width="5in", fig.cap = "Three steps of a likelihood search on simulated data (left panels) and UK measles data (right panels).
-Final log-likelihood values are plotted against starting log-likelihood for an initial search (top row), a refinement of this search (second row) and a second refinement (third row).
-Starting values for each refinement were 4 copies of the top 25\\% of end points for the previous search.
-Unit-specific parameters were $\\rho$, $S_0$, $E_0$, $I_0$ (for the simulation study, model A) and all parameters (for the data, model B).">>=
-par(mfcol=c(3,2))
-par(mai=c(0.3,0.3,0.1,0.1))
-par(omi=c(0.3,0.5,0.5,0.1))
-##
-## (1,1)
-##
-x <- v1a[,"start_loglik"]
-y <- v1a$loglik
-x_lim <- range(c(x,truth_loglik,max(y)))
-y_lim <- range(c(y,truth_loglik,max(x)))
-plot(x,y,xlim=x_lim,ylim=y_lim,
-  xlab="",
-  ylab="Initial search",
-)
-abline(a=0,b=1,lty="dashed",col="blue")
-abline(h=truth_loglik,lty="dashed",col="blue")
-abline(v=truth_loglik,lty="dashed",col="blue")
-##
-## (2,1)
-##
-x <- rep(v1a$loglik[1:18],each=2)
-y <- v2a$loglik
-x_lim <- range(c(x,truth_loglik,max(y)))
-y_lim <- range(c(y,truth_loglik,max(x)))
-plot(x,y,xlim=x_lim,ylim=y_lim,
-  xlab="",
-  ylab="1st refinement"
-)
-abline(a=0,b=1,lty="dashed",col="blue")
-abline(h=truth_loglik,lty="dashed",col="blue")
-abline(v=truth_loglik,lty="dashed",col="blue")
-##
-## (3,1)
-##
-x <- rep(v2a$loglik[1:18],each=2)
-y <- v3a$loglik
-x_lim <- range(c(x,truth_loglik,max(y)))
-y_lim <- range(c(y,truth_loglik,max(x)))
-plot(x,y,xlim=x_lim,ylim=y_lim,
-    ylab="2nd refinement",
-    xlab="Starting log-likelihood"
-)
-abline(a=0,b=1,lty="dashed",col="blue")
-abline(h=truth_loglik,lty="dashed",col="blue")
-abline(v=truth_loglik,lty="dashed",col="blue")
-##
-##
-## (1,2)
-##
-  x <- z1b[,"start_loglik"]
-  y <- z1b$loglik
-  x_lim <- range(c(x,loglik_he,max(y)))
-  y_lim <- range(c(y,loglik_he,max(x)))
-  plot(x,y,xlim=x_lim,ylim=y_lim,
-    xlab="",
-    ylab="",
-  )
-  abline(a=0,b=1,lty="dashed",col="blue")
-  abline(h=loglik_he,lty="dashed",col="blue")
-  abline(v=loglik_he,lty="dashed",col="blue")
-##
-## (2,2)
-##
-  x <- rep(z1b$loglik[1:18],each=2)
-  y <- z2b$loglik
-  x_lim <- range(c(x,loglik_he,max(y)))
-  y_lim <- range(c(y,loglik_he,max(x)))
-  plot(x,y,xlim=x_lim,ylim=y_lim,
-    xlab="",
-    ylab=""
-  )
-  abline(a=0,b=1,lty="dashed",col="blue")
-  abline(h=loglik_he,lty="dashed",col="blue")
-  abline(v=loglik_he,lty="dashed",col="blue")
-##
-## (3,2)
-##
-  x <- rep(z2b$loglik[1:18],each=2)
-  y <- z3b$loglik
-  x_lim <- range(c(x,loglik_he,max(y)))
-  y_lim <- range(c(y,loglik_he,max(x)))
-  plot(x,y,xlim=x_lim,ylim=y_lim,
-    xlab="Starting log-likelihood",
-    ylab=""
-  )
-  abline(a=0,b=1,lty="dashed",col="blue")
-  abline(h=loglik_he,lty="dashed",col="blue")
-  abline(v=loglik_he,lty="dashed",col="blue")
-
-LINE<-0.25
-CEX <- 0.9
-mtext("Initial search",side=2,line=LINE,outer=TRUE,at=0.833,cex=CEX)
-mtext("First refinement",side=2,line=LINE,outer=TRUE,at=0.5,cex=CEX)
-mtext("Second refinement",side=2,line=LINE,outer=TRUE,at=0.167,cex=CEX)
-mtext("Simulated data",side=3,line=LINE,outer=TRUE,at=0.25,cex=CEX)
-mtext("UK towns",side=3,line=LINE,outer=TRUE,at=0.75,cex=CEX)
-mtext("Log-likelihood",side=1,line=LINE,outer=TRUE,at=0.25,cex=CEX)
-mtext("Log-likelihood",side=1,line=LINE,outer=TRUE,at=0.75,cex=CEX)
-
-@
+\end{knitrout}
 
@@
 
-<<data_search_boxplot, echo=F, fig.height=4, fig.width=5.5, out.width="5.5in", fig.cap = "Fitting different models to the UK measles data using the method tested on simulated data.
-(A) $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B) $13\\times 20$ unit-specific parameters and no shared parameters.
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}
+
+\includegraphics[width=5.5in]{figure/data_search_boxplot-1} \hfill{}
+
+\caption[Fitting different models to the UK measles data using the method tested on simulated data]{Fitting different models to the UK measles data using the method tested on simulated data.
+(A) $4\times 20$ unit-specific parameters and $9$ shared parameters; (B) $13\times 20$ unit-specific parameters and no shared parameters.
 (C) Also all unit-specific, but with immigration rather than coupling, matching the model of He et al (2010).
-The horizontal dashed line denotes the likelihood obtained by He et al, using \\code{pomp}.",fig.show = "hold">>=
-a <- cbind(
-A1 = z1a$loglik,
-A2 = z2a$loglik,
-A3 = z3a$loglik,
-A4 = z4a$loglik,
-A5 = z5a$loglik
-)
+The horizontal dashed line denotes the likelihood obtained by He et al, using \code{pomp}.}\label{fig:data_search_boxplot}
+\end{figure}
 
-b <- cbind(
-B1 = z1b$loglik,
-B2 = z2b$loglik,
-B3 = z3b$loglik,
-B4 = z4b$loglik,
-B5 = z5b$loglik
-)
-
-c <- cbind(
-C1 = z1c$loglik,
-C2 = z2c$loglik,
-C3 = z3c$loglik,
-C4 = z4c$loglik,
-C5 = z5c$loglik
-)
-
-d <- cbind(
-D1 = z1d$loglik,
-D2 = z2d$loglik,
-D3 = z3d$loglik,
-D4 = z4d$loglik,
-D5 = z5d$loglik
-)
-loglik_max <- max(cbind(a,b,c),loglik_he)
-boxplot(cbind(a,b,c),ylim=loglik_max-c(5500,0),
-  ylab="Log-likelihood")
-abline(h=loglik_he,lty="dashed",col="blue")
-@
+\end{knitrout}
 
 Fig.~\ref{fig:data_search_boxplot} shows results for fitting models A, B and C to UK measles data using successive rounds of {\ibpf}, applying the same method used for the simulation study shown in Fig.~\ref{fig:sim_search_boxplot}.
-The dashed lines show the sum of the log-likelihoods obtained by \citet{he10}, $\ell_{\mathrm He}=\Sexpr{myround(loglik_he,1)}$.
+The dashed lines show the sum of the log-likelihoods obtained by \citet{he10}, $\ell_{\mathrm He}=-40345.7$.
 The value $\ell_{\mathrm He}$ corresponds to model C.
@@
 
-<<spatreg_boxplot, echo=F, fig.height=4, fig.width=5.5, out.width="5.5in", fig.pos="h", fig.cap = paste0("Varying the spatial autoregression algorithmic parameter for shared parameters with simulated measles data.
-The log-likelihood was obtained using $M=100$ iterations of IBPF starting at random parameters (median starting log-likelihood, ",signif(median_start,3), ").
-The horizontal dashed line denotes the log-likelihood at the true parameters.")>>=
-loglik_truth <- -38882.7 # from w1.r
-spat_regression <- c(0, 1/64, 1/32, 1/16, 1/8, 1/4, 1/2, 1) 
-spat_regression_char <- c("0", "1/64", "1/32", "1/16", "1/8", "1/4", "1/2", "1")
-Experiment <- length(spat_regression)
-results <- foreach(experiment=1:Experiment,.combine=rbind) %do% {
-  r <- readRDS(file=paste0("v/v6_3/global_e",experiment,".rds"))
-  cbind(r,sp_reg=spat_regression[experiment])
-}
-loglik_max <- max(c(loglik_truth,results$loglik))
-results <- as.data.frame(results)
-boxplot(loglik~sp_reg,data=results,ylim=c(-40000,loglik_max),
-  ylab="Log-likelihood", xlab="spatial autoregression parameter",
-  names=spat_regression_char)
-abline(h=loglik_truth,col="blue",lty="dotted")
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}[h]
+
+\includegraphics[width=5.5in]{figure/spatreg_boxplot-1} \hfill{}
+
+\caption[Varying the spatial autoregression algorithmic parameter for shared parameters with simulated measles data]{Varying the spatial autoregression algorithmic parameter for shared parameters with simulated measles data.
+The log-likelihood was obtained using $M=100$ iterations of IBPF starting at random parameters (median starting log-likelihood, -40800).
+The horizontal dashed line denotes the log-likelihood at the true parameters}\label{fig:spatreg_boxplot}
+\end{figure}
+
+\end{knitrout}
 

### 2022-05-31 bm2.R

@@ -1,0 +1,234 @@
+#' Brownian motion spatPomp generator with shared or unit-specific parameters
+#'
+#' Generate a class \sQuote{spatPomp} object representing a \code{U}-dimensional
+#' Brownian motion with spatial correlation decaying geometrically with
+#' distance around a circle. The model is defined in continuous time
+#' though in this case an Euler approximation is exact at the evaluation
+#' times.
+#'
+#' @param U A length-one numeric signifying dimension of the process.
+#' @param N A length-one numeric signifying the number of observation time steps to evolve the process.
+#' @param delta_t Process simulations are performed every \code{delta_t} time units
+#' whereas observations occur every one time unit
+#' @param shared_names identifies parameters that have common shared value for all units, which by default is all parameters.
+#' @param unit_specific_names determines which parameters take a different value for each unit. Cannot be specified if shared_names is specified.
+#' each unit. Other parameters are considered shared between all units.
+#' @importFrom utils data
+#' @return An object of class \sQuote{spatPomp} representing a simulation from a \code{U}-dimensional
+#' Brownian motion
+#' @examples
+#' b <- bm2(U=4, N=20)
+#' # See all the model specifications of the object
+#' spy(b)
+#' @export
+
+bm2 <- function(U=5,N=100,delta_t=0.1,
+  shared_names,unit_specific_names){
+  dist <- function(u,v,n=U) min(abs(u-v),abs(u-v+U),abs(u-v-U))
+  dmat <- matrix(0,U,U)
+  for(u in 1:U) {
+    for(v in 1:U) {
+      dmat[u,v] <- dist(u,v)
+    }
+  }
+  to_C_array <- function(v)paste0("{",paste0(v,collapse=","),"}")
+  dist_C_rows <- apply(dmat,1,to_C_array)
+  dist_C_array <- to_C_array(dist_C_rows)
+  dist_C <- paste0("const int dist[",U,"][",U,"] = ",dist_C_array,"; ")
+
+  obs_names <- paste0("U",1:U)
[…]

### 2022-05-31 bpfilter.R

@@ -1,0 +1,255 @@
+##' Block particle filter (BPF)
+##'
+##' An implementation of the block particle filter algorithm of Rebeschini and van Handel (2015), which is used to estimate the filter distribution
+##' of a spatiotemporal partially-observed Markov process.
+##' \code{bpfilter} requires a partition of the spatial units which can be provided by either the \code{block_size} or the \code{block_list} argument.
+##' The elements of the partition are called blocks. We perform resampling for each block independently based on sample weights within the block.
+##' Each resampled block only contains latent states for the spatial components within the block which allows for a ``cross-pollination" of
+##' particles where the highest weighted segments of each particle are more likely to be resampled and get combined with resampled components of
+##' other particles. The method mitigates the curse of dimensionality by resampling locally.
+##'
+##' @name bpfilter
+##' @rdname bpfilter
+##' @include spatPomp_class.R
+##' @family particle filter methods
+##' @inheritParams abf
+##'
+##' @param block_size The number of spatial units per block. If this is provided, the method subdivides units approximately evenly
+##' into blocks with size \code{block_size}.
+##' @param block_list List that specifies an exact partition of the spatial units. Each partition element, or block, is
+##' an integer vector of neighboring units.
+##'
+##' @examples
+##' # Create a simulation of a Brownian motion
+##' b <- bm(U=6, N=10)
+##'
+##' # Run BPF with the specified number of units per block
+##' bpfilterd_b1 <- bpfilter(b, Np = 100, block_size = 2)
+##'
+##' # Run BPF with the specified partition
+##' bpfilterd_b2 <- bpfilter(b,
+##'                          Np = 20,
+##'                          block_list = list(c(1,2), c(3,4), c(5,6)))
+##'
+##' # Get a likelihood estimate
+##' logLik(bpfilterd_b2)
+##'
+##' @return Upon successful completion, \code{bpfilter()} returns an object of class
+##' \sQuote{bpfilterd_spatPomp} containing the algorithmic parameters used to run \code{bpfilter()}
+##' and the estimated likelihood.
+##'
+##' @section Details:
[…]

### 2022-05-31 conv.r

@@ -1,0 +1,38 @@
+
+#file <- "v18_3/global_e5.rds"
+#file <- "v16_3/global_e3.rds"
+#file <- "v14_3/global_e3.rds"
+file <- "v15_3/global_e5.rds"
+
+conv <- function(file,sim_dat=TRUE,U=20){
+  x <- readRDS(file)
+  plotPars <- c("alpha","psi","R0","gamma","sigma","sigmaSE",
+    "cohort","amplitude","rho","g","S_0", "E_0", "I_0")
+    
+  if(sim_dat){
+    library(pomp)
+    readRDS("sim/sim.rds") -> sim
+    params <- coef(sim)[paste0(plotPars,1)]
+    names(params) <- plotPars
+  }
+
+  par(mfrow=c(length(plotPars),1))
+  par(mai=c(0.05,0.8,0.01,0.1))
+  par(omi=c(0.3,0,0.1,0))
+  for(i in seq_along(plotPars)){
+    p <- plotPars[i]
+    unit_names <- rep("",U)
+    if(i==length(plot)) unit_names <- 1:U
+    boxplot(x[,paste0(p,1:U)],names=unit_names,
+      xaxt=ifelse(i==length(plotPars),"t","n") )
+    mtext(p,side=2,line=2.6)
+    if(sim_dat) abline(h=params[p],col="red",lty="dashed")
+  }
+  mtext(1:U,line=0.7,at=seq(from=0.125,to=0.94,length=20),side=1,outer=T)
+}
+
+conv("v18_3/global_e5.rds")
+#file <- "v16_3/global_e3.rds"
+#file <- "v14_3/global_e3.rds"
+#conv("v15_3/global_e5.rds")
+

### 2022-05-31 he10.R

@@ -1,0 +1,615 @@
+# adjusted to look for higher compatibility with he10 numerical results
+library(tidyverse)
+#library(dplyr)
+library(spatPomp)
+
+#' Measles in UK: spatPomp generator with shared or unit-specific parameters
+#'
+#' Generate a spatPomp object for measles adding spatial coupling to
+#' The model and data from He et al. (2010) with gravity transport as
+#' in Park and Ionides (2020).
+#' Other transport models may be added.
+#' The structure of this spatPomp is designed to be compatible with
+#' spatiotemporal iterated filtering algorithms. Thus, shared parameters can
+#' be represented with a value for each unit, which should be the same for each
+#' unit in a valid model instance but may vary between units while optimizing.
+#'
+#' @param U A length-one numeric signifying the number of cities to be
+#' represented in the spatPomp object. Default U=20 gives all the towns
+#' studied by He et al., the 10 largest and 10 selected smaller towns.
+#' @importFrom utils data read.csv write.table
+#' @param dt a numeric (in unit of years) that is used as the Euler time-increment for simulating measles data.
+#' @param fixedParNames specifies parameters that have a shared value across
+#' units and are not being optimized so can be represented by a single value.
+#' @param sharedParNames identifies parameters that have common shared value
+#' for all units which is represented separately for each unit for optimization
+#' purposes.
+#' @param unitParNames determines which parameters take a different value for each unit. Cannot be specified if sharedParNames is specified.
+#' each unit. Other parameters are considered shared between all units.
+#' @param simulated determines whether to return a simulation from the model or the
+#' UK measles data
+#' @return An object of class \sQuote{spatPomp} representing a \code{U}-dimensional spatially coupled measles POMP model.
+#' @references
+#'
+#' \geosphere
+#'
+#' @note This function goes through a typical workflow of constructing
[…]

### 2022-05-31 ibpf.R

@@ -1,0 +1,472 @@
+##' Iterated block particle filter (IBPF)
+##'
+##' An iterated block particle filter, for both shared and unit-specific
+##' parameters. For this preliminary work, we suppose that the
+##' spatPomp has been constructed to have a unit-specific parameter "thetau"
+##' for unit u corresponding to an estimated parameter "theta", whether
+##' theta is shared or unit-specific. This permits IBPF
+##' to implement a spatiotemporal random walk to estimate theta.
+##' We require that rw.sd is positive for, and only for, all parameters
+##' of the form "thetau" if "theta" is listed in sharedParNames or
+##' unitParNames.
+##'
+##' @name ibpf
+##' @rdname ibpf
+##' @include spatPomp_class.R generics.R spatPomp.R bpfilter.R
+##' @family particle filter methods
+##' @family \pkg{spatPomp} filtering methods
+##'
+##'
+##' @inheritParams spatPomp
+##' @inheritParams pomp::mif2
+##'
+##' @param Nbpf the number of iterations of perturbed BPF.
+##'
+##' @param sharedParNames estimated parameters that are equal for each unit.
+##'
+##' @param unitParNames estimated parameters that are different for
+##' each unit.
+##'
+##' @return
+##' Upon successful completion, \code{ibpf} returns an object of class
+##' \sQuote{ibpfd_spatPomp}.
+##'
+##' @section Methods:
+##' The following methods are available for such an object:
+##' \describe{
+##' \item{\code{\link{coef}}}{ gives the Monte Carlo estimate of the maximum likelihood. }
+##' }
+##'
+
+NULL
+
+rw.sd <- pomp:::safecall
+
+setClass(
+  "ibpfd_spatPomp",
+  contains="bpfilterd_spatPomp",
+  slots=c(
+    Nbpf = 'integer',
+    spat_regression = 'numeric',
+    rw.sd = 'matrix',
+    cooling.type = 'character',
+    cooling.fraction.50 = 'numeric',
+    traces = 'matrix'
+  )
+)
+
+setGeneric(
+  "ibpf",
+  function (data, ...)
+    standardGeneric("ibpf")
+)
+
+setMethod(
+  "ibpf",
+  signature=signature(data="missing"),
+  definition=function (...) {
+    reqd_arg("ibpf","data")
[…]

### 2022-05-31 measles2.R

@@ -1,0 +1,522 @@
+#' Measles in UK: spatPomp generator with shared or unit-specific parameters
+#'
+#' Generate a spatPomp object for measles in the top-\code{U} most populous cities in England and Wales.
+#' The model is adapted from He et al. (2010) with gravity transport following Park and Ionides (2019).
+#' The data in the object is simulated using the process and measurement models of He et al. (2010).
+#' The structure of this spatPomp is designed to be compatible with
+#' spatiotemporal iterated filtering algorithms. Thus, shared parameters can
+#' be represented with a value for each unit, which should be the same for each
+#' unit in a valid model instance but may vary between units while optimizing.
+#'
+#' @param U A length-one numeric signifying the number of cities to be represented in the spatPomp object.
+#' @importFrom utils data read.csv write.table
+#' @param dt a numeric (in unit of years) that is used as the Euler time-increment for simulating measles data.
+#' @param fixedParNames specifies parameters that have a shared value across
+#' units and are not being optimized so can be represented by a single value.
+#' @param sharedParNames identifies parameters that have common shared value
+#' for all units which is represented separately for each unit for optimization
+#' purposes.
+#' @param unitParNames determines which parameters take a different value for each unit. Cannot be specified if sharedParNames is specified.
+#' each unit. Other parameters are considered shared between all units.
+#' @param simulated determines whether to return a simulation from the model or the
+#' UK measles data
+#' @return An object of class \sQuote{spatPomp} representing a \code{U}-dimensional spatially coupled measles POMP model.
+#' @references
+#'
+#' \geosphere
+#'
+#' @note This function goes through a typical workflow of constructing
+#' a typical spatPomp object (1-4 below). This allows the user to have a
[…]

### 2022-05-31 param_formats.R

@@ -1,0 +1,88 @@
+
+# Book-keeping functions to expand from a vector of shared parameters to
+# unit-specific copies, and to reverse this operation.
+
+# Only estimated shared parameters are expanded. Fixed shared parameters,
+# and unit-specific parameters that already have a copy for each unit, are
+# left unchanged.
+
+# The contracted form has a single entry for each shared parameter, and only
+# unit-specific parameters have a terminal number
+
+# The expanded form has U copies with appended 1:U for each shared or
+# unit-specific parameter. Fixed parameters (i.e., parameters that are not
+# estimated) have an appended 1. 
+
+# FixedParNames,sharedParNames and unitParNames are the names of the
+# corresponding parameters without any appended numbers.
+
+# Another form of the parameters, known as the basic form, is a reduced version
+# of the contracted form having just a single value for each parameter. It
+# corresponds to the parameter vector for a spatPomp model with all shared
+# parameters. expand_params() can also be used to generate an expanded form
+# from the basic form, if both shared and unit-specific parameters are passed
+# via the sharedParNames argument. From there, one can obtain the contracted
+# form.
+
+expand_params <- function(contracted_params,
+  fixedParNames,sharedParNames,unitParNames,U){
+  expand_shared_params <- foreach(par=sharedParNames,.combine=c)%do%{
+    x <- rep(contracted_params[par],U)
+    names(x) <- paste0(par,1:U)
+    x
+    }
+  expand_fixed_params <- contracted_params[fixedParNames]
+  names(expand_fixed_params) <- paste0(fixedParNames,"1")
+  expand_unit_params <- foreach(par=unitParNames,.combine=c)%do%{
+    contracted_params[paste0(par,1:U)]
+  }
+  c(expand_fixed_params,expand_shared_params,expand_unit_params)
+}
+
+contract_params <- function(params,
+  fixedParNames,sharedParNames,unitParNames,
+  average=FALSE,U){
+  expandedParNames <- names(params)
+  contract_fixed_params <- params[paste0(fixedParNames,"1")]
+  names(contract_fixed_params) <- fixedParNames
+  contract_unit_params <- foreach(par=unitParNames,.combine=c)%do%{
[…]

### 2022-05-31 sim.r

@@ -1,0 +1,193 @@
+# simulated data from he10 designed to choose parameters and a seed
+# such that the 20-city simulation from 1950 to 1964 looks
+# qualitatively similar to he10
+
+# when testing on simulated data, we need to check that the simulation is
+# representative of the process we want to study. For example, if simulations
+# can have 2-year or annual cycles, we may want to choose a simulation
+# with 2-year cycles for comparison with the data
+
+# also, we want to be able to test on subsets in time/space while using a
+# subset of one dataset
+
+# so, we should simulate at U=20, Tmax=1964 (CHECK THIS MATCHES HE10 EXACTLY)
+# we can fit to subsets, understanding that for spatial subsets they will
+# not exactly be a draw from the model on the subset.
+
+library(spatPomp)
+library(tidyverse)
+library(doParallel)
+library(doRNG)
+registerDoParallel()
+registerDoRNG(4454)
+
+source("ibpf.R")
+source("param_formats.R")
+load("data/twentycities.rda")
+source("he10.R")
+
+code_version <- "sim"
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+U <- 20
+
+# build a spatPomp object with all fixed parameters
+# (i.e., no unit-specific or expanded parameters)
+bake(file=paste0(out_dir,"he10.rds"),
+  he10(U=U,simulated=FALSE,  fixedParNames=c("alpha","iota","R0",
+    "cohort","amplitude","gamma",
+    "sigma","mu","sigmaSE","rho","psi","g","S_0","E_0","I_0"),
+    dt=1/365)
+) -> spo
+
+
+# N <- 20
+
+test <- 7
+
+if(test==1){
+## parameters comparable to a typical city for he10, generally more like London
+basicParams <- c(
+  alpha=0.98,
+  iota=0,
+  R0=30,
+  cohort=0.5,
+  amplitude=0.4,
+  gamma=52,
+  sigma=52,
+  mu=0.02,
+  sigmaSE=0.1,
+  rho=0.5,
+  psi=0.15,
+  g=400,
+  S_0=0.035,
+  E_0=0.00005,
[…]

### 2022-05-31 v/v6-compare.r

@@ -1,0 +1,119 @@
+
+# extension <- "_3"
+# > q75
+#        0      0.1      0.2      0.3     0.4       0.5     0.6       0.7 
+# -38662.1 -35355.7 -35250.9 -35308.9 -35350.7 -35365.4 -35284.1 -35385.0 
+#      0.8      0.9        1 
+# -35252.8 -35383.1 -35388.2 
+
+# boxplot.pdf shows that all values of spat_regression in [0.1,1] are
+# more-or-less equally good. A linear model cannot pick up a linear
+# or quadratic trend. However, spat_regression = 0 fails.
+
+# interpretation: in the context of this test (a matter of approaching the
+# mle, but not requiring to get within 100 log units) it does not matter
+# the rate at which the units talk to each other concerning shared parameters.
+# there may be a tradeoff - more diversity can lead to more information, but
+# quicker collaboration can direct the search more quickly to good places.
+
+# most runs are finding a lower-R0 regime (see R0 plot below for unit 5)
+# though some are finding values close to the truth of 30.
+
+loglik_max <- -35041 # from w1.r
+
+library(foreach)
+
+code_version <- "v6"
+
+extension <- "_3"
+
+out_dir <- paste0(code_version,extension,"/")
+# spat_regression <- c(0.3,0.4,0.5,0.6,0.7)
+spat_regression <- c(0.3,0.4,0.5,0.6,0.7,0,0.1,0.2,0.8,0.9,1)
+# Experiment <- 5 # initial test
+Experiment <- length(spat_regression)
+
+results <- foreach(experiment=1:Experiment,.combine=rbind) %do% {
+  r <- readRDS(file=paste0(out_dir,"global_",experiment,".rds"))
+  cbind(r,sp_reg=spat_regression[experiment])
+}
+
+dim(results)
+
+results <- as.data.frame(results)
+
+result_list <- split(results$loglik,results$sp_reg)
+
+
+density_list <- lapply(result_list,
+  function(x) density(x,n=1000,bw=50,from=loglik_max-1000,to=loglik_max)
+)
+density_matrix <- sapply(density_list,function(x) x$y)
+
+
+pdf(file=paste0(out_dir,"density-plot.pdf"))
[…]

### 2022-05-31 v/v6.r

@@ -1,0 +1,343 @@
+
+# experimenting with values of spat_regression.
+# such experiments could perhaps be carried out on a smaller example,
+# but we want to get experience relevant to the "target problem" of
+# data analysis on the he10 dataset.
+
+# this experiment corresponds to a broad search (starting values +/- 1 log
+# unit from the truth, rw.sd = 0.02). we could check later whether similar
+# conclusions hold for a local search
+
+# results on greatlakes, 
+# reps cores time(min) memory(GB) dir
+# 36   36    12:23     11.88/180    v6_3  with likelihood evaluation
+# 36   36    10:48     11/48/180    v6_3 without likelihood evaluation
+
+code_version <- "v6"
+run_level <- 3
+
+experiment <- 8
+spat_regression <- c(0, 1/64, 1/32, 1/16, 1/8, 1/4, 1/2, 1) 
+
+print(paste0("run_level = ",run_level))
+print(paste0("experiment = ",experiment))
+
+truth_loglik <- -38882.7 # from w1.r 
+
+# save various things of possible later interest
+# when doing the main computations
+# set to FALSE when re-running to analyze the baked results
+# save_session_data <- TRUE
+save_session_data <- experiment==1
+# save_session_data <- FALSE
+
+# do_bake = FALSE helps debug when bake triggers recomputation unintentionally
+#do_bake <- FALSE 
+do_bake <- TRUE
+
+# compute likelihood at initial guesses when we want to
+# check local optimization: how well does each search do from where it starts?
+# compute_start_likelihood <- TRUE
+compute_start_likelihood <- experiment==1
+# compute_start_likelihood <- FALSE
+
+# standard pomp pseudocode notation: U = units, N = time points,
+# J = particles, M = filter iterations.
+# Also, Reps = replicates
+# estimated is in ("mostly shared","mostly fixed",
+#   "all plausible parameters shared","all unit-specific")
+
+if(run_level==1){
+  estimated <- "mostly shared"
+  U <- 5
+  N <- 10
+  J <- 50
+  M <- 2
+  Nblocks <- U
[…]

### 2022-05-31 v/v6_3/v6.r

@@ -1,0 +1,343 @@
+
+# experimenting with values of spat_regression.
+# such experiments could perhaps be carried out on a smaller example,
+# but we want to get experience relevant to the "target problem" of
+# data analysis on the he10 dataset.
+
+# this experiment corresponds to a broad search (starting values +/- 1 log
+# unit from the truth, rw.sd = 0.02). we could check later whether similar
+# conclusions hold for a local search
+
+# results on greatlakes, 
+# reps cores time(min) memory(GB) dir
+# 36   36    12:23     11.88/180    v6_3  with likelihood evaluation
+# 36   36    10:48     11/48/180    v6_3 without likelihood evaluation
+
+code_version <- "v6"
+run_level <- 3
+
+experiment <- 1
+spat_regression <- c(0, 1/64, 1/32, 1/16, 1/8, 1/4, 1/2, 1) 
+
+print(paste0("run_level = ",run_level))
+print(paste0("experiment = ",experiment))
+
+truth_loglik <- -38882.7 # from w1.r 
+
+# save various things of possible later interest
+# when doing the main computations
+# set to FALSE when re-running to analyze the baked results
+# save_session_data <- TRUE
+save_session_data <- experiment==1
+# save_session_data <- FALSE
+
+# do_bake = FALSE helps debug when bake triggers recomputation unintentionally
+#do_bake <- FALSE 
+do_bake <- TRUE
+
+# compute likelihood at initial guesses when we want to
+# check local optimization: how well does each search do from where it starts?
+# compute_start_likelihood <- TRUE
+compute_start_likelihood <- experiment==1
+# compute_start_likelihood <- FALSE
+
+# standard pomp pseudocode notation: U = units, N = time points,
+# J = particles, M = filter iterations.
+# Also, Reps = replicates
+# estimated is in ("mostly shared","mostly fixed",
+#   "all plausible parameters shared","all unit-specific")
+
+if(run_level==1){
+  estimated <- "mostly shared"
+  U <- 5
+  N <- 10
+  J <- 50
+  M <- 2
+  Nblocks <- U
[…]

### 2022-05-31 v4/v1-analysis.r

@@ -1,0 +1,29 @@
+
+library(pomp)
+truth_eval <- readRDS("../w/w1_3/lik.rds")
+truth_loglik<-logmeanexp(truth_eval[truth_eval[,"J"]==25600,"logLik"])
+
+v1a <- readRDS(file="v1a_3/global.rds")
+v1b <- readRDS(file="v1b_3/global.rds")
+v1c <- readRDS(file="v1c_3/global.rds")
+
+a <- cbind(
+A1 = v1a$loglik
+)
+
+b <- cbind(
+B1 = v1b$loglik
+)
+
+c <- cbind(
+C1 = v1c$loglik
+)
+
+
+loglik_max <- max(cbind(a,b,c),truth_loglik)
+
+pdf(file="v1-plot.pdf")
+boxplot(cbind(a,b,c),ylim=loglik_max-c(200,0))
+abline(h=truth_loglik,lty="dashed",col="blue")
+dev.off()
+

### 2022-05-31 v4/v1a.r

@@ -1,0 +1,366 @@
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v1"
+old_version <- "NA" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- FALSE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
[…]

### 2022-05-31 v4/v1a_3/v1a.r

@@ -1,0 +1,354 @@
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+run_version <- "a"
+code_version <- paste0("v1",run_version)
+run_level <- 3
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10") # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+if(continuation){
+  # old results used for starting the new search
+  old_results <- read.csv(file=paste0(out_dir,"global.csv"))
+}
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- 0.1
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
+print(paste0("experiment = ",experiment))
+
+# save various things of possible later interest
+# when doing the main computations
+# set to FALSE when re-running to analyze the baked results
+save_session_data <- TRUE
+# save_session_data <- FALSE
+
+# do_bake = FALSE helps debug when bake triggers recomputation unintentionally
+#do_bake <- FALSE 
+do_bake <- TRUE
+
[…]

### 2022-05-31 v4/v1b.r

@@ -1,0 +1,366 @@
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v1"
+old_version <- "NA" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- FALSE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
[…]

### 2022-05-31 v4/v1b_3/v1b.r

@@ -1,0 +1,354 @@
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+run_version <- "b"
+code_version <- paste0("v1",run_version)
+run_level <- 3
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10") # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+if(continuation){
+  # old results used for starting the new search
+  old_results <- read.csv(file=paste0(out_dir,"global.csv"))
+}
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- 0.1
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
+print(paste0("experiment = ",experiment))
+
+# save various things of possible later interest
+# when doing the main computations
+# set to FALSE when re-running to analyze the baked results
+save_session_data <- TRUE
+# save_session_data <- FALSE
+
+# do_bake = FALSE helps debug when bake triggers recomputation unintentionally
+#do_bake <- FALSE 
+do_bake <- TRUE
+
[…]

### 2022-05-31 v4/v1c.r

@@ -1,0 +1,366 @@
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v1"
+old_version <- "NA" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+code_version <- paste0(devel_version,run_version)
+run_level <- 1
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- FALSE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
[…]

### 2022-05-31 v4/v1c_3/v1c.r

@@ -1,0 +1,354 @@
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+run_version <- "c"
+code_version <- paste0("v1",run_version)
+run_level <- 3
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10") # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+if(continuation){
+  # old results used for starting the new search
+  old_results <- read.csv(file=paste0(out_dir,"global.csv"))
+}
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- 0.1
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
+print(paste0("experiment = ",experiment))
+
+# save various things of possible later interest
+# when doing the main computations
+# set to FALSE when re-running to analyze the baked results
+save_session_data <- TRUE
+# save_session_data <- FALSE
+
+# do_bake = FALSE helps debug when bake triggers recomputation unintentionally
+#do_bake <- FALSE 
+do_bake <- TRUE
+
[…]

### 2022-05-31 v4/v2-analysis.r

@@ -1,0 +1,95 @@
+
+library(pomp)
+truth_eval <- readRDS("../w/w1_3/lik.rds")
+truth_loglik<-logmeanexp(truth_eval[truth_eval[,"J"]==25600,"logLik"])
+
+# global.rds is ordered by starting value enumeration
+# global.csv is ordered by likelihood
+LIK_ORDER <- TRUE
+#LIK_ORDER <- FALSE
+if(LIK_ORDER){
+  v1a <- read.csv(file="v1a_3/global.csv")
+  v1b <- read.csv(file="v1b_3/global.csv")
+  v1c <- read.csv(file="v1c_3/global.csv")
+  v2a <- read.csv(file="v2a_3/global.csv")
+  v2b <- read.csv(file="v2b_3/global.csv")
+  v2c <- read.csv(file="v2c_3/global.csv")
+} else {
+  v1a <- readRDS(file="v1a_3/global.rds")
+  v1b <- readRDS(file="v1b_3/global.rds")
+  v1c <- readRDS(file="v1c_3/global.rds")
+  v2a <- readRDS(file="v2a_3/global.rds")
+  v2b <- readRDS(file="v2b_3/global.rds")
+  v2c <- readRDS(file="v2c_3/global.rds")
+}
+
+REMAKE=TRUE
+if(REMAKE){
+
+  pdf(file="v1a-vs-initial.pdf")
+  x <- v1a[,"start_loglik"]
+  y <- v1a$loglik
+  x_lim <- range(c(x,truth_loglik,max(y)))
+  y_lim <- range(c(y,truth_loglik,max(x)))
+  plot(x,y,xlim=x_lim,ylim=y_lim,
+    xlab="starting log lik",
+    ylab="v1 log lik"
+  )
+  abline(a=0,b=1,lty="dashed",col="blue")
+  abline(h=truth_loglik,lty="dashed",col="blue")
+  abline(v=truth_loglik,lty="dashed",col="blue")
+  dev.off()
+
+  pdf(file="v2a-vs-v1a.pdf")
+  x <- rep(v1a$loglik[1:18],each=2)
+  y <- v2a$loglik
+  x_lim <- range(c(x,truth_loglik,max(y)))
+  y_lim <- range(c(y,truth_loglik,max(x)))
+  plot(x,y,xlim=x_lim,ylim=y_lim,
+    xlab="v1 log lik",
[…]

### 2022-05-31 v4/v2a.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v2"
+old_version <- "v1" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+seed_offset <- 1
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v2a_3/v2a.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v2"
+old_version <- "v1" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+seed_offset <- 1
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v2b.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v2"
+old_version <- "v1" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+seed_offset <- 1
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v2c.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v2"
+old_version <- "v1" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+seed_offset <- 1
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v2c_3/v2c.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v2"
+old_version <- "v1" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+seed_offset <- 1
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v3-analysis.r

@@ -1,0 +1,129 @@
+
+library(pomp)
+truth_eval <- readRDS("../w/w1_3/lik.rds")
+truth_loglik<-logmeanexp(truth_eval[truth_eval[,"J"]==25600,"logLik"])
+
+# global.rds is ordered by starting value enumeration
+# global.csv is ordered by likelihood
+LIK_ORDER <- TRUE 
+if(LIK_ORDER){
+  v1a <- read.csv(file="v1a_3/global.csv")
+  v1b <- read.csv(file="v1b_3/global.csv")
+  v1c <- read.csv(file="v1c_3/global.csv")
+  v2a <- read.csv(file="v2a_3/global.csv")
+  v2b <- read.csv(file="v2b_3/global.csv")
+  v2c <- read.csv(file="v2c_3/global.csv")
+  v3a <- read.csv(file="v3a_3/global.csv")
+  v3b <- read.csv(file="v3b_3/global.csv")
+  v3c <- read.csv(file="v3c_3/global.csv")
+} else {
+  v1a <- readRDS(file="v1a_3/global.rds")
+  v1b <- readRDS(file="v1b_3/global.rds")
+  v1c <- readRDS(file="v1c_3/global.rds")
+  v2a <- readRDS(file="v2a_3/global.rds")
+  v2b <- readRDS(file="v2b_3/global.rds")
+  v2c <- readRDS(file="v2c_3/global.rds")
+}
+
+REMAKE=TRUE
+if(REMAKE){
+
+  pdf(file="v1a-vs-initial.pdf")
+  x <- v1a[,"start_loglik"]
+  y <- v1a$loglik
+  x_lim <- range(c(x,truth_loglik,max(y)))
+  y_lim <- range(c(y,truth_loglik,max(x)))
+  plot(x,y,xlim=x_lim,ylim=y_lim,
+    xlab="starting log lik",
+    ylab="v1 log lik"
+  )
+  abline(a=0,b=1,lty="dashed",col="blue")
+  abline(h=truth_loglik,lty="dashed",col="blue")
+  abline(v=truth_loglik,lty="dashed",col="blue")
+  dev.off()
+
+  pdf(file="v2a-vs-v1a.pdf")
+  x <- rep(v1a$loglik[1:18],each=2)
+  y <- v2a$loglik
[…]

### 2022-05-31 v4/v3a.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v3"
+old_version <- "v2" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+seed_offset <- 3
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v3b.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v3"
+old_version <- "v2" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+seed_offset <- 3
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v3c.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v3"
+old_version <- "v2" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+seed_offset <- 3
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v4-analysis.r

@@ -1,0 +1,163 @@
+
+library(pomp)
+truth_eval <- readRDS("../w/w1_3/lik.rds")
+truth_loglik<-logmeanexp(truth_eval[truth_eval[,"J"]==25600,"logLik"])
+
+# global.rds is ordered by starting value enumeration
+# global.csv is ordered by likelihood
+LIK_ORDER <- TRUE 
+if(LIK_ORDER){
+  v1a <- read.csv(file="v1a_3/global.csv")
+  v1b <- read.csv(file="v1b_3/global.csv")
+  v1c <- read.csv(file="v1c_3/global.csv")
+  v2a <- read.csv(file="v2a_3/global.csv")
+  v2b <- read.csv(file="v2b_3/global.csv")
+  v2c <- read.csv(file="v2c_3/global.csv")
+  v3a <- read.csv(file="v3a_3/global.csv")
+  v3b <- read.csv(file="v3b_3/global.csv")
+  v3c <- read.csv(file="v3c_3/global.csv")
+  v4a <- read.csv(file="v4a_3/global.csv")
+  v4b <- read.csv(file="v4b_3/global.csv")
+  v4c <- read.csv(file="v4c_3/global.csv")
+} else {
+  v1a <- readRDS(file="v1a_3/global.rds")
+  v1b <- readRDS(file="v1b_3/global.rds")
+  v1c <- readRDS(file="v1c_3/global.rds")
+  v2a <- readRDS(file="v2a_3/global.rds")
+  v2b <- readRDS(file="v2b_3/global.rds")
+  v2c <- readRDS(file="v2c_3/global.rds")
+}
+
+# REMAKE=TRUE
+REMAKE = FALSE
+if(REMAKE){
+
+  pdf(file="v1a-vs-initial.pdf")
+  x <- v1a[,"start_loglik"]
+  y <- v1a$loglik
+  x_lim <- range(c(x,truth_loglik,max(y)))
+  y_lim <- range(c(y,truth_loglik,max(x)))
+  plot(x,y,xlim=x_lim,ylim=y_lim,
+    xlab="starting log lik",
+    ylab="v1 log lik"
+  )
+  abline(a=0,b=1,lty="dashed",col="blue")
+  abline(h=truth_loglik,lty="dashed",col="blue")
[…]

### 2022-05-31 v4/v4a_3/v4a.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v4"
+old_version <- "v3" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+seed_offset <- 4
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v4b_3/v4b.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v4"
+old_version <- "v3" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+seed_offset <- 4
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v4c_3/v4c.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v4"
+old_version <- "v3" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+seed_offset <- 4
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v5-analysis.r

@@ -1,0 +1,198 @@
+
+library(pomp)
+truth_eval <- readRDS("../w/w1_3/lik.rds")
+truth_loglik<-logmeanexp(truth_eval[truth_eval[,"J"]==25600,"logLik"])
+
+# global.rds is ordered by starting value enumeration
+# global.csv is ordered by likelihood
+LIK_ORDER <- TRUE 
+if(LIK_ORDER){
+  v1a <- read.csv(file="v1a_3/global.csv")
+  v1b <- read.csv(file="v1b_3/global.csv")
+  v1c <- read.csv(file="v1c_3/global.csv")
+  v2a <- read.csv(file="v2a_3/global.csv")
+  v2b <- read.csv(file="v2b_3/global.csv")
+  v2c <- read.csv(file="v2c_3/global.csv")
+  v3a <- read.csv(file="v3a_3/global.csv")
+  v3b <- read.csv(file="v3b_3/global.csv")
+  v3c <- read.csv(file="v3c_3/global.csv")
+  v4a <- read.csv(file="v4a_3/global.csv")
+  v4b <- read.csv(file="v4b_3/global.csv")
+  v4c <- read.csv(file="v4c_3/global.csv")
+  v5a <- read.csv(file="v5a_3/global.csv")
+  v5b <- read.csv(file="v5b_3/global.csv")
+  v5c <- read.csv(file="v5c_3/global.csv")
+} else {
+  v1a <- readRDS(file="v1a_3/global.rds")
+  v1b <- readRDS(file="v1b_3/global.rds")
+  v1c <- readRDS(file="v1c_3/global.rds")
+  v2a <- readRDS(file="v2a_3/global.rds")
+  v2b <- readRDS(file="v2b_3/global.rds")
+  v2c <- readRDS(file="v2c_3/global.rds")
+}
+
+# REMAKE=TRUE
+REMAKE = FALSE
+if(REMAKE){
+
+  pdf(file="v1a-vs-initial.pdf")
+  x <- v1a[,"start_loglik"]
+  y <- v1a$loglik
+  x_lim <- range(c(x,truth_loglik,max(y)))
+  y_lim <- range(c(y,truth_loglik,max(x)))
+  plot(x,y,xlim=x_lim,ylim=y_lim,
+    xlab="starting log lik",
+    ylab="v1 log lik"
+  )
[…]

### 2022-05-31 v4/v5a_3/v5a.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v5"
+old_version <- "v4" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+seed_offset <- 5
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v5b.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v5"
+old_version <- "v4" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+seed_offset <- 5
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 v4/v5c.r

@@ -1,0 +1,363 @@
+# v4 is the same as v3 except restarting the top 25% not top 50%
+# v3 is a new attempt at maximizing simulated data from the he10 model
+
+# the goal is to work out what it takes to make reasonable progress on the
+# last 100 units or so of log likelihood, which can be problematic when
+# there are many parameters.
+
+devel_version <- "v5"
+old_version <- "v4" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+seed_offset <- 5
+
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- TRUE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 5
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
[…]

### 2022-05-31 w/w1_3/w1.r

@@ -1,0 +1,143 @@
+
+# evaluating the likelihood at the truth on simulated data
+# how many particles / replicates do we need?
+
+# conclusions: 20 evaluations with 5000 particles
+# for J=6400, we obtained one evaluation with considerably higher likelihood,
+# leading to a large se. this suggests that there may be more subtlety to
+# the evaluation.
+
+# estimated time for run level 3: 8hr on greatlakes
+
+#             400           800          1600          3200          6400
+#   -39016.501733 -38945.237090 -38911.400026 -3.890015e+04 -38882.711908
+#se      8.883076      1.915404      1.107329  9.755192e-01      9.926747
+
+code_version <- "w1"
+run_level <- 3
+save_session_data <- TRUE
+
+print(paste0("run_level = ",run_level))
+
+library(spatPomp)
+library(tidyverse)
+source("../ibpf.R")
+source("../param_formats.R")
+source("../he10.R")
+
+library(doParallel)
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores)) cores <- detectCores()  
+registerDoParallel(cores)
+
+library(doRNG)
+registerDoRNG(4454)
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+sbat_file <- paste0(code_version,"_",run_level,".sbat")
+
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# standard pomp pseudocode notation: U = units, N = time points,
+# J = particles, M = filter iterations.
+# Also, Reps = replicates
+
+# the number of estimated parameters should be irrelevant for likelihood
+# evaluation. it does affect the extended model, but that is
+# checked elsewhere to give rise to the same simulations, so
+# here we use sim_test which has all fixed parameters
+
+
+## parameters comparable to a typical city for he10, generally more like London
+## selected and tested in sim.r
+sim <- readRDS(file="../sim/sim.rds")
+
+# here, we're interested in how to do this for the he10 data
+U <- nrow(obs(sim)) # i.e., 20
+
[…]

### 2022-05-31 w/w2-analysis.r

@@ -1,0 +1,17 @@
+
+he10mle <- read.csv("../data/he10mle.csv")
+loglik_he10 <- sum(he10mle$loglik)
+
+results <- readRDS("w2_3/lik.rds")
+
+library(pomp)
+library(tidyverse)
+
+logliks <- split(results[,"logLik"],results[,"J"]) %>% lapply(logmeanexp)
+
+logmeanexp(results[results[,"J"]==25600,"logLik"])
+
+unlist(c(he10=loglik_he10,logliks))
+#      he10      1600      3200      6400     12800     25600 
+# -40345.70 -40400.84 -40369.14 -40354.10 -40346.10 -40338.72 
+

### 2022-05-31 w/w2.r

@@ -1,0 +1,157 @@
+
+# evaluates the likelihood for the measles data
+# how many particles / replicates do we need?
+
+# we use the mle from he10, so this is a test of the block particle filter
+# in a situation where there is no coupling
+
+# estimates time for run level 3: 8 hr on greatlakes
+
+code_version <- "w2"
+run_level <- 3
+#save_session_data <- TRUE
+save_session_data <- FALSE
+
+print(paste0("run_level = ",run_level))
+
+library(spatPomp)
+library(tidyverse)
+source("../ibpf.R")
+source("../param_formats.R")
+load("../data/twentycities.rda")
+source("../he10.R")
+
+library(doParallel)
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores)) cores <- detectCores()  
+registerDoParallel(cores)
+
+library(doRNG)
+registerDoRNG(4454)
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+sbat_file <- paste0(code_version,"_",run_level,".sbat")
+
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# standard pomp pseudocode notation: U = units, N = time points,
+# J = particles, M = filter iterations.
+# Also, Reps = replicates
+
+
+parNames <- c("S_0","E_0","I_0","alpha","R0","psi","g","sigma","gamma","amplitude","cohort","sigmaSE","rho","mu","iota")
+U <- 20
+N <- 730
+sharedParNames <- NULL
+fixedParNames <- c("mu","g")
+unitParNames <- setdiff(parNames,c(sharedParNames,fixedParNames))
+estParNames <- c(sharedParNames,unitParNames)
+ivpParNames <- c("S_0","E_0","I_0")
+ivpEstParNames <- intersect(ivpParNames,estParNames)
+regEstParNames <- setdiff(estParNames,ivpParNames)
+
+Tmax <- min(1950 + N/52, 1964)
+
+spo <- he10(U=U, Tmax=Tmax,
+  sharedParNames=sharedParNames,
+  fixedParNames=fixedParNames,
+  unitParNames=unitParNames,
+  simulated=FALSE,
+  dt=1/365
+)
[…]

### 2022-05-31 w/w3.r

@@ -1,0 +1,92 @@
+# cross-checking he10.R against he10 logliks
+# using bpfilter
+
+# CHECKS OKAY!
+
+code_version <- "w3"
+out_dir <- paste0(code_version,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+library(spatPomp)
+library(tidyverse)
+source("../ibpf.R")
+source("../param_formats.R")
+load("../data/twentycities.rda")
+source("../he10.R")
+
+parNames <- c("S_0","E_0","I_0","alpha","R0","psi","g","sigma","gamma","amplitude","cohort","sigmaSE","rho","mu","iota")
+#for(u in 1:20){
+  town_vec <- c(1:20) # note: town_vec should be an increasing sequence
+  U <- length(town_vec)
+  N <- 730
+  sharedParNames <- NULL
+  fixedParNames <- c("mu","g")
+  unitParNames <- setdiff(parNames,c(sharedParNames,fixedParNames))
+  estParNames <- c(sharedParNames,unitParNames)
+  ivpParNames <- c("S_0","E_0","I_0")
+  ivpEstParNames <- intersect(ivpParNames,estParNames)
+  regEstParNames <- setdiff(estParNames,ivpParNames)
+
+  Tmax <- min(1950 + N/52, 1964)
+
+  spo <- he10(U=U, Tmax=Tmax,
+    sharedParNames=sharedParNames,
+    fixedParNames=fixedParNames,
+    unitParNames=unitParNames,
+    simulated=FALSE,
+    dt=1/365,
+    town_vec=town_vec
+  )
+
+  he10mle <- read.csv("../data/he10mle.csv")
+  theta <- rep(NA,length(coef(spo)))
+  names(theta) <- names(coef(spo))
+
+  for(u in 1:U){
+    he_u <- which(he10mle[,1]==spo@unit_names[u])
+    theta[paste0(unitParNames,u)]<- as.numeric(he10mle[he_u,unitParNames])  
+  }
+
+  theta["g1"]<-0
+  theta["mu1"] <- 0.02
+
+  coef(spo) <- theta
+  
+  bpf1 <- bake(file=paste0(out_dir,"bpf1.rds"),{
+    bpfilter(spo,Np=10000,block_size=1)
+  })
+
+  print(bpf1@unit_names)
+  print(logLik(bpf1))
[…]

### 2022-05-31 z3/z1-analysis.r

@@ -1,0 +1,28 @@
+
+library(pomp)
+loglik_he <- sum(read.csv("../data/he10mle.csv")$loglik)
+# -40345.7
+
+a <- cbind(
+A1 = readRDS(file="z1a_3/global.rds")$loglik
+)
+
+b <- cbind(
+B1 = readRDS(file="z1b_3/global.rds")$loglik
+)
+
+c <- cbind(
+C1 = readRDS(file="z1c_3/global.rds")$loglik
+)
+
+d <- cbind(
+D1 = readRDS(file="z1d_3/global.rds")$loglik
+)
+
+loglik_max <- max(cbind(a,b,c,d),loglik_he)
+
+pdf(file="z1-plot.pdf")
+boxplot(cbind(a,b,c,d),ylim=loglik_max-c(5000,0))
+abline(h=loglik_he,lty="dashed",col="blue")
+dev.off()
+

### 2022-05-31 z3/z1a.r

@@ -1,0 +1,378 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z1"
+old_version <- "NA" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+code_version <- paste0(devel_version,run_version)
+run_level <- 1
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+ do_bake <- TRUE
+
+continuation <- FALSE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+if(continuation){
+  # old results used for starting the new search
+  old_results <- read.csv(file=paste0(out_dir,"global.csv"))
+}
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
[…]

### 2022-05-31 z3/z1a_3/z1a.r

@@ -1,0 +1,370 @@
+# z2 is a new attempt at maximizing likelihood for data & model of he10
+
+devel_version <- "z1"
+old_version <- "NA" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- FALSE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+if(continuation){
+  # old results used for starting the new search
+  old_results <- read.csv(file=paste0(out_dir,"global.csv"))
+}
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
[…]

### 2022-05-31 z3/z1b_3/z1b.r

@@ -1,0 +1,370 @@
+# z2 is a new attempt at maximizing likelihood for data & model of he10
+
+devel_version <- "z1"
+old_version <- "NA" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- FALSE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+if(continuation){
+  # old results used for starting the new search
+  old_results <- read.csv(file=paste0(out_dir,"global.csv"))
+}
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
[…]

### 2022-05-31 z3/z1c_3/z1c.r

@@ -1,0 +1,370 @@
+# z2 is a new attempt at maximizing likelihood for data & model of he10
+
+devel_version <- "z1"
+old_version <- "NA" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- FALSE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+if(continuation){
+  # old results used for starting the new search
+  old_results <- read.csv(file=paste0(out_dir,"global.csv"))
+}
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
[…]

### 2022-05-31 z3/z1d_3/z1d.r

@@ -1,0 +1,370 @@
+# z2 is a new attempt at maximizing likelihood for data & model of he10
+
+devel_version <- "z1"
+old_version <- "NA" # used to read in starting values when continuation=TRUE
+run_version <- "d"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+continuation <- FALSE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+continuation <- FALSE # set to be false when starting a new search
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+if(continuation){
+  # old results used for starting the new search
+  old_results <- read.csv(file=paste0(out_dir,"global.csv"))
+}
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
[…]

### 2022-05-31 z3/z2-analysis.r

@@ -1,0 +1,98 @@
+
+library(pomp)
+loglik_he <- sum(read.csv("../data/he10mle.csv")$loglik)
+# -40345.7
+
+# global.rds is ordered by starting value enumeration
+# global.csv is ordered by likelihood
+LIK_ORDER <- TRUE 
+if(LIK_ORDER){
+  z1a <- read.csv(file="z1a_3/global.csv")
+  z1b <- read.csv(file="z1b_3/global.csv")
+  z1c <- read.csv(file="z1c_3/global.csv")
+  z1d <- read.csv(file="z1d_3/global.csv")
+  z2a <- read.csv(file="z2a_3/global.csv")
+  z2b <- read.csv(file="z2b_3/global.csv")
+  z2c <- read.csv(file="z2c_3/global.csv")
+  z2d <- read.csv(file="z2d_3/global.csv")
+} else {
+  z1a <- readRDS(file="z1a_3/global.rds")
+  z1b <- readRDS(file="z1b_3/global.rds")
+  z1c <- readRDS(file="z1c_3/global.rds")
+  z1d <- readRDS(file="z1d_3/global.rds")
+  z2a <- readRDS(file="z2a_3/global.rds")
+  z2b <- readRDS(file="z2b_3/global.rds")
+  z2c <- readRDS(file="z2c_3/global.rds")
+  z2d <- readRDS(file="z2d_3/global.rds")
+}
+
+pdf(file="z1a-vs-initial.pdf")
+x <- z1a[,"start_loglik"]
+y <- z1a$loglik
+x_lim <- range(c(x,loglik_he,max(y)))
+y_lim <- range(c(y,loglik_he,max(x)))
+plot(x,y,xlim=x_lim,ylim=y_lim,
+  xlab="starting log lik",
+  ylab="z1 log lik"
+)
+abline(a=0,b=1,lty="dashed",col="blue")
+abline(h=loglik_he,lty="dashed",col="blue")
+abline(v=loglik_he,lty="dashed",col="blue")
+dev.off()
+
+pdf(file="z2a-vs-z1a.pdf")
+x <- rep(z1a$loglik[1:18],each=2)
+y <- z2a$loglik
+x_lim <- range(c(x,loglik_he,max(y)))
+y_lim <- range(c(y,loglik_he,max(x)))
[…]

### 2022-05-31 z3/z2a_3/z2a.r

@@ -1,0 +1,370 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z2"
+old_version <- "z1" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z2b.r

@@ -1,0 +1,370 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z2"
+old_version <- "z1" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z2c_3/z2c.r

@@ -1,0 +1,370 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z2"
+old_version <- "z1" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z2d_3/z2d.r

@@ -1,0 +1,370 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z2"
+old_version <- "z1" # used to read in starting values when continuation=TRUE
+run_version <- "d"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z3-analysis.r

@@ -1,0 +1,140 @@
+
+library(pomp)
+loglik_he <- sum(read.csv("../data/he10mle.csv")$loglik)
+# -40345.7
+
+# global.rds is ordered by starting value enumeration
+# global.csv is ordered by likelihood
+LIK_ORDER <- TRUE 
+if(LIK_ORDER){
+  z1a <- read.csv(file="z1a_3/global.csv")
+  z1b <- read.csv(file="z1b_3/global.csv")
+  z1c <- read.csv(file="z1c_3/global.csv")
+  z1d <- read.csv(file="z1d_3/global.csv")
+  z2a <- read.csv(file="z2a_3/global.csv")
+  z2b <- read.csv(file="z2b_3/global.csv")
+  z2c <- read.csv(file="z2c_3/global.csv")
+  z2d <- read.csv(file="z2d_3/global.csv")
+  z3a <- read.csv(file="z3a_3/global.csv")
+  z3b <- read.csv(file="z3b_3/global.csv")
+  z3c <- read.csv(file="z3c_3/global.csv")
+  z3d <- read.csv(file="z3d_3/global.csv")
+} else {
+  z1a <- readRDS(file="z1a_3/global.rds")
+  z1b <- readRDS(file="z1b_3/global.rds")
+  z1c <- readRDS(file="z1c_3/global.rds")
+  z1d <- readRDS(file="z1d_3/global.rds")
+  z2a <- readRDS(file="z2a_3/global.rds")
+  z2b <- readRDS(file="z2b_3/global.rds")
+  z2c <- readRDS(file="z2c_3/global.rds")
+  z2d <- readRDS(file="z3d_3/global.rds")
+  z3a <- readRDS(file="z3a_3/global.rds")
+  z3b <- readRDS(file="z3b_3/global.rds")
+  z3c <- readRDS(file="z3c_3/global.rds")
+  z3d <- readRDS(file="z3d_3/global.rds")
+}
+
+REMAKE=FALSE
+if(REMAKE){
+  pdf(file="z1a-vs-initial.pdf")
+  x <- z1a[,"start_loglik"]
+  y <- z1a$loglik
+  x_lim <- range(c(x,loglik_he,max(y)))
+  y_lim <- range(c(y,loglik_he,max(x)))
+  plot(x,y,xlim=x_lim,ylim=y_lim,
[…]

### 2022-05-31 z3/z3a.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z3"
+old_version <- "z2" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 1
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z3b.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z3"
+old_version <- "z2" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 1
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z3c_3/z3c.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z3"
+old_version <- "z2" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 1
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z3d.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z3"
+old_version <- "z2" # used to read in starting values when continuation=TRUE
+run_version <- "d"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 1
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z4-analysis.r

@@ -1,0 +1,175 @@
+
+library(pomp)
+loglik_he <- sum(read.csv("../data/he10mle.csv")$loglik)
+# -40345.7
+
+# global.rds is ordered by starting value enumeration
+# global.csv is ordered by likelihood
+LIK_ORDER <- TRUE 
+if(LIK_ORDER){
+  z1a <- read.csv(file="z1a_3/global.csv")
+  z1b <- read.csv(file="z1b_3/global.csv")
+  z1c <- read.csv(file="z1c_3/global.csv")
+  z1d <- read.csv(file="z1d_3/global.csv")
+  z2a <- read.csv(file="z2a_3/global.csv")
+  z2b <- read.csv(file="z2b_3/global.csv")
+  z2c <- read.csv(file="z2c_3/global.csv")
+  z2d <- read.csv(file="z2d_3/global.csv")
+  z3a <- read.csv(file="z3a_3/global.csv")
+  z3b <- read.csv(file="z3b_3/global.csv")
+  z3c <- read.csv(file="z3c_3/global.csv")
+  z3d <- read.csv(file="z3d_3/global.csv")
+  z4a <- read.csv(file="z4a_3/global.csv")
+  z4b <- read.csv(file="z4b_3/global.csv")
+  z4c <- read.csv(file="z4c_3/global.csv")
+  z4d <- read.csv(file="z4d_3/global.csv")
+} else {
+  z1a <- readRDS(file="z1a_3/global.rds")
+  z1b <- readRDS(file="z1b_3/global.rds")
+  z1c <- readRDS(file="z1c_3/global.rds")
+  z1d <- readRDS(file="z1d_3/global.rds")
+  z2a <- readRDS(file="z2a_3/global.rds")
+  z2b <- readRDS(file="z2b_3/global.rds")
+  z2c <- readRDS(file="z2c_3/global.rds")
+  z2d <- readRDS(file="z3d_3/global.rds")
+  z3a <- readRDS(file="z3a_3/global.rds")
+  z3b <- readRDS(file="z3b_3/global.rds")
+  z3c <- readRDS(file="z3c_3/global.rds")
+  z3d <- readRDS(file="z3d_3/global.rds")
+}
+
+REMAKE=FALSE
+if(REMAKE){
[…]

### 2022-05-31 z3/z4a.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z4"
+old_version <- "z3" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 1
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z4b_3/z4b.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z4"
+old_version <- "z3" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 1
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z4c_3/z4c.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z4"
+old_version <- "z3" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 1
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z4d.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z4"
+old_version <- "z3" # used to read in starting values when continuation=TRUE
+run_version <- "d"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 1
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z5-analysis.r

@@ -1,0 +1,244 @@
+
+library(pomp)
+loglik_he <- sum(read.csv("../data/he10mle.csv")$loglik)
+# -40345.7
+
+# global.rds is ordered by starting value enumeration
+# global.csv is ordered by likelihood
+LIK_ORDER <- TRUE 
+if(LIK_ORDER){
+  z1a <- read.csv(file="z1a_3/global.csv")
+  z1b <- read.csv(file="z1b_3/global.csv")
+  z1c <- read.csv(file="z1c_3/global.csv")
+  z1d <- read.csv(file="z1d_3/global.csv")
+  z2a <- read.csv(file="z2a_3/global.csv")
+  z2b <- read.csv(file="z2b_3/global.csv")
+  z2c <- read.csv(file="z2c_3/global.csv")
+  z2d <- read.csv(file="z2d_3/global.csv")
+  z3a <- read.csv(file="z3a_3/global.csv")
+  z3b <- read.csv(file="z3b_3/global.csv")
+  z3c <- read.csv(file="z3c_3/global.csv")
+  z3d <- read.csv(file="z3d_3/global.csv")
+  z4a <- read.csv(file="z4a_3/global.csv")
+  z4b <- read.csv(file="z4b_3/global.csv")
+  z4c <- read.csv(file="z4c_3/global.csv")
+  z4d <- read.csv(file="z4d_3/global.csv")
+  z5a <- read.csv(file="z5a_3/global.csv")
+  z5b <- read.csv(file="z5b_3/global.csv")
+  z5c <- read.csv(file="z5c_3/global.csv")
+  z5d <- read.csv(file="z5d_3/global.csv")
+} else {
+  z1a <- readRDS(file="z1a_3/global.rds")
+  z1b <- readRDS(file="z1b_3/global.rds")
+  z1c <- readRDS(file="z1c_3/global.rds")
+  z1d <- readRDS(file="z1d_3/global.rds")
+  z2a <- readRDS(file="z2a_3/global.rds")
+  z2b <- readRDS(file="z2b_3/global.rds")
+  z2c <- readRDS(file="z2c_3/global.rds")
+  z2d <- readRDS(file="z3d_3/global.rds")
+  z3a <- readRDS(file="z3a_3/global.rds")
[…]

### 2022-05-31 z3/z5a_3/z5a.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z5"
+old_version <- "z4" # used to read in starting values when continuation=TRUE
+run_version <- "a"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 2
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z5b_3/z5b.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z5"
+old_version <- "z4" # used to read in starting values when continuation=TRUE
+run_version <- "b"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 2
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z5c.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z5"
+old_version <- "z4" # used to read in starting values when continuation=TRUE
+run_version <- "c"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 2
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-05-31 z3/z5d_3/z5d.r

@@ -1,0 +1,373 @@
+# z3 is an attempt at maximizing likelihood for data & model of he10
+# z3 modifies z2 by restarting the top 25% of searches rather than 50%
+
+devel_version <- "z5"
+old_version <- "z4" # used to read in starting values when continuation=TRUE
+run_version <- "d"
+code_version <- paste0(devel_version,run_version)
+run_level <- 3
+seed_offset <- 2
+
+
+code_file <- paste0(code_version,".r")
+out_dir <- paste0(code_version,"_",run_level,"/")
+if(!dir.exists(out_dir)) dir.create(out_dir)
+
+# estimated time on greatlakes at run level 3: 24hr with 4000 particles
+
+# do_bake = FALSE manually avoids unintentional recomputation, e.g.,
+# when editing on a different machine.
+# do_bake <- FALSE 
+do_bake <- TRUE
+
+continuation <- TRUE # set to be false when starting a new search
+if(continuation){
+  old_code_version <- paste0(old_version,run_version)
+  old_out_dir <- paste0(old_code_version,"_",run_level,"/")
+  old_results <- read.csv(file=paste0(old_out_dir,"global.csv"))
+}
+
+sim_data <- FALSE # fit simulated data rather than the actual data
+
+estimated <- switch(run_version,
+  a="mostly shared",
+  b="all unit-specific",
+  c="he10",
+  d="mostly shared"
+) # he10 has all unit-specific with immigration but not coupling
+
+rw_r_vec <- c(0.02,0.01,0.005, 0.0025, 0.00125)
+rw_ivp_vec <- c(0.05,0.025,0.01, 0.005, 0.0025)
+# set rw.sd smaller for alpha since it has uncertainty +/- 5 percent
+# whereas +/- 50 percent is more generally the case.
+rw_alpha_frac <- 0.1
+
+# rate of autoregression of means (zero for uncoupled)
+spat_reg <- switch(run_version,
+  "a" = 0.1,
+  "b" = 0.1,
+  "c" = 0.1,
+  "d" = 0.001
+)
+
+experiment <- 3
+rw_r <- rw_r_vec[experiment]
+rw_ivp <- rw_ivp_vec[experiment]
+rw_alpha <- rw_r * rw_alpha_frac
+
+print(paste0("code version =",code_version))
+print(paste0("run_level = ",run_level))
[…]

### 2022-06-01 ms.Rnw
ready for submission

@@
 
-\renewcommand{\baselinestretch}{1.5}
+\renewcommand{\baselinestretch}{1.45}
 
@@
 \vspace{.4cm} 
-\centerline{E. L. Ionides, J. Wheeler and N. Ning}
+\centerline{Edward. L. Ionides, Ning Ning and Jesse Wheeler}
 \vspace{.4cm} 
-\centerline{University of Michigan, department of Statistics,}
+\centerline{University of Michigan, Department of Statistics,}
 
@@
 Population models in ecology and epidemiology can be characterized by high levels of stochasticity, nonlinearity, measurement error, and model uncertainty, leading to challenges of a somewhat different nature than geophysical models.
-In addition, biological population systems may sometimes have low population due to local introductions or fade-outs of one or more constituent species.
+In addition, biological population systems may sometimes have a low population count due to a local introduction or fade-out of one or more constituent species.
 Low population situations may require consideration of integer-valued counts rather than continuous population approximations.
@@
 
-Statistical inference for partially observed nonlinear biological systems was an open methodological challenge until recently even in the time series case \citep{bjornstad01}.
+Statistical inference for partially observed nonlinear biological systems was, until recently, an open methodological challenge even in the time series case \citep{bjornstad01}.
 Advances in Monte Carlo methods based on the particle filter have made inference accessible on many low-dimensional problems, but the curse of dimensionality \citep{bengtsson08} prevents applicability of the basic particle filter on metapopulations with more than a few units.
@@
 Situations where previous Monte Carlo methods can provably beat the curse of dimensionality may have limited applicability \citep{beskos17,park20,ionides21}.
-Consequently, state-of-the-art analysis has depended on ensemble Kalman filter (EnKF) methods \citep{li20}.
+Consequently, state-of-the-art scientific analysis for metapopulation dynamics has depended on ensemble Kalman filter (EnKF) methods \citep{li20}.
 EnKF algorithms scale well but are founded on an approximation that can be unsuitable for discrete populations with fade-out and re-introduction dynamics and other highly non-Gaussian features \citep{ionides21}.
@@
 IBPF extends the block particle filter (BPF) of \citet{rebeschini15} to carry out maximum likelihood parameter estimation.
-A previous IBPF algorithm was developed by \citet{ning21-ibpf} for the particular case that all estimated parameters are unit-specific, which is to say that the dynamics and measurement process for a unit $u$ are determined by a set of parameters $\psi_u$ specific to unit $u$.
+A previous IBPF algorithm was developed by \citet{ning21-ibpf} for the particular case that all estimated parameters are unit-specific, which is to say that the dynamics and measurement process for a unit $u$ are determined by a vector of parameters $\psi_u$ specific to unit $u$.
 The formal meaning of this assertion is given, together with the pseudocode for our algorithm, in Sec.~\ref{sec:alg}.
 We propose an extension of their IBPF which additionally permits the estimation of a vector of shared parameters, $\phi$.
-In this case, the full parameter vector is $\theta=(\phi,\psi_{1:U})$ where the $U$ units are labeled $\{1,\dots,\Unit\}$ which we denote by $\seq{1}{\Unit}$.
+In this case, the full parameter vector is $\theta=(\phi,\psi_{1:U})$ where the $U$ units are named $\{1,\dots,\Unit\}$ which we denote by $\seq{1}{\Unit}$.
 \citet{ning21-ibpf} developed theoretical justification for their algorithm, but our extension of IBPF currently carries only empirical support.
@@
 Another relevant possibility is that a parameter may differ between units as a shared function of unit-specific covariates, which formally is a special case of a shared parameter.
-Addressing these issues is also prerequisite for studying questions about coupling of metapopulation systems via model-based inference from spatiotemporal data.
+Addressing these issues is also a prerequisite for studying questions about the coupling of metapopulation systems via model-based inference from spatiotemporal data.
 We will show empirically, in Sec.~\ref{sec:data}, that our {\ibpf} algorithm is applicable to an inference challenge in epidemiological metapopulation dynamics.
@@
 There are intermediate possibilities, where a parameter shared for only a subset of all units, but such situations are not explored further here.
-The special case where all parameters are unit-specific may arise sometimes, but models typically have some shared parameters which arise in transition densities and/or measurement densities for multiple units.
+The special case where all parameters are unit-specific may occur sometimes, but models typically have some shared parameters which arise in transition densities and/or measurement densities for multiple units.
 
-Our approach to iterated filtering for shared parameters involves constructing an extended model within which the a shared parameters is represented as a unit-specific parameter that happens to be constant across units.
-We construct a spatiotemporally extended model by supposing a numerical solution for the transition from time $t$ to time $t+\delta$ for each unit $u$ with the form
+Our approach to iterated filtering for shared parameters involves constructing an extended model within which the shared parameters is represented as a unit-specific parameter that happens to be constant across units.
+We construct a spatiotemporally extended model by supposing that a numerical solution for the transition from time $t$ to time $t+\delta$ for each unit $u$ has a functional form
 \begin{equation}
 \label{eq:extension}
-  X_{\unit}(t+\delta) = X_\unit(t) + Q_u(\myvec{X}(t),\myvec{\eta}_t,\phi,\psi_{\unit},t,\delta),
+  X_{\unit}(t+\delta) = X_\unit(t) + Q_u
+    \big(\myvec{X}(t),\myvec{\eta}_t,\phi,\psi_{\unit},t,\delta \big),
 \end{equation}
 where the random vector $\myvec{\eta}_t=\eta_{1:\Unit,t}$ is shared for all $u\in \seq{1}{U}$ and does not depend on $\theta$.
-If a representation \ref{eq:extension} exists, an extended model is defined by replacing $\phi$ with $\phi_{\unit}(t)$ and $\psi_u$ with $\psi_{u}(t)$.
-Equation \ref{eq:extension} implicitly defines a continuous-time extended model by the limit as $\delta\to 0$, when that limit exists, but for simulation-based methods the numerical solution is of more immediate concern than this limit.
+If a representation \eqref{eq:extension} exists, an extended model is defined by replacing $\phi$ with $\phi_{\unit}(t)$ and $\psi_u$ with $\psi_{u}(t)$.
+Equation \eqref{eq:extension} implicitly defines a continuous-time extended model by the limit as $\delta\to 0$, when that limit exists, but for simulation-based methods the numerical solution is of more immediate concern than this limit.
 Admitting a minor abuse of notation, we subsequently use a density $f$ to denote both the original model and its extension for spatiotemporally varying parameters, with the context determining which one is intended.
@@
 $\sigma_\time$ is a $D\times D$ diagonal matrix with entries $\sigma_{d,\time}$.
+The blocks, $\blocklist_{1:\Block}$, are a partition of $\seq{1}{\Unit}$.
 
-The pseudocode in Alg.~\ref{alg:ibpfilter} represents our implementation of {\ibpf} as the R function \code{ibpf} in the open-source package \code{spatPomp} \citep{asfaw21github,asfaw21arxiv}.
+The pseudocode in Alg.~\ref{alg:ibpfilter} represents our implementation of {\ibpf} as the R function \code{ibpf} extending the open-source package \code{spatPomp} \citep{asfaw21github,asfaw21arxiv}.
 We will contribute \code{ibpf} to \code{spatPomp}.
-Aadditionally, the source code for all the results in this article is available at \url{https://github.com:ionides/ibpf_article}.
+Additionally, the source code for all the results in this article is available at \url{https://github.com/ionides/ibpf_article}.
 Various generalizations of this implementation are possible.
@@
 Bearing all this in mind, measles provides a natural testing ground for our new methodologies.
-We demonstrate that we now have the tools to carry out likelihood maximization and profile likelihood confidence interval construction on mechanistic statistical models that are appropriate for spatiotemporal metapopulation disease data.
+We demonstrate that we now have the tools to carry out likelihood maximization (and therefore, in principle, profile likelihood confidence interval construction and likelihood-based model selection) on mechanistic statistical models that are appropriate for spatiotemporal metapopulation disease data.
 
 We specifically set ourselves the task of fitting a spatiotemporal model to the case reports for 20 towns studied by \citet{he10}.
-We require ability to handle discrete case counts which vary from zero to thousands of cases per week.
+We require the ability to handle discrete case counts which vary from zero to thousands of cases per week.
 We consider a model with the same structure as \citet{he10}, namely a Markov chain with Gamma noise on the infection rate, but with the addition of a term for transmission between cities.
@@
 
-<<data-plot, echo=F, fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Weekly measles case reports for 20 UK towns.",fig.show = "hold">>=
+<<data-plot, echo=F, fig.height=6.5, fig.width=6.5, out.width="5.5in", fig.cap = "Weekly measles case reports for 20 UK towns.",fig.show = "hold">>=
 library(spatPomp)
@@
 
-<<sim-plot, echo=F, fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Simulated weekly measles case reports.",fig.show = "hold">>=
+<<sim-plot, echo=F, fig.height=6.5, fig.width=6.5, out.width="5.5in", fig.cap = "Simulated weekly measles case reports.",fig.show = "hold">>=
 library(spatPomp)
@@
 Parameters for the simulated model were based on the analysis of individual towns by \citet{he10} and adjusted to provide a visual match in the simulations.
-A goal of the simulation study is to obtain appropriate algorithmic choices for the data analysis, and to develop methodology which can be successful when the truth is known before applying it to data.
+A goal of the simulation study is to obtain appropriate algorithmic choices for the data analysis.
+We seek to develop methodology which is demonstrably successful when the truth is known, before applying it to data.
 We present the simulation study from this perspective, to describe the process of developing our data analysis, though we acknowledge one could revisit the simulation study based on the data analysis in Sec.~\ref{sec:data} using maximum likelihood estimates of the parameters.
 
-A key data analysis question is to decide which parameters should be shared between units and which should be unit-specific, taking distinct values on each unit.
-For a general SpatPOMP parameter, it is not necessarily well-defined how to extend the model so that each unit has a distinct parameter value, however our measles model falls into the structure defined in \eqref{eq:extension} and we suspect that similar considerations may arise in other situations.
+A key data analysis question is to decide which parameters should be shared between units and which should be unit-specific.
+For a general SpatPOMP parameter, it is not necessarily well-defined how to extend the model so that each unit has a distinct parameter value, however our measles model falls into the structure defined in \eqref{eq:extension} and we suspect that similar considerations may arise in other applications.
 Different numerical issues may arise as the number of unit-specific parameters increases.
@@
 In the presence of weak identifiability, small differences in likelihood could lead to large differences in parameter estimates.
-In such sitatuations, a scientist may choose to investigate what functions of the parameters can be inferred accurately without adding additional assumptions, or alternatively they may choose to investigate the consequences of placing constraints on some parameters to improve identifiability of the remainder.
+In such situations, a scientist may choose to investigate what functions of the parameters can be inferred accurately without adding additional assumptions, or alternatively they may choose to investigate the consequences of placing constraints on some parameters to improve identifiability of the remainder.
 Likelihood maximization permits such investigations, but they are beyond the scope of this article.
+Parameter values used for the simulation study and parameter values obtained by the likelihood maximization searches are provided in the supplementary code repository, for the interested reader.
 
@@
 That analysis used a customized treatment for the unit-specific initial value parameters and did not attempt to estimate other unit-specific parameters.
-IGIRF is sensitive to the choice of the guide function, and the model-specific implementation used by \citet{park20} performed better than the generic implementation of IGIRF currently available in \code{spatPomp}.
+IGIRF is sensitive to the choice of the guide function, and the model-specific implementation used by \citet{park20} performed better than the generic implementation of IGIRF which is currently available in \code{spatPomp}.
 To our knowledge, {\ibpf} is therefore advancing the current limits for scalability of likelihood-based inference for metapopulation dynamics.
@@
 Wilks's theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+9)/2=44.5$ for A and $(13\times 20)/2=130$ for B.
-We find log-likelihoods exceeding the truth by less than this, indicating some imperfection to the Monte Carlo maximization so far as Wilks's asymptotic result holds in this situation. 
+The $y$-axis values in Fig.~\ref{fig:sim_search_boxplot} show log-likelihoods exceeding the truth by less than this, indicating some imperfection to the Monte Carlo maximization so far as Wilks's asymptotic result holds in this situation.
 Despite this limitation, searches that exceed the likelihood at the truth have found inferentially plausible sets of parameters which can be used to study the likelihood surface around the maximum.
@@
 Final log-likelihood values are plotted against starting log-likelihood for an initial search (top row), a refinement of this search (second row) and a second refinement (third row).
-Starting values for each refinement were 4 copies of the top 25\\% of end points for the previous search.
 Unit-specific parameters were $\\rho$, $S_0$, $E_0$, $I_0$ (for the simulation study, model A) and all parameters (for the data, model B).">>=
@@
 The data we consider here are weekly case reports for twenty towns, identical to those studied by \citet{he10}.
-Some previous analyses have used counts aggregated over two week intervals \citep{park20,ionides21} since these were available for more cities, but our goal here is to extend the analysis of \citet{he10}.
+Some previous analyses have used counts aggregated over two-week intervals \citep{park20,ionides21} since these were available for more cities, but our goal here is to extend the analysis of \citet{he10}.
 Apart from this, our model matches \citet{ionides21} and for completeness we repeat the description here.
@@
 The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{\bullet D,\unit}(t)$ are modeled as conditional Poisson process with per-capita rates $\mu_{EI}$, $\mu_{IR}$ and $\mu_{\bullet D}$ respectively, and we fix $\mu_{\bullet D}=(50 \mbox{ year})^{-1}$.
-The birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson processes with rate $\mu_{BS,\unit}(t)$, given by interpolated census data.
+The birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson process with rate $\mu_{BS,\unit}(t)$, given by interpolated census data.
 
@@
 BPF with each town forming a separate block, applied to an uncoupled model such as C, is equivalent to carrying out independent particle filters for each unit.
-Indeed, if the published parameter values from \citet{he10} are inserted into the \code{spatPomp} object for model C, then the \code{bpfilter()} function can be used to carry out BPF, which retrieves $\ell_{\mathrm He}$, up to Monte Carlo error.
+Indeed, if the published parameter values from \citet{he10} are inserted into the \code{spatPomp} object for model C, then the \code{bpfilter} function can be used to carry out BPF, which retrieves $\ell_{\mathrm He}$, up to Monte Carlo error.
 Each of models A, B and C in Fig.~\ref{fig:data_search_boxplot} shows shortfall relative to $\ell_{\mathrm He}$.
@@
 
-Theoretical interest in BPF was inspired by \citet{rebeschini15} who showed mathematically that BPF can, under particular conditions, enjoy linear scaling properties.
+Theoretical interest in BPF was inspired by \citet{rebeschini15}, who showed mathematically that BPF can enjoy linear scaling properties under suitable conditions.
 \citet{rebeschini15} expressed pessimism about the practical applicability of their algorithm, which may help to explain why the practical use of this algorithm has been limited.
@@
 In particular, we do not expect applicability to situations where it is necessary to consider $10^3$ or more spatial units.
-In such situations, it may be necessary to make additional approximations such as those inherent in the ensemble Kalman filter \citep{evensen09,katzfuss19} or other numerical filtering techniques \citep{whitehouse22}.
+In such situations, it may be necessary to make additional approximations such as those inherent in the ensemble Kalman filter \citep{evensen09book,katzfuss19} or other numerical filtering techniques \citep{whitehouse22}.
 However, we anticipate that many practical metapopulation analyses can be addressed within the scope we have demonstrated.
@@
 
-\noindent The data we analyze are in the public domain, and are available together with source code at \url{https://github.com:ionides/ibpf_article}.
+\noindent The data we analyze are in the public domain, and are available together with source code at \url{https://github.com/ionides/ibpf_article}.
 
@@
 
-\noindent This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108, and National Institutes of Health grant 1-R01-AI143852.
+\noindent This work was supported by National Science Foundation grant DMS-1761603 and DMS 1646108.
 We acknowledge an anonymous referee from a previous article who recommended that we should try implementing a block particle filter.

### 2022-06-01 ms.tex
ready for submission

@@
 
-\includegraphics[width=6.5in]{figure/data-plot-1} \hfill{}
+\includegraphics[width=5.5in]{figure/data-plot-1} \hfill{}
 
@@
 
-\includegraphics[width=6.5in]{figure/sim-plot-1} \hfill{}
+\includegraphics[width=5.5in]{figure/sim-plot-1} \hfill{}
 
@@
 Final log-likelihood values are plotted against starting log-likelihood for an initial search (top row), a refinement of this search (second row) and a second refinement (third row).
-Starting values for each refinement were 4 copies of the top 25\% of end points for the previous search.
 Unit-specific parameters were $\rho$, $S_0$, $E_0$, $I_0$ (for the simulation study, model A) and all parameters (for the data, model B).}\label{fig:search_diagnostics}
@@
 The log-likelihood was obtained using $M=100$ iterations of IBPF starting at random parameters (median starting log-likelihood, -40800).
-The horizontal dashed line denotes the log-likelihood at the true parameters}\label{fig:spatreg_boxplot}
+The horizontal dashed line denotes the log-likelihood at the true parameters.}\label{fig:spatreg_boxplot}
 \end{figure}

### 2022-06-01 ms.Rnw
a few minor changes

@@
 \vspace{.4cm} 
-\centerline{University of Michigan, Department of Statistics,}
+\centerline{University of Michigan, Department of Statistics}
 
@@
 IGIRF is sensitive to the choice of the guide function, and the model-specific implementation used by \citet{park20} performed better than the generic implementation of IGIRF which is currently available in \code{spatPomp}.
-To our knowledge, {\ibpf} is therefore advancing the current limits for scalability of likelihood-based inference for metapopulation dynamics.
+To our knowledge, {\ibpf} is therefore advancing the current limits for scalability of simulation-based maximum-likelihood inference for metapopulation dynamics.
 
@@
 Apart from this, our model matches \citet{ionides21} and for completeness we repeat the description here.
-We compartmentalize the population of each city into susceptible ($S$), exposed ($E$), infectious ($I$), and recovered/removed ($R$) categories.
-The numbers of individuals in each compartment for city $\unit$ at time $t$ are denoted by integer-valued random variables $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$.
-The population dynamics are written in terms of counting processes $N_{AB,\unit}(t)$ enumerating cumulative transitions in city $\unit$, up to time $t$, between compartments $A$ and $B$.
+We compartmentalize the population of each town into susceptible ($S$), exposed ($E$), infectious ($I$), and recovered/removed ($R$) categories.
+The numbers of individuals in each compartment for town $\unit$ at time $t$ are denoted by integer-valued random variables $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$.
+The population dynamics are written in terms of counting processes $N_{AB,\unit}(t)$ enumerating cumulative transitions in town $\unit$, up to time $t$, between compartments $A$ and $B$.
 Here, $A,B\in \{S,E,I,R,B,D\}$ with $B$ denoting a source compartment for immigration or birth, and $D$ a sink compartment for emigration or death.
-We enumerate the $U=20$ cities studied by \citet{he10} in decreasing size, so that $\unit=1$ corresponds to London.
+We enumerate the $U=20$ towns studied by \citet{he10} in decreasing size, so that $\unit=1$ corresponds to London.
 Our model is described by the following system of stochastic differential equations, for $\unit=1,\dots, \Unit$,
@@
 The total population $P_\unit(t)=S_\unit(t)+E_\unit(t)+I_\unit(t)+R_\unit(t)$ is calculated by smoothing census data and is treated as known.
-The number of recovered individuals $R_\unit(t)$ in city $\unit$ is therefore defined implicitly.
+The number of recovered individuals $R_\unit(t)$ in town $\unit$ is therefore defined implicitly.
 $N_{SE,\unit}(t)$ is modeled as negative binomial death processes \citep{breto09,breto11}
@@
 with $p = 0.759$ being the proportion of the year taken up by the school terms, $\meanBeta$ is the mean transmission rate, and $\amplitude$ measures the reduction of transmission during school holidays.
-In \eqref{eq:dEdt}, $\alpha$ is a mixing exponent modeling inhomogeneous contact rates within a city, and $\iota$ models immigration of infected individuals which is appropriate when analyzing a subset of cities that cannot be treated as a closed system.
-The number of travelers from city $\unit$ to $\altUnit$ is denoted by $v_{\unit\altUnit}$. 
+In \eqref{eq:dEdt}, $\alpha$ is a mixing exponent modeling inhomogeneous contact rates within a town, and $\iota$ models immigration of infected individuals which is appropriate when analyzing a subset of cities that cannot be treated as a closed system.
+The number of travelers from town $\unit$ to $\altUnit$ is denoted by $v_{\unit\altUnit}$. 
 Here, $v_{\unit\altUnit}$ is constructed using the gravity model of \cite{xia04}, 
@@
 \]
-where $\dist(\unit,\altUnit)$ denotes the distance between city $\unit$ and city $\altUnit$, $P_\unit$ is the average population for city $\unit$ across time, $\bar{P}$ is the average population across cities, and $\overline{\dist}$ is the average distance between a randomly chosen pair of cities.
-Here, we model $v_{\unit\altUnit}$ as fixed through time and symmetric between any two arbitrary cities, though a natural extension would allow for temporal variation and asymmetric movement between two cities.
+where $\dist(\unit,\altUnit)$ denotes the distance between town $\unit$ and town $\altUnit$, $P_\unit$ is the average population for town $\unit$ across time, $\bar{P}$ is the average population across towns, and $\overline{\dist}$ is the average distance between a randomly chosen pair of towns.
+Here, we model $v_{\unit\altUnit}$ as fixed through time and symmetric between any two arbitrary towns, though a natural extension would allow for temporal variation and asymmetric movement between two towns.
 The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{\bullet D,\unit}(t)$ are modeled as conditional Poisson process with per-capita rates $\mu_{EI}$, $\mu_{IR}$ and $\mu_{\bullet D}$ respectively, and we fix $\mu_{\bullet D}=(50 \mbox{ year})^{-1}$.
@@
 
-\noindent This work was supported by National Science Foundation grant DMS-1761603 and DMS 1646108.
+\noindent This work was supported by National Science Foundation grants DMS-1761603 and DMS 1646108.
 We acknowledge an anonymous referee from a previous article who recommended that we should try implementing a block particle filter.

### 2022-11-04 ms.Rnw
accepted version at Statistica Sinica

@@
+\documentclass[11pt]{article}
+\usepackage{fullpage}
+
+%\documentclass[12pt]{article}
 %\input{stat-sinica-header.tex}
 
-\documentclass[12pt]{article}
+\usepackage[sectionbib]{natbib}
 
-\setlength{\abovecaptionskip}{-2mm}
-
-%%%%%%% START OF STAT SINICA SETTINGS %%%%%%%%%%
-
-\textwidth=31.9pc
-\textheight=46.5pc
-\oddsidemargin=1pc
-\evensidemargin=1pc
-\headsep=15pt
-\topmargin=.6cm
-\parindent=1.7pc
-\parskip=0pt
-
-\usepackage[sectionbib]{natbib}
-\usepackage{array,epsfig,fancyheadings,rotating}
-%\usepackage[]{hyperref}  %<----modified by Ivan
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage{sectsty, secdot}
-\sectionfont{\fontsize{12}{15}\selectfont}
-\sectionfont{\fontsize{12}{14pt plus.8pt minus .6pt}\selectfont}
-\renewcommand{\theequation}{\thesection\arabic{equation}}
-\subsectionfont{\fontsize{12}{14pt plus.8pt minus .6pt}\selectfont}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\pagestyle{fancy}
-\def\n{\noindent}
-\lhead[\fancyplain{} \leftmark]{}
-\chead[]{}
-\rhead[]{\fancyplain{}\rightmark}
-\cfoot{}
-
-%%%%%% END OF STAT SINICA SETTINGS %%%%%%%%5
+%%% Statistica Sinica formatting
+%\newcommand\statSinica[1]{#1}
+ \newcommand\statSinica[1]{}
+%%% arXiv formatting
+ \newcommand\arxiv[1]{#1}
+%\newcommand\arxiv[1]{}
 
@@
 \usepackage{amsmath,amssymb}
-%\usepackage{graphicx,psfrag,epsf}
 \usepackage{graphicx}
@@
 \usepackage{url} % not crucial - just used below for the URL
-
-%\newcommand\jpc[1]{{\textcolor{orange}{#1}}}
-\newcommand\aak[1]{{\textcolor{blue}{#1}}}
 
@@
 
-\renewcommand{\baselinestretch}{1.45}
+\statSinica{\statSinicaSetup}
 
-\markright{ \hbox{\footnotesize\rm Statistica Sinica
-%{\footnotesize\bf 24} (201?), 000-000
-}\hfill\\[-13pt]
-\hbox{\footnotesize\rm
-%\href{http://dx.doi.org/10.5705/ss.20??.???}{doi:http://dx.doi.org/10.5705/ss.20??.???}
-}\hfill }
+\centerline{\statSinica{\large}\arxiv{\Large}\bf An iterated block particle filter for inference on coupled dynamic}
 
-\markboth{\hfill{\footnotesize\rm E. L. Ionides et al.} \hfill}
-{\hfill {\footnotesize\rm An iterated block particle filter} \hfill}
+\arxiv{\vspace{1mm}}
 
-\renewcommand{\thefootnote}{}
-$\ $\par
+\centerline{\statSinica{\large}\arxiv{\Large}\bf systems with shared and unit-specific parameters}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\arxiv{\vspace{2mm}}
 
-\fontsize{12}{14pt plus.8pt minus .6pt}\selectfont \vspace{0.8pc}
-\centerline{\large\bf An iterated block particle filter for inference on coupled dynamic}
-\vspace{2pt} 
-\centerline{\large\bf systems with shared and unit-specific parameters}
-\vspace{.4cm} 
-\centerline{Edward. L. Ionides, Ning Ning and Jesse Wheeler}
-\vspace{.4cm} 
-\centerline{University of Michigan, Department of Statistics}
+\centerline{Edward L. Ionides$^1$, Ning Ning$^2$ and Jesse Wheeler$^1$}
 
- \vspace{.55cm} \fontsize{9}{11.5pt plus.8pt minus.6pt}\selectfont
+\arxiv{\vspace{1mm}}
 
-\centerline{Submitted to {\it Statistica Sinica} special edition on}
-\centerline{Recent developments of sequential Monte Carlo methods}
-\centerline{and their applications}
+{\small
+\centerline{\small University of Michigan, Department of Statistics$^1$}
+\centerline{\small Texas A\&M University, Department of Statistics$^2$}
+}
+ \vspace{.25cm} \fontsize{9}{11.5pt plus.8pt minus.6pt}\selectfont
+
+\statSinica{\statSinicaSubmission}
 
@@
 Iterated filtering algorithms can facilitate likelihood maximization for simulation-based filters.
-We introduce a new iterated block particle filter algorithm applicable when parameters are unit-specific or shared between units.
+We introduce an iterated block particle filter applicable when parameters are unit-specific or shared between units.
 We demonstrate this algorithm by performing inference on a coupled epidemiological model describing spatiotemporal measles case report data for twenty towns.
 
-\vspace{9pt}
-\noindent {\it Key words and phrases:}
-sequential Monte Carlo; metapopulation; spatiotemporal; maximum likelihood estimation; partially observed Markov process.
-\par
+\statSinica{\statSinicaKeywords}
+
 \end{quotation}\par
@@
 
+\statSinica{
   \fontsize{12}{14pt plus.8pt minus .6pt}\selectfont
-
 %%  \def\spacingset#1{\renewcommand{\baselinestretch}%
 %%  {#1}\small\normalsize} \spacingset{1}
+}
 
 \date{}
-
-%% \spacingset{1.45}
 
@@
 Statistical inference for partially observed nonlinear biological systems was, until recently, an open methodological challenge even in the time series case \citep{bjornstad01}.
-Advances in Monte Carlo methods based on the particle filter have made inference accessible on many low-dimensional problems, but the curse of dimensionality \citep{bengtsson08} prevents applicability of the basic particle filter on metapopulations with more than a few units.
+Advances in Monte Carlo methods based on the particle filter have made inference accessible on many low-dimensional problems \citep{doucet11,kantas15,king16} but the curse of dimensionality \citep{bengtsson08} prevents applicability of the basic particle filter on metapopulations with more than a few units.
 Methods based on improving the proposal distribution for the particle filter may not fully resolve the curse of dimensionality \citep{snyder15}.
@@
 In Sec.~\ref{sec:alg}, we propose an algorithm for inference on metapopulation dynamics which we call an iterated block particle filter (IBPF).
-IBPF extends the block particle filter (BPF) of \citet{rebeschini15} to carry out maximum likelihood parameter estimation.
+IBPF combines the iterated filtering likelihood maximization technique of \citet{ionides15} with the block particle filter (BFP) of \citet{rebeschini15}.
+Iterated filtering algorithms use parameter perturbations to coerce a filtering algorithm into exploring the parameter space.
+BPF addresses the curse of dimensionality by modifying the resampling step of a particle filter to resample independently on blocks that form a partition of the collection of units.
 A previous IBPF algorithm was developed by \citet{ning21-ibpf} for the particular case that all estimated parameters are unit-specific, which is to say that the dynamics and measurement process for a unit $u$ are determined by a vector of parameters $\psi_u$ specific to unit $u$.
@@
 Measles case reports have been a longstanding benchmark problem for inference on biological dynamics, motivating the development of time series methodology and, more recently, the progression from single populations to metapopulation systems.
-Unlike previous attempts on sequential Monte Carlo inference for metapopulation models, we show that our algorithm can provide practical inference when parameters are either shared between units or differ between units.
+Unlike previous attempts on sequential Monte Carlo inference for metapopulation models, we show that our algorithm can provide practical plug-and-play, likelihood-based inference when parameters are either shared between units or differ between units.
+Demonstrating a solution to this open problem provides numerical evidence substituting for numerical comparisons against alternative methods.
+
 Our data analysis results do not fully resolve open questions about what models for coupling between towns are supported by the data, and which parameters should be modeled as unit-specific.
@@
 We suppose the existence of a joint density $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$  for $X_{1:\Unit,0:\Time}$ and $Y_{1:\Unit,1:\Time}$ with respect to some appropriate measure, following a notational convention that the subscripts of $f$ denote the joint or conditional density under consideration.
-We suppose that $f$ depends on a parameter vector $\theta=(\phi,\psi_{1:U})$.
+We suppose that $f$ depends on a real-valued parameter vector $\theta=(\phi,\psi_{1:U})$, which we write as $\theta=\theta_{1:D}$ when we wish to concatenate the component shared and unit-specific parameters into a single vector of length $D$.
 The data are $\data{y}_{\unit\comma\time}$ for unit $\unit$ at time $\time$.
@@
 Numerical evaluation of the likelihood function is closely related to the filtering problem of evaluating $f_{\myvec{X}_n|\myvec{Y}_{1:n}}(\myvec{x}_n\given \data{\myvec{y}}_{1:n}\giventh \theta)$.
-Including parameters into the latent space of a dynamic model can coerce a filter into calculating a Bayesian posterior distribution, though regularization is required to make the calculation numerically tractable using Monte Carlo methods \citep{kitagawa98,janeliu01}.
+If the dynamic model is extended to include the parameters as latent variables, the filtering problem leads to the Bayesian posterior distribution, though regularization is required to make the calculation numerically tractable using Monte Carlo methods \citep{kitagawa98,janeliu01}.
 Iterating this Bayesian calculation recursively targets a maximum likelihood estimate (MLE), a strategy known as data cloning \citep{lele07,lele10}.
-Adding noise to perturb the parameters at each time point can stabilize the numerics while retaining the capability to approximate the MLE \citep{ionides06-pnas,ionides11,ionides15}.
+Adding noise to perturb the parameters in the extended model at each time point can stabilize the numerics while retaining the capability to approximate the MLE \citep{ionides06-pnas,ionides11,ionides15}.
 Many variations on this idea have been developed using different filter methods \citep{park20,li20,ionides21,manoli15} or employing different perturbation systems \citep{doucet13,nguyen17}.
 
-Previous approaches to iterated perturbed particle filtering are not immediately applicable to BPF.
-In the special case where each parameter is localized to an individual unit (i.e., all parameters are unit-specific) one can resample perturbed parameters at the block level and obtain a theoretical guarantee of approximating the MLE \citep{ning21-ibpf}.
+A direct approach to iterating BPF for parameter estimation is to resample the extended model independently on each block, giving rise to separate collections of parameters for each block.
+\citet{ning21-ibpf} proved that this IBPF algorithm targets the MLE for the special case where each parameter is localized to an individual unit, i.e., when all parameters are unit-specific.
 Formally, we say that a parameter for a discrete-time SpatPOMP is specific to a unit $u$ if it is involved in specifying the measurement density $f_{Y_{u,n}|X_{u,n}}$ or transition density $f_{X_{u,n+1}|\myvec{X}_n}$ for some $n$, and it is not involved in $f_{Y_{v,m}|X_{v,m}}$ or $f_{X_{v,m+1}|\myvec{X}_m}$ for any $v\neq u$ and any $m$.
@@
 A parameter which is not unit-specific is said to be shared between units.
-There are intermediate possibilities, where a parameter shared for only a subset of all units, but such situations are not explored further here.
+There are intermediate possibilities, where a parameter is shared for only a subset of all units, but such situations are not explored further here.
 The special case where all parameters are unit-specific may occur sometimes, but models typically have some shared parameters which arise in transition densities and/or measurement densities for multiple units.
@@
 
-The IBPF algorithm described in Alg.~\ref{alg:ibpfilter} carries out a block particle filter on this extended model.
-The extended parameters are given independent perturbations, but a spatial autoregressive step pulls the values of shared parameters toward their mean over the units.
-Indeed, this autoregressive step is the only difference between  Alg.~\ref{alg:ibpfilter} and the algorithm for unit-specific parameters proposed by \citet{ning21-ibpf}.
-This algorithm, in turn, is essentially the IF2 algorithm of \citet{ionides15} with the particle filter replaced by BPF.
-Alg.~\ref{alg:ibpfilter} assumes implicit loops over $\np\ \text{in}\ \seq{1}{\Np}$ and $\unit\ \text{in}\ \seq{1}{\Unit}$, 
-$\normal(\mu,\Sigma)$ denotes the normal distribution with mean $\mu$ and variance matrix $\Sigma$, and
-$\sigma_\time$ is a $D\times D$ diagonal matrix with entries $\sigma_{d,\time}$.
-The blocks, $\blocklist_{1:\Block}$, are a partition of $\seq{1}{\Unit}$.
-
-The pseudocode in Alg.~\ref{alg:ibpfilter} represents our implementation of {\ibpf} as the R function \code{ibpf} extending the open-source package \code{spatPomp} \citep{asfaw21github,asfaw21arxiv}.
-We will contribute \code{ibpf} to \code{spatPomp}.
-Additionally, the source code for all the results in this article is available at \url{https://github.com/ionides/ibpf_article}.
-Various generalizations of this implementation are possible.
-For example, iterated filtering theory does not rely on parameter perturbations following the normal distribution \citep{ionides15} though in practice we transform parameters to facilitate this convenient choice.
-Also, the pseudocode does not include parameter-dependent random walk variance, or a different perturbation schedule for initial value parameters, both of which could be included.
-
 %%%%%%% ibpf BBBBBBBBBBBBBBBBBBBBBBBBBBBBBBBBB %%%%%%%%%%%%%%%%%
@@
 \begin{algorithm}[H]
-   \caption{\textbf{IBPF}. \\
+\setstretch{1.35}
+  \caption{
+  \textbf{IBPF}. \newline
   {\bf Inputs}:
-   simulator for the extended model, $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1}\giventh\myvec{\theta})$;
+  simulator for the extended model, $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1}\giventh\myvec{\theta})$ and initialization, $f_{\myvec{X}_0}(\myvec{x}_0\giventh\myvec{\theta})$;
   evaluator for $f_{{Y}_{\unit,\time}|{X}_{\unit,\time}}({y}_{\unit,\time}\given {x}_{\unit,\time}\giventh\theta)$;
-    simulator for extended initialization, $f_{\myvec{X}_0}(\myvec{x}_0\giventh\myvec{\theta})$;
   data, $\data{\myvec{y}}_{1:\Time}$;
@@
   initial parameter swarm with decomposition into shared and unit-specific parameters, $\Theta^{0,\np}_{\unit}=(\Phi^{0,\np}_{\unit},\Psi^{0,\np}_{\unit})$;
-    random walk perturbation, $\sigma_{d,\time}$.
-%{\bf \code{spatPomp} implementation}:	
-%     \code{ibpf(P,\,
-%        params{\argequals}$\theta_0${\argequals}$(\phi_0,\psi_{1:\Unit,0})$,
-%        Nbpf{\argequals}$\Nit$,
-%        Np{\argequals}$\Np\!$,
-%        rw.sd{\argequals}$\sigma$,
-%        cooling.fraction.50{\argequals}$a$,
-%        spat.regression{\argequals}$\spatReg$,
-%        block\_list{\argequals}$\blocklist$\,)}	
-%    where \code{P} is a	
-%     \class{\spatPomp} object \citep{asfaw21arxiv}.
-%
+  random walk perturbation, $\sigma_{d,\time}$; cooling rate, $a$; number of iterations, $M$; spatial autoregression, $\spatReg$.
 }\label{alg:ibpfilter}
@@
 
+The IBPF algorithm described in Alg.~\ref{alg:ibpfilter} carries out a block particle filter on this extended model.
+The extended parameters are given independent perturbations, but a spatial autoregressive step pulls the values of shared parameters toward their mean over the units.
+Indeed, this autoregressive step is the only difference between  Alg.~\ref{alg:ibpfilter} and the algorithm for unit-specific parameters proposed by \citet{ning21-ibpf}.
+This algorithm, in turn, is essentially the IF2 algorithm of \citet{ionides15} with the particle filter replaced by BPF.
+
+Alg.~\ref{alg:ibpfilter} assumes implicit loops over $\np\ \text{in}\ \seq{1}{\Np}$ and $\unit\ \text{in}\ \seq{1}{\Unit}$,
+$\normal(\mu,\Sigma)$ denotes the normal distribution with mean $\mu$ and variance matrix $\Sigma$, and
+$\sigma_\time$ is a $D\times D$ diagonal matrix with entries $\sigma_{d,\time}$.
+The blocks, $\blocklist_{1:\Block}$, are a partition of $\seq{1}{\Unit}$.
+
+The pseudocode in Alg.~\ref{alg:ibpfilter} represents our implementation of {\ibpf} as the R function \code{ibpf} which we have contributed to the open-source package \code{spatPomp} \citep{asfaw21github,asfaw21arxiv}.
+Additionally, the source code for all the results in this article is available at \url{https://github.com/ionides/ibpf_article}.
+Various generalizations of this implementation are possible.
+For example, iterated filtering theory does not rely on parameter perturbations following the normal distribution \citep{ionides15} though in practice we transform parameters to facilitate this convenient choice (Sec.~\supSecAlg).
+
 \subsection{Algorithmic parameters}
@@
 
-We set the cooling rate parameter to be $a=0.5$, corresponding to a 1\% reduction in the random walk standard deviation at each iteration.
+We used $J=4000$ particles, and we set the cooling rate parameter to be $a=0.5$, corresponding to a 1\% reduction in the random walk standard deviation at each iteration.
+We set $M=100$ optimizations iterations, chosen as an empirically assessed compromise between effort spent on each search and the number of searches conducted.
+Additional discussion of algorithmic parameters is given in Sec.~\supSecAlg.
 The only additional algorithmic parameter over previous iterated filtering algorithms is the spatial autoregressive parameter, $\spatReg$.
-Numerical experimentation suggests that the performance is not sensitive to the choice of $\spatReg>0$ (see Fig~\ref{fig:spatreg_boxplot} in Appendix A).
+Numerical experimentation suggests that the performance is not sensitive to the choice of $\spatReg>0$ (Supplementary Fig.~\supFigSpatReg).
 Subsequently, we used $\spatReg=0.1$.
@@
 Before tackling these issues, we use simulated data to demonstrate the capabilities of the methodology on a correctly specified model.
-
-\section{Testing {\ibpf} on a measles transmission model: simulation results}
-\label{sec:sim}
-
-Measles transmission has provided a useful example of epidemiological dynamics (and therefore also ecological dynamics, for a host-pathogen ecosystem), with plentiful case report data and relatively simple biology.
-Model-based analysis of measles time series data led to advances in understanding of the seasonality of infectious disease \citep{fine82}, critical community size \citep{bartlett57}, the recognition that relatively simple mechanistic models can provide a remarkably good description of the dynamics \citep{earn00}, and much other foundational research on disease dynamics.
-Some progress has been on building and fitting spatiotemporal models for measles; e.g., \citet{xia04,eggo11,bjornstad19,becker20}.
-However, the lack of suitable methodology to fit and assess a flexible class of coupled models is an obstacle to further progress \citep{becker20}.
-Previous methodological work has used spatiotemporal measles models as a test problem \citep{park20,ionides21}.
-However, these methods have fallen short as tools for data analysis due to numerical considerations.
-Bearing all this in mind, measles provides a natural testing ground for our new methodologies.
-We demonstrate that we now have the tools to carry out likelihood maximization (and therefore, in principle, profile likelihood confidence interval construction and likelihood-based model selection) on mechanistic statistical models that are appropriate for spatiotemporal metapopulation disease data.
-
-We specifically set ourselves the task of fitting a spatiotemporal model to the case reports for 20 towns studied by \citet{he10}.
-We require the ability to handle discrete case counts which vary from zero to thousands of cases per week.
-We consider a model with the same structure as \citet{he10}, namely a Markov chain with Gamma noise on the infection rate, but with the addition of a term for transmission between cities.
-This requirement limits us to plug-and-play methodologies, which are those that require simulation from the latent process model but not the ability to  evaluate transition densities \citep{breto09,he10}.
-The model equations are given in Sec.~\ref{sec:model}.
 
@@
 @
+
+\section{Testing {\ibpf} on a measles transmission model}
+\label{sec:measles}
+
+Measles transmission has provided a useful example of epidemiological dynamics (and therefore also ecological dynamics, for a host-pathogen ecosystem), with plentiful case report data and relatively simple biology.
+Model-based analysis of measles time series data led to advances in understanding of the seasonality of infectious disease \citep{fine82}, critical community size \citep{bartlett57}, the recognition that relatively simple mechanistic models can provide a remarkably good description of the dynamics \citep{earn00}, and much other foundational research on disease dynamics.
+Some progress has been on building and fitting spatiotemporal models for measles; e.g., \citet{xia04,eggo11,bjornstad19,becker20}.
+However, the lack of suitable methodology to fit and assess a flexible class of coupled models is an obstacle to further progress \citep{becker20}.
+Previous methodological work has used spatiotemporal measles models as a test problem \citep{park20,ionides21}, but these methods have fallen short as tools for data analysis due to numerical considerations.
+Bearing all this in mind, measles provides a natural testing ground for our new methodologies.
+We demonstrate that we now have the tools to carry out likelihood maximization (and therefore, in principle, profile likelihood confidence interval construction and likelihood-based model selection) on mechanistic statistical models that are appropriate for spatiotemporal metapopulation disease data.
+
+\subsection{The measles data, a model and three submodels}
+\label{sec:model}
+
+We set ourselves the task of fitting a spatiotemporal model to the case reports for 20 towns studied by \citet{he10}.
+We require the ability to handle discrete case counts which vary from zero to thousands of cases per week.
+We consider a model with the same structure as \citet{he10}, namely a Markov chain with Gamma noise on the infection rate, but with the addition of a term for transmission between cities.
+This requirement limits us to plug-and-play methodologies, which are those that require simulation from the latent process model but not the ability to  evaluate transition densities \citep{breto09,he10}.
+
+Some previous analyses have used counts aggregated over two-week intervals \citep{park20,ionides21} since these were available for more cities, but our goal here is to extend the analysis of \citet{he10}.
+Apart from this, our model matches \citet{ionides21} and for completeness we repeat the description here.
+We compartmentalize the population of each town into susceptible ($S$), exposed ($E$), infectious ($I$), and recovered/removed ($R$) categories.
+The numbers of individuals in each compartment for town $\unit$ at time $t$ are denoted by integer-valued random variables $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$.
+The population dynamics are written in terms of counting processes $N_{Q_1Q_2,\unit}(t)$ enumerating cumulative transitions in town $\unit$, up to time $t$, from compartment $Q_1$ to $Q_2$.
+Here, $Q_1,Q_2\in \{S,E,I,R,B,D\}$ with $B$ denoting a source compartment for immigration or birth, and $D$ a sink compartment for emigration or death.
+We enumerate the $U=20$ towns studied by \citet{he10} in decreasing size, so that $\unit=1$ corresponds to London.
+Our model is described by the following system of stochastic differential equations, for $\unit=1,\dots, \Unit$,
+\begin{equation}
+\nonumber
+%\label{eq:measles:system}
+\begin{array}{lllllll}
+\displaystyle dS_\unit(t) &=& dN_{BS,\unit}(t) &-& dN_{SE,\unit}(t) &-& dN_{SD,\unit}(t), \\
+\displaystyle dE_\unit(t) &=& dN_{SE,\unit}(t) &-& dN_{EI,\unit}(t) &-& dN_{ED,\unit}(t), \\
+\displaystyle dI_\unit(t) &=& dN_{EI,\unit}(t) &-& dN_{IR,\unit}(t) &-& dN_{ID,\unit}(t).
+\end{array}
+\end{equation}
+The total population $P_\unit(t)=S_\unit(t)+E_\unit(t)+I_\unit(t)+R_\unit(t)$ is calculated by smoothing census data and is treated as known.
+The number of recovered individuals $R_\unit(t)$ in town $\unit$ is therefore defined implicitly.
+$N_{SE,\unit}(t)$ is modeled as a negative binomial death process \citep{breto09,breto11} defined by a rate, $\mu_{SE,\unit}(t)$, specified as
+%\begin{eqnarray}
+\begin{equation}
+\nonumber
+\mu_{SE,\unit}=
+\beta_{\unit}
+\left[
+  \left( \frac{I_\unit+\iota_u}{P_\unit} \right)^{\alpha_{\unit}}
+\hspace{-1mm}
+ + \sum_{\altUnit \neq \unit} \frac{v_{\unit\altUnit}}{P_\unit}
+  \left\{
+    \left(
+      \frac{ I_{\altUnit} }{ P_{\altUnit} }
+    \right)^{\alpha_{\unit}} -
+    \hspace{-1mm}
+    \left(
+      \frac{I_\unit}{P_\unit}
+    \right)^{\alpha_{\unit}}
+  \right\}
+\right] \frac{d\Gamma_{SE,\unit}}{dt}.
+%\label{eqn:transmissionrate}
+%\end{eqnarray}
+\end{equation}
+Here, the time dependence of $\mu_{SE,u}$, $\beta_u$, $I_u$ and $P_u$ is suppressed;
+$\alpha_{\unit}$ is a mixing exponent modeling inhomogeneous contact rates within a town;
+$\iota_{\unit}$ models immigration of infected individuals;
+${d\Gamma_{SE,\unit}}/{dt}$ is gamma white noise, with intensity parameter $\sigma_{SE.\unit}$;
+$\beta_{\unit}$, models the seasonal transmission driven by high contact rates between children at school,
+\begin{equation}
+\nonumber
+  \beta_{\unit}(t)=\Bigg\{
+  \begin{array}{ll}
+\big(1+\amplitude_{\unit}(1-\schoolTermFrac)\schoolTermFrac^{-1} \big)\, \meanBeta_{\unit} & \mbox{ during school term},\\
+\big( 1-\amplitude_\unit\big) \, \meanBeta_{\unit} & \mbox{ during vacation},
+  \end{array}
+\label{eq:term}
+\end{equation}
+where $\schoolTermFrac = 0.759$ is the proportion of the year taken up by the school terms, $\meanBeta_\unit$ is the mean transmission rate, and $\amplitude_{\unit}$ measures the reduction of transmission during school holidays.
+
+The number of travelers from town $\unit$ to $\altUnit$ is denoted by $v_{\unit\altUnit}$.
+Here, $v_{\unit\altUnit}$ is constructed using the gravity movement model of \cite{xia04} given by
+\[
+v_{\unit\altUnit} = \gravity_{\unit} \cdot \frac{\hspace{2mm} \overline{\dist}\hspace{2mm}}{\hspace{1.5mm}\overline{\pop}^2\hspace{1mm}} \cdot \frac{\pop_\unit \cdot \pop_{\altUnit}}{\dist_{\unit\altUnit}},
+\]
+where $\gravity_u$ is the {\it gravitational constant}, $\dist_{\unit\altUnit}$ is the distance between town $\unit$ and town $\altUnit$, $\pop_\unit$ is the average of $P_{\unit}(t)$ across time, $\overline{\pop}$ is the average of $\pop_{\unit}$ across towns, and $\overline{\dist}$ is the average of $\dist_{\unit\altUnit}$ across all pairs of towns.
+The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{QD,\unit}(t)$, for $Q\in\{S,E,I,R\}$,  are modeled as conditional Poisson process with per-capita rates $\mu_{EI,\unit}$, $\mu_{IR,\unit}$ and $\mu_{QD,\unit}$ respectively, and we fix $\mu_{QD,\unit}=(50 \mbox{ year})^{-1}$.
+The birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson process with rate $\mu_{BS,\unit}(t)$, given by interpolated census data.
+
+To describe the measurement process, let $Z_{\unit\comma\time}=N_{IR\comma\unit}(t_\time)-N_{IR\comma\unit}(t_{\time-1})$ be the number of removed infected individuals in the $n$th reporting interval.
+Suppose that cases are quarantined once they are identified, so that reported cases comprise a fraction $\rho$ of these removal events.
+The case report $\data{y}_{\unit\comma\time}$ is modeled as a realization of a discretized conditionally Gaussian random variable $Y_{\unit\comma\time}$, defined for $y>0$ via
+\begin{eqnarray}
+\nonumber
+\prob\big[Y_{\unit\comma\time}{=}y\mid Z_{\unit\comma\time}{=}z\big] &=& \varPhi\big(y+0.5; \rho_{\unit} z,\rho_{\unit}(1-\rho_{\unit})z+\measOD_{\unit}^2\rho_{\unit}^2z^2\big)
+\\
+&&\hspace{-20mm}
+- \varPhi\big(y-0.5; \rho_{\unit} z,\rho_{\unit}(1-\rho_{\unit})z+\measOD_{\unit}^2\rho^2z^2\big)
+\label{eq:obs}
+\end{eqnarray}
+where $\varPhi(\cdot;\mu,\sigma^2)$ is the $\normal(\mu,\sigma^2)$ cumulative distribution function, and $\measOD$ models overdispersion relative to the binomial distribution.
+For $y=0$, we replace $y-0.5$ by $-\infty$ in \eqref{eq:obs}.
+Three data points were treated as missing by \citet{he10} due to presumed recording errors.
+We followed that decision in the measurement model by including a special value $\mathrm{NA}$ in $\Yspace$ and setting $Y_{\unit,n}$ to be $\mathrm{NA}$ with probability one when $\data{y}_{\unit,n}$ is missing.
+
+We have written the model with all parameters unit-specific, so $\theta=\psi_{1:U}$ with
+\arxiv{\begin{equation*}}
+\statSinica{$}
+\psi_u=\big(\meanBeta_{\unit},\mu_{EI,\unit}, \mu_{IR,\unit},\rho_{\unit},\measOD_{\unit},\sigma_{SE,\unit},\gravity_{\unit},\iota_{\unit},\amplitude_{\unit},\alpha_{\unit}, S_{0,\unit}, E_{0,\unit},I_{0,\unit}\big)
+\arxiv{\end{equation*}}
+\statSinica{$}
+for $u\in\seq{1}{U}$.
+This defines an extended model for implementing IBPF with shared parameters, as in \eqref{eq:extension}.
+It also lets us address a key data analysis question of which parameters should be shared between units and which should be unit-specific.
 
@@
 The data are shown in Fig~\ref{fig:data-plot} and simulations from the model are shown in Fig~\ref{fig:sim-plot}.
-Parameters for the simulated model were based on the analysis of individual towns by \citet{he10} and adjusted to provide a visual match in the simulations.
-A goal of the simulation study is to obtain appropriate algorithmic choices for the data analysis.
-We seek to develop methodology which is demonstrably successful when the truth is known, before applying it to data.
-We present the simulation study from this perspective, to describe the process of developing our data analysis, though we acknowledge one could revisit the simulation study based on the data analysis in Sec.~\ref{sec:data} using maximum likelihood estimates of the parameters.
-
-A key data analysis question is to decide which parameters should be shared between units and which should be unit-specific.
-For a general SpatPOMP parameter, it is not necessarily well-defined how to extend the model so that each unit has a distinct parameter value, however our measles model falls into the structure defined in \eqref{eq:extension} and we suspect that similar considerations may arise in other applications.
-Different numerical issues may arise as the number of unit-specific parameters increases.
-Each unit-specific parameter contributes $U$ dimensions to the parameter space, potentially contributing difficulties associated with high-dimensional or weakly identified parameters.
-On the other hand, information about a unit-specific parameters may be localized in neighboring units which could align conveniently with the structure of the BPF.
-Estimation of shared parameters, by contrast, has the task of harvesting information from across all the units, which may have its own difficulties.
+Parameters for the simulated model were based on the analysis of individual towns by \citet{he10}.
+In order to investigate estimation of either shared or unit-specific parameters, the simulation was conducted with all parameters being shared.
+To find a shared parameter vector capable of providing a reasonable representation of all the towns simultaneously, we experimented to look for a visual match between the data and the simulations.
+The parameters used are reported in the supplement (Sec.~\supSecMeaslesParams).
 
@@
 To facilitate data analysis, we wish to have methodology that operates across the full spectrum of decisions on shared versus unit-specific parameter designations.
-We therefore test our method on two cases: (A) mostly shared parameters, choosing only the initial values and reporting rate to be unit-specific; (B) every parameter is unit-specific.
-We also consider a third case; (C) every parameter is unit-specific and the dynamic coupling (in our context, movement of infected individuals between towns) is replaced by an external forcing of each unit (in our context, an immigration rate of infected individuals from outside the study population).
-Case (C) provides a useful point of reference since it is a special case of a PanelPOMP model \citep{breto19} and also can be analyzed as a collection of separate POMP models.
-We set up our model so that in case (C) we retrieve the analysis of \citet{he10}.
-One may expect that methods which take advantage of the special structure of (C) should out-perform more general methods that permit coupling between units, so we expect the SpatPOMP methods to be less numerically efficient than applying POMP methods separately to each unit.
+We therefore test our method on three submodels: {\modelA} has mostly shared parameters, choosing only the initial values and reporting rate to be unit-specific; {\modelB} has every parameter unit-specific; {\modelC} has every parameter is unit-specific and the dynamic coupling (in our context, movement of infected individuals between towns) is replaced by an external forcing of each unit (in our context, an immigration rate of infected individuals from outside the study population).
+Sec.~{\supSecMeaslesParams} defines the submodels in more detail.
+Submodel {\modelC} provides a useful point of reference since it is a special case of a PanelPOMP model \citep{breto19} and also can be analyzed as a collection of separate POMP models.
+We set up our model so that {\modelC} matches the analysis of \citet{he10}.
+One may expect that methods which take advantage of the special structure of {\modelC} should out-perform more general methods that permit coupling between units, so we expect the SpatPOMP methods to be less numerically efficient than applying POMP methods separately to each unit.
 Questions at stake are: How much less efficient are the SpatPOMP methods? Are simulation-based SpatPOMP inference methods practical for situations such as the measles model of \citet{he10}?
 
-The simulated model is drawn from model (A), which is nested within model (B) but not within (C).
-For models B and C, with all parameters are unit-specific, we must estimate $20\times 13=260$ parameters.
+The simulated model is drawn from model {\modelA}, which is nested within model {\modelB} but not within {\modelC}.
+For models {\modelB} and {\modelC}, with all parameters are unit-specific, we must estimate $20\times 12=240$ parameters.
 This greatly exceeds the 7 shared parameters fitted by \citet{ionides21} for a measles metapopulation model using a large computational effort.
@@
 Whether this is sufficient to obtain a good approximation to the filter distribution is situation-dependent, but for the specific case of metapopulation models (for which the strongest coupling is within units rather than between units) the approximation can be empirically successful.
-In principle, the approximation error can be reduced by increasing block size, but in practice the additional Monte Carlo variance acquired by doing this is not worthwhile in situations where the coupling is relatively weak.
+In principle, the approximation error can be reduced by increasing the number of units in each block, but in practice the additional Monte Carlo variance acquired by doing this is not worthwhile in situations where the coupling is relatively weak.
 The BPF log-likelihood evaluation at the true parameter value is $\loglik_{\mathrm{true}}=\Sexpr{myround(loglik_truth[1],1)}$ with a Monte Carlo standard error of $\Sexpr{myround(loglik_truth[2],1)}$.
 
-In Fig.~\ref{fig:sim_search_boxplot} we investigate a sequence of successive searches for the MLE for the models A, B and C described above.
-Each search was replicated $36$ times.
-Search~1 was started with each parameter adjusted by a uniform $[-0.1,0.1]$ random perturbation on an appropriate dimensionless scale (log for non-negative parameters, logit for $[0,1]$ valued parameters).
-This is a fairly small perturbation, but nevertheless sufficient to knock the likelihood of the parameter vector around 300 log units below the MLE (shown in Fig.~\ref{fig:search_diagnostics}).
-The goal of this study is to show that the algorithm can succeed reliably on a relatively easy local optimization task.
-Subsequent searches were started with four copies of each parameter vector with estimated likelihood in the top 25\% for the previous search.
+\subsection{IBPF on simulated data}
+\label{sec:sim}
 
-The MLE is not known exactly in this case.
-Wilks's theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+9)/2=44.5$ for A and $(13\times 20)/2=130$ for B.
-The $y$-axis values in Fig.~\ref{fig:sim_search_boxplot} show log-likelihoods exceeding the truth by less than this, indicating some imperfection to the Monte Carlo maximization so far as Wilks's asymptotic result holds in this situation.
-Despite this limitation, searches that exceed the likelihood at the truth have found inferentially plausible sets of parameters which can be used to study the likelihood surface around the maximum.
-For example, Monte Carlo profile confidence intervals can give proper coverage even in the presence of considerable Monte Carlo error \citep{ionides17,ning21}.
-
-<<v6_start,echo=F>>=
+<<v4_start,echo=F>>=
 readRDS("v4/v1a_3/initial_lik.rds") -> loglik_start
@@
 
-<<sim_search_boxplot, echo=F, fig.height=4, fig.width=5.5, out.width="5.5in", fig.cap = "Fitting different models to the simulated measles data, using an initial search and four further refinement searches.
-(A) $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B) $13\\times 20$ unit-specific parameters and no shared parameters.
-(C) Also all unit-specific, but with immigration rather than coupling, matching the model of He et al (2010).
-The horizontal dashed line denotes the likelihood obtained by He et al, using \\code{pomp}.
-Here, $J=4000$, and the random walk standard deviation (on log- and logit-transformed parameters) is 0.005 for the initial search and 0.00125 for subsequent refinements.",fig.show = "hold">>=
+<<sim_search_boxplot, echo=F, fig.height=3, fig.width=5.5, out.width="5in", fig.cap = "Fitting simulated measles data, using an initial search and four refinement steps.
+(A1--A5) model \\modelA, $4\\times 20$ unit-specific parameters and $8$ shared parameters; (B1--B5) model \\modelB, $12\\times 20$ unit-specific parameters and no shared parameters;  (C1--C5) model \\modelC, also all unit-specific, but with immigration rather than coupling, matching He et al (2010).
+The horizontal dashed line is the log-likelihood at the true parameters, evaluated by BPF.",fig.show = "hold">>=
 a <- cbind(
@@
 
+par(mai=c(0.4,0.6,0.1,0.1))
 loglik_max <- max(cbind(a,b,c),truth_loglik)
@@
 
-<<search_diagnostics, echo=F, fig.height=6.5, fig.width=4.5, out.width="5in", fig.cap = "Three steps of a likelihood search on simulated data (left panels) and UK measles data (right panels).
-Final log-likelihood values are plotted against starting log-likelihood for an initial search (top row), a refinement of this search (second row) and a second refinement (third row).
-Unit-specific parameters were $\\rho$, $S_0$, $E_0$, $I_0$ (for the simulation study, model A) and all parameters (for the data, model B).">>=
+A goal of this simulation study is to obtain appropriate algorithmic choices for the data analysis.
+We seek to develop methodology which is demonstrably successful when the truth is known, before applying it to data.
+One could revisit the simulation study based on the data analysis in Sec.~\ref{sec:data} using maximum likelihood estimates of the parameters.
+
+In Fig.~\ref{fig:sim_search_boxplot} we investigate a sequence of successive searches for the MLE for the models {\modelA}, {\modelB} and {\modelC} described above.
+Each search was replicated $36$ times.
+Search~1 was started with each parameter adjusted by a uniform $[-0.1,0.1]$ random perturbation on an appropriate dimensionless scale (log for non-negative parameters, logit for $[0,1]$ valued parameters, see Sec.~{\supSecAlg}).
+This is a fairly small perturbation, but nevertheless sufficient to knock the likelihood of the parameter vector around 200 log units below the MLE (shown in Fig.~\ref{fig:search_diagnostics}).
+The goal of this study is to show that the algorithm can succeed reliably on a relatively easy local optimization task.
+Subsequent searches were started with four copies of each parameter vector with estimated likelihood in the top 25\% for the previous search.
+
+The MLE is not known exactly in this case.
+Wilks' theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+8)/2=44$ for {\modelA} and $(12\times 20)/2=120$ for {\modelB}.
+The $y$-axis values in Fig.~\ref{fig:sim_search_boxplot} show log-likelihoods exceeding the truth by less than this, indicating some imperfection to the Monte Carlo maximization so far as Wilks' asymptotic result holds in this situation.
+Despite this limitation, searches that exceed the likelihood at the truth have found inferentially plausible sets of parameters which can be used to study the likelihood surface around the maximum.
+For example, Monte Carlo profile confidence intervals can give proper coverage even in the presence of considerable Monte Carlo error \citep{ionides17,ning21}.
+
+<<search_diagnostics, echo=F, fig.height=6.5, fig.width=4.5, out.width="5in", fig.cap = "Three steps of a likelihood search on simulated data (left panel, model \\modelA) and UK measles data (right panel, model \\modelB).Dashed lines parallel to the axes show the log-likelihood at the truth (left panel) and the \\citet{he10} value (right panel).
+Points above the diagonal dashed line show improvement due to the search step.">>=
 par(mfcol=c(3,2))
@@
 
-In Fig.~\ref{fig:search_diagnostics} we take a closer look at convergence diagnostics corresponding to A for the simulated data (the first column) and B for the UK measles data (the second column).
+In Fig.~\ref{fig:search_diagnostics} we take a closer look at convergence diagnostics corresponding to {\modelA} for the simulated data (the first column) and {\modelB} for the UK measles data (the second column).
 For now, we focus on the first column.
@@
 
-\subsection{The measles data and model}
-\label{sec:model}
-
-The data we consider here are weekly case reports for twenty towns, identical to those studied by \citet{he10}.
-Some previous analyses have used counts aggregated over two-week intervals \citep{park20,ionides21} since these were available for more cities, but our goal here is to extend the analysis of \citet{he10}.
-Apart from this, our model matches \citet{ionides21} and for completeness we repeat the description here.
-We compartmentalize the population of each town into susceptible ($S$), exposed ($E$), infectious ($I$), and recovered/removed ($R$) categories.
-The numbers of individuals in each compartment for town $\unit$ at time $t$ are denoted by integer-valued random variables $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$.
-The population dynamics are written in terms of counting processes $N_{AB,\unit}(t)$ enumerating cumulative transitions in town $\unit$, up to time $t$, between compartments $A$ and $B$.
-Here, $A,B\in \{S,E,I,R,B,D\}$ with $B$ denoting a source compartment for immigration or birth, and $D$ a sink compartment for emigration or death.
-We enumerate the $U=20$ towns studied by \citet{he10} in decreasing size, so that $\unit=1$ corresponds to London.
-Our model is described by the following system of stochastic differential equations, for $\unit=1,\dots, \Unit$,
-\begin{equation}
-\nonumber
-%\label{eq:measles:system}
-\begin{array}{lllllll}
-\displaystyle dS_\unit(t) &=& dN_{BS,\unit}(t) &-& dN_{SE,\unit}(t) &-& dN_{SD,\unit}(t), \\
-\displaystyle dE_\unit(t) &=& dN_{SE,\unit}(t) &-& dN_{EI,\unit}(t) &-& dN_{ED,\unit}(t), \\
-\displaystyle dI_\unit(t) &=& dN_{EI,\unit}(t) &-& dN_{IR,\unit}(t) &-& dN_{ID,\unit}(t). 
-\end{array}
-\end{equation}
-The total population $P_\unit(t)=S_\unit(t)+E_\unit(t)+I_\unit(t)+R_\unit(t)$ is calculated by smoothing census data and is treated as known.
-The number of recovered individuals $R_\unit(t)$ in town $\unit$ is therefore defined implicitly.
-$N_{SE,\unit}(t)$ is modeled as negative binomial death processes \citep{breto09,breto11}
-with over-dispersion parameter $\sigma_{SE}$, and rate given by
-\begin{eqnarray}
-\nonumber
-\mathbb{E} \big[ N_{SE,\unit}(t+dt) - N_{SE,\unit}(t) \big] 
-&=& 
-\beta(t) \, S_\unit(t) 
-\Big[ 
-  \left( \frac{I_\unit+\iota}{P_\unit} \right)^\alpha
-\\
-\label{eq:dEdt}
-&& \hspace{-3.8cm}
- + \sum_{\altUnit \neq \unit} \frac{v_{\unit\altUnit}}{P_\unit} 
-  \left\{ 
-    \left(
-      \frac{ I_{\altUnit} }{ P_{\altUnit} }
-    \right)^\alpha - 
-    \left(
-      \frac{I_\unit}{P_\unit}  
-    \right)^\alpha
-  \right\}
-\Big] dt + o(dt),\label{eqn:transmissionrate}
-\end{eqnarray}
-where $\beta(t)$ models seasonality driven by high contact rates between children at school, described by
-\begin{equation}
-\nonumber
-  \beta(t)=\begin{cases}
-\big(1+\amplitude(1-p)p^{-1} \big)\, \meanBeta & \mbox{ during school term},\\
-\big( 1-\amplitude\big) \, \meanBeta& \mbox{ during vacation},
-  \end{cases} \label{eq:term}
-\end{equation}
-with $p = 0.759$ being the proportion of the year taken up by the school terms, $\meanBeta$ is the mean transmission rate, and $\amplitude$ measures the reduction of transmission during school holidays.
-In \eqref{eq:dEdt}, $\alpha$ is a mixing exponent modeling inhomogeneous contact rates within a town, and $\iota$ models immigration of infected individuals which is appropriate when analyzing a subset of cities that cannot be treated as a closed system.
-The number of travelers from town $\unit$ to $\altUnit$ is denoted by $v_{\unit\altUnit}$. 
-Here, $v_{\unit\altUnit}$ is constructed using the gravity model of \cite{xia04}, 
-\[
-v_{\unit\altUnit} = \gravity \cdot \frac{\;\overline{\dist}\;}{\bar{P}^2} \cdot \frac{P_\unit \cdot P_{\altUnit}}{\dist(\unit,\altUnit)},
-\]
-where $\dist(\unit,\altUnit)$ denotes the distance between town $\unit$ and town $\altUnit$, $P_\unit$ is the average population for town $\unit$ across time, $\bar{P}$ is the average population across towns, and $\overline{\dist}$ is the average distance between a randomly chosen pair of towns.
-Here, we model $v_{\unit\altUnit}$ as fixed through time and symmetric between any two arbitrary towns, though a natural extension would allow for temporal variation and asymmetric movement between two towns.
-The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{\bullet D,\unit}(t)$ are modeled as conditional Poisson process with per-capita rates $\mu_{EI}$, $\mu_{IR}$ and $\mu_{\bullet D}$ respectively, and we fix $\mu_{\bullet D}=(50 \mbox{ year})^{-1}$.
-The birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson process with rate $\mu_{BS,\unit}(t)$, given by interpolated census data.
-
-To describe the measurement process, let $Z_{\unit\comma\time}=N_{IR\comma\unit}(t_\time)-N_{IR\comma\unit}(t_{\time-1})$ be the number of removed infected individuals in the $n$th reporting interval.
-Suppose that cases are quarantined once they are identified, so that reported cases comprise a fraction $\rho$ of these removal events.
-The case report $\data{y}_{\unit\comma\time}$ is modeled as a realization of a discretized conditionally Gaussian random variable $Y_{\unit\comma\time}$, defined for $y>0$ via
-\begin{eqnarray}
-\nonumber
-\prob\big[Y_{\unit\comma\time}{=}y\mid Z_{\unit\comma\time}{=}z\big] &=& \Phi\big(y+0.5; \rho z,\rho(1-\rho)z+\psi^2\rho^2z^2\big)
-\\
-&&\hspace{-20mm}
-- \Phi\big(y-0.5; \rho z,\rho(1-\rho)z+\psi^2\rho^2z^2\big)
-\label{eq:obs}
-\end{eqnarray}
-where $\Phi(\cdot;\mu,\sigma^2)$ is the $\normal(\mu,\sigma^2)$ cumulative distribution function, and $\psi$ models overdispersion relative to the binomial distribution.
-For $y=0$, we replace $y-0.5$ by $-\infty$ in \eqref{eq:obs}.
-
-Three data points were treated as missing by \citet{he10} due to presumed recording errors, and we followed that decision.
-The general SpatPOMP framework has flexibility that permits modeling of missing data in various ways. 
-We chose to include a special value $\mathrm{NA}$ in $\Yspace$ and set $Y_{u,n}$ to be $\mathrm{NA}$ with probability one when $\data{y}_{u,n}$ is missing.
-
-\section{Demonstration on data}
+\subsection{IBPF applied to data}
 \label{sec:data}
 
-<<data_search_boxplot, echo=F, fig.height=4, fig.width=5.5, out.width="5.5in", fig.cap = "Fitting different models to the UK measles data using the method tested on simulated data.
+<<data_search_boxplot, echo=F,  fig.height=3, fig.width=5.5, out.width="5in", fig.cap = "Fitting different models to the UK measles data using the method tested on simulated data.
 (A) $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B) $13\\times 20$ unit-specific parameters and no shared parameters.
-(C) Also all unit-specific, but with immigration rather than coupling, matching the model of He et al (2010).
-The horizontal dashed line denotes the likelihood obtained by He et al, using \\code{pomp}.",fig.show = "hold">>=
+(C) Also all unit-specific, but with immigration rather than coupling.
+The horizontal dashed line is the likelihood from \\citet{he10}.">>=
 a <- cbind(
@@
 loglik_max <- max(cbind(a,b,c),loglik_he)
+
+par(mai=c(0.5,0.8,0.1,0.1))
 boxplot(cbind(a,b,c),ylim=loglik_max-c(5500,0),
@@
 
-Fig.~\ref{fig:data_search_boxplot} shows results for fitting models A, B and C to UK measles data using successive rounds of {\ibpf}, applying the same method used for the simulation study shown in Fig.~\ref{fig:sim_search_boxplot}.
+Fig.~\ref{fig:data_search_boxplot} shows results for fitting models {\modelA}, {\modelB} and {\modelC} to UK measles data using successive rounds of {\ibpf}, applying the same method used for the simulation study shown in Fig.~\ref{fig:sim_search_boxplot}.
 The dashed lines show the sum of the log-likelihoods obtained by \citet{he10}, $\ell_{\mathrm He}=\Sexpr{myround(loglik_he,1)}$.
-The value $\ell_{\mathrm He}$ corresponds to model C.
-BPF with each town forming a separate block, applied to an uncoupled model such as C, is equivalent to carrying out independent particle filters for each unit.
-Indeed, if the published parameter values from \citet{he10} are inserted into the \code{spatPomp} object for model C, then the \code{bpfilter} function can be used to carry out BPF, which retrieves $\ell_{\mathrm He}$, up to Monte Carlo error.
-Each of models A, B and C in Fig.~\ref{fig:data_search_boxplot} shows shortfall relative to $\ell_{\mathrm He}$.
+The value $\ell_{\mathrm He}$ corresponds to model {\modelC}.
+BPF with each town forming a separate block, applied to an uncoupled model such as {\modelC}, is equivalent to carrying out independent particle filters for each unit.
+Indeed, if the published parameter values from \citet{he10} are inserted into the \code{spatPomp} object for model {\modelC}, then the \code{bpfilter} function can be used to carry out BPF, which retrieves $\ell_{\mathrm He}$, up to Monte Carlo error.
+Each of models {\modelA}, {\modelB} and {\modelC} in Fig.~\ref{fig:data_search_boxplot} shows shortfall relative to $\ell_{\mathrm He}$.
 
-The largest shortfall is for A, and an explanation is that A has more shared parameters than the evidence in the data supports.
+The largest shortfall is for {\modelA}, and an explanation is that {\modelA} has more shared parameters than the evidence in the data supports.
 The results of \citet{he10} suggest that the data are explained better when various parameters are a function of the town population.
@@
 
-Models B and C give rise to comparable likelihoods, which is in contrast to Fig.~\ref{fig:sim_search_boxplot}.
-Fitting to simulated data from B (in the special case where all unit-specific parameters are equal across all units) the shortfall for C in Fig.~\ref{fig:sim_search_boxplot} indicates that we obtain a substantially worse fit by approximating coupling by an uncoupled reservoir of infection.
-If the actual data were also explained substantially better by a coupled model, we would expect to see comparable results in the data analysis.
+Models {\modelB} and {\modelC} give rise to comparable likelihoods, which is in contrast to Fig.~\ref{fig:sim_search_boxplot}.
+Fitting to simulated data from {\modelB} (in the special case where all unit-specific parameters are equal across all units) the shortfall for {\modelC} in Fig.~\ref{fig:sim_search_boxplot} indicates that we obtain a substantially worse fit by approximating coupling by an uncoupled reservoir of infection.
+If the actual data were also explained substantially better by the coupled model, we would expect to see comparable results in the data analysis.
 Since we do not, we conclude that this coupled model is not a substantially better explanation than the uncoupled model.
@@
 Instead, \citet{bjornstad19} looked at summary statistics based on local fade-outs and re-introductions.
-Further investigation is needed to determine whether other models for coupling can show improvement over this gravity model on these data.
-This should be carried out in conjunction with research into the specification of shared and unit-specific parameters.
 
-We notice from Fig.~\ref{fig:data_search_boxplot} that model B has a small but distinct shortfall compared to $\ell_{\mathrm He}$.
+We notice from Fig.~\ref{fig:data_search_boxplot} that model {\modelB} has a small but distinct shortfall compared to $\ell_{\mathrm He}$.
 On the simulated data, we do not observe this shortfall, and we deduce that the real data provides a more challenging optimization environment.
@@
 
+  We have demonstrated a workflow that led to a likelihood-based assessment of measles metapopulation models ({\modelA} and {\modelB}) with the possibility of finding evidence that they out-perform the uncoupled model, {\modelC}.
+  Our methodology showed the potential to successfully refute {\modelC} on simulated data, when the truth was within {\modelA} and {\modelB}, but no advantage was found for these coupled models when the same comparison was carried out on the data.
+Likelihood maximization for the measles data fits the {\it common task framework} described by \citet[][Sec.~6]{donoho17}, with the likelihood value for {\modelC} obtained by \citet{he10} providing a benchmark challenge.
+Future improvements in models, perhaps facilitated by the open-source models and methods accompanying this article, may obtain metapopulation models that convincingly beat model {\modelC}.
+
 Computation time for Fig.~\ref{fig:sim_search_boxplot} and Fig.~\ref{fig:data_search_boxplot} was approximately 24hr on a single core of a compute node for 4000 particles iterated 100 times over the 730 time points (weekly data for 14 years) for 20 cities.
 Each box in these figures involves 36 replications and thus took 24hr on all cores of one node on a computing cluster.
-This level of effort is tolerable, but does suggest that the methods will not readily extend to tasks that are considerably larger. 
-In particular, we do not expect applicability to situations where it is necessary to consider $10^3$ or more spatial units.
-In such situations, it may be necessary to make additional approximations such as those inherent in the ensemble Kalman filter \citep{evensen09book,katzfuss19} or other numerical filtering techniques \citep{whitehouse22}.
+A measure of computational efficiency is the size of problem that can be solved on this timescale, discussed further in Sec.~{\supSecEfficiency}.
+Tasks that are considerably larger, perhaps $10^3$ or more spatial units, may require additional approximations such as those inherent in the ensemble Kalman filter \citep{evensen09book,katzfuss19} or other numerical filtering techniques \citep{whitehouse22}.
 However, we anticipate that many practical metapopulation analyses can be addressed within the scope we have demonstrated.
@@
 
-\noindent {\bf Code and data availability}
-
-\vspace{1mm}
-
-\noindent The data we analyze are in the public domain, and are available together with source code at \url{https://github.com/ionides/ibpf_article}.
-
-\vspace{3mm}
-
 \noindent {\bf Acknowledgments}
@@
 
-\noindent This work was supported by National Science Foundation grants DMS-1761603 and DMS 1646108.
-We acknowledge an anonymous referee from a previous article who recommended that we should try implementing a block particle filter.
+\noindent This work was supported by National Science Foundation grants DMS-1761603 and DMS 1646108, and a Texas A\&M University Seed Fund Grant.
+We acknowledge an anonymous referee from a previous article who recommended implementing a block particle filter.
+We are grateful to two anonymous referees whose constructive feedback has improved this article.
 
 \bibliography{bib-ibpf}
+%\InputIfFileExists{ms.bbl}
 
-% \vspace{8mm}
-\newpage
+\arxiv{
+\setcounter{section}{0}
+\setcounter{equation}{0}
+\def\theequation{S\arabic{section}.\arabic{equation}}
+%\def\thesection{S\arabic{section}}
+\renewcommand\thefigure{S\arabic{figure}}
+\renewcommand\thetable{S\arabic{table}}
+\def\thesection{S\arabic{section}}
 
-\noindent{\bf \Large Appendix A}
+<<read-sim-model,echo=F,eval=T,results="hide",warning=FALSE,message=FALSE>>=
+library(spatPomp)
+po <- readRDS("sim/he10.rds")
+theta <- coef(po)
+myround<- function (x, digits = 1) {
+  # adapted from the broman package
+  # solves the bug that round() kills significant trailing zeros
+  if (length(digits) > 1) {
+    digits <- digits[1]
+    warning("Using only digits[1]")
+  }
+  if (digits < 1) {
+    as.character(round(x,digits))
+  } else {
+    tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
+    zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
+    tmp[tmp == paste0("-", zero)] <- zero
+    tmp
+  }
+}
 
-\vspace{3mm}
+mysignif <- function (x, digits = 1) {
+  myround(x, digits - ceiling(log10(abs(x))))
+}
+@
 
-\noindent{\bf \large Varying the IBPF spatial autoregression parameter, $\spatReg$}
+\section{Parameters for the measles model}
 
-\vspace{-5mm}
+\newcommand\measlesShared{shared}
+\newcommand\measlesUnitSpecific{unit-specific}
 
-<<spatreg_boxplot, echo=F, fig.height=4, fig.width=5.5, out.width="5.5in", fig.pos="h", fig.cap = paste0("Varying the spatial autoregression algorithmic parameter for shared parameters with simulated measles data.
-The log-likelihood was obtained using $M=100$ iterations of IBPF starting at random parameters (median starting log-likelihood, ",signif(median_start,3), ").
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|}
+\hline
+Parameter & \modelA & \modelB & \modelC & Simulation \\
+\hline
+$\pinit_{S,u}$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["S_01"],3)}$ \\
+$\pinit_{E,u}$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["E_01"],5)}$ \\
+$\pinit_{I,u}$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["I_01"],5)}$ \\
+$\rho_u$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["rho1"],1)}$ \\
+$\measOD_{\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["psi1"],2)}$ \\  
+$\mu_{EI,\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["sigma1"]/52,0)} \, \mathrm{week}^{-1}$ \\
+$\mu_{IR}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["gamma1"]/52,0)} \, \mathrm{week}^{-1}$ \\
+$\meanBeta_\unit$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["gamma1"]*(theta["R01"]+theta["mu1"])/52,0)} \, \mathrm{week}^{-1}$ \\
+$\sigma_{SE,\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["sigmaSE1"],2)} \, \mathrm{year}^{1/2}$ \\
+$\amplitude_{\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["amplitude1"],1)}$ \\
+$\alpha_{\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["alpha1"],0)}$ \\
+$\gravity_{\unit}$ & \measlesShared & \measlesUnitSpecific & 0 &
+  $\Sexpr{myround(theta["g1"],0)}$ \\
+$\iota_{\unit}$ & 0 & 0 & \measlesUnitSpecific &
+  0 \\
+\hline
+\end{tabular}
+\caption{Parameters for the measles model.
+Sub-model {\modelA} has 4 unit-specific parameters and 8 shared parameters, with movement between units following a gravity equation.
+Sub-model {\modelB} has 12 unit-specific parameters and no shared parameters, with gravity movement.
+Sub-model {\modelC} has 12 unit-specific parameters and no shared parameters, with independent immigration of infections rather than movement between units.
+The last column shows the parameter values used for the simulated data in Figure~2 of the main text.
+}\label{tab:params}
+\end{center}
+\end{table}
+
+\noindent Notes on Table~\ref{tab:params}:
+\begin{enumerate}
+\item In the model formulation, all parameters are written as unit-specific, i.e., with a $u$ subscript. Shared parameters take equal values across all units.
+\item For the simulation, all parameters are shared, so one can assess the inferential consequences of estimating models with some or all parameters unit-specific.
+Establishing a good choice of parameters to be shared or unit-specific for the data is a data analysis goal.
+\item The initial values of the latent states are parameterized as fractions of the total population, so
+$S_{0,u}=\pinit_{S,u}P_u(t_0)$,
+$E_{0,u}=\pinit_{E,u}P_u(t_0)$,
+$I_{0,u}=\pinit_{I,u}P_u(t_0)$,
+and $R_{0,u}=\big(1-\pinit_{S,u}-\pinit_{E,u}-\pinit_{I,u}\big)P_u(t_0)$.
+\item The parameterization used in the numerical implementation replaces $\meanBeta_{\unit}$ with the basic reproductive number, a dimensionless ratio defined as
+$\mathscr{R}_{0,\unit}= \meanBeta_{\unit} \big(\mu_{IR,\unit}+\mu_D\big)^{-1}$.
+For the simulation,  $\mathscr{R}_{0,\unit}=\Sexpr{myround(theta["R01"],0)}$.
+\end{enumerate}
+
+%\newpage
+
+\section{Algorithmic parameters and transformations}
+
+%% values taken from z3/z1a.r and v4/v4a.r
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|}
+\hline
+Parameter & Value \\
+\hline
+$J$ & 4000 \\
+$M$ & 100 \\
+$\breve\sigma$ & 0.005 (0.00125 on some searches for the simulated data)\\
+$\spatReg$ & 0.1 \\
+$a$ & 0.5 \\
+\hline
+\end{tabular}
+\caption{Algorithmic parameters used for applying IBPF to the measles model. The same values were used for each of the sub-models {\modelA}, {\modelB} and {\modelC}.}\label{tab:algpars}
+\end{center}
+\end{table}
+
+\noindent Notes on Table~\ref{tab:algpars}:
+\begin{enumerate}
+\item The full $D\times \Time$ matrix of perturbation standard deviations with entries $\sigma_{d,n}$ was reduced to a single algorithmic parameter, $\breve\sigma$, via
+\begin{equation}
+\nonumber
+\sigma_{d,n} =
+  \left\{ \begin{array}{ll}
+    \breve\sigma & \mbox{if $\theta_d$ is not an IVP and not $\alpha_{\unit}$}
+    \\
+    0.1\, \breve\sigma & \mbox{if $\theta_d$ is $\alpha_{\unit}$}
+    \\
+    2\, \breve\sigma & \mbox{if $\theta_d$ is an IVP, and $n=0$}
+    \\
+    0 & \mbox{if $\theta_d$ is an IVP, and $n\ge 1$}
+  \end{array}\right.
+\end{equation}
+where the initial value parameters (IVPs) are $\pinit_{S,u}$, $\pinit_{E,u}$ and $\pinit_{I,u}$.
+
+\item Parameters perturbations were carried out on a transformed scale.
+A logit transform was used for $\pinit_{S,u}$, $\pinit_{E,u}$, $\pinit_{I,u}$, $\rho_{\unit}$.
+A log transformation was used for $\measOD_{\unit}$, $\sigma_{SE,u}$, $\mu_{EI,u}$, $\mu_{IR,\unit}$, $\meanBeta_{\unit}$, $\alpha_{\unit}$, $\gravity_{\unit}$, $\iota_{\unit}$.
+No transformation was used for $\amplitude_{\unit}$.
+
+\item The algorithmic parameters do not have any scientific significance once successful maximization has been demonstrated.
+They may affect the ease of successful maximization, or even the ability to attain this within an acceptable level of Monte Carlo uncertainty.
+
+\item $M=100$ was chosen empirically.
+For a computationally challenging maximization problem, we expect to carry out many searches from a variety of starting values, and we conduct further experiments following up on successful leads.
+In this case, it is enough to choose $M$ so that each search has a fair chance of finding a higher likelihood when there are local improvements to be made.
+
+\item Numerical experiments were carried out to choose $J$.
+We look for the smallest $J$ such that there is not much to be gained by making $J$ larger.
+This is carried out by looking for evidence about the bias and variance (discussed further in Sec.~\supSecEfficiency) which both need consideration when carrying out likelihood ratio tests and Monte Carlo adjusted profile confidence intervals \citep{ionides17,ning21}.
+
+\item Each iteration of IBPF produces a log-likelihood estimate corresponding to the extended model with dynamically perturbed parameters.
+At the end of the search, the log-likelihood was re-evaluated using BFP, with 10 replications at $J=8000$ particles.
+A high level of effort on likelihood evaluation assists the task of building understanding about the likelihood surface from repeated Monte Carlo searches.
+
+\item The exact record of all our computations is the source code for our numerical results, which is available at \url{https://github.com/ionides/ibpf_article}.
+
+\end{enumerate}
+
+%\newpage
+
+\section{Computational efficiency}
+
+Numerical error of Monte Carlo methods can be decomposed as bias and variance, and efficiency corresponds to how error scales with the amount of computational resources used.
+Expended resources can be quantified by objective metrics such as joules or dollars, but in practice we usually assess resources in terms of computational time on the machines that we personally have available.
+
+Here, our main goals are evaluation and maximization of the log-likelihood.
+Variance in Monte Carlo likelihood estimates results in negative bias on the log-likelihood due to Jensen's inequality.
+Approximations involved in constructing a filter provide another source of bias for estimating the log-likelihood.
+This Monte Carlo approximation bias has negative expectation, over when the model is correct, since log-likelihood is a proper scoring rule \citep{gneiting07}.
+Monte Carlo maximization error (meaning the difference between the unknown, exact log-likelihood at the Monte Carlo MLE compared to the unknown, exact MLE) can only be negative.
+Based on these considerations, we seek methods giving high average Monte Carlo log-likelihood at a Monte Carlo MLE, and our practical goal is to obtain reliably high log-likelihoods using calculations taking no longer than a day or so on one 36-core computing node.
+
+%\newpage
+
+\section{Varying the spatial autoregression parameter, $\spatReg$}
+
+\fontsize{9}{11.5pt plus.8pt minus .6pt}\selectfont
+
+<<v-v6_start,echo=F>>=
+readRDS("v/v6_3/initial_lik.rds") -> loglik_start_spatreg
+median_start_spatreg <- median(loglik_start_spatreg[,"start_loglik"])
+@
+
+<<spatreg_boxplot, echo=F, fig.height=3.6, fig.width=5.5, out.width="4.5in", fig.pos="h", fig.cap = paste0("Varying the IBPF spatial autoregression algorithmic parameter for shared parameters with simulated measles data.
+The log-likelihood was obtained using $M=100$ iterations of IBPF starting at random parameters (median starting log-likelihood, ",signif(median_start_spatreg,3), ").
 The horizontal dashed line denotes the log-likelihood at the true parameters.")>>=
+library(foreach)
 loglik_truth <- -38882.7 # from w1.r
@@
 results <- as.data.frame(results)
+par(mai=c(0.6,0.8,0.1,0.1))
 boxplot(loglik~sp_reg,data=results,ylim=c(-40000,loglik_max),
@@
 
+}
+
 \end{document}

### 2022-12-29 ms.Rnw
minor correction incorporated into accepted stat sinica version

@@
 %%% Statistica Sinica formatting
-%\newcommand\statSinica[1]{#1}
+% \newcommand\statSinica[1]{#1}
  \newcommand\statSinica[1]{}
@@
  \newcommand\arxiv[1]{#1}
-%\newcommand\arxiv[1]{}
+% \newcommand\arxiv[1]{}
 
@@
 The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{QD,\unit}(t)$, for $Q\in\{S,E,I,R\}$,  are modeled as conditional Poisson process with per-capita rates $\mu_{EI,\unit}$, $\mu_{IR,\unit}$ and $\mu_{QD,\unit}$ respectively, and we fix $\mu_{QD,\unit}=(50 \mbox{ year})^{-1}$.
-The birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson process with rate $\mu_{BS,\unit}(t)$, given by interpolated census data.
+A fraction $\cohort$ of births enter the susceptible cohort on the school admission day, and hence the birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson process with rate
+%\begin{equation}\nonumber
+$\mu_{BS,\unit}(t-\birthdelay) \big[
+  (1-\cohort) + \cohort \delta(t-t_a) \big]$,
+%\end{equation}
+where $\mu_{BS,\unit}(t)$ is specified by interpolated census data, $t_a=\lfloor t \rfloor + 252/365$ is the admission date for the year containing $t$, $\delta$ is a Dirac delta function, and $\birthdelay=4\, \mbox{ year}$ is a fixed delay between birth and entry into a high-transmission school community.
 
@@
 The simulated model is drawn from model {\modelA}, which is nested within model {\modelB} but not within {\modelC}.
-For models {\modelB} and {\modelC}, with all parameters are unit-specific, we must estimate $20\times 12=240$ parameters.
+For models {\modelB} and {\modelC}, with all parameters are unit-specific, we must estimate $20\times 13=260$ parameters.
 This greatly exceeds the 7 shared parameters fitted by \citet{ionides21} for a measles metapopulation model using a large computational effort.
@@
 <<sim_search_boxplot, echo=F, fig.height=3, fig.width=5.5, out.width="5in", fig.cap = "Fitting simulated measles data, using an initial search and four refinement steps.
-(A1--A5) model \\modelA, $4\\times 20$ unit-specific parameters and $8$ shared parameters; (B1--B5) model \\modelB, $12\\times 20$ unit-specific parameters and no shared parameters;  (C1--C5) model \\modelC, also all unit-specific, but with immigration rather than coupling, matching He et al (2010).
+(A1--A5) model \\modelA, $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B1--B5) model \\modelB, $13\\times 20$ unit-specific parameters and no shared parameters;  (C1--C5) model \\modelC, also all unit-specific, but with immigration rather than coupling, matching He et al (2010).
 The horizontal dashed line is the log-likelihood at the true parameters, evaluated by BPF.",fig.show = "hold">>=
@@
 The MLE is not known exactly in this case.
-Wilks' theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+8)/2=44$ for {\modelA} and $(12\times 20)/2=120$ for {\modelB}.
+Wilks' theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+9)/2=44.5$ for {\modelA} and $(13\times 20)/2=130$ for {\modelB}.
 The $y$-axis values in Fig.~\ref{fig:sim_search_boxplot} show log-likelihoods exceeding the truth by less than this, indicating some imperfection to the Monte Carlo maximization so far as Wilks' asymptotic result holds in this situation.
@@
 
+\arxiv{
+\vspace{2mm}
+\noindent{\bf Version notes}
+
+\vspace{1mm}
+
+\noindent Version [v1] was submitted to {\it Statistica Sinica}, and [v2] was accepted for publication.
+[v3] corrects for an omission of the cohort effect in the description of the model in Section~\ref{sec:model}.
+}
+
 \bibliography{bib-ibpf}
@@
 \arxiv{
+\newpage
+
 \setcounter{section}{0}
@@
   $\Sexpr{myround(theta["alpha1"],0)}$ \\
+$\cohort$& \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
+  $\Sexpr{myround(theta["cohort1"],0)}$ \\  
 $\gravity_{\unit}$ & \measlesShared & \measlesUnitSpecific & 0 &
@@
 \caption{Parameters for the measles model.
-Sub-model {\modelA} has 4 unit-specific parameters and 8 shared parameters, with movement between units following a gravity equation.
-Sub-model {\modelB} has 12 unit-specific parameters and no shared parameters, with gravity movement.
-Sub-model {\modelC} has 12 unit-specific parameters and no shared parameters, with independent immigration of infections rather than movement between units.
+Sub-model {\modelA} has 4 unit-specific parameters and 9 shared parameters, with movement between units following a gravity equation.
+Sub-model {\modelB} has 13 unit-specific parameters and no shared parameters, with gravity movement.
+Sub-model {\modelC} has 13 unit-specific parameters and no shared parameters, with independent immigration of infections rather than movement between units.
 The last column shows the parameter values used for the simulated data in Figure~2 of the main text.

### 2023-02-04 talk/talk.tex
starting a talk for Oberwolfach

%\documentclass[handout]{beamer}
\documentclass{beamer}

\usepackage{amsmath,amsfonts,amssymb,graphicx} 
\input{header.tex}
\input{header-ms.tex}
\input{theorems1.tex}

\begin{document}

\begin{frame}

\begin{center}
  {\Large\bf An iterated block particle filter for inference on coupled dynamic systems with shared and unit-specific parameters}

%% ABSTRACT: We consider inference for a collection of partially observed, stochastic, interacting, nonlinear dynamic processes. Each process is called a unit, and our primary motivation arises in biological metapopulation systems where a unit is a spatially distinct sub-population. Block particle filters are an effective tool for simulation-based likelihood evaluation for these systems, which are strongly dependent through time on a single unit and relatively weakly coupled between units. Iterated filtering algorithms can facilitate likelihood maximization for simulation-based filters. We introduce a new iterated block particle filter algorithm applicable to parameters that are either unit-specific or shared between units. We demonstrate this algorithm to carry out inference on a coupled epidemiological model for spatiotemporal measles case report data in twenty towns.

\vspace{2mm}

Edward Ionides\\
University of Michigan, Department of Statistics

\vspace{8mm}

MFO worksho on \\
Design and Analysis of Infectious Disease Studies
\\
Feb 23, 2023

\hspace{3mm}

Slides are at \url{https://ionides.github.io/talks/mfo23.pdf}
%% 8:15am EST

\vspace{8mm}

Joint work with\\
Kidus Asfaw, Ning Ning, Joonha Park, Jesse Wheeler and Aaron King

\end{center}

\end{frame}

\begin{frame}{Questions to be addressed}

  \begin{enumerate}
  \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models?
  \item When should we want to?
  \item Is there an intersection between scientific problems of interest and the capabilities of the \texttt{spatPomp} R package?
  \end{enumerate}

  \vspace{3mm}

  We introduce the iterated block particle filter because it currently appears the most powerful algorithm available in \texttt{spatPomp}.

\end{frame}

\newcommand\challengeSep{\vspace{3mm}}

\begin{frame}{Inference challenges in population dynamics}

  \begin{enumerate}
\item Combining measurement noise and process noise.
\item Including covariates in mechanistically plausible ways.
\item  Continuous time models.
\item  Modeling and estimating interactions in coupled systems.
\item  Dealing with unobserved variables.
\item  \myemph{Modeling spatial-temporal dynamics}.
\item  \myemph{Studying population dynamics via genetic sequence data}.
  \end{enumerate}

  \vspace{3mm}

  1--6 are from Bjornstad \& Grenfell ({\it Science}, 2001).\\
  7 is from Grenfell et al ({\it Science}, 2004).\\
  1--5 are largely solved, from a methodological perspective.

\end{frame}

\begin{frame}{Example: Pre-vaccination measles in England \& Wales}

\vspace{-2mm}

\begin{center}
\includegraphics[width=8.4cm]{he10-data.pdf}

\end{center}

\vspace{-2mm}

\end{frame}

\begin{frame}{Time series data, panel data \& spatiotemporal data}

  \bi
\item Looking at one unit (town) is \myemph{time series analysis}.

  \item Joint modeling of a few units (say, 2 or 3) is \myemph{multivariate time series analysis}.

\item Analysis of many time series, without consideration of dynamic interactions, is \myemph{panel data analysis}.

\item Allowing for coupling between units, we get \myemph{spatiotemporal analysis}, which in our context is \myemph{metapopulation analysis}.

  \ei

Question: When should we avoid inference for spatiotemporal models? When do we need to consider coupling? How?

\end{frame}

\begin{frame}{Desiderata}

  \begin{itemize}
    \item We want to be able to fit arbitrary dynamic models. The limitations should be our scientific creativity and the information in the data.

    \item In practice, that means using \myemph{plug-and-play} methods which need a simulator from the model but not nice closed-form expressions for densities.

    \item We want statistically efficient inference, to extract all the information in the data.

    \item In practice, that means using likelihood-based methods.

      \item  In the time series case, iterated particle filtering (IF2) implemented in the R package \texttt{pomp} enables Masters-level statisticians to do this (\url{https://ionides.github.io/531w22/}). The science may be hard, but the statistics is becoming routine.
      \end{itemize}
  \end{frame}

\begin{frame}{Panel data}

\bi
\item To investigate epidemiological dynamics in multiple cities, one can consider each city independently, perhaps modeling a background immigration rate of infections for each city.

\item \myemph{Decoupling} leads to panel data analysis, by assumption. Iterated filtering methods extend to panel data (Breto et al, {\it Journal of the American Statistical Association}, 2019).

\item We must decide which parameters should be modeled as \myemph{shared} vs \myemph{unit-specific}.

\item The consequences of decoupling are becoming easier to study with the development of statistical inference methods for coupled systems, i.e., metapopulation dynamics.

  \ei

  \end{frame}

\begin{frame}{The curse of dimensionality}

  \bi
  \item
    Particle filter (PF) methods are effective for inference on low-dimensional nonlinear partially observed stochastic dynamic systems. They scale exponentially badly.

\item Extending the successes of particle filter methods from time series data to metapopulation data is becoming possible.

\item Algorithms under consideration:\\
  {\bf
  Bagged filters (BF, IBF)\\
  Ensemble Kalman filter (EnKF, IEnKF)\\
  Guided intermediate resampling filter (GIRF, IGIRF)\\
  Block particle filter (BPF, IBPF)\\
  }

\item Filters estimate latent states and evaluate the likelihood.
\item Each filter has an iterated version which estimates parameters by repeated filtering using stochastic parameter perturbations.

\item These algorithms are all implemented in an R package, \texttt{spatPomp}.

  \ei

\end{frame}

\begin{frame}{Filtering $U$ units of a coupled measles SEIR model}

\vspace{-3mm}

\begin{center}
\includegraphics[width=10cm]{mscale_loglik_plot-1.pdf}

\end{center}

\vspace{-2mm}

Simulated data using a gravity model with geography, demography and transmssion parameters corresponding to UK pre-vaccination measles (Ionides et al, {JASA}, 2021).

%\end{center}

\end{frame}

\begin{frame}
\frametitle{$U=40$ units for a coupled measles SEIR model}

\vspace{-2.7mm}

\begin{center}
\includegraphics[width=8cm]{slice_image_plot-1.pdf}
\end{center}

\vspace{-3mm}

{\bf A}. Simulated Susceptible-Exposed-Infected-Recovered dynamics coupled with a gravity model (log of biweekly reported cases).

{\bf B}. Measles UK pre-vaccination case reports for the 40 largest cities.

%\end{center}

\end{frame}

\begin{frame}{Parameters for the measles model}
  \bi
\item Seasonal tranmission: mean and amplitude, using school term for contact rate.
  \item Durations of latency and infectious period.
\item Initial values: fraction susceptible, latent and infectious.
\item Cohort effect: all births in an age cohort start school in September.
\item Inhomogenous mixing coefficient.
\item Measurement fraction.
\item Transport model gravity constant.
\item Dynamic noise (process overdispersion).
\item Measurement overdispersion.

\ei

  \end{frame}

\begin{frame}{More on the block particle filter}

\bi
\item BPF worked quickly, easily and reliably on our measles model filtering experiments.

\item This motivated us to develop an IBPF for parameter estimation.

\item BPF has theoretical support in some situations (Rebeschini \& Van Handel, {\it Annals of Applied Probability}, 2015).

\item BPF was independently proposed as the ``factored particle filter'' by Ng et al (2002, {\it Proc. 18th Conference on Uncertainty and Artificial Intelligence}) but not widely popularized.

\ei

\end{frame}

\begin{frame}{Particle filter (PF)}

  \begin{columns}
    \begin{column}{0.48\linewidth}
      \begin{center}
      {\bf \textcolor{blue}{Evolutionary analogy}}

      \vspace{5mm}

      {\bf Mutation}

      $\downarrow$

      {\bf Fitness}

      $\downarrow$

      {\bf Natural selection}

      \end{center}
    \end{column}
     \begin{column}{0.48\linewidth}
      \begin{center}
      {\bf \textcolor{blue}{Particle filter algorithm}}

      \vspace{5mm}

      {\bf Predict: stochastic dynamics}

      $\downarrow$

      {\bf Measurement: weight}

      $\downarrow$

      {\bf Filter: resample}
      \end{center}
    \end{column}
  \end{columns}

  \vspace{15mm}

    \begin{myitemize}
  \item PF is an evolutionary algorithm with good mathematical properties: an unbiased likelihood estimate and consistent latent state distribution.
  \end{myitemize}

\end{frame}

\begin{frame}{Block particle filter (BPF)}

  \begin{columns}
    \begin{column}{0.48\linewidth}
      \begin{center}
      {\bf \textcolor{blue}{Evolutionary analogy}}

      \vspace{5mm}

      {\bf Mutation}

      $\downarrow$

      {\bf Fitness\\
      for each chromosome}

      $\downarrow$

      {\bf Natural selection\\
      for each chromosome}

      $\downarrow$

      {\bf Recombine chromosomes}

      \end{center}
    \end{column}
     \begin{column}{0.48\linewidth}
      \begin{center}
      {\bf \textcolor{blue}{Block particle filter}}

      \vspace{5mm}

      {\bf Predict: stochastic dynamics}

      $\downarrow$

      {\bf Measurement: weight\\
      for each block}

      $\downarrow$

      {\bf Filter: resample\\
      for each block}

      $\downarrow$

      {\bf Recombine blocks}
      \end{center}
    \end{column}
  \end{columns}

  \vspace{5mm}

    \begin{myitemize}
    \item Blocks in BPF allow recombination (reassortment of chromosomes in sexual reproduction) in the evolutionary analogy.
    \item Blocks are a partition of the metapopulation units. Our experiments suggest treating each sub-population (i.e., city) as a block.

  \end{myitemize}

\end{frame}

\begin{frame}

\frametitle{Measles likelihood slices for coupling parameter, $G$}

\vspace{-5mm}

\begin{columns}[T] % align columns
\begin{column}{.55\textwidth}
  \includegraphics[width=6cm]{slice_combined_plot-1.pdf}
\end{column}
\begin{column}{.52\textwidth}

  \vspace{6mm}

  Simulating $15$ year of data from $U=40$ cities for the measles model.
  Slice likelihood, varying $G$ with other paramters fixed at the truth.

  \vspace{3mm}

  {\bf A}. Evaluation using adapted bagged filter (ABF).

    \vspace{3mm}

  {\bf B}. Evaluation using block particle filter (BPF).

    \vspace{3mm}

  {\bf C}. Evaluation using EnKF.

  \vspace{3mm}

Results from Ionides et al (2021, {\it JASA}). We computed a slice due to lack of good optimization algorithms to compute a profile.

\end{column}
\end{columns}

\end{frame}

\begin{frame}{Scalability needed for practical inference}

Large numbers of parameters
  \begin{itemize}
  \item  Initial conditions will typically have to be estimated for each unit.
  \item Various dynamic parameters and measurement parameters (e.g., reporting rate) may also need to be unit-specific to obtain a statistical fit to the data.
\item Working with hundreds of estimated parameters raises additional challenges on top of the high-dimensional coupled dynamics.
  \end{itemize}

\vspace{5mm}

A moderate numbers of spatial units is enough to open new possibilities.

  \begin{itemize}
  \item As soon as dimension exceeds capabilities of a particle filter (say, $U=5$) we are in situations where likelihood-based inference for general models has been inaccessible.

  \item 10-100 coupled units is our target problem size.

  \item Larger problems will need numerical approximations (e.g., EnKF). Exact Monte Carlo methods help study the effect of these approximations.

  \end{itemize}

  \end{frame}

\begin{frame}{An iterated block particle filter for parameter estimation}

  \begin{center}

  \includegraphics[trim={0 0 0 10mm},clip,width=9cm]{IBPF_workflow.pdf}

  \end{center}

\end{frame}

\begin{frame}

\vspace{-6mm}

\begin{columns}[T] % align columns
\begin{column}{.75\textwidth}
  \includegraphics[width=6.8cm]{ibpf/search_diagnostics-1.pdf}
\end{column}

\hspace{-15mm}
\begin{column}{.35\textwidth}
  \vspace{10mm}

$\bullet$    IBPF applied to simulations and data.

    \vspace{5mm}

$\bullet$    Multiple searches from random starting points.

    \vspace{5mm}

$\bullet$    Top 25\% of searches are subsequently refined.

  \end{column}
\end{columns}

\end{frame}

\begin{frame}

\begin{columns}[T] % align columns
\begin{column}{.75\textwidth}
  \includegraphics[width=7.5cm]{ibpf/sim_search_boxplot-1.pdf}

    \includegraphics[width=7.5cm]{ibpf/data_search_boxplot-1.pdf}

  \end{column}

\hspace{-15mm}
\begin{column}{.35\textwidth}
  \vspace{5mm}

$\bullet$    IBPF applied to simulations (top panel) and data (bottom panel).

    \vspace{5mm}

$\bullet$    Multiple search refinements.

    \vspace{5mm}

    (A) Mostly shared parameters.

    \vspace{5mm}

    (B) All unit-specific parameters

    \vspace{5mm}

    (C) all unit-specific, without coupling

  \end{column}
\end{columns}

\end{frame}

\begin{frame}
\frametitle{Filtering $U$-dimensional correlated Brownian motion}

\vspace{-3mm}

\begin{center}
\includegraphics[width=10cm]{bm_alt_plot-1.pdf}

\vspace{-1mm}

$\cov\big(X_{\unit,\time}-X_{\unit,\time-1},X_{\altUnit,\time}-X_{\altUnit,\time-1}\big) \sim 0.4^{|\unit-\altUnit|}_{}$

\end{center}

\end{frame}

\begin{frame}
\frametitle{Filtering $U$ units of Lorenz 96 toy atmospheric model}

\vspace{-3mm}

\begin{center}
\includegraphics[width=10cm]{lz_loglik_plot-1.pdf}

\vspace{-1mm}

$dX_{\unit}(t) = \big \{  X_{\unit-1}(t) \big(X_{\unit+1}(t) - X_{\unit-2}(t)\big) - X_{\unit}(t) + F \big\} dt + \sigma \, dB_{\unit}(t)$

\end{center}

\end{frame}

\begin{frame}{Future work}

  \newcommand\futuresep{\vspace{3mm}}

  \begin{myitemize}
  \item We are reaching to the point where we can carry out likelihood-based inference for a flexible class of metapopulation models for measles.
Flexibility supports generation and testing of scientific hypotheses.

    \futuresep

%  \item Measles was previously a motivating model system for POMP methods for single populations.

    \futuresep

  \item Many systems in ecology, epidemiology and elsewhere could be studied in a SpatPOMP framework. Including microbiomes?

     \futuresep

   \item Modeling and inference for nonlinear stochastic dynamics is hard. But, if you can't build a quantitative statistical model then you don't understand it and you can't control it?

\end{myitemize}

\end{frame}

\nocite{bjornstad01,grenfell04,breto19,rebeschini15,ng02,ionides21}

\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{bib}
\end{frame}

%%% extra material

\end{document}

### 2023-02-04 talk/theorems1.tex
starting a talk for Oberwolfach

\newcommand\PartOfBoundOffDiagonal{3\efourB+4(K + \dmax)\,\efive}
\newcommand\boundOffDiagonal{\efourA + \PartOfBoundOffDiagonal}
\newcommand\K{K^\prime}
\newcommand\ThmOneVarBound{Q^{4b} \, \Unit\Time  \big(\altb + \etwo \, (\Unit\Time-\altb) \big)}

\newcommand\BvConstant{\mathcal{C}_0}
\newcommand\dmax{d_{\max}}
\newcommand\boundLemma{\efourB+(K+d_{\unit,\time})\efive}
\newcommand\boundLemmaUniform{\efourB+(K+\dmax)\efive}

%\newcommand\altB{\tilde{B}}
\newcommand\altB{C}
%\newcommand\el{\epsilon_\ell}
\newcommand\el{\epsilon}   %%% epsilon for Thm 1
%\newcommand\eone{\epsilon_1}
%\newcommand\eone{\epsilon^{}_{\!\mathrm{\ref{A1}}}}
\newcommand\eone{\epsilon^{}_{\!\mathrm{A1}}}
%\newcommand\eone{\epsilon^{A}_{1}}
%\newcommand\etwo{\epsilon^{}_{\!\mathrm{\ref{A:unconditional:mix}}}}
\newcommand\etwo{\epsilon^{}_{\!\mathrm{A4}}}
%\newcommand\altb{\tilde{b}}
\newcommand\altb{c}
\newcommand\Ai{
\begin{assumption}\label{A1}
There is an $\eone>0$ and a collection of neighborhoods $\{B_{\unit\comma\time}\subset A_{\unit\comma\time}, \unit\in 1\mycolon\Unit,\time\in 1\mycolon\Time\}$ such that, for all  $\unit$, $\time$, $\Unit$, $\Time$,  any  bounded real-valued function $|h(x)|\le 1$, and any value of $x_{B^c_{\unit\comma\time}}$, 
%\begin{equation}
\begin{eqnarray}
\nonumber
&& \hspace{-12mm} \Bigg| \int h(x_{\unit\comma\time}) f_{X_{\unit\comma\time}|Y_{B_{\unit\comma\time}},X_{B^c_{\unit\comma\time}}}(x_{\unit\comma\time}\given  \data{y}_{B_{\unit\comma\time}}, x_{B^c_{\unit\comma\time}}) \, dx_{\unit\comma\time}
\\
\siOnly{\label{eq:weak_coupling2}}
\msOnly{\nonumber}
&& \hspace{-3mm} - \int h(x_{\unit\comma\time})f_{X_{\unit\comma\time}|Y_{B_{\unit\comma\time}}}(x_{\unit\comma\time}\given \data{y}_{B_{\unit\comma\time}})
\, dx_{\unit\comma\time}
\Bigg| \hspace{1mm} < \hspace{1mm} \eone.
%\end{equation}
\end{eqnarray}
\end{assumption}
}

\newcommand\Aii{
\begin{assumption}\label{A1b}
For the collection of neighborhoods in Assumption~\ref{A1}, with $B^+_{\unit\comma\time}=B^{}_{\unit\comma\time}\cup (\unit,\time)$, there is a constant $\Bsize$, depending on $\eone$ but not on $\Unit$ and $\Time$, such that
\begin{equation}
\nonumber
\sup_{\unit\in 1:\Unit, \, \time\in 1:\Time} \big| B^+_{\unit\comma\time}\big| \le \Bsize.
\end{equation}
\end{assumption}
}

\newcommand\Aiii{
\begin{assumption}\label{A2}
There is a constant $Q$ such that, for all $\unit$ and $\time$, $\Unit$ and $\Time$,
\begin{equation}
\nonumber
Q^{-1} <f_{Y_{\unit\comma\time}|X_{\unit\comma\time}}(\data{y}_{\unit\comma\time}\given x_{\unit\comma\time}) < Q
\end{equation}
\end{assumption}
}

\newcommand\Aiv{
  \begin{assumption}\label{A:unconditional:mix}
    There exists $\etwo > 0$ such that the following holds.
    For each $\unit, \time$, a set $\altB^{}_{\unit\comma\time} \subset (1\mycolon U)\times(0\mycolon N)$ exists such that $(\altUnit, \altTime) \notin \altB_{\unit\comma\time}$ implies $B_{\unit\comma\time}^+ \cap B_{\altUnit\comma\altTime}^+ = \emptyset$ and
      \begin{equation} 
\siOnly{\label{eq:A:unconditional:mix}}
\msOnly{\nonumber}
        \big| f_{X_{B^{+}_{\altUnit\comma\altTime}}|X^{}_{B^{+}_{\unit\comma\time}}} - f_{X_{B^{+}_{\altUnit\comma\altTime}}} \big|
        < \etwo \, f_{X_{B^{+}_{\altUnit\comma\altTime}}}
      \end{equation}
      Further, there is a uniform bound $|\altB_{\unit\comma\time}| \le \altb$.
\end{assumption}
}

\newcommand\TheoremI{
\begin{theorem}  \label{thm:tif}
Let $\MC{\loglik}$ denote the Monte Carlo likelihood approximation constructed by {\TIF}.
Consider a limit with a growing number of islands, $\Island\to\infty$, and suppose assumptions~\ref{A1},~\ref{A1b} and~\ref{A2}.
There are quantities $\el(\Unit,\Time)$ and $V(\Unit,\Time)$, with bounds $|\el|<\eone Q^{2}$ and $V < Q^{4\Bsize} \, U^2 N^2$, such that
\begin{equation}
\label{th1:lik:bound}
\Island^{1/2}\big[ \MC{\loglik}-\loglik-\el\Unit\Time \big]  \xrightarrow[\Island \rightarrow \infty]{d} \normal\big[0,V\big],
\end{equation}
where $\xrightarrow[\Island \rightarrow \infty]{d}$ denotes convergence in distribution and $\normal[\mu,\Sigma]$ is the normal distribution with mean $\mu$ and variance $\Sigma$.
If additionally Assumption~\ref{A:unconditional:mix} holds, we obtain an improved variance bound
\begin{equation}
\label{th1:lik:bound2}
V < \ThmOneVarBound.
\end{equation}
\end{theorem}
}

%%% 2222222222222222222222

%\newcommand\ethree{\epsilon^{}_{\mathrm{\ref{B1}}}}
\newcommand\ethree{\epsilon^{}_{\mathrm{B1}}}
%\newcommand\efourA{\epsilon^{}_{\mathrm{\ref{B:unconditional:mix}}}}
\newcommand\efourA{\epsilon^{}_{\mathrm{B4}}}
%\newcommand\efourB{\epsilon^{}_{\mathrm{\ref{B:temporal:mix}}}}
\newcommand\efourB{\epsilon^{}_{\mathrm{B5}}}
%\newcommand\efive{\epsilon^{}_{\mathrm{\ref{B:girf}}}}
\newcommand\efive{\epsilon_{\mathrm{B6}}}

\newcommand\Bi{

\begin{assumptionB}\label{B1}
  There is an $\ethree>0$ and a collection of neighborhoods $\{B_{\unit\comma\time}\subset A_{\unit\comma\time}, \unit\in 1\mycolon\Unit,\time\in 1\mycolon\Time\}$ such that the following holds for all $\unit$, $\time$, $\Unit$, $\Time$, and any bounded real-valued function $|h(x)|\le 1$\emph{:} if we write $A=A_{u,n}$, $B=B_{u,n}$, $f_{A}(x_A) = f_{Y_{A}|X_{A}}(\data y_{A}|x_A)$, and $f_{B}(x_B) = f_{Y_{B}|X_{B}}(\data y_{B}|x_B)$,
%\begin{equation}
\begin{eqnarray}
\nonumber
%\label{asif:eq:weak_coupling2}
&&\hspace{-7mm}
\left| 
\int \hspace{-1mm} h(x)
\left\{
\frac{
  \E_{g} \! \big[ 
    f_{A}\big(X_{A}^P\big) f_{X_{u,n}|X_{A^{[n]}},\vec X_{n-1}} \big(x | X_{A^{[n]}}^P, \vec X_{n-1}\big) 
  \big]
} {
  \E_{g} \! \big[f_A\big(X_A^P\big)\big]
} - 
\right.
\right.
\\
\nonumber
&& 
\siOnly{\hspace{20mm}}
\hspace{-6.5mm} 
% - \hspace{-1mm} 
\siOnly{\hspace{2mm}}
\left.
\left.
\frac{
  \E_{g} \! \big[
    f_{B}\big(X_{B}^P\big) f_{X_{u,n}|X_{B^{[n]}},\vec X_{n-1}}\big(x | X_{B^{[n]}}^P, \vec X_{n-1}\big) \big] 
} {
  \E_{g} \! \big[f_B\big(X_B^P\big)\big]
} 
\!
\right\}
\!
dx
\right| \hspace{0mm} < \hspace{0mm} \ethree.
%\end{equation}
\end{eqnarray}
\end{assumptionB}
}

\newcommand\Bii{
\begin{assumptionB}\label{B1b}
The bound $\sup_{\unit\in 1:\Unit,\time\in 1:\Time} \big| B^+_{\unit\comma\time}\big| \le \Bsize$ in Assumption~\ref{A1b} applies for the neighborhoods defined in Assumption~\ref{B1}. 
This also implies there is a finite maximum temporal depth for the collection of neighborhoods, defined as
\begin{equation}
\nonumber
\dmax=\sup_{(\unit,\time)}\hspace{1mm} \sup_{(\altUnit,\altTime)\in B_{\unit,\time}}|\time-\altTime|.
\end{equation}
\end{assumptionB}
}

\newcommand\Biii{
\begin{assumptionB}\label{B2} Identically to Assumption~\ref{A2}, 
$Q^{-1} <f_{Y_{\unit\comma\time}|X_{\unit\comma\time}}(\data{y}_{\unit\comma\time}\given x_{\unit\comma\time}) < Q$.
\end{assumptionB}
}

\newcommand\BivA{
\begin{assumptionB}\label{B:unconditional:mix}
Define $g_{{X}_{A}|{X}_{B}}$ via \myeqref{eq:g}.
Suppose there is an $\efourA$ such that the following holds.
For each $\unit, \time$, a set $\altB^{}_{\unit\comma\time} \subset (1\mycolon U)\times(0\mycolon N)$ exists such that $(\altUnit, \altTime) \notin \altB_{\unit\comma\time}$ implies $B_{\unit\comma\time}^+ \cap B_{\altUnit\comma\altTime}^+ = \emptyset$ and
%      \begin{eqnarray} 
%\label{eq:B:unconditional:mix:1}
\begin{equation}
\nonumber
\big| 
   g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime}}{X}^{P}_{B^{}_{\unit\comma\time}}} 
   - g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime}}} \, g^{}_{{X}^{P}_{B^{}_{\unit\comma\time}}} 
\big|
<
%&\hspace{-3mm}<\hspace{-3mm} & 
%\\
%&&\hspace{-30mm}
(1/2) \, \efourA \, \,  g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime}}} \, g^{}_{{X}^{P}_{B^{}_{\unit\comma\time}}}
%\\
%\label{eq:B:unconditional:mix:2}
\end{equation}
\begin{eqnarray}
\nonumber
\big|
g^{}_{X^P_{B^{}_{\altUnit\comma\altTime}}|\vec{X}_{0:\Time}} 
\, g^{}_{X^P_{B^{}_{\unit\comma\time}}|\vec{X}_{0:\Time}} 
-
g^{}_{X^P_{B^{}_{\altUnit\comma\altTime}},X^P_{B^{}_{\unit\comma\time}}|\vec{X}_{0:\Time}}
\big|
&&
%&\hspace{-3mm}<\hspace{-3mm}& 
\\
\nonumber
%&&\hspace{-30mm}(1/2) \, \efourA \, \, 
&&\hspace{-30mm} < (1/2) \, \efourA \, \, 
 g^{}_{X^P_{B^{}_{\altUnit\comma\altTime}},X^P_{B^{}_{\unit\comma\time}}|\vec{X}_{0:\Time}}
      \end{eqnarray}
      Further, there is a uniform bound $|\altB_{\unit\comma\time}| \le \altb$.
\end{assumptionB}
}
\newcommand\BivB{
\begin{assumptionB}\label{B:temporal:mix}
There is a constant $K$ such that, for any set $D\subset (\UnitSet)\times(n{\mycolon}n-d)$ and for all $\time\ge K+d$, $\vec{x}^{(1)}_{\time-d-K}$, $\vec{x}^{(2)}_{\time-d-K}$, and all $x^{}_{D_n}$,
\begin{eqnarray}
\nonumber
&&\hspace{-6mm} \big|
g^{}_{{X}_{D}|\vec{X}_{\time-d-K}}(x^{}_{D}\given\vec{x}^{(1)}_{\time-d-K})
-
g^{}_{{X}_{D}|\vec{X}_{\time-d-K}}(x^{}_{D}\given\vec{x}^{(2)}_{\time-d-K})
\big|
\\
\nonumber
&& \siOnly{\hspace{10mm}} \hspace{4mm}
< 
\efourB  \,  \,
g^{}_{{X}_{D}|\vec{X}_{\time-d-K}}(x^{}_{D_n}\given\vec{x}^{(1)}_{\time-d-K})
\end{eqnarray}

\end{assumptionB}
}

\newcommand\Bv{
\begin{assumptionB}\label{B:girf}
Let $h$ be a bounded function with $|h(x)|\le 1$. 
Let $\vec{X}^{\GR}_{\time,\Ninter,\np,\island}$ be the Monte Carlo quantity constructed in ASIF-IR, conditional on $\vec{X}^{\IF}_{\time-1,\Ninter,\island}=\vec{x}^{\IF}_{\time-1,\Ninter,\island}$.
There is a constant $\BvConstant(\Unit,\Time,\Ninter)$ such that, for all $\efive^{}>0$ and $\vec{x}^{\IF}_{\time-1,\Ninter,\island}$, whenever the number of particles satisfies $J > \BvConstant(\Unit,\Time,\Ninter)/\efive^{4}$,
\begin{equation}
\nonumber
\hspace{-0.5mm}
\left|  
\, 
  \EMC\Big[ 
    \frac{1}{\Np} \sum_{\np=1}^{\Np} h(\vec{X}^{\GR}_{\time,\Ninter,\np,\island}) 
   \Big] \!
-  \E_g \! \big[h(\vec{X}_\time)\given \vec{X}_{\time-1}=\vec{x}^{\IF}_{\time-1,\Ninter,\island}\big] 
\right| \!
< \efive^{}.
\end{equation}
\end{assumptionB}
}

\newcommand\Bvi{
\begin{assumptionB}\label{B:XG_XA_ind}
For $1\leq \time \leq N$, the Monte Carlo random variable $X^A_{\time,i}$ is independent of $w^M_{\unit,\time,i,j}$ conditional on $X^A_{\time-1,i}$.
\end{assumptionB}
}

\newcommand\TheoremII{
\begin{theorem}  \label{thm:asif}
Let $\MC{\loglik}$ denote the Monte Carlo likelihood approximation constructed by ASIF-IR, or by ASIF since this is the special case of ASIF-IR with $\Ninter=1$.
Consider a limit with a growing number of islands, $\Island\to\infty$, and suppose assumptions~\ref{B1},~\ref{B1b}, \ref{B2}, \ref{B:temporal:mix}, \ref{B:girf} and~\ref{B:XG_XA_ind}. 
Suppose the number of particles $\Np$ exceeds the requirement for~\ref{B:girf}.
There are quantities $\el(\Unit,\Time)$ and $V(\Unit,\Time)$ with $|\el|<
Q^2\ethree + 2 Q^{2\Bsize}\big(\boundLemma\big)$
and $V<Q^{4\Bsize}\Unit^2\Time^2$ 
such that
\begin{equation}
\siOnly{\label{th:asif:lik:bound}}
\msOnly{\nonumber}
\Island^{1/2}
  \big[ 
    \MC{\loglik}-\loglik-\el\Unit\Time 
  \big]  
  \xrightarrow[\Island \rightarrow \infty]{d} 
  \normal\big[ 0,V \big].
\end{equation}
If additionally Assumption~\ref{B:unconditional:mix} holds, we obtain an improved rate of
\begin{equation}
\siOnly{\label{th:asif:lik:bound2}}
\msOnly{\nonumber}
V < Q^{4\Bsize}\Time\Unit
\big\{ c + \big(\boundOffDiagonal \big) \big(\Time\Unit-c\big)
\big\}
\end{equation}
\end{theorem}
}

### 2023-02-11 talk/talk.tex
a few typos and tweaks

@@
 
-MFO worksho on \\
+MFO workshop on \\
 Design and Analysis of Infectious Disease Studies
@@
 
-  \vspace{3mm}
+  \vspace{10mm}
 
-  We introduce the iterated block particle filter because it currently appears the most powerful algorithm available in \texttt{spatPomp}.
+  We introduce the {\bf iterated block particle filter} because it is currently the most powerful algorithm available in \texttt{spatPomp}.
 
@@
 \item  \myemph{Modeling spatial-temporal dynamics}.
-\item  \myemph{Studying population dynamics via genetic sequence data}.
+\item  Studying population dynamics via genetic sequence data.
   \end{enumerate}
@@
   1--6 are from Bjornstad \& Grenfell ({\it Science}, 2001).\\
-  7 is from Grenfell et al ({\it Science}, 2004).\\
+  7 is an active topic, but not the focus of this talk.\\
   1--5 are largely solved, from a methodological perspective.

### 2023-02-12 talk/talk.tex
start to add haiti slides
haiti results

@@
 
+\begin{frame}{Cholera in Haiti, 2010-2019}
+
+% https://web.archive.org/web/20151106110552/http://www.ihsi.ht/pdf/projection/Estimat_PopTotal_18ans_Menag2015.pdf
+%Map	Department	Capital	Area (km2)	Population	Density (Pop./km2)	Planning Region
+%1	Artibonite	Gonaïves	4,987	1,727,524	350	Central
+%2	Centre	        Hinche	        3,487	746,236	        210	Central
+%3	Grand'Anse	Jérémie	        1,912	468,301	        240	South
+%4	Nippes	        Miragoâne	1,268	342,525	        270	South
+%5	Nord	        Cap-Haïtien	2,115	1,067,177	500	North
+%6	Nord-Est	Fort-Liberté	1,623	393,967	        240	North
+%7	Nord-Ouest	Port-de-Paix	2,103	728,807	        350	Central
+%8	Ouest	        Port-au-Prince	4,983	4,029,705	810	West
+%9	Sud-Est	        Jacmel	        2,034	632,601	        310	West
+%10	Sud	        Les Cayes	2,654	774,976	        290	South
+
+  % Hurricane Matthew: October 2016
+
+\begin{columns}[T] % align columns
+
+%\hspace{-15mm}
+\begin{column}{.35\textwidth}
+  \vspace{5mm}
+
+$\bullet$  820,000 reported cases and nearly 10,000 deaths
+
+    \vspace{5mm}
+
+$\bullet$  Population of 10,000,000 in 10 d\'{e}partements
+
+    \vspace{5mm}
+
+    $\bullet$  Lee et al (2020) developed 3 models to guide vaccination plans:
+
+    \vspace{1mm}
+
+    {\bf 1}. stochastic, national \\
+    {\bf 2}. ODE, spatial \\
+    %d\'{e}partment \\
+    {\bf 3}. stochastic, spatial
+    %d\'{e}partment.
+
+\end{column}
+
+\begin{column}{.62\textwidth}
+  \includegraphics[width=7.5cm]{haiti/haiti.png}
+
+  \vspace{-3mm}
+
+  \begin{center}
+    {  \tiny \url{https://doi.org/10.5194/nhess-20-471-2020}}
+  \end{center}
+
+\end{column}
+
+\end{columns}
+
+  \vspace{4mm}
+
+   $\bullet$ Wheeler et al (2023) continued the data analysis
+
+\end{frame}
+
+\begin{frame}{Haiti cholera weekly reports 2010-2019 (log scale)}
+\includegraphics[width=12cm]{haiti/Plot_Reported_Cases-1.pdf}
+\end{frame}
+
+\begin{frame}{Log-likelihoods of models and an ARMA benchmark}
+
+  \begin{center}
+\begin{tabular}{|l|rc|rc|rc|}
+\hline
+& Model~1& (p) & Model~2 & (p) & Model~3 & (p)
+\\
+\hline
+Wheeler et al
+& { -2731.3} & (15)
+& -21957.3 & (6)
+& { -17850.4} & (35)
+\\
+Lee et al
+& -3030.9 & (20)
+& -29367.4 & (6)
+& -31840.8 & (29)
+\\
+Log-ARMA(2,1)
+& -2803.7 & (4)  %% -0.5*5615.3+4
+& { -18027.0} & (40)  %% -0.5*36133.9+40
+& -18027.0 & (40)
+\\
+\hline
+\end{tabular}
+  \end{center}
+
+  \vspace{3mm}
+
+  Model 1 (stochastic, aggregated): \\
+  We add process overdispersion, and beat the ARMA benchmark.
+
+  \vspace{5mm}
+
+  Model 2 (deterministic, spatial): \\
+  We use a log-normal measurement model. ODE models fall far short of the ARMA benchmark.
+
+    \vspace{5mm}
+
+    Model 3 (stochastic, spatial): \\
+    (We fit the model using IBPF, and beat the ARMA benchmark.
+
+\end{frame}
+
 \begin{frame}
@@
 
-\nocite{bjornstad01,grenfell04,breto19,rebeschini15,ng02,ionides21}
+\nocite{bjornstad01,grenfell04,breto19,rebeschini15,ng02,ionides21,wheeler23,lee20,asfaw23arxiv}
 

### 2023-02-13 talk/talk.tex
edit measles slides and start covid

@@
 Joint work with\\
-Kidus Asfaw, Ning Ning, Joonha Park, Jesse Wheeler and Aaron King
+Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King
 
@@
   \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models?
+
+    \vspace{2mm}
+
   \item When should we want to?
+
+    \vspace{2mm}
+
   \item Is there an intersection between scientific problems of interest and the capabilities of the \texttt{spatPomp} R package?
@@
 \item  Dealing with unobserved variables.
-\item  \myemph{Modeling spatial-temporal dynamics}.
+\item  \myemph{Modeling spatiotemporal dynamics}.
 \item  Studying population dynamics via genetic sequence data.
@@
 
-  \vspace{3mm}
+  \vspace{4mm}
 
-  1--6 are from Bjornstad \& Grenfell ({\it Science}, 2001).\\
-  7 is an active topic, but not the focus of this talk.\\
-  1--5 are largely solved, from a methodological perspective.
+  1--5 are largely solved, from a methodological perspective.\\
+  7 is exciting but not the focus of this talk.\\
+  6 is our topic.
+
+  \vspace{4mm}
+
+  \resizebox{12cm}{!}{Reviews: Bjornstad \& Grenfell ({\it Science}, 2001); Grenfell et al ({\it Science}, 2004)}
 
@@
 
+\vspace{3mm}
+
 Question: When should we avoid inference for spatiotemporal models? When do we need to consider coupling? How?
@@
   \begin{itemize}
-    \item We want to be able to fit arbitrary dynamic models. The limitations should be our scientific creativity and the information in the data.
+    \item Consideration of arbitrary dynamic models. The limitations should be our scientific creativity and the information in the data.
 
-    \item In practice, that means using \myemph{plug-and-play} methods which need a simulator from the model but not nice closed-form expressions for densities.
+      \vspace{2mm}
 
-    \item We want statistically efficient inference, to extract all the information in the data.
+     Hence, \myemph{plug-and-play} methods which need a simulator from the model but not nice closed-form expressions for densities.
 
-    \item In practice, that means using likelihood-based methods.
+    \item Statistically efficient inference, to extract all the information in the data.
+
+      \vspace{2mm}
+
+    Hence, likelihood-based methods.
 
@@
 \item Measurement fraction.
-\item Transport model gravity constant.
+\item Population movement model: gravity constant.
 \item Dynamic noise (process overdispersion).
@@
 
-\begin{frame}{More on the block particle filter}
+\begin{frame}{More on the block particle filter (BPF)}
 
@@
 \end{column}
+
+\hspace{-5mm}
 \begin{column}{.52\textwidth}
@@
 
-  Simulating $15$ year of data from $U=40$ cities for the measles model.
-  Slice likelihood, varying $G$ with other paramters fixed at the truth.
+  Simulating $15$ year of data from $U=40$ cities.
+  Slice likelihood, varying $G$ with other parameters fixed.\\
+   (Ionides et al, {\it JASA}, 2021).
 
@@
   \vspace{3mm}
-
-Results from Ionides et al (2021, {\it JASA}). We computed a slice due to lack of good optimization algorithms to compute a profile.
+  A much easier task than computing a profile likelihood for real data.
 
@@
 \end{columns}
+
+\end{frame}
+
+\begin{frame}{Comments on the Ensemble Kalman Filter (EnKF)}
+
+  \begin{itemize}
+  \item EnKF is more dependent on approximate Gaussianity than is sometimes supposed.
+
+    \vspace{2mm}
+
+  \item The Gaussian-inspired update rule is similar to the extended Kalman filter (EKF), which has largely been superseded by particle filter methods for low-dimensional nonlinear biological dynamics.
+
+    \vspace{2mm}
+
+\item Simple systems can defeat EnKF: the linear Gaussian update is helpless when data inform the conditional variance rather than the conditional mean.
+
+  \vspace{2mm}
+
+    \item Big systems need computationally tractable analysis. EnKF may sometimes be the best solution available, but be aware of its limitations.
+
+  \end{itemize}
 
@@
 
-  \item Larger problems will need numerical approximations (e.g., EnKF). Exact Monte Carlo methods help study the effect of these approximations.
+  \item Larger problems will need numerical approximations (e.g., EnKF).
 
@@
 \begin{columns}[T] % align columns
-\begin{column}{.75\textwidth}
+\hspace{-4mm}\begin{column}{.75\textwidth}
   \includegraphics[width=7.5cm]{ibpf/sim_search_boxplot-1.pdf}
@@
 
-\hspace{-15mm}
-\begin{column}{.35\textwidth}
+\hspace{-18mm}
+\begin{column}{.4\textwidth}
   \vspace{5mm}
 
-$\bullet$    IBPF applied to simulations (top panel) and data (bottom panel).
+  {\bf IBPF for measles, 20 cities}\\
+
+  \vspace{1mm}
+
+  $\bullet$ simulations (top panel)\\
+  dash line: likelihood at truth
+
+  \vspace{2mm}
+
+  $\bullet$ data (bottom panel)\\
+  dash line: MLE for case C.
 
@@
 
-$\bullet$    Multiple search refinements.
+%$\bullet$    Multiple search refinements.
+
+%    \vspace{5mm}
+
+    (A) 9 shared parameters, $4\times 20$ unit-specific parameters
 
@@
 
-    (A) Mostly shared parameters.
+    (B) all parameters unit-specific
 
@@
 
-    (B) All unit-specific parameters
+    (C) all unit-specific, with constant immigration instead of coupling
+
+  \end{column}
+\end{columns}
+
+\end{frame}
+
+\begin{frame}{Inferences from the previous slide}
+
+  \begin{itemize}
+    %$\bullet$
+  \item
+IBPF can work for simulated data, since it finds likelihoods higher than the truth when the model is well-specified
 
@@
 
-    (C) all unit-specific, without coupling
+  \item
+    %$\bullet$
+    When coupled by a gravity model, simulated data show evidence against a constant immigration model.
 
-  \end{column}
-\end{columns}
+    \vspace{5mm}
+
+  \item 
+    %$\bullet$
+    Real data do not favor this gravity model over a constant immigration model.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{He et al (2010) parameters; $r$ is rank correlation with $N_{1950}$}
+
+\resizebox{12cm}{!}{
+\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r}
+\hline
+  & $N_{1950}$ & $R_0$ & IP & LP & $\alpha$ & $a$ & $\iota$ & $\psi$ & $\rho$ & $\sigma_{SE}$\\
+\hline
+Halesworth & 2200 & 33.00 & 2.30 & 7.90 & 0.95 & 0.38 & 0.0091 & 0.64 & 0.75 & 0.075\\
+\hline
+Lees & 4200 & 30.00 & 2.10 & 8.50 & 0.97 & 0.15 & 0.0310 & 0.68 & 0.61 & 0.080\\
+\hline
+Mold & 6400 & 21.00 & 1.80 & 5.90 & 1.00 & 0.27 & 0.0140 & 2.90 & 0.13 & 0.054\\
+\hline
+Dalton in Furness & 11000 & 28.00 & 2.00 & 5.50 & 0.99 & 0.20 & 0.0390 & 0.82 & 0.46 & 0.078\\
+\hline
+Oswestry & 11000 & 53.00 & 2.70 & 10.00 & 1.00 & 0.34 & 0.0300 & 0.48 & 0.63 & 0.070\\
+\hline
+Northwich & 18000 & 30.00 & 3.00 & 8.50 & 0.95 & 0.42 & 0.0600 & 0.40 & 0.80 & 0.086\\
+\hline
+Bedwellty & 29000 & 25.00 & 3.00 & 6.80 & 0.94 & 0.16 & 0.0400 & 0.95 & 0.31 & 0.061\\
+\hline
+Consett & 39000 & 36.00 & 2.70 & 9.10 & 1.00 & 0.20 & 0.0730 & 0.41 & 0.65 & 0.071\\
+\hline
+Hastings & 66000 & 34.00 & 5.40 & 7.00 & 1.00 & 0.30 & 0.1900 & 0.40 & 0.70 & 0.096\\
+\hline
+Cardiff & 240000 & 34.00 & 3.10 & 9.90 & 1.00 & 0.22 & 0.1400 & 0.27 & 0.60 & 0.054\\
+\hline
+Bradford & 290000 & 32.00 & 3.40 & 8.50 & 0.99 & 0.24 & 0.2400 & 0.19 & 0.60 & 0.045\\
+\hline
+Hull & 300000 & 39.00 & 5.50 & 9.20 & 0.97 & 0.22 & 0.1400 & 0.26 & 0.58 & 0.064\\
+\hline
+Nottingham & 310000 & 23.00 & 3.70 & 5.70 & 0.98 & 0.16 & 0.1700 & 0.26 & 0.61 & 0.038\\
+\hline
+Bristol & 440000 & 27.00 & 4.90 & 6.20 & 1.00 & 0.20 & 0.4400 & 0.20 & 0.63 & 0.039\\
+\hline
+Leeds & 510000 & 48.00 & 11.00 & 9.50 & 1.00 & 0.27 & 1.2000 & 0.17 & 0.67 & 0.078\\
+\hline
+Sheffield & 520000 & 33.00 & 6.40 & 7.20 & 1.00 & 0.31 & 0.8500 & 0.18 & 0.65 & 0.043\\
+\hline
+Manchester & 700000 & 33.00 & 6.90 & 11.00 & 0.96 & 0.29 & 0.5900 & 0.16 & 0.55 & 0.055\\
+\hline
+Liverpool & 800000 & 48.00 & 9.80 & 7.90 & 0.98 & 0.30 & 0.2600 & 0.14 & 0.49 & 0.053\\
+\hline
+Birmingham & 1100000 & 43.00 & 12.00 & 8.50 & 1.00 & 0.43 & 0.3400 & 0.18 & 0.54 & 0.061\\
+\hline
+London & 3400000 & 57.00 & 13.00 & 13.00 & 0.98 & 0.55 & 2.9000 & 0.12 & 0.49 & 0.088\\
+\hline
+$\qquad r$ & 1 & 0.46 & 0.95 & 0.32 & 0.11 & 0.30 & 0.9300 & -0.93 & -0.20 & -0.330\\
+\hline
+\end{tabular}
+}
 
@@
 
+% From Li et al (2020) page S-14:
+% According to official report, around 5 million people left Wuhan city before January 23.
+The total population in Wuhan city after January 23 (when travel restrictions were imposed) is therefore around 6 million.
+
+\begin{frame}{COVID in 375 cities in China, 10-23 January, 2020}
+
+  \begin{itemize}
+  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness, using an iterated EnKF \\
+    (Li et al, {\it Science}, May 2020).
+   \item
+    The time interval predates Wuhan travel restrictions.
+    %
+%     load("data/incidence.rda")
+%    li20dat <- incidence[1:14,-1]
+%     tot <- sort(apply(li20dat,2,sum))
+%     sum(tot)         
+   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
+   \item 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
+\end{itemize}
+
+\end{frame}
+
 \begin{frame}{Future work}
 
-  \newcommand\futuresep{\vspace{3mm}}
+  \newcommand\futuresep{\vspace{5mm}}
 
   \begin{myitemize}
-  \item We are reaching to the point where we can carry out likelihood-based inference for a flexible class of metapopulation models for measles.
-Flexibility supports generation and testing of scientific hypotheses.
+
+  \item We anticipate continued progress in modeling and inference for nonlinear spatiotemporal partially observed stochastic dynamic systems.
 
@@
 
-%  \item Measles was previously a motivating model system for POMP methods for single populations.
+  \item The \texttt{spatPomp} package facilitates data analysis.
 
@@
 
-  \item Many systems in ecology, epidemiology and elsewhere could be studied in a SpatPOMP framework. Including microbiomes?
-
-     \futuresep
-
-   \item Modeling and inference for nonlinear stochastic dynamics is hard. But, if you can't build a quantitative statistical model then you don't understand it and you can't control it?
+    \item Questions and comments?
 

### 2023-02-14 talk/talk.tex
various talk updates

@@
 
-Joint work with\\
+Joint work with
 Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King
@@
   1--5 are largely solved, from a methodological perspective.\\
-  7 is exciting but not the focus of this talk.\\
-  6 is our topic.
+  6 is our immediate topic.\\
+  7 is exciting but not the focus of this talk.
 
@@
 
-\begin{frame}{Example: Pre-vaccination measles in England \& Wales}
+\begin{frame}{Pre-vaccination weakly measles in England \& Wales}
 
@@
 
+     \vspace{2mm}
+
     \item Statistically efficient inference, to extract all the information in the data.
@@
 
-    Hence, likelihood-based methods.
+    Hence, \myemph{likelihood-based} methods.
 
-      \item  In the time series case, iterated particle filtering (IF2) implemented in the R package \texttt{pomp} enables Masters-level statisticians to do this (\url{https://ionides.github.io/531w22/}).
+    \vspace{2mm}
+
+      \item  For time series analysis, iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do this (\url{https://ionides.github.io/531w22/}).
 The science may be hard, but the statistics is becoming routine.
@@
 
-\item \myemph{Decoupling} leads to panel data analysis, by assumption. Iterated filtering methods extend to panel data (Breto et al, {\it Journal of the American Statistical Association}, 2019).
+  \vspace{2mm}
 
-\item We must decide which parameters should be modeled as \myemph{shared} vs \myemph{unit-specific}.
+\item \myemph{Decoupling} leads to panel data analysis, by assumption.
 
-\item The consequences of decoupling are becoming easier to study with the development of statistical inference methods for coupled systems, i.e., metapopulation dynamics.
+  \vspace{2mm}
+
+  \item Iterated filtering methods extend to panel data (Breto et al, {\it Journal of the American Statistical Association}, 2019).
+
+    \vspace{2mm}
+
+\item We must decide which parameters should be modeled as \myemph{shared} (i.e., the same for each unit) vs \myemph{unit-specific}.
+
+%\item The consequences of decoupling are becoming easier to study with the development of statistical inference methods for coupled systems, i.e., metapopulation dynamics.
 
@@
   \item
-    Particle filter (PF) methods are effective for inference on low-dimensional nonlinear partially observed stochastic dynamic systems. They scale exponentially badly.
+    Particle filter (PF) methods fail for high-dimensional systems. They scale exponentially badly.
 
-\item Extending the successes of particle filter methods from time series data to metapopulation data is becoming possible.
+    \vspace{2mm}
 
-\item Algorithms under consideration:\\
+  \item Algorithms with improved scalability include:\\
+
+    \vspace{1mm}
+
   {\bf
@@
 
+      \vspace{2mm}
+
 \item Filters estimate latent states and evaluate the likelihood.
+
+    \vspace{2mm}
+
   \item Each filter has an iterated version which estimates parameters by repeated filtering using stochastic parameter perturbations.
 
-\item These algorithms are all implemented in an R package, \texttt{spatPomp}.
+        \vspace{2mm}
+
+\item These algorithms are all implemented in the \texttt{spatPomp} R package.
 
@@
 
+        \vspace{2mm}
+
+\item BPF has theoretical support in some situations (Rebeschini \& Van Handel, {\it Annals of Applied Probability}, 2015).
+
+        \vspace{2mm}
+
       \item This motivated us to develop an IBPF for parameter estimation.
 
-\item BPF has theoretical support in some situations (Rebeschini \& Van Handel, {\it Annals of Applied Probability}, 2015).
+        \vspace{2mm}
+
+      \item IBPF has theoretical guarantees similar to BPF (Ning \& Ionides, {\it arXiv}, 2023).
+
+        \vspace{2mm}
 
@@
 
-\begin{frame}{He et al (2010) parameters; $r$ is rank correlation with $N_{1950}$}
+\begin{frame}{He et al (2010) estimates. $r$ is rank correlation with $N_{1950}$}
 
@@
 \hline
-$\qquad r$ & 1 & 0.46 & 0.95 & 0.32 & 0.11 & 0.30 & 0.9300 & -0.93 & -0.20 & -0.330\\
+$\qquad r$ & 1 & 0.46 & \cellcolor{yellow!50} 0.95 & 0.32 & 0.11 & 0.30 & 0.9300 & -0.93 & -0.20 & -0.330\\
 \hline
@@
 Wheeler et al
-& { -2731.3} & (15)
+& \cellcolor{yellow!50} -2731.3 & \cellcolor{yellow!50}(15)
 & -21957.3 & (6)
-& { -17850.4} & (35)
+& \cellcolor{yellow!50} -17850.4 & \cellcolor{yellow!50}(35)
 \\
@@
 \\
-Log-ARMA(2,1)
+Log-ARMA(2,1) \hspace{-2.5mm}
 & -2803.7 & (4)  %% -0.5*5615.3+4
-& { -18027.0} & (40)  %% -0.5*36133.9+40
+& \cellcolor{yellow!50} { -18027.0} & \cellcolor{yellow!50} (40)  %% -0.5*36133.9+40
 & -18027.0 & (40)
@@
     Model 3 (stochastic, spatial): \\
-    (We fit the model using IBPF, and beat the ARMA benchmark.
+    We fit the model using IBPF, and beat the ARMA benchmark.
 
+\end{frame}
+
+\begin{frame}{Evidence for a decreasing transmission rate in Model 1}
+\includegraphics[width=12cm]{haiti/Beta_trend_Figure-1.pdf}
 \end{frame}
@@
 
-\nocite{bjornstad01,grenfell04,breto19,rebeschini15,ng02,ionides21,wheeler23,lee20,asfaw23arxiv}
+\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23}
 

### 2023-02-15 talk/talk.tex
more on covid example

@@
 
-\vspace{-3mm}
+\vspace{-1mm}
 
@@
 
-\vspace{-2mm}
+\vspace{-4mm}
 
@@
     The time interval predates Wuhan travel restrictions.
-    %
-%     load("data/incidence.rda")
-%    li20dat <- incidence[1:14,-1]
-%     tot <- sort(apply(li20dat,2,sum))
-%     sum(tot)         
+%% see README for R code
    \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
-   \item 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
+   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
+   \item{} 0 cases reported for all cities during Jan 10--15.
+
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Top: data. Middle: IBPF fit. Bottom: IEnKF fit} 
+\includegraphics[width=12cm]{covid/3-panel.pdf}
+\end{frame}
+
+\begin{frame}{Technical issues with EnKF}
+
+  \begin{itemize}
+  \item EnKF is based on a continuous Gaussian approximation.
+  \item Log-likelihoods with respect to counting measure are never positive, and cannot properly be compared with log-likelihoods corresponding to continuous densities.
+    \item For data with many zeros, the unbounded EnKF likelihood can substantially bias the MLE.
+    \item Adapting EnKF for count data is non-trivial (Katzfuss et al, {\it JASA}, 2019).
+    \item Li et al avoided the issue by using a ``heuristic observation error variance'' within EnKF.
+The corresponding ``log-likelihood'' is not the log-likelihood of any model.
+In particular, it cannot validly be compared againt simple statistical benchmarks.
+  \item Data and models will never be perfect. We want to reveal, rather than conceal, limitations.
   \end{itemize}
@@
 
-\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23}
+\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23}
 

### 2023-02-15 talk/covid/p.r

@@ -1,0 +1,121 @@
+
+# mle from metapop/code_and_data/plot/mif_global_nv1.csv
+par1 <- c(
+  alpha_be1=0.093898173,
+  Beta_be1=0.779439748,
+  alpha_af1=0.461829062,
+  Beta_af1=0.134595027,
+  mu_be1=0.966823556,
+  Z_be1=0.116508318,
+  D_be1=0.934673622,
+  mu_af1=0.686401734,
+  Z_af1=0.044448782,
+  D_af1=0.126463381,
+  theta1=0.89626861,
+  tau1=0.917180577,
+  size_a1=0.550023483/6, # adaptation for new code
+  size_b1=0.056420797/6, # adaptation for new code
+  Td_be1=0.8,
+  Td_af1=0.2,
+  E_01=0.44362243,
+  Iu_01=0.984098384
+)
+
+parli1 <- c(
+  alpha_be1=0.14,
+  Beta_be1=1.12,
+  alpha_af1=0.00001,
+  Beta_af1=0.000001,
+  mu_be1=0.55,
+  Z_be1=3.69,
+  D_be1=3.47,
+  mu_af1=0.0000001,
+  Z_af1=0.000001,
+  D_af1=0.000001,
+  theta1=1.36,
+  tau1=0.0000001,
+  size_a1=0.550023483/6, # adaptation for new code
+  size_b1=0.056420797/6, # adaptation for new code
+  Td_be1=0.8,
+  Td_af1=0.2,
+  E_01=0.5,
+  Iu_01=0.5,
+)
+
+#run_level <- 1
+#Np <-switch(run_level,5, 500, 1e3)
+#mif <-switch(run_level, 3, 20, 40)
+#Nbpf <-switch(run_level, 3, 20, 40)
+# Nrep <-switch(run_level, 3,100,100)
+# U <-switch(run_level, 5,50,375)
+
+library(metapoppkg)
+library(doParallel)
+# plot covid model
+
+library(doRNG)
+library(patchwork)
+library(ggplot2)
+
+m1=li20(U=375,overdispersion_dynamic = T, overdispersion_measurement = T)
+
+## early epidemic
+if(0){
+  e1 <- li20(U=100,overdispersion_dynamic = T, overdispersion_measurement = T)
+  e1@times <- e1@times + 9 # to match days in January
+  plot(window(e1,16,23),ty="h",log=T,plot_unit=F)
[…]

### 2023-02-23 talk/talk.tex
talk for mfo23

@@
   \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models?
-
-    \vspace{2mm}
-
-  \item When should we want to?
+(Of particular interest are metapopulation models.)
 
@@
   \item Is there an intersection between scientific problems of interest and the capabilities of the \texttt{spatPomp} R package?
+
+        \vspace{2mm}
+
+\item We introduce the {\bf iterated block particle filter} because it is currently the most powerful algorithm available in \texttt{spatPomp}.
+
+  \vspace{2mm}
+
+\item Bonus question: How do we know if our model is statistically adequate, or needs more work?
+
     \end{enumerate}
-
-  \vspace{10mm}
-
-  We introduce the {\bf iterated block particle filter} because it is currently the most powerful algorithm available in \texttt{spatPomp}.
-
 \end{frame}
@@
 
+  %% full weekly data now available:
+  %% Korevaar, Hannah, C.
+Jessica Metcalf, and Bryan T.
+Grenfell.
+"Structure, space and size: competing drivers of variation in urban and rural measles transmission." Journal of the Royal Society Interface 17, no.
+168 (2020): 20200010.
+  %% https://royalsocietypublishing.org/doi/pdf/10.1098/rsif.2020.0010
+  %% data at https://rs.figshare.com/collections/Supplementary_material_from_Structure_space_and_size_competing_drivers_of_variation_in_urban_and_rural_measles_transmission_/5036567/1
+
 \vspace{-2mm}
@@
 
-\vspace{3mm}
+%\vspace{3mm}
 
-Question: When should we avoid inference for spatiotemporal models? When do we need to consider coupling? How?
+%Question: When should we avoid inference for spatiotemporal models? When do we need to consider coupling? How?
 
@@
 
-    \vspace{2mm}
-
-      \item  For time series analysis, iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do this (\url{https://ionides.github.io/531w22/}).
-The science may be hard, but the statistics is becoming routine.
       \end{itemize}
@@
 
-\begin{frame}{Panel data}
+\begin{frame}{Fitting mechanistic models to time series}
+
+    \vspace{8mm}
+
+    \begin{itemize}
+    \item  Iiterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference:
+
+      \vspace{2mm}
+
+      \url{https://ionides.github.io/531w22/}
+
+      \vspace{5mm}
+      \item
+      The science may be hard, but the statistics is becoming routine.
+
+    % examples: COVID etc, bison in Yellowstone National Park, volatility in the stock market, gun purchases, human proximal small intestine pH
+
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{Fitting mechanistic models to panel data}
 
@@
 
-  \item Iterated filtering methods extend to panel data (Breto et al, {\it Journal of the American Statistical Association}, 2019).
+  \item Iterated filtering methods extend to panel data (\texttt{panelPomp}, Breto et al, {\it JASA}, 2019).
 
@@
 
-  \item Each filter has an iterated version which estimates parameters by repeated filtering using stochastic parameter perturbations.
+  \item Iterated filters estimate parameters using stochastic parameter perturbations.
 
@@
   \bi
-\item Seasonal tranmission: mean and amplitude, using school term for contact rate.
+\item Seasonal transmission: mean and amplitude, using school term for contact rate.
   \item Durations of latency and infectious period.
@@
 \item Cohort effect: all births in an age cohort start school in September.
-\item Inhomogenous mixing coefficient.
+\item Inhomogenous mixing exponent, $I^{\alpha}$.
 \item Measurement fraction.
 \item Population movement model: gravity constant.
-\item Dynamic noise (process overdispersion).
+\item Dynamic noise: process overdispersion.
 \item Measurement overdispersion.
@@
 \ei
+\vspace{3mm}
+(Adding gravity coupling to the model of He et al, {\it JRSI}, 2010)
+%\vspace{5mm}
+%Plug-and-play methods let you add or subtract features
 
@@
 
-\item BPF was independently proposed as the ``factored particle filter'' by Ng et al (2002, {\it Proc. 18th Conference on Uncertainty and Artificial Intelligence}) but not widely popularized.
+\item BPF was independently proposed as the ``factored particle filter'' by Ng et al (2002).
 
@@
 
+  \begin{myitemize}
+    \item Blocks are a partition of the metapopulation units.
+    \item For measles, we use each city as a block.
+  \end{myitemize}
+
   \begin{columns}
@@
     \begin{myitemize}
-    \item Blocks in BPF allow recombination (reassortment of chromosomes in sexual reproduction) in the evolutionary analogy.
-    \item Blocks are a partition of the metapopulation units. Our experiments suggest treating each sub-population (i.e., city) as a block.
+    \item Blocks are segments of the full state which can be reassorted between particles at the resampling step.
 
@@
 
-  \item 10-100 coupled units is our target problem size.
+  \item 5-500 coupled units is our target problem size.
 
@@
 \begin{frame}
+
+%%% A: 9 shared parameters; 4 unit-specific
+%%% B: all unit-specific
 
@@
 
+%%% Wilks' Theorem: improvement on truth is about (1/2) chisq(p)
+
 \begin{itemize}
-    %$\bullet$
   \item
@@
   \item
-    %$\bullet$
     When coupled by a gravity model, simulated data show evidence against a constant immigration model.
@@
 \item 
-    %$\bullet$
-    Real data do not favor this gravity model over a constant immigration model.
+Data do not favor this gravity model over a constant immigration model.
+
+%%% An underwhelming sounding result, but a necessary step to make progress,
+%%% and previously we did not know how to do this.
+
 \end{itemize}
@@
 \begin{column}{.35\textwidth}
-  \vspace{5mm}
+  \vspace{2mm}
 
-$\bullet$  820,000 reported cases and nearly 10,000 deaths
+  $\bullet$  820,000 reported cases
 
-    \vspace{5mm}
+  \vspace{4mm}
+
+  $\bullet$ nearly 10,000 deaths
+
+    \vspace{4mm}
 
@@
 
-    \vspace{5mm}
+    \vspace{4mm}
 
-    $\bullet$  Lee et al (2020) developed 3 models to guide vaccination plans:
+    $\bullet$  Lee et al (2020) developed 4 models to guide vaccination plans:
+
+    %%% Elizabeth Lee, Johns Hopkins
 
@@
     {\bf 2}. ODE, spatial \\
-    %d\'{e}partment \\
-    {\bf 3}. stochastic, spatial
-    %d\'{e}partment.
+    {\bf 3}. stochastic, spatial\\
+    {\bf 4}. agent based
 
@@
 
-\begin{frame}
-\frametitle{Filtering $U$-dimensional correlated Brownian motion}
-
-\vspace{-3mm}
-
-\begin{center}
-\includegraphics[width=10cm]{bm_alt_plot-1.pdf}
-
-\vspace{-1mm}
-
-$\cov\big(X_{\unit,\time}-X_{\unit,\time-1},X_{\altUnit,\time}-X_{\altUnit,\time-1}\big) \sim 0.4^{|\unit-\altUnit|}_{}$
-
-\end{center}
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Filtering $U$ units of Lorenz 96 toy atmospheric model}
-
-\vspace{-3mm}
-
-\begin{center}
-\includegraphics[width=10cm]{lz_loglik_plot-1.pdf}
-
-\vspace{-1mm}
-
-$dX_{\unit}(t) = \big \{  X_{\unit-1}(t) \big(X_{\unit+1}(t) - X_{\unit-2}(t)\big) - X_{\unit}(t) + F \big\} dt + \sigma \, dB_{\unit}(t)$
-
-\end{center}
-
-\end{frame}
-
 % From Li et al (2020) page S-14:
+% Ruiyun Li, Imperial (Oslo as of 2021)
 % According to official report, around 5 million people left Wuhan city before January 23.
@@
   \begin{itemize}
-  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness, using an iterated EnKF \\
+  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
     (Li et al, {\it Science}, May 2020).
+  \item SEIR model with asymptomatics, reporting delay, and coupling based on cell phone data.
+  \item Inference using iterated EnKF.
+
+    \vspace{5mm}
    \item
@@
 
+\begin{frame}
+\frametitle{Filtering $U$-dimensional correlated Brownian motion}
+
+\vspace{-3mm}
+
+\begin{center}
+\includegraphics[width=10cm]{bm_alt_plot-1.pdf}
+
+\vspace{-1mm}
+
+$\cov\big(X_{\unit,\time}-X_{\unit,\time-1},X_{\altUnit,\time}-X_{\altUnit,\time-1}\big) \sim 0.4^{|\unit-\altUnit|}_{}$
+
+\end{center}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Filtering $U$ units of Lorenz 96 toy atmospheric model}
+
+\vspace{-3mm}
+
+\begin{center}
+\includegraphics[width=10cm]{lz_loglik_plot-1.pdf}
+
+\vspace{-1mm}
+
+$dX_{\unit}(t) = \big \{  X_{\unit-1}(t) \big(X_{\unit+1}(t) - X_{\unit-2}(t)\big) - X_{\unit}(t) + F \big\} dt + \sigma \, dB_{\unit}(t)$
+
+\end{center}
+
+\end{frame}
+
 \nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23}

### 2023-02-23 talk/talk.tex
a few talk edits

@@
 
-\begin{frame}{Pre-vaccination weakly measles in England \& Wales}
+\begin{frame}{Pre-vaccination weekly measles in England \& Wales}
 
@@
 
+  %%% Haiti and Dominican Republic share Hispaniola
+
 % https://web.archive.org/web/20151106110552/http://www.ihsi.ht/pdf/projection/Estimat_PopTotal_18ans_Menag2015.pdf
@@
     \item Adapting EnKF for count data is non-trivial (Katzfuss et al, {\it JASA}, 2019).
-    \item Li et al avoided the issue by using a ``heuristic observation error variance'' within EnKF.
+%    \item Li et al avoided the issue by using a ``heuristic observation error variance'' within EnKF.
 The corresponding ``log-likelihood'' is not the log-likelihood of any model.
 In particular, it cannot validly be compared againt simple statistical benchmarks.
-  \item Data and models will never be perfect. We want to reveal, rather than conceal, limitations.
+%  \item Data and models will never be perfect. We want to reveal, rather than conceal, limitations.
   \end{itemize}
@@
 
-  \item The \texttt{spatPomp} package facilitates data analysis.
+  \item As long as the \texttt{spatPomp} package facilitates data analysis, we expect continuing package development.
 
-    \futuresep
+%    \futuresep
 
-    \item Questions and comments?
+ %   \item Questions and comments?
 

### 2023-03-30 talk/talk.tex
draft of tamu talk
typo

@@
 We demonstrate this algorithm to carry out inference on a coupled epidemiological model for spatiotemporal measles case report data in twenty towns.
+We discuss other applications, to cholera in Haiti and COVID-19 in China.
 
@@
 
-MFO workshop on \\
-Design and Analysis of Infectious Disease Studies
-\\
-Feb 23, 2023
+Texas A\&M Statistics Colloquium\\
+Apr 21, 2023
 
@@
 
-Slides are at \url{https://ionides.github.io/talks/mfo23.pdf}
-%% 8:15am EST
+Slides are at \url{https://ionides.github.io/talks/tamu23.pdf}
 
@@
 
-      \item IBPF has theoretical guarantees similar to BPF (Ning \& Ionides, {\it arXiv}, 2023).
+      \item IBPF has theoretical guarantees similar to BPF (Ning \& Ionides, {\it JMLR}, 2023).
 

### 2023-04-03 ms.Rnw
copy editing updates

@@
 \noindent {\it Abstract:}
-We consider inference for a collection of partially observed, stochastic, interacting, nonlinear dynamic processes.
-Each process is identified with a label called its unit, and our primary motivation arises in biological metapopulation systems where a unit corresponds to a spatially distinct sub-population.
-Metapopulation systems are characterized by strong dependence through time within a single unit and relatively weak interactions between units, and these properties make block particle filters an effective tool for simulation-based likelihood evaluation.
+We consider inference for a collection of partially observed stochastic interacting nonlinear dynamic processes.
+Each process is identified with a label, called its unit.
+Here, our primary motivation arises in biological metapopulation systems, in which a unit corresponds to a spatially distinct sub-population.
+Metapopulation systems are characterized by strong dependence over time within a single unit, and relatively weak interactions between units.
+These properties make block particle filters effective for simulation-based likelihood evaluation.
 Iterated filtering algorithms can facilitate likelihood maximization for simulation-based filters.
-We introduce an iterated block particle filter applicable when parameters are unit-specific or shared between units.
-We demonstrate this algorithm by performing inference on a coupled epidemiological model describing spatiotemporal measles case report data for twenty towns.
+Here, we introduce an iterated block particle filter that can be applied when parameters are unit-specific or shared between units.
+We demonstrate the proposed algorithm by performing inference on a coupled epidemiological model describing spatiotemporal measles case report data for 20 towns.
 
@@
 
-Statistical inference for high-dimensional partially-observed nonlinear dynamic systems arises in various scientific contexts.
-Massive models and datasets are considered in the geophysical sciences, carried out under the name of data assimilation \citep{evensen09book}.
-Population models in ecology and epidemiology can be characterized by high levels of stochasticity, nonlinearity, measurement error, and model uncertainty, leading to challenges of a somewhat different nature than geophysical models.
-In addition, biological population systems may sometimes have a low population count due to a local introduction or fade-out of one or more constituent species.
-Low population situations may require consideration of integer-valued counts rather than continuous population approximations.
+Statistical inference for high-dimensional partially observed nonlinear dynamic systems arises in various scientific contexts.
+Massive models and data sets are considered in the geophysical sciences, carried out under the name of data assimilation \citep{evensen09book}.
+Population models in ecology and epidemiology can be characterized by high levels of stochasticity, nonlinearity, measurement error, and model uncertainty, leading to challenges of a somewhat different nature to those of geophysical models.
+In addition, biological population systems may have a low population count, owing to a local introduction or fade-out of one or more constituent species.
+Such situations may require models with integer-valued counts, rather than continuous population approximations.
 Collections of biological populations measured at different spatial locations may have spatial interactions in addition to local population dynamics; such collections are called a metapopulation.
@@
 
-Statistical inference for partially observed nonlinear biological systems was, until recently, an open methodological challenge even in the time series case \citep{bjornstad01}.
-Advances in Monte Carlo methods based on the particle filter have made inference accessible on many low-dimensional problems \citep{doucet11,kantas15,king16} but the curse of dimensionality \citep{bengtsson08} prevents applicability of the basic particle filter on metapopulations with more than a few units.
+Until recently, statistical inference for partially observed nonlinear biological systems was an open methodological challenge, even in the time series case \citep{bjornstad01}.
+Advances in Monte Carlo methods based on particle filters have made inference accessible in many low-dimensional problems \citep{doucet11,kantas15,king16}.
+However, the curse of dimensionality \citep{bengtsson08} prevents application of the basic particle filter on metapopulations with more than a few units.
 Methods based on improving the proposal distribution for the particle filter may not fully resolve the curse of dimensionality \citep{snyder15}.
-Situations where previous Monte Carlo methods can provably beat the curse of dimensionality may have limited applicability \citep{beskos17,park20,ionides21}.
-Consequently, state-of-the-art scientific analysis for metapopulation dynamics has depended on ensemble Kalman filter (EnKF) methods \citep{li20}.
-EnKF algorithms scale well but are founded on an approximation that can be unsuitable for discrete populations with fade-out and re-introduction dynamics and other highly non-Gaussian features \citep{ionides21}.
-
-In Sec.~\ref{sec:alg}, we propose an algorithm for inference on metapopulation dynamics which we call an iterated block particle filter (IBPF).
-IBPF combines the iterated filtering likelihood maximization technique of \citet{ionides15} with the block particle filter (BFP) of \citet{rebeschini15}.
+Previous Monte Carlo methods which can provably beat the curse of dimensionality may have limited applicability \citep{beskos17,park20,ionides21}.
+Consequently, state-of-the-art scientific analysis for metapopulation dynamics depends on ensemble Kalman filter (EnKF) methods \citep{li20}.
+EnKF algorithms scale well, but are founded on an approximation that can be unsuitable for discrete populations with fade-out and re-introduction dynamics and other highly non-Gaussian features \citep{ionides21}.
+
+In Section~\ref{sec:alg}, we propose an algorithm for inference on metapopulation dynamics, which we call an iterated block particle filter (IBPF).
+IBPF algorithms combine an iterated filtering likelihood maximization technique \citep{ionides15} with a block particle filter (BPF) \citep{rebeschini15}.
 Iterated filtering algorithms use parameter perturbations to coerce a filtering algorithm into exploring the parameter space.
-BPF addresses the curse of dimensionality by modifying the resampling step of a particle filter to resample independently on blocks that form a partition of the collection of units.
-A previous IBPF algorithm was developed by \citet{ning21-ibpf} for the particular case that all estimated parameters are unit-specific, which is to say that the dynamics and measurement process for a unit $u$ are determined by a vector of parameters $\psi_u$ specific to unit $u$.
-The formal meaning of this assertion is given, together with the pseudocode for our algorithm, in Sec.~\ref{sec:alg}.
-We propose an extension of their IBPF which additionally permits the estimation of a vector of shared parameters, $\phi$.
-In this case, the full parameter vector is $\theta=(\phi,\psi_{1:U})$ where the $U$ units are named $\{1,\dots,\Unit\}$ which we denote by $\seq{1}{\Unit}$.
-\citet{ning21-ibpf} developed theoretical justification for their algorithm, but our extension of IBPF currently carries only empirical support.
-
-There may be scientific interest concerning which parameters in metapopulation systems are best understood as unit-specific, and which can reasonably be modeled as shared between units.
-Another relevant possibility is that a parameter may differ between units as a shared function of unit-specific covariates, which formally is a special case of a shared parameter.
-Addressing these issues is also a prerequisite for studying questions about the coupling of metapopulation systems via model-based inference from spatiotemporal data.
-We will show empirically, in Sec.~\ref{sec:data}, that our {\ibpf} algorithm is applicable to an inference challenge in epidemiological metapopulation dynamics.
-Our demonstration focuses on a dataset for weekly measles incidence in 20 towns in the United Kingdom (UK) during the pre-vaccination era \citep{he10} modeled using a previously studied metapopulation model \citep{park20,ionides21}.
-Measles case reports have been a longstanding benchmark problem for inference on biological dynamics, motivating the development of time series methodology and, more recently, the progression from single populations to metapopulation systems.
-Unlike previous attempts on sequential Monte Carlo inference for metapopulation models, we show that our algorithm can provide practical plug-and-play, likelihood-based inference when parameters are either shared between units or differ between units.
-Demonstrating a solution to this open problem provides numerical evidence substituting for numerical comparisons against alternative methods.
+BPF algorithms address the curse of dimensionality by modifying the resampling step of a particle filter to resample independently on blocks that form a partition of the collection of units.
+A previous IBPF algorithm was developed by \citet{ning21-ibpf} for the particular case in which all estimated parameters are unit-specific, that is, the dynamics and measurement process for a unit $u$ are determined by a vector of parameters $\psi_u$ specific to unit $u$.
+In Section~\ref{sec:alg}, we provide a formal meaning of this assertion, together with the pseudocode for our algorithm.
+We propose an extension of the aforementioned IBPF which additionally allows us to estimate a vector of shared parameters, $\phi$.
+In this case, the full parameter vector is $\theta=(\phi,\psi_{1:U})$, where the $U$ units are named $\{1,\dots,\Unit\}$, which we denote by $\seq{1}{\Unit}$.
+\citet{ning21-ibpf} develop a theoretical justification for their algorithm; however, our new IBPF currently relies on empirical support only.
+
+There may be scientific interest in which parameters in a metapopulation system are best understood as being unit-specific, and which can reasonably be modeled as being shared between units.
+Another relevant possibility is that a parameter may differ between units as a shared function of unit-specific covariates.
+Formally, this is a special case of a shared parameter.
+Addressing these issues is also a prerequisite for studying questions about the coupling of metapopulation systems using model-based inference from spatiotemporal data.
+In Section~\ref{sec:data}, we will show empirically that our {\ibpf} algorithm is applies to an inference challenge in epidemiological metapopulation dynamics.
+Our demonstration focuses on a data set of weekly measles incidence in 20 towns in the United Kingdom (UK) during the pre-vaccination era \citep{he10}, modeled using a previously studied metapopulation model \citep{park20,ionides21}.
+Measles case reports are a longstanding benchmark problem for inference on biological dynamics, motivating the development of time series methodologies and, more recently, the progression from single populations to metapopulation systems.
+Unlike previous attempts on sequential Monte Carlo inference for metapopulation models, we show that our algorithm can provide practical plug-and-play, likelihood-based inference when the parameters are either shared between units or differ between units.
+Demonstrating a solution to this open problem provides numerical evidence substituting for numerical comparisons with alternative methods.
 
@@
 Rather, we demonstrate steps toward this research goal.
-We use a simulation study, in Sec.~\ref{sec:sim}, to show that our methodology can deliver a good approximation to the maximum likelihood estimate when fitting the model used to simulate the data.
-This reassurance allows us to interpret our data analysis results as evidence of model misspecification, providing a guide for future investigations of these data as well as tools to carry out those investigations.
-
-Optimization of high-dimensional, non-convex and potentially multi-modal functions, evaluated using stochastic methods, is not straightforward even when it is possible to evaluate the function within an acceptable level of error.
-Therefore, we discuss approaches that assist the noisy likelihood searches, and suggest diagnostic plots to assess their success.
-
-\section{The {\ibpf} algorithm for likelihood maximization}
+We use a simulation study, discussed in Section~\ref{sec:sim}, to show that our methodology can deliver a good approximation to the maximum likelihood estimate (MLE) when fitting the model used to simulate the data.
+This allows us to interpret our data analysis results as evidence of model misspecification, providing a guide for future investigations of these data, as well as tools to carry out such investigations.
+
+Optimization of high-dimensional, non-convex and potentially multi-modal functions, evaluated using stochastic methods, is not straightforward, even when it is possible to evaluate the function within an acceptable level of error.
+Therefore, we discuss approaches that assist noisy likelihood searches, and suggest diagnostic plots to assess their success.
+
+\section{An {\ibpf} algorithm for likelihood maximization}
 \label{sec:alg}
@@
 We define set-valued subscripts by $X_A=\{X_a,a\in A\}$ and $X_{A,B}=\{X_{a,b}, a\in A, b\in B\}$.
-The discrete time process $\myvec{X}_{0:\Time}$ may arise from a continuous time Markov process $\{\myvec{X}(t), t_0\le t\le t_{\Time} \}$ observed at times $t_{1:\Time}$, and in this case we set $\myvec{X}_\time=\myvec{X}(t_\time)$.
+The discrete time process $\myvec{X}_{0:\Time}$ may arise from a continuous-time Markov process $\{\myvec{X}(t), t_0\le t\le t_{\Time} \}$ observed at times $t_{1:\Time}$, in which case, we set $\myvec{X}_\time=\myvec{X}(t_\time)$.
 The initial value $\myvec{X}_{0}$ may be stochastic or deterministic.
-Observations are made on each unit, modeled by an observable process $\myvec{Y}_{1:\Time}=Y_{1:\Unit,1:\Time}$ which takes values  at each time $\time$ in a product space $\Yspace^\Unit$.
-Observations are modeled as conditionally independent given the latent process.
-The conditional independence of measurements applies over both time and the unit structure, so the collection $\big\{Y_{\unit\comma\time},\unit\in\seq{1}{\Unit},\time\in\seq{1}{\Time}\big\}$ is conditionally independent given
+Observations are made on each unit, modeled by an observable process $\myvec{Y}_{1:\Time}=Y_{1:\Unit,1:\Time}$ that takes values  at each time $\time$ in a product space $\Yspace^\Unit$.
+Observations are modeled as conditionally independent, given the latent process.
+The conditional independence of measurements applies over both time and the unit structure; thus, the collection $\big\{Y_{\unit\comma\time},\unit\in\seq{1}{\Unit},\time\in\seq{1}{\Time}\big\}$ is conditionally independent, given
 $\big\{X_{\unit\comma\time},\unit\in\seq{1}{\Unit},\time\in\seq{1}{\Time}\big\}$.
 We suppose the existence of a joint density $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$  for $X_{1:\Unit,0:\Time}$ and $Y_{1:\Unit,1:\Time}$ with respect to some appropriate measure, following a notational convention that the subscripts of $f$ denote the joint or conditional density under consideration.
-We suppose that $f$ depends on a real-valued parameter vector $\theta=(\phi,\psi_{1:U})$, which we write as $\theta=\theta_{1:D}$ when we wish to concatenate the component shared and unit-specific parameters into a single vector of length $D$.
-The data are $\data{y}_{\unit\comma\time}$ for unit $\unit$ at time $\time$.
-This model is a special case of a  partially observed Markov process (POMP) also known as a state space model or hidden Markov model.
-The additional unit structure, not generally required for a POMP, is appropriate for modeling interactions between units characterized by a spatial location, and so we call the model a SpatPOMP.
-For metapopulation models, the units are not generally arranged on a spatial grid but comprise a collection of spatially distributed population centers.
-
-A numerical challenge of fundamental statistical relevance is the maximization of the log-likelihood function of the data given the model, $\loglik(\theta)=\log f_{\myvec{Y}_{1:\Time}}(\data{\myvec{y}}_{1:\Time}\giventh \theta)$.
+We suppose that $f$ depends on a real-valued parameter vector $\theta=(\phi,\psi_{1:U})$, which we write as $\theta=\theta_{1:D}$ when we wish to concatenate the shared and unit-specific parameters into a single vector of length $D$.
+The data are $\data{y}_{\unit\comma\time}$, for unit $\unit$ at time $\time$.
+This model is a special case of a  partially observed Markov process (POMP), also known as a state-space model or hidden Markov model.
+The additional unit structure, not generally required for a POMP, is appropriate for modeling interactions between units characterized by a spatial location; thus, we call the model a SpatPOMP.
+For metapopulation models, the units are not, in general, arranged on a spatial grid, but instead comprise a collection of spatially distributed population centers.
+
+A numerical challenge of fundamental statistical relevance is maximizing the log-likelihood function of the data, given the model, $\loglik(\theta)=\log f_{\myvec{Y}_{1:\Time}}(\data{\myvec{y}}_{1:\Time}\giventh \theta)$.
 Numerical evaluation of the likelihood function is closely related to the filtering problem of evaluating $f_{\myvec{X}_n|\myvec{Y}_{1:n}}(\myvec{x}_n\given \data{\myvec{y}}_{1:n}\giventh \theta)$.
 If the dynamic model is extended to include the parameters as latent variables, the filtering problem leads to the Bayesian posterior distribution, though regularization is required to make the calculation numerically tractable using Monte Carlo methods \citep{kitagawa98,janeliu01}.
-Iterating this Bayesian calculation recursively targets a maximum likelihood estimate (MLE), a strategy known as data cloning \citep{lele07,lele10}.
-Adding noise to perturb the parameters in the extended model at each time point can stabilize the numerics while retaining the capability to approximate the MLE \citep{ionides06-pnas,ionides11,ionides15}.
+Iterating this Bayesian calculation recursively targets an MLE, a strategy known as data cloning \citep{lele07,lele10}.
+Adding noise to perturb the parameters in the extended model at each time point can stabilize the numerics, while still being able to approximate the MLE \citep{ionides06-pnas,ionides11,ionides15}.
 Many variations on this idea have been developed using different filter methods \citep{park20,li20,ionides21,manoli15} or employing different perturbation systems \citep{doucet13,nguyen17}.
 
-A direct approach to iterating BPF for parameter estimation is to resample the extended model independently on each block, giving rise to separate collections of parameters for each block.
-\citet{ning21-ibpf} proved that this IBPF algorithm targets the MLE for the special case where each parameter is localized to an individual unit, i.e., when all parameters are unit-specific.
-Formally, we say that a parameter for a discrete-time SpatPOMP is specific to a unit $u$ if it is involved in specifying the measurement density $f_{Y_{u,n}|X_{u,n}}$ or transition density $f_{X_{u,n+1}|\myvec{X}_n}$ for some $n$, and it is not involved in $f_{Y_{v,m}|X_{v,m}}$ or $f_{X_{v,m+1}|\myvec{X}_m}$ for any $v\neq u$ and any $m$.
-For a continuous-time SpatPOMP, we replace the requirement on $f_{X_{u,n+1}|\myvec{X}_n}$ with an equivalent requirement on a numerical solution over a small time increment $\delta$, as in equation \eqref{eq:extension} below.
-A parameter which is not unit-specific is said to be shared between units.
-There are intermediate possibilities, where a parameter is shared for only a subset of all units, but such situations are not explored further here.
-The special case where all parameters are unit-specific may occur sometimes, but models typically have some shared parameters which arise in transition densities and/or measurement densities for multiple units.
-
-Our approach to iterated filtering for shared parameters involves constructing an extended model within which the shared parameters is represented as a unit-specific parameter that happens to be constant across units.
-We construct a spatiotemporally extended model by supposing that a numerical solution for the transition from time $t$ to time $t+\delta$ for each unit $u$ has a functional form
+A direct approach to iterating a BPF for parameter estimation is to resample the extended model independently on each block, yielding separate collections of parameters for each block.
+\citet{ning21-ibpf} prove that this IBPF algorithm targets the MLE for the special case where each parameter is localized to an individual unit, that is, when all parameters are unit-specific.
+Formally, we say that a parameter for a discrete-time SpatPOMP is specific to a unit $u$ if it is involved in specifying the measurement density $f_{Y_{u,n}|X_{u,n}}$ or the transition density $f_{X_{u,n+1}|\myvec{X}_n}$, for some $n$, and it is not involved in $f_{Y_{v,m}|X_{v,m}}$ or $f_{X_{v,m+1}|\myvec{X}_m}$, for any $v\neq u$ and any $m$.
+For a continuous-time SpatPOMP, we replace the requirement on $f_{X_{u,n+1}|\myvec{X}_n}$ with an equivalent requirement on a numerical solution over a small time increment $\delta$, as in equation \eqref{eq:extension}.
+A parameter that is not unit-specific is said to be shared between units.
+There are intermediate possibilities, where a parameter is shared for only a subset of all units, but such a parameter is formally classified as shared.
+The special case in which all parameters are unit-specific may occur, but models typically have some shared parameters that arise in transition densities and/or measurement densities for multiple units.
+
+In our approach to iterated filtering for shared parameters, we construct an extended model, within which, the shared parameters are represented as a unit-specific parameter that happens to be constant across units.
+We construct a spatiotemporally extended model by supposing that a numerical solution for the transition from time $t$ to time $t+\delta$ for each unit $u$ has the functional form
 \begin{equation}
@@
 \end{equation}
-where the random vector $\myvec{\eta}_t=\eta_{1:\Unit,t}$ is shared for all $u\in \seq{1}{U}$ and does not depend on $\theta$.
+where the random vector $\myvec{\eta}_t=\eta_{1:\Unit,t}$ is shared, for all $u\in \seq{1}{U}$, and does not depend on $\theta$.
 If a representation \eqref{eq:extension} exists, an extended model is defined by replacing $\phi$ with $\phi_{\unit}(t)$ and $\psi_u$ with $\psi_{u}(t)$.
-Equation \eqref{eq:extension} implicitly defines a continuous-time extended model by the limit as $\delta\to 0$, when that limit exists, but for simulation-based methods the numerical solution is of more immediate concern than this limit.
-Admitting a minor abuse of notation, we subsequently use a density $f$ to denote both the original model and its extension for spatiotemporally varying parameters, with the context determining which one is intended.
-
-In some situations, the extended model could have problematic properties.
-For example, it could break conservation laws obeyed by the original system.
-In other situations, the extended model may make scientific sense in its own right.
-For example, in biological metapopulation systems it might be scientifically meaningful to consider a model where there is variation over space and time in the parameters describing the local dynamics.
-Here, we are focusing on the hypothesis that there are some parameters that are fixed across space and time, but the specification in \eqref{eq:extension} requires this hypothesis to be nested within a more flexible alternative.
+Equation \eqref{eq:extension} implicitly defines a continuous-time extended model by the limit as $\delta\to 0$, when that limit exists, but for simulation-based methods, a numerical solution is of more immediate concern than this limit.
+Admitting a minor abuse of notation, we subsequently use the density $f$ to denote both the original model and its extension for spatiotemporally varying parameters, with the context determining which one is intended.
+
+In some situations, the extended model may be problematic; for example, it could break conservation laws obeyed by the original system.
+In other situations, the extended model may make scientific sense in its own right; for example, in biological metapopulation systems, it might be scientifically meaningful to consider a model in which there is variation over space and time in the parameters that describe the local dynamics.
+Here, we focus on the hypothesis that some parameters are fixed across space and time, but the specification in \eqref{eq:extension} requires that this hypothesis be nested within a more flexible alternative.
 
@@
   {\bf Inputs}:
-  simulator for the extended model, $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1}\giventh\myvec{\theta})$ and initialization, $f_{\myvec{X}_0}(\myvec{x}_0\giventh\myvec{\theta})$;
+  simulator for the extended model, $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1}\giventh\myvec{\theta})$, and initialization, $f_{\myvec{X}_0}(\myvec{x}_0\giventh\myvec{\theta})$;
   evaluator for $f_{{Y}_{\unit,\time}|{X}_{\unit,\time}}({y}_{\unit,\time}\given {x}_{\unit,\time}\giventh\theta)$;
@@
 
-The IBPF algorithm described in Alg.~\ref{alg:ibpfilter} carries out a block particle filter on this extended model.
-The extended parameters are given independent perturbations, but a spatial autoregressive step pulls the values of shared parameters toward their mean over the units.
-Indeed, this autoregressive step is the only difference between  Alg.~\ref{alg:ibpfilter} and the algorithm for unit-specific parameters proposed by \citet{ning21-ibpf}.
-This algorithm, in turn, is essentially the IF2 algorithm of \citet{ionides15} with the particle filter replaced by BPF.
-
-Alg.~\ref{alg:ibpfilter} assumes implicit loops over $\np\ \text{in}\ \seq{1}{\Np}$ and $\unit\ \text{in}\ \seq{1}{\Unit}$,
+The IBPF algorithm described in Algorithm~\ref{alg:ibpfilter} carries out a BPF on this extended model.
+The extended parameters are given independent perturbations, but a spatial autoregressive step pulls the values of the shared parameters toward their mean over the units.
+Indeed, this autoregressive step is the only difference between  Algorithm~\ref{alg:ibpfilter} and the algorithm for unit-specific parameters proposed by \citet{ning21-ibpf}.
+This algorithm, in turn, is essentially the IF2 algorithm of \citet{ionides15}, with the particle filter replaced with the BPF of \citet{rebeschini15}.
+
+Algorithm~\ref{alg:ibpfilter} assumes implicit loops over $\np\ \text{in}\ \seq{1}{\Np}$ and $\unit\ \text{in}\ \seq{1}{\Unit}$,
 $\normal(\mu,\Sigma)$ denotes the normal distribution with mean $\mu$ and variance matrix $\Sigma$, and
 $\sigma_\time$ is a $D\times D$ diagonal matrix with entries $\sigma_{d,\time}$.
-The blocks, $\blocklist_{1:\Block}$, are a partition of $\seq{1}{\Unit}$.
-
-The pseudocode in Alg.~\ref{alg:ibpfilter} represents our implementation of {\ibpf} as the R function \code{ibpf} which we have contributed to the open-source package \code{spatPomp} \citep{asfaw21github,asfaw21arxiv}.
-Additionally, the source code for all the results in this article is available at \url{https://github.com/ionides/ibpf_article}.
+The blocks $\blocklist_{1:\Block}$ are a partition of $\seq{1}{\Unit}$.
+
+The pseudocode in Algorithm~\ref{alg:ibpfilter} represents our implementation of an {\ibpf} as the R function \code{ibpf}, which we have contributed to the open-source package \code{spatPomp} \citep{asfaw21github,asfaw21arxiv}.
+Additionally, the source code for all results presented here is available at \url{https://github.com/ionides/ibpf_article}.
 Various generalizations of this implementation are possible.
-For example, iterated filtering theory does not rely on parameter perturbations following the normal distribution \citep{ionides15} though in practice we transform parameters to facilitate this convenient choice (Sec.~\supSecAlg).
+For example, iterated filtering theory does not rely on parameter perturbations following the normal distribution \citep{ionides15}, though, in practice, we transform the parameters to facilitate this convenient choice (see Section~\supSecAlg).
 
@@
 
-Model parameters are optimized on a transformed scale for which unit variation is scientifically meaningful.
-In practice, this means working with positive parameters on a log scale, and $(0,1)$ interval-valued parameters on a logistic scale.
+The model parameters are optimized on a transformed scale for which unit variation is scientifically meaningful.
+In practice, this means working with positive parameters on a log scale, and with $(0,1)$ interval-valued parameters on a logistic scale.
 We follow standard iterated filtering practice by using independent random walks for each parameter on this transformed scale \citep{king16}.
-We discovered that the large number of parameters following a random walk, in the presence of unit-specific parameters, can require considerably smaller random walk standard deviations than the values around $\sigma_{d,n}=0.02$ (i.e., 2\% perturbation per time point) that have been employed for iterated filtering of time series models.
-After experimentation, we used $\sigma_{d,n}=0.005$ for the initial search and $\sigma_{d,n}=0.00125$ for subsequent refinements.
-Occasionally, a parameter which can be estimated precisely from the data can benefit substantially from a smaller perturbation.
-This is the case for one parameter in our measles analysis, an exponent $\alpha$ for which the scale of uncertainty is an order of magnitude smaller than other parameters; therefore, $\sigma_{d,n}$ for this parameter was scaled accordingly.
-In principle, $\sigma_{d,n}$ can be a function of $n$ as well as the parameter, $d$.
-The most common reason for using this flexibility is to avoid perturbing parameters during time intervals when there is no information about them: following an evolutionary analogy, evolution cannot operate effectively when there is mutation but no selection.
-For an initial value parameter, meaning one which specifies only the latent process value at the initial time $t_0$, we use $\sigma_{d,n}=0$ for $n\ge 1$ and we doubled the value of $\sigma_{d,n}$ for $n=0$.
-
-We used $J=4000$ particles, and we set the cooling rate parameter to be $a=0.5$, corresponding to a 1\% reduction in the random walk standard deviation at each iteration.
-We set $M=100$ optimizations iterations, chosen as an empirically assessed compromise between effort spent on each search and the number of searches conducted.
-Additional discussion of algorithmic parameters is given in Sec.~\supSecAlg.
+We find that the large number of parameters following a random walk, in the presence of unit-specific parameters, can require considerably smaller random walk standard deviations than the values around $\sigma_{d,n}=0.02$ (i.e., 2\% perturbation per time point) that have been employed for iterated filtering of time series models.
+After experimentation, we used $\sigma_{d,n}=0.005$ for the initial search, and $\sigma_{d,n}=0.00125$ for subsequent refinements.
+Occasionally, a parameter that can be estimated precisely from the data can benefit substantially from a smaller perturbation.
+This is the case for one parameter in our measles analysis, an exponent $\alpha$, for which the scale of uncertainty is an order of magnitude smaller than that of the other parameters; therefore, we scaled $\sigma_{d,n}$ for this parameter accordingly.
+In principle, $\sigma_{d,n}$ can be a function of $n$ and the parameter, $d$.
+The most common reason for using this flexibility is to avoid perturbing parameters during time intervals in which we have no information about these parameters.
+Following an evolutionary analogy, evolution cannot operate effectively if there is mutation, but no selection.
+For an initial value parameter, that specifies only the latent process value at the initial time $t_0$, we use $\sigma_{d,n}=0$ for $n\ge 1$, and we double the value of $\sigma_{d,n}$, for $n=0$.
+
+We use $J=4000$ particles, and set the cooling rate parameter to $a=0.5$, corresponding to a 1\% reduction in the random walk standard deviation at each iteration.
+We set $M=100$ optimization iterations, chosen as an empirically assessed compromise between the effort spent on each search and the number of searches conducted.
+An additional discussion of algorithmic parameters is given in Section~\supSecAlg.
 The only additional algorithmic parameter over previous iterated filtering algorithms is the spatial autoregressive parameter, $\spatReg$.
-Numerical experimentation suggests that the performance is not sensitive to the choice of $\spatReg>0$ (Supplementary Fig.~\supFigSpatReg).
-Subsequently, we used $\spatReg=0.1$.
-
-Applying IBPF to real data forces us to address issues of model development, model misspecification, and performance in the presence of outliers.
-Before tackling these issues, we use simulated data to demonstrate the capabilities of the methodology on a correctly specified model.
+The results of our numerical experiments suggest that the performance is not sensitive to the choice of $\spatReg>0$ (see the Supplementary Material, Figure~\supFigSpatReg).
+Consequently, here we use $\spatReg=0.1$.
+
+Applying an IBPF to real data forces us to address problems related to model development, model misspecification, and performance in the presence of outliers.
+Before doing so, we use simulated data to demonstrate the capabilities of the methodology on a correctly specified model.
 
@@
   ggplot2::ylab("log10(reported cases)")+
-  ggplot2::xlab("Year")
+  ggplot2::xlab("Year")+
+  theme(text=element_text(family="serif"))
 print(g)
@@
 
-\section{Testing {\ibpf} on a measles transmission model}
+\section{Testing an IBPF on a measles transmission model}
 \label{sec:measles}
 
-Measles transmission has provided a useful example of epidemiological dynamics (and therefore also ecological dynamics, for a host-pathogen ecosystem), with plentiful case report data and relatively simple biology.
-Model-based analysis of measles time series data led to advances in understanding of the seasonality of infectious disease \citep{fine82}, critical community size \citep{bartlett57}, the recognition that relatively simple mechanistic models can provide a remarkably good description of the dynamics \citep{earn00}, and much other foundational research on disease dynamics.
-Some progress has been on building and fitting spatiotemporal models for measles; e.g., \citet{xia04,eggo11,bjornstad19,becker20}.
+Measles transmission is a useful example of epidemiological dynamics (and therefore also ecological dynamics, for a host-pathogen ecosystem), with plentiful case report data and relatively simple biology.
+Model-based analyses of measles time series data have led to a better understanding of the seasonality of infectious diseases \citep{fine82}, critical community size \citep{bartlett57}, the recognition that relatively simple mechanistic models can provide a remarkably good description of the dynamics \citep{earn00}, and other foundational research on disease dynamics.
+Some progress has been made on building and fitting spatiotemporal models for measles; see, for example, \citet{xia04}, \citet{eggo11}, \citet{bjornstad19}, and \citet{becker20}.
 However, the lack of suitable methodology to fit and assess a flexible class of coupled models is an obstacle to further progress \citep{becker20}.
-Previous methodological work has used spatiotemporal measles models as a test problem \citep{park20,ionides21}, but these methods have fallen short as tools for data analysis due to numerical considerations.
+Previous methodological research has used spatiotemporal measles models as a test problem \citep{park20,ionides21}.
+However, these methods have fallen short as tools for data analysis, owing to numerical considerations.
 Bearing all this in mind, measles provides a natural testing ground for our new methodologies.
@@
 
-\subsection{The measles data, a model and three submodels}
+\subsection{The measles data, a model, and three submodels}
 \label{sec:model}
 
+\setcounter{equation}{0}
+
 We set ourselves the task of fitting a spatiotemporal model to the case reports for 20 towns studied by \citet{he10}.
-We require the ability to handle discrete case counts which vary from zero to thousands of cases per week.
-We consider a model with the same structure as \citet{he10}, namely a Markov chain with Gamma noise on the infection rate, but with the addition of a term for transmission between cities.
-This requirement limits us to plug-and-play methodologies, which are those that require simulation from the latent process model but not the ability to  evaluate transition densities \citep{breto09,he10}.
-
-Some previous analyses have used counts aggregated over two-week intervals \citep{park20,ionides21} since these were available for more cities, but our goal here is to extend the analysis of \citet{he10}.
-Apart from this, our model matches \citet{ionides21} and for completeness we repeat the description here.
+We need to be able to handle discrete case counts, which vary from zero to thousands of cases per week.
+We consider a model with the same structure as that of \citet{he10}, namely, a Markov chain with gamma noise on the infection rate, but with an additional term for the transmission between cities.
+This requirement limits us to plug-and-play methodologies, which are those that require simulation from the latent process model, but not the ability to  evaluate transition densities \citep{breto09,he10}.
+
+Some previous analyses have used counts aggregated over two-week intervals \citep{park20,ionides21}, because these were available for more cities, but our goal here is to extend the analysis of \citet{he10}.
+Apart from this, our model matches that of \citet{ionides21} and, for completeness, we repeat the description here.
 We compartmentalize the population of each town into susceptible ($S$), exposed ($E$), infectious ($I$), and recovered/removed ($R$) categories.
-The numbers of individuals in each compartment for town $\unit$ at time $t$ are denoted by integer-valued random variables $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$.
-The population dynamics are written in terms of counting processes $N_{Q_1Q_2,\unit}(t)$ enumerating cumulative transitions in town $\unit$, up to time $t$, from compartment $Q_1$ to $Q_2$.
-Here, $Q_1,Q_2\in \{S,E,I,R,B,D\}$ with $B$ denoting a source compartment for immigration or birth, and $D$ a sink compartment for emigration or death.
+The numbers of individuals in each compartment for town $\unit$ at time $t$ are denoted by integer-valued random variables $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$, respectively.
+The population dynamics are written in terms of the counting processes $N_{Q_1Q_2,\unit}(t)$ enumerating the cumulative transitions in town $\unit$, up to time $t$, from compartment $Q_1$ to $Q_2$.
+Here, $Q_1,Q_2\in \{S,E,I,R,B,D\}$, with $B$ denoting a source compartment for immigration or birth, and $D$ denoting a sink compartment for emigration or death.
 We enumerate the $U=20$ towns studied by \citet{he10} in decreasing size, so that $\unit=1$ corresponds to London.
-Our model is described by the following system of stochastic differential equations, for $\unit=1,\dots, \Unit$,
+Our model is described by the following system of stochastic differential equations, for $\unit=1,\dots, \Unit$:
 \begin{equation}
@@
 \end{equation}
-The total population $P_\unit(t)=S_\unit(t)+E_\unit(t)+I_\unit(t)+R_\unit(t)$ is calculated by smoothing census data and is treated as known.
+The total population $P_\unit(t)=S_\unit(t)+E_\unit(t)+I_\unit(t)+R_\unit(t)$ is calculated by smoothing census data, and is treated as known.
 The number of recovered individuals $R_\unit(t)$ in town $\unit$ is therefore defined implicitly.
-$N_{SE,\unit}(t)$ is modeled as a negative binomial death process \citep{breto09,breto11} defined by a rate, $\mu_{SE,\unit}(t)$, specified as
-%\begin{eqnarray}
+Furthermore, $N_{SE,\unit}(t)$ is modeled as a negative binomial death process \citep{breto09,breto11}, defined by a rate, $\mu_{SE,\unit}(t)$, specified as
 \begin{equation}
@@
 %\label{eqn:transmissionrate}
-%\end{eqnarray}
 \end{equation}
-Here, the time dependence of $\mu_{SE,u}$, $\beta_u$, $I_u$ and $P_u$ is suppressed;
+Here, the time dependence of $\mu_{SE,u}$, $\beta_u$, $I_u$, and $P_u$ is suppressed;
 $\alpha_{\unit}$ is a mixing exponent modeling inhomogeneous contact rates within a town;
@@
 ${d\Gamma_{SE,\unit}}/{dt}$ is gamma white noise, with intensity parameter $\sigma_{SE.\unit}$;
-$\beta_{\unit}$, models the seasonal transmission driven by high contact rates between children at school,
+and $\beta_{\unit}$ models the seasonal transmission driven by high contact rates between children at school,
 \begin{equation}
@@
 \end{equation}
-where $\schoolTermFrac = 0.759$ is the proportion of the year taken up by the school terms, $\meanBeta_\unit$ is the mean transmission rate, and $\amplitude_{\unit}$ measures the reduction of transmission during school holidays.
+where $\schoolTermFrac = 0.759$ is the proportion of the year taken up by the school terms, $\meanBeta_\unit$ is the mean transmission rate, and $\amplitude_{\unit}$ measures the reduction in transmission during school holidays.
 
 The number of travelers from town $\unit$ to $\altUnit$ is denoted by $v_{\unit\altUnit}$.
-Here, $v_{\unit\altUnit}$ is constructed using the gravity movement model of \cite{xia04} given by
-\[
+Here, $v_{\unit\altUnit}$ is constructed using the gravity movement model of \cite{xia04}, given by
+\begin{equation}
+\nonumber
 v_{\unit\altUnit} = \gravity_{\unit} \cdot \frac{\hspace{2mm} \overline{\dist}\hspace{2mm}}{\hspace{1.5mm}\overline{\pop}^2\hspace{1mm}} \cdot \frac{\pop_\unit \cdot \pop_{\altUnit}}{\dist_{\unit\altUnit}},
-\]
-where $\gravity_u$ is the {\it gravitational constant}, $\dist_{\unit\altUnit}$ is the distance between town $\unit$ and town $\altUnit$, $\pop_\unit$ is the average of $P_{\unit}(t)$ across time, $\overline{\pop}$ is the average of $\pop_{\unit}$ across towns, and $\overline{\dist}$ is the average of $\dist_{\unit\altUnit}$ across all pairs of towns.
-The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{QD,\unit}(t)$, for $Q\in\{S,E,I,R\}$,  are modeled as conditional Poisson process with per-capita rates $\mu_{EI,\unit}$, $\mu_{IR,\unit}$ and $\mu_{QD,\unit}$ respectively, and we fix $\mu_{QD,\unit}=(50 \mbox{ year})^{-1}$.
+\end{equation}
+where $\gravity_u$ is the {\it gravitational constant}, $\dist_{\unit\altUnit}$ is the distance between town $\unit$ and town $\altUnit$, $\pop_\unit$ is the average of $P_{\unit}(t)$ over time, $\overline{\pop}$ is the average of $\pop_{\unit}$ across towns, and $\overline{\dist}$ is the average of $\dist_{\unit\altUnit}$ across all pairs of towns.
+The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$, and $N_{QD,\unit}(t)$, for $Q\in\{S,E,I,R\}$,  are modeled as conditional Poisson processes with per-capita rates $\mu_{EI,\unit}$, $\mu_{IR,\unit}$, and $\mu_{QD,\unit}$, respectively, and we fix $\mu_{QD,\unit}=(50 \mbox{ year})^{-1}$.
 A fraction $\cohort$ of births enter the susceptible cohort on the school admission day, and hence the birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson process with rate
-%\begin{equation}\nonumber
 $\mu_{BS,\unit}(t-\birthdelay) \big[
   (1-\cohort) + \cohort \delta(t-t_a) \big]$,
-%\end{equation}
 where $\mu_{BS,\unit}(t)$ is specified by interpolated census data, $t_a=\lfloor t \rfloor + 252/365$ is the admission date for the year containing $t$, $\delta$ is a Dirac delta function, and $\birthdelay=4\, \mbox{ year}$ is a fixed delay between birth and entry into a high-transmission school community.
@@
 &&\hspace{-20mm}
-- \varPhi\big(y-0.5; \rho_{\unit} z,\rho_{\unit}(1-\rho_{\unit})z+\measOD_{\unit}^2\rho^2z^2\big)
+- \varPhi\big(y-0.5; \rho_{\unit} z,\rho_{\unit}(1-\rho_{\unit})z+\measOD_{\unit}^2\rho^2z^2\big),
 \label{eq:obs}
@@
 where $\varPhi(\cdot;\mu,\sigma^2)$ is the $\normal(\mu,\sigma^2)$ cumulative distribution function, and $\measOD$ models overdispersion relative to the binomial distribution.
-For $y=0$, we replace $y-0.5$ by $-\infty$ in \eqref{eq:obs}.
-Three data points were treated as missing by \citet{he10} due to presumed recording errors.
-We followed that decision in the measurement model by including a special value $\mathrm{NA}$ in $\Yspace$ and setting $Y_{\unit,n}$ to be $\mathrm{NA}$ with probability one when $\data{y}_{\unit,n}$ is missing.
+For $y=0$, we replace $y-0.5$ with $-\infty$ in \eqref{eq:obs}.
+Three data points are treated as missing by \citet{he10}, owing to presumed recording errors.
+We follow this decision, implemented by including a special value $\mathrm{NA}$ in $\Yspace$, and setting $Y_{\unit,n}$ to $\mathrm{NA}$ with probability one when $\data{y}_{\unit,n}$ is missing.
 
@@
 \statSinica{$}
-\psi_u=\big(\meanBeta_{\unit},\mu_{EI,\unit}, \mu_{IR,\unit},\rho_{\unit},\measOD_{\unit},\sigma_{SE,\unit},\gravity_{\unit},\iota_{\unit},\amplitude_{\unit},\alpha_{\unit}, S_{0,\unit}, E_{0,\unit},I_{0,\unit}\big)
+\psi_u=\big(\meanBeta_{\unit},\mu_{EI,\unit}, \mu_{IR,\unit},\rho_{\unit},\measOD_{\unit},\sigma_{SE,\unit},\gravity_{\unit},\iota_{\unit},\amplitude_{\unit},\alpha_{\unit}, S_{0,\unit}, E_{0,\unit},I_{0,\unit}\big),
 \arxiv{\end{equation*}}
@@
 for $u\in\seq{1}{U}$.
-This defines an extended model for implementing IBPF with shared parameters, as in \eqref{eq:extension}.
-It also lets us address a key data analysis question of which parameters should be shared between units and which should be unit-specific.
+This defines an extended model for implementing an IBPF with shared parameters, as in \eqref{eq:extension}.
+It also serves another function by allowing us address a key data analysis question of which parameters should be shared between units and which should be unit-specific.
 
@@
   ggplot2::ylab("log10(simulated cases)")+
-  ggplot2::xlab("Year")
+  ggplot2::xlab("Year")+
+  theme(text=element_text(family="serif"))
 print(g)
@@
 
-The data are shown in Fig~\ref{fig:data-plot} and simulations from the model are shown in Fig~\ref{fig:sim-plot}.
+The data are shown in Fig~\ref{fig:data-plot}, and simulations from the model are shown in Fig~\ref{fig:sim-plot}.
 Parameters for the simulated model were based on the analysis of individual towns by \citet{he10}.
-In order to investigate estimation of either shared or unit-specific parameters, the simulation was conducted with all parameters being shared.
+In order to investigate estimation of either shared or unit-specific parameters, we conduct the simulation with all parameters being shared.
 To find a shared parameter vector capable of providing a reasonable representation of all the towns simultaneously, we experimented to look for a visual match between the data and the simulations.
-The parameters used are reported in the supplement (Sec.~\supSecMeaslesParams).
+The parameters used are reported in the Supplementary Material (Section~\supSecMeaslesParams).
 
@@
 In the presence of weak identifiability, small differences in likelihood could lead to large differences in parameter estimates.
-In such situations, a scientist may choose to investigate what functions of the parameters can be inferred accurately without adding additional assumptions, or alternatively they may choose to investigate the consequences of placing constraints on some parameters to improve identifiability of the remainder.
-Likelihood maximization permits such investigations, but they are beyond the scope of this article.
-Parameter values used for the simulation study and parameter values obtained by the likelihood maximization searches are provided in the supplementary code repository, for the interested reader.
-
-To facilitate data analysis, we wish to have methodology that operates across the full spectrum of decisions on shared versus unit-specific parameter designations.
-We therefore test our method on three submodels: {\modelA} has mostly shared parameters, choosing only the initial values and reporting rate to be unit-specific; {\modelB} has every parameter unit-specific; {\modelC} has every parameter is unit-specific and the dynamic coupling (in our context, movement of infected individuals between towns) is replaced by an external forcing of each unit (in our context, an immigration rate of infected individuals from outside the study population).
-Sec.~{\supSecMeaslesParams} defines the submodels in more detail.
-Submodel {\modelC} provides a useful point of reference since it is a special case of a PanelPOMP model \citep{breto19} and also can be analyzed as a collection of separate POMP models.
+In such situations, a scientist may choose to investigate what functions of the parameters can be inferred accurately without adding additional assumptions.
+Alternatively, they may choose to investigate the consequences of placing constraints on some parameters in order to improve the identifiability of the remainder.
+Likelihood maximization permits such investigations, but is beyond the scope of this study.
+In the Supplementary Material, we provide the parameter values used for the simulation study and those obtained by the likelihood maximization.
+
+To facilitate data analysis, our methodology needs to operate across the full spectrum of decisions on shared versus unit-specific parameter designations.
+We therefore test our method on three submodels: {\modelA} has mostly shared parameters, with only the initial values and the reporting rate unit-specific; {\modelB} has every parameter unit-specific; and {\modelC} has every parameter unit-specific, and the dynamic coupling (in our context, the movement of infected individuals between towns) is replaced with an external forcing of each unit (in our context, the immigration rate of infected individuals from outside the study population).
+Section~{\supSecMeaslesParams} defines the submodels in further detail.
+Submodel {\modelC} provides a useful point of reference, because it is a special case of a PanelPOMP model \citep{breto19}, and also can be analyzed as a collection of separate POMP models.
 We set up our model so that {\modelC} matches the analysis of \citet{he10}.
-One may expect that methods which take advantage of the special structure of {\modelC} should out-perform more general methods that permit coupling between units, so we expect the SpatPOMP methods to be less numerically efficient than applying POMP methods separately to each unit.
-Questions at stake are: How much less efficient are the SpatPOMP methods? Are simulation-based SpatPOMP inference methods practical for situations such as the measles model of \citet{he10}?
-
-The simulated model is drawn from model {\modelA}, which is nested within model {\modelB} but not within {\modelC}.
-For models {\modelB} and {\modelC}, with all parameters are unit-specific, we must estimate $20\times 13=260$ parameters.
-This greatly exceeds the 7 shared parameters fitted by \citet{ionides21} for a measles metapopulation model using a large computational effort.
-An iterated guided intermediate resampling filter (IGIRF) algorithm was used to fit 9 shared parameters and 3 unit-specific initial value parameters in a measles metapopulation model \citep{park20}.
-That analysis used a customized treatment for the unit-specific initial value parameters and did not attempt to estimate other unit-specific parameters.
-IGIRF is sensitive to the choice of the guide function, and the model-specific implementation used by \citet{park20} performed better than the generic implementation of IGIRF which is currently available in \code{spatPomp}.
-To our knowledge, {\ibpf} is therefore advancing the current limits for scalability of simulation-based maximum-likelihood inference for metapopulation dynamics.
+One may expect that methods that take advantage of the special structure of {\modelC} should outperform more general methods that permit coupling between units.
+Thus, we expect the SpatPOMP methods to be less efficient numerically than application of POMP methods separately to each unit.
+Here we answer the following questions: How much less efficient are the SpatPOMP methods?
+Are simulation-based SpatPOMP inference methods practical for situations such as the measles model of \citet{he10}?
+
+The simulated model is drawn from model {\modelA}, which is nested within model {\modelB}, but not within {\modelC}.
+For models {\modelB} and {\modelC}, in which all parameters are unit-specific, we must estimate $20\times 13=260$ parameters.
+This greatly exceeds the seven shared parameters fitted by \citet{ionides21} for a measles metapopulation model, which required significant computational effort.
+An iterated guided intermediate resampling filter (IGIRF) algorithm was used to fit nine shared parameters and three unit-specific initial value parameters in a measles metapopulation model \citep{park20}.
+The latter analysis uses a customized treatment for the unit-specific initial value parameters, and does not attempt to estimate other unit-specific parameters.
+IGIRF is sensitive to the choice of guide function, and the model-specific implementation of \citet{park20} outperforms the current generic implementation of IGIRF in \code{spatPomp}.
+Thus, to the best of our knowledge, the {\ibpf} presented in Algorithm~\ref{alg:ibpfilter} advances the current limits on the scalability of simulation-based maximum-likelihood inference for metapopulation dynamics.
 
@@
 
-Before engaging in likelihood maximization, one first wants to validate likelihood evaluation.
-Likelihood evaluation for metapopulation models was the main topic of \citet{ionides21} and we do not repeat that analysis here.
-In brief, the basic particle filter provides a consistent evaluation of the log-likelihood in a limit where the number of particles is large enough to make the Monte Carlo standard deviation small.
-This is practical only when $U$ is small (say, $U\le 5$) but that can be used to calibrate the bias induced by BPF, which turns out to be small for our multi-town measles model when each town is its own block.
-It may be surprising that resampling independently on each block (which is what the block particle filter does) is able to capture the dependence.
-Heuristically, we note that the dynamic dependence between blocks is maintained by BPF, which updates particles according to the full coupled dynamics.
-Whether this is sufficient to obtain a good approximation to the filter distribution is situation-dependent, but for the specific case of metapopulation models (for which the strongest coupling is within units rather than between units) the approximation can be empirically successful.
-In principle, the approximation error can be reduced by increasing the number of units in each block, but in practice the additional Monte Carlo variance acquired by doing this is not worthwhile in situations where the coupling is relatively weak.
-The BPF log-likelihood evaluation at the true parameter value is $\loglik_{\mathrm{true}}=\Sexpr{myround(loglik_truth[1],1)}$ with a Monte Carlo standard error of $\Sexpr{myround(loglik_truth[2],1)}$.
+Before engaging in likelihood maximization, we first validate the likelihood evaluation; see \citet{ionides21} for likelihood evaluation for metapopulation models.
+Briefly, the basic particle filter provides a consistent evaluation of the log-likelihood in a limit with sufficient particles to make the Monte Carlo standard deviation small.
+This is practical only when $U$ is small (say, $U\le 5$), but this situation can be used to calibrate the bias induced by a BPF, which turns out to be small for our multi-town measles model when each town is its own block.
+It may be surprising that resampling independently on each block (which is what a BPF does) is able to capture the dependence.
+Heuristically, note that the dynamic dependence between blocks is maintained by a BPF, which updates particles according to the full coupled dynamics.
+Whether this is sufficient to obtain a good approximation to the filter distribution depends on the situation, but for the specific case of metapopulation models (for which the strongest coupling is within units, rather than between units), the approximation can be empirically successful.
+In principle, the approximation error can be reduced by increasing the number of units in each block.
+However, in practice, the additional Monte Carlo variance acquired by doing so is not worthwhile when the coupling is relatively weak.
+The BPF log-likelihood evaluation at the true parameter value is $\loglik_{\mathrm{true}}=\Sexpr{myround(loglik_truth[1],1)}$, with a Monte Carlo standard error of $\Sexpr{myround(loglik_truth[2],1)}$.
 
@@
 <<sim_search_boxplot, echo=F, fig.height=3, fig.width=5.5, out.width="5in", fig.cap = "Fitting simulated measles data, using an initial search and four refinement steps.
-(A1--A5) model \\modelA, $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B1--B5) model \\modelB, $13\\times 20$ unit-specific parameters and no shared parameters;  (C1--C5) model \\modelC, also all unit-specific, but with immigration rather than coupling, matching He et al (2010).
-The horizontal dashed line is the log-likelihood at the true parameters, evaluated by BPF.",fig.show = "hold">>=
+(A1--A5) model \\modelA, $4\\times 20$ unit-specific parameters and nine shared parameters; (B1--B5) model \\modelB, $13\\times 20$ unit-specific parameters and no shared parameters;  (C1--C5) model \\modelC, all unit-specific, but with immigration rather than coupling, matching He et al.~(2010).
+The horizontal dashed line is the log-likelihood at the true parameters, evaluated using BPF.",fig.show = "hold">>=
 a <- cbind(
@@
 par(mai=c(0.4,0.6,0.1,0.1))
+par(family="serif")
 loglik_max <- max(cbind(a,b,c),truth_loglik)
@@
 
-A goal of this simulation study is to obtain appropriate algorithmic choices for the data analysis.
-We seek to develop methodology which is demonstrably successful when the truth is known, before applying it to data.
-One could revisit the simulation study based on the data analysis in Sec.~\ref{sec:data} using maximum likelihood estimates of the parameters.
-
-In Fig.~\ref{fig:sim_search_boxplot} we investigate a sequence of successive searches for the MLE for the models {\modelA}, {\modelB} and {\modelC} described above.
-Each search was replicated $36$ times.
-Search~1 was started with each parameter adjusted by a uniform $[-0.1,0.1]$ random perturbation on an appropriate dimensionless scale (log for non-negative parameters, logit for $[0,1]$ valued parameters, see Sec.~{\supSecAlg}).
-This is a fairly small perturbation, but nevertheless sufficient to knock the likelihood of the parameter vector around 200 log units below the MLE (shown in Fig.~\ref{fig:search_diagnostics}).
-The goal of this study is to show that the algorithm can succeed reliably on a relatively easy local optimization task.
-Subsequent searches were started with four copies of each parameter vector with estimated likelihood in the top 25\% for the previous search.
+One of our goals is to obtain appropriate algorithmic choices for a data analysis.
+Thus, we seek to develop methodology that is demonstrably successful when the truth is known, before applying it to data.
+One could revisit the simulation study based on the data analysis in Section~\ref{sec:data} using MLEs of the parameters.
+
+In Figure~\ref{fig:sim_search_boxplot} we investigate a sequence of successive searches for the MLE for the models {\modelA}, {\modelB}, and {\modelC} described previously.
+Each search is replicated $36$ times.
+Search~1 was started with each parameter adjusted by a uniform $[-0.1,0.1]$ random perturbation on an appropriate dimensionless scale (log for nonnegative parameters, and logit for $[0,1]$ valued parameters; see Section~{\supSecAlg}).
+This is a fairly small perturbation, but it is nevertheless sufficient to knock the likelihood of the parameter vector around 200 log units below the MLE (shown in Figure~\ref{fig:search_diagnostics}).
+Our goal is to show that the algorithm can succeed reliably on a relatively easy local optimization task.
+Subsequent searches were started with four copies of each parameter vector with an estimated likelihood in the top 25\% for the previous search.
 
 The MLE is not known exactly in this case.
-Wilks' theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+9)/2=44.5$ for {\modelA} and $(13\times 20)/2=130$ for {\modelB}.
-The $y$-axis values in Fig.~\ref{fig:sim_search_boxplot} show log-likelihoods exceeding the truth by less than this, indicating some imperfection to the Monte Carlo maximization so far as Wilks' asymptotic result holds in this situation.
-Despite this limitation, searches that exceed the likelihood at the truth have found inferentially plausible sets of parameters which can be used to study the likelihood surface around the maximum.
-For example, Monte Carlo profile confidence intervals can give proper coverage even in the presence of considerable Monte Carlo error \citep{ionides17,ning21}.
-
-<<search_diagnostics, echo=F, fig.height=6.5, fig.width=4.5, out.width="5in", fig.cap = "Three steps of a likelihood search on simulated data (left panel, model \\modelA) and UK measles data (right panel, model \\modelB).Dashed lines parallel to the axes show the log-likelihood at the truth (left panel) and the \\citet{he10} value (right panel).
-Points above the diagonal dashed line show improvement due to the search step.">>=
+Wilks' theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+9)/2=44.5$ for {\modelA}, and $(13\times 20)/2=130$ for {\modelB}.
+The $y$-axis values in Figure~\ref{fig:sim_search_boxplot} show log-likelihoods exceeding the truth by less than this, indicating some imperfection in the Monte Carlo maximization so far as Wilks' asymptotic result holds.
+Despite this limitation, searches that exceed the likelihood at the truth have found inferentially plausible sets of parameters that can be used to study the likelihood surface around the maximum.
+For example, Monte Carlo profile confidence intervals can give proper coverage, even in the presence of considerable Monte Carlo error \citep{ionides17,ning21}.
+
+<<search_diagnostics, echo=F, fig.height=6.5, fig.width=4.5, out.width="5in", fig.cap = "Three steps of a likelihood search on simulated data (left panel, model \\modelA) and UK measles data (right panel, model \\modelB).
+Dashed lines parallel to the axes show the log-likelihood at the truth (left panel) and the \\citet{he10} value (right panel).
+Points above the diagonal dashed line show improvement owing to the search step.">>=
 par(mfcol=c(3,2))
@@
 par(omi=c(0.3,0.5,0.5,0.1))
+par(family = "serif")
 ##
@@
 
-In Fig.~\ref{fig:search_diagnostics} we take a closer look at convergence diagnostics corresponding to {\modelA} for the simulated data (the first column) and {\modelB} for the UK measles data (the second column).
+Figure~\ref{fig:search_diagnostics} shows the convergence diagnostics corresponding to {\modelA} for the simulated data (the first column) and {\modelB} for the UK measles data (the second column).
 For now, we focus on the first column.
@@
 We find that {\ibpf} robustly and rapidly approaches a neighborhood of the MLE, as measured by likelihood.
-The second row shows that further investigation of the more successful searches can reliably obtain likelihood values higher than those at the truth.
-However, for a Monte Carlo search based on Monte Carlo likelihood evaluation, it may be anticipated that pinpointing the exact maximum in a high-dimensional space is problematic.
+The second row shows that further investigation of more successful searches can reliably obtain likelihood values higher than those at the truth.
+However, for a Monte Carlo search based on a Monte Carlo likelihood evaluation, it may be problematic to pinpoint the exact maximum in a high-dimensional space.
 The third row shows that continued searching does not lead to substantially better outcomes.
@@
 
-The results in Fig.~\ref{fig:sim_search_boxplot} and the first column of  Fig.~\ref{fig:search_diagnostics} document that {\ibpf} can be effective on simulated data, but do not guarantee comparable performance for our data analysis.
-Indeed, model misspecification that is inevitable for data analysis may be expected to add difficulties to filtering and therefore to numerical methods based on filtering.
-Rather, we view the simulation study as a lower bound on the effort required to carry out effective inference on data, and a starting point for investigating effective ways to proceed with the data analysis.
+The results in Figure~\ref{fig:sim_search_boxplot} and the first column of  Figure~\ref{fig:search_diagnostics} show that {\ibpf} can be effective for simulated data, but do not guarantee comparable performance for our data analysis.
+Indeed, model misspecification, which is inevitable for data analysis, may be expected to add difficulties to filtering and therefore to numerical methods based on filtering.
+Rather, we view the simulation study as a lower bound on the effort required to perform effective inference on data, and therefore a starting point for investigating how to proceed with a data analysis.
 Before moving on to the data analysis, we briefly describe the details of the model.
@@
 <<data_search_boxplot, echo=F,  fig.height=3, fig.width=5.5, out.width="5in", fig.cap = "Fitting different models to the UK measles data using the method tested on simulated data.
-(A) $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B) $13\\times 20$ unit-specific parameters and no shared parameters.
-(C) Also all unit-specific, but with immigration rather than coupling.
+(A) $4\\times 20$ unit-specific parameters and nine shared parameters; (B) $13\\times 20$ unit-specific parameters and no shared parameters; (C) all unit-specific, but with immigration rather than coupling.
 The horizontal dashed line is the likelihood from \\citet{he10}.">>=
@@
 par(mai=c(0.5,0.8,0.1,0.1))
+par(family="serif")
 boxplot(cbind(a,b,c),ylim=loglik_max-c(5500,0),
@@
 
-Fig.~\ref{fig:data_search_boxplot} shows results for fitting models {\modelA}, {\modelB} and {\modelC} to UK measles data using successive rounds of {\ibpf}, applying the same method used for the simulation study shown in Fig.~\ref{fig:sim_search_boxplot}.
+Figure~\ref{fig:data_search_boxplot} shows the results from fitting models {\modelA}, {\modelB}, and {\modelC} to the UK measles data using successive rounds of {\ibpf}, applying the same method used for the simulation study shown in Figure~\ref{fig:sim_search_boxplot}.
 The dashed lines show the sum of the log-likelihoods obtained by \citet{he10}, $\ell_{\mathrm He}=\Sexpr{myround(loglik_he,1)}$.
@@
 BPF with each town forming a separate block, applied to an uncoupled model such as {\modelC}, is equivalent to carrying out independent particle filters for each unit.
-Indeed, if the published parameter values from \citet{he10} are inserted into the \code{spatPomp} object for model {\modelC}, then the \code{bpfilter} function can be used to carry out BPF, which retrieves $\ell_{\mathrm He}$, up to Monte Carlo error.
-Each of models {\modelA}, {\modelB} and {\modelC} in Fig.~\ref{fig:data_search_boxplot} shows shortfall relative to $\ell_{\mathrm He}$.
-
-The largest shortfall is for {\modelA}, and an explanation is that {\modelA} has more shared parameters than the evidence in the data supports.
+Indeed, if we assign the published parameter values from \citet{he10} to the \code{spatPomp} object for model {\modelC}, and apply the \code{bpfilter} function to carry out BPF, we retrieve $\ell_{\mathrm He}$, up to a Monte Carlo error.
+Models {\modelA}, {\modelB}, and {\modelC} in Figure~\ref{fig:data_search_boxplot} each show a shortfall relative to $\ell_{\mathrm He}$.
+
+The largest shortfall is for {\modelA}, perhaps because {\modelA} has more shared parameters than the evidence in the data supports.
 The results of \citet{he10} suggest that the data are explained better when various parameters are a function of the town population.
-However, it remains an open problem to determine suitable functional forms for this relationship, and to establish regularities across towns that can be represented by shared parameters.
-This may be investigated using panel methods such as PanelPOMP models \citep{breto19} in addition to consideration of SpatPOMP models.
-
-Models {\modelB} and {\modelC} give rise to comparable likelihoods, which is in contrast to Fig.~\ref{fig:sim_search_boxplot}.
-Fitting to simulated data from {\modelB} (in the special case where all unit-specific parameters are equal across all units) the shortfall for {\modelC} in Fig.~\ref{fig:sim_search_boxplot} indicates that we obtain a substantially worse fit by approximating coupling by an uncoupled reservoir of infection.
+However, determining suitable functional forms for this relationship, and establishing regularities across towns that can be represented by shared parameters, remains an open problem.
+This may be investigated using panel methods, such as PanelPOMP models \citep{breto19}, in addition to consideration of SpatPOMP models.
+
+Models {\modelB} and {\modelC} yield comparable likelihoods, which is in contrast to the results shown in Figure~\ref{fig:sim_search_boxplot}.
+Fitting to simulated data from {\modelB} (in the special case in which all unit-specific parameters are equal across all units), the shortfall for {\modelC} in Figure~\ref{fig:sim_search_boxplot} indicates that we obtain a substantially worse fit when we approximate coupling using an uncoupled reservoir of infection.
 If the actual data were also explained substantially better by the coupled model, we would expect to see comparable results in the data analysis.
-Since we do not, we conclude that this coupled model is not a substantially better explanation than the uncoupled model.
-Various other candidate coupling mechanisms have been proposed \citep{bjornstad19} but they have not yet been fitted to the full data, suggesting a lack of appropriate methodology to do so.
-Instead, \citet{bjornstad19} looked at summary statistics based on local fade-outs and re-introductions.
-
-We notice from Fig.~\ref{fig:data_search_boxplot} that model {\modelB} has a small but distinct shortfall compared to $\ell_{\mathrm He}$.
-On the simulated data, we do not observe this shortfall, and we deduce that the real data provides a more challenging optimization environment.
-When carrying out hard optimization problems, it may be possible to develop helpful strategies specific to the model and data in question.
+However, because we do not, we conclude that this coupled model does not provide a substantially better explanation than the uncoupled model.
+Although various other candidate coupling mechanisms have been proposed \citep{bjornstad19}, they have not yet been fitted to the full data, suggesting a lack of appropriate methodology to do so.
+Instead, \citet{bjornstad19} consider summary statistics based on local fade-outs and re-introductions.
+
+Figure~\ref{fig:data_search_boxplot} shows that model {\modelB} has a small, but distinct shortfall compared with $\ell_{\mathrm He}$.
+For the simulated data, we do not observe this shortfall, and thus we deduce that the real data provide a more challenging optimization environment.
+When performing difficult optimization problems, it may be possible to develop helpful strategies specific to the model and data in question.
 For example, one could try merging unit-specific parameters from different searches, using the likelihood component for each unit to assess successful choices.
-Optimization heuristics such as this do not have general theoretical support; if they obtain higher likelihoods that is sufficient justification.
-For analysis of panel time series data using PanelPOMP models, such methods have clearer theoretical support and have been found useful \citep{breto19}.
+However, optimization heuristics such as this do not have general theoretical support; if they obtain higher likelihoods, that is sufficient justification.
+For analyses of panel time series data using PanelPOMP models, such methods have clearer theoretical support and have been found to be useful \citep{breto19}.
 
@@
 
-Theoretical interest in BPF was inspired by \citet{rebeschini15}, who showed mathematically that BPF can enjoy linear scaling properties under suitable conditions.
-\citet{rebeschini15} expressed pessimism about the practical applicability of their algorithm, which may help to explain why the practical use of this algorithm has been limited.
-The algorithm was independently invented under the name of a factored particle filter by \citet{ng02} who offered an empirical justification.
-However, the proposal of \citet{ng02} also saw only limited use.
-The numerical results in Fig.~3 of \citet{ionides21} suggest that BPF is particularly well suited to metapopulation models.
-In a metapopulation model, one expects most of the population dynamics to occur at a local level, among the individuals at one spatial unit.
-As a broad generalization, in ecological systems dispersion of individuals between spatial units is rare but dynamically important.
-Edge effects between blocks, which may be a serious problem for BPF in a system with stronger spatial coupling \citep{ionides21} can therefore be a relatively minor concern for metapopulation models.
+Theoretical interest in BPF algorithms has been inspired by the work of \citet{rebeschini15}, who showed mathematically that a BPF enjoys linear scaling properties under suitable conditions.
+However, they expressed pessimism about the practical applicability of their algorithm, which may help to explain its limited practical use.
+A similar algorithm was proposed independently  under the name of a factored particle filter by \citet{ng02}, who offered an empirical justification.
+However, the latter algorithm has also seen only limited use.
+The numerical results in Figure~3 of \citet{ionides21} suggest that BPF is particularly well suited to metapopulation models, in which we expect most of the population dynamics to occur at a local level, among individuals at one spatial unit.
+As a broad generalization, in ecological systems, the dispersion of individuals between spatial units is rare but dynamically important.
+Therefore, edge effects between blocks, which may be a serious problem for a BPF in a system with stronger spatial coupling \citep{ionides21} is a relatively minor concern in metapopulation models.
 Blocks of size one unit are therefore a natural choice for block filtering of metapopulation models when the constituent populations are spatially distinct.
 
-This article has investigated a likelihood maximization approach to inference.
-Much research has been done on inference for high-dimensional partially observed stochastic dynamic systems, and we have cited above only the most directly relevant work.
-A Bayesian inference approach based on the ensemble Kalman filter was developed by \citet{katzfuss19}.
-An expectation-maximization approach based on a block particle filter was presented by \citet{finke17}.
-Spatiotemporal models having a convenient factorization across units were studied by \citet{beskos17} and \citet{xu19}.
-
-  We have demonstrated a workflow that led to a likelihood-based assessment of measles metapopulation models ({\modelA} and {\modelB}) with the possibility of finding evidence that they out-perform the uncoupled model, {\modelC}.
-  Our methodology showed the potential to successfully refute {\modelC} on simulated data, when the truth was within {\modelA} and {\modelB}, but no advantage was found for these coupled models when the same comparison was carried out on the data.
-Likelihood maximization for the measles data fits the {\it common task framework} described by \citet[][Sec.~6]{donoho17}, with the likelihood value for {\modelC} obtained by \citet{he10} providing a benchmark challenge.
+We have proposed a likelihood-maximization approach to inference.
+Much research has been done on inference for high-dimensional partially observed stochastic dynamic systems, and we have cited only the most directly relevant work.
+A Bayesian inference approach based on the ensemble Kalman filter is proposed by \citet{katzfuss19}.
+An expectation-maximization approach based on a BPF is presented by \citet{finke17}.
+Spatiotemporal models with a convenient factorization across units are studied by \citet{beskos17} and \citet{xu19}.
+
+  We have demonstrated a workflow that led to a likelihood-based assessment of measles metapopulation models ({\modelA} and {\modelB}), with the possibility of finding evidence that they outperform the uncoupled model, {\modelC}.
+  Our methodology successfully refutes {\modelC} on simulated data, when the truth is within {\modelA} and {\modelB}.
+  However, there was no evidence of an advantage for these coupled models when the same comparison was carried out on the data.
+Likelihood maximization for the measles data fits the {\it common task framework} described by \citet[][Section~6]{donoho17}, with the likelihood value for {\modelC} obtained by \citet{he10} providing a benchmark challenge.
 Future improvements in models, perhaps facilitated by the open-source models and methods accompanying this article, may obtain metapopulation models that convincingly beat model {\modelC}.
 
-Computation time for Fig.~\ref{fig:sim_search_boxplot} and Fig.~\ref{fig:data_search_boxplot} was approximately 24hr on a single core of a compute node for 4000 particles iterated 100 times over the 730 time points (weekly data for 14 years) for 20 cities.
-Each box in these figures involves 36 replications and thus took 24hr on all cores of one node on a computing cluster.
-A measure of computational efficiency is the size of problem that can be solved on this timescale, discussed further in Sec.~{\supSecEfficiency}.
-Tasks that are considerably larger, perhaps $10^3$ or more spatial units, may require additional approximations such as those inherent in the ensemble Kalman filter \citep{evensen09book,katzfuss19} or other numerical filtering techniques \citep{whitehouse22}.
+The computation time for Figures~\ref{fig:sim_search_boxplot} and~\ref{fig:data_search_boxplot} was approximately 24hr on a single core of a computing node, for 4000 particles iterated 100 times over 730 time points (weekly data for 14 years) for 20 cities.
+Each box in these figures involves 36 replications, and thus took 24 hours on all cores of one node on a computing cluster.
+A measure of computational efficiency is the size of a problem that can be solved on this timescale, as discussed further in Section~{\supSecEfficiency}.
+Tasks that are considerably larger, perhaps $10^3$ or more spatial units, may require additional approximations, such as those inherent in the ensemble Kalman filter \citep{evensen09book,katzfuss19} or other numerical filtering techniques \citep{whitehouse22}.
 However, we anticipate that many practical metapopulation analyses can be addressed within the scope we have demonstrated.
-The majority of the computational effort is spent on simulating from the dynamic model.
-In the \code{spatPomp} R package, as in \code{pomp}, the user supplies a snippet of C code for simulating a single particle between a single arbitrary pair of times, and the package provides a vectorized form of this computation which is used by the inference methodologies within the package.
-As a consequence, competitive computational performance is obtained even though the majority of the package is written in R \citep[Table~1]{FitzJohn20}.
+The majority of the computational effort lies in simulating from the dynamic model.
+In the \code{spatPomp} R package, as in \code{pomp}, the user supplies a snippet of C code to simulate a single particle between a single arbitrary pair of times, and the package provides a vectorized form of this computation, which is used by the inference methodologies within the package.
+As a result, the computational performance is competitive even though the majority of the package is written in R \citep[Table~1]{FitzJohn20}.
+
+\vspace{3mm}
+
+%\newpage
+
+\noindent {\bf Supplementary material}
+
+\vspace{1mm}
+
+\noindent Supplementary text presents the following: S1, parameters for the measles model; S2, algorithmic parameters and transformations; S3, computational efficiency; S4, varying the spatial autoregression parameter, $r$.
+Source code is at \url{https://github.com/ionides/ibpf_article}.
 
@@
 We acknowledge an anonymous referee from a previous article who recommended implementing a block particle filter.
-We are grateful to two anonymous referees whose constructive feedback has improved this article.
+We are also grateful for constructive feedback from two anonymous referees.
 
@@
 We look for the smallest $J$ such that there is not much to be gained by making $J$ larger.
-This is carried out by looking for evidence about the bias and variance (discussed further in Sec.~\supSecEfficiency) which both need consideration when carrying out likelihood ratio tests and Monte Carlo adjusted profile confidence intervals \citep{ionides17,ning21}.
+This is carried out by looking for evidence about the bias and variance (discussed further in Section~\supSecEfficiency) which both need consideration when carrying out likelihood ratio tests and Monte Carlo adjusted profile confidence intervals \citep{ionides17,ning21}.
 
@@
 par(mai=c(0.6,0.8,0.1,0.1))
+par(family="serif")
 boxplot(loglik~sp_reg,data=results,ylim=c(-40000,loglik_max),

### 2023-12-27 talk/aprm24.tex
draft of aprm24 talk

--- 2023-03-30 talk/talk.tex
@@
 
+%% for flow diagrams
+\usepackage{tikz}
+\usetikzlibrary{positioning}
+\usetikzlibrary {arrows.meta}
+\usetikzlibrary{shapes.geometric}
+
 \begin{document}
@@
 \begin{center}
-  {\Large\bf An iterated block particle filter for inference on coupled dynamic systems with shared and unit-specific parameters}
+  {\Large\bf An iterated block particle filter for inference on coupled dynamic systems}
 
-%% ABSTRACT: We consider inference for a collection of partially observed, stochastic, interacting, nonlinear dynamic processes.
+% ABSTRACT
+% We consider inference for a collection of partially observed, stochastic, interacting, nonlinear dynamic processes.
 Each process is called a unit, and our primary motivation arises in biological metapopulation systems where a unit is a spatially distinct sub-population.
@@
 We introduce a new iterated block particle filter algorithm applicable to parameters that are either unit-specific or shared between units.
-We demonstrate this algorithm to carry out inference on a coupled epidemiological model for spatiotemporal measles case report data in twenty towns.
-We discuss other applications, to cholera in Haiti and COVID-19 in China.
+We demonstrate this algorithm to carry out inference on a coupled epidemiological model for spatiotemporal COVID-19 case report data in 373 cities.
 
@@
 
-Texas A\&M Statistics Colloquium\\
-Apr 21, 2023
+IMS Asia Pacific Rim Meeting\\
+Jan 5, 2024
 
@@
 
-Slides are at \url{https://ionides.github.io/talks/tamu23.pdf}
+Slides are at \url{https://ionides.github.io/talks/aprm24.pdf}
 
@@
 Joint work with
-Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King
+Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King, Mercedes Pascual
 
@@
   \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models?
-(Of particular interest are metapopulation models.)
+In particular, models for networks of interacting biological population dynamics.
 
@@
 
-\item We introduce the {\bf iterated block particle filter} because it is currently the most powerful algorithm available in \texttt{spatPomp}.
+\item We introduce the {\bf iterated block particle filter}, currently the most powerful algorithm available in \texttt{spatPomp}.
 
@@
 
-\begin{frame}{Pre-vaccination weekly measles in England \& Wales}
-
-  %% full weekly data now available:
-  %% Korevaar, Hannah, C.
-Jessica Metcalf, and Bryan T.
-Grenfell.
-"Structure, space and size: competing drivers of variation in urban and rural measles transmission." Journal of the Royal Society Interface 17, no.
-168 (2020): 20200010.
-  %% https://royalsocietypublishing.org/doi/pdf/10.1098/rsif.2020.0010
-  %% data at https://rs.figshare.com/collections/Supplementary_material_from_Structure_space_and_size_competing_drivers_of_variation_in_urban_and_rural_measles_transmission_/5036567/1
-
-\vspace{-2mm}
-
-\begin{center}
-\includegraphics[width=8.4cm]{he10-data.pdf}
-
-\end{center}
-
-\vspace{-2mm}
-
-\end{frame}
-
-\begin{frame}{Time series data, panel data \& spatiotemporal data}
-
-  \bi
-\item Looking at one unit (town) is \myemph{time series analysis}.
-
-  \item Joint modeling of a few units (say, 2 or 3) is \myemph{multivariate time series analysis}.
-
-\item Analysis of many time series, without consideration of dynamic interactions, is \myemph{panel data analysis}.
-
-\item Allowing for coupling between units, we get \myemph{spatiotemporal analysis}, which in our context is \myemph{metapopulation analysis}.
-
-  \ei
-
-%\vspace{3mm}
-
-%Question: When should we avoid inference for spatiotemporal models? When do we need to consider coupling? How?
-
-\end{frame}
-
 \begin{frame}{Desiderata}
@@
     \begin{itemize}
-    \item  Iiterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference:
+    \item  Iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference:
 
@@
     \end{itemize}
-\end{frame}
-
-\begin{frame}{Fitting mechanistic models to panel data}
-
-\bi
-\item To investigate epidemiological dynamics in multiple cities, one can consider each city independently, perhaps modeling a background immigration rate of infections for each city.
-
-  \vspace{2mm}
-
-\item \myemph{Decoupling} leads to panel data analysis, by assumption.
-
-  \vspace{2mm}
-
-  \item Iterated filtering methods extend to panel data (\texttt{panelPomp}, Breto et al, {\it JASA}, 2019).
-
-    \vspace{2mm}
-
-\item We must decide which parameters should be modeled as \myemph{shared} (i.e., the same for each unit) vs \myemph{unit-specific}.
-
-%\item The consequences of decoupling are becoming easier to study with the development of statistical inference methods for coupled systems, i.e., metapopulation dynamics.
-
-  \ei
-
 \end{frame}
@@
 
-\begin{frame}{Filtering $U$ units of a coupled measles SEIR model}
+\begin{frame}{COVID in 375 cities in China, 10-23 January, 2020}
 
-\vspace{-1mm}
+  \begin{itemize}
+  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
+    (Li et al, {\it Science}, May 2020).
+  \item SEIR (susceptible-exposed-infected-removed) model with asymptomatics, reporting delay, and coupling based on cell phone data.
+  \item Li et al (2020) used iterated EnKF for inference.
+  \item The time interval covers the initial China lockdown.
+  \item Here, we summarize our re-analysis of this model and data, recently posted on arXiv.
+%% see README for R code
+%   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
+%   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
+%   \item{} 0 cases reported for all cities during Jan 10--15.
 
-\begin{center}
-\includegraphics[width=10cm]{mscale_loglik_plot-1.pdf}
-
-\end{center}
-
-\vspace{-4mm}
-
-Simulated data using a gravity model with geography, demography and transmssion parameters corresponding to UK pre-vaccination measles (Ionides et al, {JASA}, 2021).
-
-%\end{center}
+\end{itemize}
 
@@
 \begin{frame}
-\frametitle{$U=40$ units for a coupled measles SEIR model}
+  %%{An SEAIR metapopulation model for COVID-19}
+\begin{center}
+%%%%% SEAIR diagram
+\resizebox{11cm}{!}{
+\begin{tikzpicture}[
+  square/.style={rectangle, draw=black, minimum width=0.8cm, minimum height=0.8cm, rounded corners=.1cm, fill=blue!8},
+  travel/.style={circle, draw=black, minimum width=0.85cm, minimum height=0.8cm, fill=green!8},
+  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.9cm,inner sep=0cm},
+  bendy/.style={bend left=25},
+  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
+  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
+]
+\tikzset{>={}}; % this is needed to implement the arrow redefinition
 
-\vspace{-2.7mm}
+\node (S) at (-0.5,0) [square] {S$_u$};
+\node (E) at (2,0) [square] {E$_u$};
+\node (A) at (4.5,0) [square] {A$_u$};
+\node (I) at (4.5,-1.75) [square] {I$_u$};
+\node (R) at (7,0) [square] {R$_u$};
 
-\begin{center}
-\includegraphics[width=8cm]{slice_image_plot-1.pdf}
+\node (T1) at (-0.5,1.75) [travel] {T};
+\node (T2) at (2,1.75) [travel] {T};
+\node (T3) at (4.5,1.75) [travel] {T};
+
+\node (Ca) at (4.5,-3.5) [report] {C$^a_u$};
+\node (Cb) at (7,-3.5) [report] {C$^b_u$};
+\node (C) at (9.5,-3.5) [report] {C$_u$};
+
+\node (V1) at (2.5,-1.75) {};
+\node (V2) at (3,-3.5) {};
+\draw [->] (E) -- (Ca -| V2) -- (Ca);
+\draw [->] (I -| V1) -- (I);
+
+\draw [->] (S) -- (E);
+\draw [->] (A) -- (R);
+
+\draw [->] (Ca) -- (Cb);
+\draw [->] (Cb) -- (C);
+
+\draw [->] (I.east) -- (R);
+\draw [->] (E) -- (A);
+
+\draw [->, bendy] (S) to (T1);
+\draw [->, bendy] (T1) to (S);
+
+\draw [->, bendy] (E) to (T2);
+\draw [->, bendy] (T2) to (E);
+
+\draw [->, bendy] (A) to (T3);
+\draw [->, bendy] (T3) to (A);
+
+\end{tikzpicture}
+}
 \end{center}
-
-\vspace{-3mm}
-
-{\bf A}. Simulated Susceptible-Exposed-Infected-Recovered dynamics coupled with a gravity model (log of biweekly reported cases).
-
-{\bf B}. Measles UK pre-vaccination case reports for the 40 largest cities.
-
-%\end{center}
+\bi
+\item 
+  Reportably infectious individuals, $\mathrm{I}_u$ for city $u$, are included in the delayed reporting compartment, $\mathrm{C^a_u}$.
+  \item 
+An individual arriving at $\mathrm{C}_u$ is a case report for city $u$.
+\item
+  Individuals in $\mathrm{A}_u$ are not reportable and transmit at a reduced rate.
+  \item 
+Travel occurs to and from $\mathrm{T}$, based on 2018 data from Tencent.
+%\item Many models have been proposed. We want a framework to implement and evaluate various scientific hypotheses.
+\ei
 
@@
 
-\begin{frame}{Parameters for the measles model}
-  \bi
-\item Seasonal transmission: mean and amplitude, using school term for contact rate.
-  \item Durations of latency and infectious period.
-\item Initial values: fraction susceptible, latent and infectious.
-\item Cohort effect: all births in an age cohort start school in September.
-\item Inhomogenous mixing exponent, $I^{\alpha}$.
-\item Measurement fraction.
-\item Population movement model: gravity constant.
-\item Dynamic noise: process overdispersion.
-\item Measurement overdispersion.
+\begin{frame}
+  \begin{center}
+    \includegraphics[height=9.5cm]{covid/panel_plot-1.png}
+    \end{center}
+\end{frame}
 
-\ei
-\vspace{3mm}
-(Adding gravity coupling to the model of He et al, {\it JRSI}, 2010)
-%\vspace{5mm}
-%Plug-and-play methods let you add or subtract features
+\begin{frame}
+ \includegraphics[width=12cm]{covid/filter_tests10.png}
+
+  \begin{itemize}
+\item BPF is the best available filter for this COVID model.
+\item Similar results hold for other models (Ionides et al, {\it JASA}, 2023).
+  \end{itemize}
 
@@
 \bi
-\item BPF worked quickly, easily and reliably on our measles model filtering experiments.
+\item BPF also worked quickly, easily and reliably on a measles metapopulation, as well as various toy benchmark problems (Ionides et al, {\it JASA}, 2023).
 
@@
 
-\begin{frame}
-
-\frametitle{Measles likelihood slices for coupling parameter, $G$}
-
-\vspace{-5mm}
-
-\begin{columns}[T] % align columns
-\begin{column}{.55\textwidth}
-  \includegraphics[width=6cm]{slice_combined_plot-1.pdf}
-\end{column}
-
-\hspace{-5mm}
-\begin{column}{.52\textwidth}
-
-  \vspace{6mm}
-
-  Simulating $15$ year of data from $U=40$ cities.
-  Slice likelihood, varying $G$ with other parameters fixed.\\
-   (Ionides et al, {\it JASA}, 2021).
-
-  \vspace{3mm}
-
-  {\bf A}. Evaluation using adapted bagged filter (ABF).
-
-    \vspace{3mm}
-
-  {\bf B}. Evaluation using block particle filter (BPF).
-
-    \vspace{3mm}
-
-  {\bf C}. Evaluation using EnKF.
-
-  \vspace{3mm}
-  A much easier task than computing a profile likelihood for real data.
-
-\end{column}
-\end{columns}
-
-\end{frame}
-
 \begin{frame}{Comments on the Ensemble Kalman Filter (EnKF)}
@@
 
-\begin{frame}{Scalability needed for practical inference}
-
-Large numbers of parameters
-  \begin{itemize}
-  \item  Initial conditions will typically have to be estimated for each unit.
-  \item Various dynamic parameters and measurement parameters (e.g., reporting rate) may also need to be unit-specific to obtain a statistical fit to the data.
-\item Working with hundreds of estimated parameters raises additional challenges on top of the high-dimensional coupled dynamics.
-  \end{itemize}
-
-\vspace{5mm}
-
-A moderate numbers of spatial units is enough to open new possibilities.
-
-  \begin{itemize}
-  \item As soon as dimension exceeds capabilities of a particle filter (say, $U=5$) we are in situations where likelihood-based inference for general models has been inaccessible.
-
-  \item 5-500 coupled units is our target problem size.
-
-  \item Larger problems will need numerical approximations (e.g., EnKF).
-
-  \end{itemize}
-
-  \end{frame}
-
 \begin{frame}{An iterated block particle filter for parameter estimation}
@@
 
-\begin{frame}
+\begin{frame}{Practical inference using IBPF}
+  \bi
+  \item to do
 
-%%% A: 9 shared parameters; 4 unit-specific
-%%% B: all unit-specific
-
-\vspace{-6mm}
-
-\begin{columns}[T] % align columns
-\begin{column}{.75\textwidth}
-  \includegraphics[width=6.8cm]{ibpf/search_diagnostics-1.pdf}
-\end{column}
-
-\hspace{-15mm}
-\begin{column}{.35\textwidth}
-  \vspace{10mm}
-
-$\bullet$    IBPF applied to simulations and data.
-
-    \vspace{5mm}
-
-$\bullet$    Multiple searches from random starting points.
-
-    \vspace{5mm}
-
-$\bullet$    Top 25\% of searches are subsequently refined.
-
-  \end{column}
-\end{columns}
+  \ei
 
@@
 
-\begin{frame}
+%% \begin{frame}{Technical issues with EnKF}
 
-\begin{columns}[T] % align columns
-\hspace{-4mm}\begin{column}{.75\textwidth}
-  \includegraphics[width=7.5cm]{ibpf/sim_search_boxplot-1.pdf}
-
-    \includegraphics[width=7.5cm]{ibpf/data_search_boxplot-1.pdf}
-
-  \end{column}
-
-\hspace{-18mm}
-\begin{column}{.4\textwidth}
-  \vspace{5mm}
-
-  {\bf IBPF for measles, 20 cities}\\
-
-  \vspace{1mm}
-
-  $\bullet$ simulations (top panel)\\
-  dash line: likelihood at truth
-
-  \vspace{2mm}
-
-  $\bullet$ data (bottom panel)\\
-  dash line: MLE for case C.
-
-    \vspace{5mm}
-
-%$\bullet$    Multiple search refinements.
-
-%    \vspace{5mm}
-
-    (A) 9 shared parameters, $4\times 20$ unit-specific parameters
-
-    \vspace{5mm}
-
-    (B) all parameters unit-specific
-
-    \vspace{5mm}
-
-    (C) all unit-specific, with constant immigration instead of coupling
-
-  \end{column}
-\end{columns}
-
-\end{frame}
-
-\begin{frame}{Inferences from the previous slide}
-
-%%% Wilks' Theorem: improvement on truth is about (1/2) chisq(p)
-
-\begin{itemize}
-  \item
-IBPF can work for simulated data, since it finds likelihoods higher than the truth when the model is well-specified
-
-  \vspace{5mm}
-
-  \item
-    When coupled by a gravity model, simulated data show evidence against a constant immigration model.
-
-    \vspace{5mm}
-
-\item 
-Data do not favor this gravity model over a constant immigration model.
-
-%%% An underwhelming sounding result, but a necessary step to make progress,
-%%% and previously we did not know how to do this.
-
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}{He et al (2010) estimates. $r$ is rank correlation with $N_{1950}$}
-
-\resizebox{12cm}{!}{
-\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r}
-\hline
-  & $N_{1950}$ & $R_0$ & IP & LP & $\alpha$ & $a$ & $\iota$ & $\psi$ & $\rho$ & $\sigma_{SE}$\\
-\hline
-Halesworth & 2200 & 33.00 & 2.30 & 7.90 & 0.95 & 0.38 & 0.0091 & 0.64 & 0.75 & 0.075\\
-\hline
-Lees & 4200 & 30.00 & 2.10 & 8.50 & 0.97 & 0.15 & 0.0310 & 0.68 & 0.61 & 0.080\\
-\hline
-Mold & 6400 & 21.00 & 1.80 & 5.90 & 1.00 & 0.27 & 0.0140 & 2.90 & 0.13 & 0.054\\
-\hline
-Dalton in Furness & 11000 & 28.00 & 2.00 & 5.50 & 0.99 & 0.20 & 0.0390 & 0.82 & 0.46 & 0.078\\
-\hline
-Oswestry & 11000 & 53.00 & 2.70 & 10.00 & 1.00 & 0.34 & 0.0300 & 0.48 & 0.63 & 0.070\\
-\hline
-Northwich & 18000 & 30.00 & 3.00 & 8.50 & 0.95 & 0.42 & 0.0600 & 0.40 & 0.80 & 0.086\\
-\hline
-Bedwellty & 29000 & 25.00 & 3.00 & 6.80 & 0.94 & 0.16 & 0.0400 & 0.95 & 0.31 & 0.061\\
-\hline
-Consett & 39000 & 36.00 & 2.70 & 9.10 & 1.00 & 0.20 & 0.0730 & 0.41 & 0.65 & 0.071\\
-\hline
-Hastings & 66000 & 34.00 & 5.40 & 7.00 & 1.00 & 0.30 & 0.1900 & 0.40 & 0.70 & 0.096\\
-\hline
-Cardiff & 240000 & 34.00 & 3.10 & 9.90 & 1.00 & 0.22 & 0.1400 & 0.27 & 0.60 & 0.054\\
-\hline
-Bradford & 290000 & 32.00 & 3.40 & 8.50 & 0.99 & 0.24 & 0.2400 & 0.19 & 0.60 & 0.045\\
-\hline
-Hull & 300000 & 39.00 & 5.50 & 9.20 & 0.97 & 0.22 & 0.1400 & 0.26 & 0.58 & 0.064\\
-\hline
-Nottingham & 310000 & 23.00 & 3.70 & 5.70 & 0.98 & 0.16 & 0.1700 & 0.26 & 0.61 & 0.038\\
-\hline
-Bristol & 440000 & 27.00 & 4.90 & 6.20 & 1.00 & 0.20 & 0.4400 & 0.20 & 0.63 & 0.039\\
-\hline
-Leeds & 510000 & 48.00 & 11.00 & 9.50 & 1.00 & 0.27 & 1.2000 & 0.17 & 0.67 & 0.078\\
-\hline
-Sheffield & 520000 & 33.00 & 6.40 & 7.20 & 1.00 & 0.31 & 0.8500 & 0.18 & 0.65 & 0.043\\
-\hline
-Manchester & 700000 & 33.00 & 6.90 & 11.00 & 0.96 & 0.29 & 0.5900 & 0.16 & 0.55 & 0.055\\
-\hline
-Liverpool & 800000 & 48.00 & 9.80 & 7.90 & 0.98 & 0.30 & 0.2600 & 0.14 & 0.49 & 0.053\\
-\hline
-Birmingham & 1100000 & 43.00 & 12.00 & 8.50 & 1.00 & 0.43 & 0.3400 & 0.18 & 0.54 & 0.061\\
-\hline
-London & 3400000 & 57.00 & 13.00 & 13.00 & 0.98 & 0.55 & 2.9000 & 0.12 & 0.49 & 0.088\\
-\hline
-$\qquad r$ & 1 & 0.46 & \cellcolor{yellow!50} 0.95 & 0.32 & 0.11 & 0.30 & 0.9300 & -0.93 & -0.20 & -0.330\\
-\hline
-\end{tabular}
-}
-
-\end{frame}
-
-\begin{frame}{Cholera in Haiti, 2010-2019}
-
-  %%% Haiti and Dominican Republic share Hispaniola
-
-% https://web.archive.org/web/20151106110552/http://www.ihsi.ht/pdf/projection/Estimat_PopTotal_18ans_Menag2015.pdf
-%Map	Department	Capital	Area (km2)	Population	Density (Pop./km2)	Planning Region
-%1	Artibonite	Gonaïves	4,987	1,727,524	350	Central
-%2	Centre	        Hinche	        3,487	746,236	        210	Central
-%3	Grand'Anse	Jérémie	        1,912	468,301	        240	South
-%4	Nippes	        Miragoâne	1,268	342,525	        270	South
-%5	Nord	        Cap-Haïtien	2,115	1,067,177	500	North
-%6	Nord-Est	Fort-Liberté	1,623	393,967	        240	North
-%7	Nord-Ouest	Port-de-Paix	2,103	728,807	        350	Central
-%8	Ouest	        Port-au-Prince	4,983	4,029,705	810	West
-%9	Sud-Est	        Jacmel	        2,034	632,601	        310	West
-%10	Sud	        Les Cayes	2,654	774,976	        290	South
-
-  % Hurricane Matthew: October 2016
-
-\begin{columns}[T] % align columns
-
-%\hspace{-15mm}
-\begin{column}{.35\textwidth}
-  \vspace{2mm}
-
-  $\bullet$  820,000 reported cases
-
-  \vspace{4mm}
-
-  $\bullet$ nearly 10,000 deaths
-
-    \vspace{4mm}
-
-$\bullet$  Population of 10,000,000 in 10 d\'{e}partements
-
-    \vspace{4mm}
-
-    $\bullet$  Lee et al (2020) developed 4 models to guide vaccination plans:
-
-    %%% Elizabeth Lee, Johns Hopkins
-
-    \vspace{1mm}
-
-    {\bf 1}. stochastic, national \\
-    {\bf 2}. ODE, spatial \\
-    {\bf 3}. stochastic, spatial\\
-    {\bf 4}. agent based
-
-\end{column}
-
-\begin{column}{.62\textwidth}
-  \includegraphics[width=7.5cm]{haiti/haiti.png}
-
-  \vspace{-3mm}
-
-  \begin{center}
-    {  \tiny \url{https://doi.org/10.5194/nhess-20-471-2020}}
-  \end{center}
-
-\end{column}
-
-\end{columns}
-
-  \vspace{4mm}
-
-   $\bullet$ Wheeler et al (2023) continued the data analysis
-
-\end{frame}
-
-\begin{frame}{Haiti cholera weekly reports 2010-2019 (log scale)}
-\includegraphics[width=12cm]{haiti/Plot_Reported_Cases-1.pdf}
-\end{frame}
-
-\begin{frame}{Log-likelihoods of models and an ARMA benchmark}
-
-  \begin{center}
-\begin{tabular}{|l|rc|rc|rc|}
-\hline
-& Model~1& (p) & Model~2 & (p) & Model~3 & (p)
-\\
-\hline
-Wheeler et al
-& \cellcolor{yellow!50} -2731.3 & \cellcolor{yellow!50}(15)
-& -21957.3 & (6)
-& \cellcolor{yellow!50} -17850.4 & \cellcolor{yellow!50}(35)
-\\
-Lee et al
-& -3030.9 & (20)
-& -29367.4 & (6)
-& -31840.8 & (29)
-\\
-Log-ARMA(2,1) \hspace{-2.5mm}
-& -2803.7 & (4)  %% -0.5*5615.3+4
-& \cellcolor{yellow!50} { -18027.0} & \cellcolor{yellow!50} (40)  %% -0.5*36133.9+40
-& -18027.0 & (40)
-\\
-\hline
-\end{tabular}
-  \end{center}
-
-  \vspace{3mm}
-
-  Model 1 (stochastic, aggregated): \\
-  We add process overdispersion, and beat the ARMA benchmark.
-
-  \vspace{5mm}
-
-  Model 2 (deterministic, spatial): \\
-  We use a log-normal measurement model. ODE models fall far short of the ARMA benchmark.
-
-    \vspace{5mm}
-
-    Model 3 (stochastic, spatial): \\
-    We fit the model using IBPF, and beat the ARMA benchmark.
-
-\end{frame}
-
-\begin{frame}{Evidence for a decreasing transmission rate in Model 1}
-\includegraphics[width=12cm]{haiti/Beta_trend_Figure-1.pdf}
-\end{frame}
-
-% From Li et al (2020) page S-14:
-% Ruiyun Li, Imperial (Oslo as of 2021)
-% According to official report, around 5 million people left Wuhan city before January 23.
-The total population in Wuhan city after January 23 (when travel restrictions were imposed) is therefore around 6 million.
-
-\begin{frame}{COVID in 375 cities in China, 10-23 January, 2020}
-
-  \begin{itemize}
-  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
-    (Li et al, {\it Science}, May 2020).
-  \item SEIR model with asymptomatics, reporting delay, and coupling based on cell phone data.
-  \item Inference using iterated EnKF.
-
-    \vspace{5mm}
-   \item
-    The time interval predates Wuhan travel restrictions.
-%% see README for R code
-   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
-   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
-   \item{} 0 cases reported for all cities during Jan 10--15.
-
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}{Top: data. Middle: IBPF fit. Bottom: IEnKF fit} 
-\includegraphics[width=12cm]{covid/3-panel.pdf}
-\end{frame}
-
-\begin{frame}{Technical issues with EnKF}
-
-  \begin{itemize}
-  \item EnKF is based on a continuous Gaussian approximation.
-  \item Log-likelihoods with respect to counting measure are never positive, and cannot properly be compared with log-likelihoods corresponding to continuous densities.
-    \item For data with many zeros, the unbounded EnKF likelihood can substantially bias the MLE.
-    \item Adapting EnKF for count data is non-trivial (Katzfuss et al, {\it JASA}, 2019).
-%    \item Li et al avoided the issue by using a ``heuristic observation error variance'' within EnKF.
+%%   \begin{itemize}
+%%   \item EnKF is based on a continuous Gaussian approximation.
+%%   \item Log-likelihoods with respect to counting measure are never positive, and cannot properly be compared with log-likelihoods corresponding to continuous densities.
+%%     \item For data with many zeros, the unbounded EnKF likelihood can substantially bias the MLE.
+%%     \item Adapting EnKF for count data is non-trivial (Katzfuss et al, {\it JASA}, 2019).
+%% %    \item Li et al avoided the issue by using a ``heuristic observation error variance'' within EnKF.
 The corresponding ``log-likelihood'' is not the log-likelihood of any model.
 In particular, it cannot validly be compared againt simple statistical benchmarks.
-%  \item Data and models will never be perfect. We want to reveal, rather than conceal, limitations.
-  \end{itemize}
+%% %  \item Data and models will never be perfect. We want to reveal, rather than conceal, limitations.
+%%   \end{itemize}
 
-  \end{frame}
+%%   \end{frame}
 
-\begin{frame}{Future work}
+%% \begin{frame}{Future work}
 
-  \newcommand\futuresep{\vspace{5mm}}
+%%   \newcommand\futuresep{\vspace{5mm}}
 
-  \begin{myitemize}
+%%   \begin{myitemize}
 
-  \item We anticipate continued progress in modeling and inference for nonlinear spatiotemporal partially observed stochastic dynamic systems.
+%%   \item We anticipate continued progress in modeling and inference for nonlinear spatiotemporal partially observed stochastic dynamic systems.
 
-    \futuresep
+%%     \futuresep
 
-  \item As long as the \texttt{spatPomp} package facilitates data analysis, we expect continuing package development.
+%%   \item As long as the \texttt{spatPomp} package facilitates data analysis, we expect continuing package development.
 
-%    \futuresep
+%% %    \futuresep
 
- %   \item Questions and comments?
+%%  %   \item Questions and comments?
 
-\end{myitemize}
+%% \end{myitemize}
 
-\end{frame}
+%% \end{frame}
 

### 2023-12-30 talk/aprm24.tex
edits to aprm24

@@
 
-\begin{frame}{Questions to be addressed}
+\begin{frame}{Question to be addressed}
 
   \begin{enumerate}
-  \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models?
+  \item When can we carry out full-information likelihood-based inference on nonlinear non-Gaussian spatiotemporal partially observed Markov process (SpatPOMP) models?
 In particular, models for networks of interacting biological population dynamics.
@@
 
-  \item Is there an intersection between scientific problems of interest and the capabilities of the \texttt{spatPomp} R package?
+\item We introduce the {\bf iterated block particle filter}, currently the most powerful algorithm available in the \texttt{spatPomp} R package.
 
-        \vspace{2mm}
+%  \vspace{2mm}
 
-\item We introduce the {\bf iterated block particle filter}, currently the most powerful algorithm available in \texttt{spatPomp}.
-
-  \vspace{2mm}
-
-\item Bonus question: How do we know if our model is statistically adequate, or needs more work?
+%  \item Bonus question: How do we know if our model is statistically adequate, or needs more work?
 
@@
 \begin{frame}{Practical inference using IBPF}
-  \bi
-  \item to do
+  \begin{enumerate}
+  \item Monte Carlo adjusted profile likelihood \citep{ionides17profile} obtains confidence intervals that accommodate Monte Carlo error.
 
-  \ei
+  \item Comparing the log-likelihood with an autoregressive model (or other simple statistical model) provides a check of model fit.
+
+  \item Comparing the block log-likelihood against the benchmark provides insight into problematic units.
+
+    \item Comparing the conditional log-likelihood for each observation against the benchmark helps to identify outliers.
+
+  \end{enumerate}
 
@@
 %% \end{frame}
+
+\begin{frame}{Filtering $U$ units of a coupled measles SEIR model}
+
+\vspace{-1mm}
+
+\begin{center}
+\includegraphics[width=10cm]{mscale_loglik_plot-1.pdf}
+
+\end{center}
+
+\vspace{-4mm}
+
+Simulated data using a gravity model with geography, demography and transmssion parameters corresponding to UK pre-vaccination measles (Ionides et al, {JASA}, 2023).
+
+%\end{center}
+
+\end{frame}
 

### 2023-12-30 talk/aprm24.tex
aprm talk submitted

@@
 
-\item We introduce the {\bf iterated block particle filter}, currently the most powerful algorithm available in the \texttt{spatPomp} R package.
+\item We introduce the {\bf iterated block particle filter}, currently the most effective algorithm in the \texttt{spatPomp} R package.
 
@@
     \begin{itemize}
-    \item  Iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference:
+    \item  Iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference for nonlinear, non-Gaussian, partially observed dynamic systems:
 
@@
 
-\begin{frame}{COVID in 375 cities in China, 10-23 January, 2020}
+\begin{frame}{COVID in 373 cities in China, Jan 10 to Feb 8, 2020}
 
@@
   \item The time interval covers the initial China lockdown.
-  \item Here, we summarize our re-analysis of this model and data, recently posted on arXiv.
+  \item We present re-analysis of this model and data, recently posted on arXiv.
 %% see README for R code
@@
 
-      \item This motivated us to develop an IBPF for parameter estimation.
+      \item This motivated us to develop an iterated BPF (IBPF) for parameter estimation.
 
@@
 
+    \item Two recent case studies (Wheeler et al, 2023; Li et al, 2023) demonstrate data analysis using IBPF. Code and data are provided via R packages extending \code{spatPomp}.
+
   \end{enumerate}
@@
   \end{frame}
-
-%% \begin{frame}{Technical issues with EnKF}
-
-%%   \begin{itemize}
-%%   \item EnKF is based on a continuous Gaussian approximation.
-%%   \item Log-likelihoods with respect to counting measure are never positive, and cannot properly be compared with log-likelihoods corresponding to continuous densities.
-%%     \item For data with many zeros, the unbounded EnKF likelihood can substantially bias the MLE.
-%%     \item Adapting EnKF for count data is non-trivial (Katzfuss et al, {\it JASA}, 2019).
-%% %    \item Li et al avoided the issue by using a ``heuristic observation error variance'' within EnKF.
-The corresponding ``log-likelihood'' is not the log-likelihood of any model.
-In particular, it cannot validly be compared againt simple statistical benchmarks.
-%% %  \item Data and models will never be perfect. We want to reveal, rather than conceal, limitations.
-%%   \end{itemize}
-
-%%   \end{frame}
-
-%% \begin{frame}{Future work}
-
-%%   \newcommand\futuresep{\vspace{5mm}}
-
-%%   \begin{myitemize}
-
-%%   \item We anticipate continued progress in modeling and inference for nonlinear spatiotemporal partially observed stochastic dynamic systems.
-
-%%     \futuresep
-
-%%   \item As long as the \texttt{spatPomp} package facilitates data analysis, we expect continuing package development.
-
-%% %    \futuresep
-
-%%  %   \item Questions and comments?
-
-%% \end{myitemize}
-
-%% \end{frame}
 
@@
 
-\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23}
+\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23,li23}
 

### 2025-06-11 talk/cmo25.tex
add cmo25 talk

--- 2023-12-30 talk/aprm24.tex
@@
 \begin{center}
-  {\Large\bf An iterated block particle filter for inference on coupled dynamic systems}
+  {\Large\bf Inference on spatiotemporal infectious disease dynamics}
 
-% ABSTRACT
-% We consider inference for a collection of partially observed, stochastic, interacting, nonlinear dynamic processes.
-Each process is called a unit, and our primary motivation arises in biological metapopulation systems where a unit is a spatially distinct sub-population.
-Block particle filters are an effective tool for simulation-based likelihood evaluation for these systems, which are strongly dependent through time on a single unit and relatively weakly coupled between units.
-Iterated filtering algorithms can facilitate likelihood maximization for simulation-based filters.
-We introduce a new iterated block particle filter algorithm applicable to parameters that are either unit-specific or shared between units.
-We demonstrate this algorithm to carry out inference on a coupled epidemiological model for spatiotemporal COVID-19 case report data in 373 cities.
+%ABSTRACT: Epidemiological models must be consistent with observed data in order to generate reliable knowledge and evidence-based policy.
+Metapopulation systems, which consist of a network of connected sub-populations, pose technical challenges in statistical inference owing to nonlinear, stochastic interactions.
+Numerical difficulties encountered in conducting inference can obstruct the core scientific questions concerning the link between the mathematical models and the data.
+Recently, an algorithm has been proposed that enables computationally tractable likelihood-based inference for high-dimensional partially observed stochastic dynamic models of metapopulation systems.
+We use this algorithm to build a statistically principled data analysis workflow for metapopulation systems.
+Via a case study of COVID-19, we show how this workflow addresses the limitations of previous approaches.
+The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we revisit an influential metapopulation model used to inform basic epidemiological understanding early in the pandemic.
+Our methods support self-critical data analysis, enabling us to identify and address model weaknesses, leading to a new model with substantially improved statistical fit and parameter identifiability.
+Our results suggest that the lockdown initiated on 23 January 2020 in China was more effective than previously thought.
 
@@
 
-IMS Asia Pacific Rim Meeting\\
-Jan 5, 2024
+Workshop on {\it Mathematical and Statistical Challenges in \\
+  Post-Pandemic Epidemiology and Public Health}\\
+Casa Matem\'{a}tica Oaxaca\\
+June 16, 2025
 
@@
 
-Slides are at \url{https://ionides.github.io/talks/aprm24.pdf}
+Slides are at \url{https://ionides.github.io/talks/cmo25.pdf}
 
@@
 Joint work with
-Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King, Mercedes Pascual
+Ning (Patricia) Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King, Mercedes Pascual
 
@@
 
-%  \vspace{2mm}
+\vspace{2mm}
 
-%  \item Bonus question: How do we know if our model is statistically adequate, or needs more work?
+\item Bonus question: How do we know if our model is statistically adequate, or needs more work?
+Hint: we need {\bf performance benchmarks}, i.e., comparison with non-mechanistic machine learning methods.
+If there is performance shortfall, we need diagnostics to figure out where and how.
 
@@
 
-      \url{https://ionides.github.io/531w22/}
+      \url{https://ionides.github.io/531w24/}
 
@@
 
+\begin{frame}{\texttt{spatPomp}: an R package for SpatPOMP models}
+
+  Asfaw et al ({\it J. Open Source Software}, 2024)
+  \url{https://github.com/spatPomp-org}
+
+  \bi
+\item Permits specification of arbitrary non-linear non-Gaussian spatiotemporal POMP models.
+\item Adds spatial structure to \texttt{pomp} \citep{king16}.
+\item Monte Carlo algorithms designed to address the \myemph{curse of dimensionality}.
+  \item Algorithms compile via C to assist performance.
+\item Statistically efficient, likelihood-based inference despite methods being simulation-based.
+  \ei
+
+  Future goals: add GPU and autodiff support via \texttt{pypomp}, \url{https://github.com/pypomp}
+
+\end{frame}
+
 \begin{frame}{COVID in 373 cities in China, Jan 10 to Feb 8, 2020}
@@
   \item The time interval covers the initial China lockdown.
-  \item We present re-analysis of this model and data, recently posted on arXiv.
+  \item We present re-analysis of this model and data (Li et al, {J. Roy. Soc. Interface}, 2024).
 %% see README for R code
@@
   \begin{center}
-%%%%% SEAIR diagram
-\resizebox{11cm}{!}{
-\begin{tikzpicture}[
-  square/.style={rectangle, draw=black, minimum width=0.8cm, minimum height=0.8cm, rounded corners=.1cm, fill=blue!8},
-  travel/.style={circle, draw=black, minimum width=0.85cm, minimum height=0.8cm, fill=green!8},
-  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.9cm,inner sep=0cm},
-  bendy/.style={bend left=25},
-  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
-  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
-]
-\tikzset{>={}}; % this is needed to implement the arrow redefinition
+      \begin{center}
+    \includegraphics[height=5cm]{seair.png}
+    \end{center}
 
-\node (S) at (-0.5,0) [square] {S$_u$};
-\node (E) at (2,0) [square] {E$_u$};
-\node (A) at (4.5,0) [square] {A$_u$};
-\node (I) at (4.5,-1.75) [square] {I$_u$};
-\node (R) at (7,0) [square] {R$_u$};
-
-\node (T1) at (-0.5,1.75) [travel] {T};
-\node (T2) at (2,1.75) [travel] {T};
-\node (T3) at (4.5,1.75) [travel] {T};
-
-\node (Ca) at (4.5,-3.5) [report] {C$^a_u$};
-\node (Cb) at (7,-3.5) [report] {C$^b_u$};
-\node (C) at (9.5,-3.5) [report] {C$_u$};
-
-\node (V1) at (2.5,-1.75) {};
-\node (V2) at (3,-3.5) {};
-\draw [->] (E) -- (Ca -| V2) -- (Ca);
-\draw [->] (I -| V1) -- (I);
-
-\draw [->] (S) -- (E);
-\draw [->] (A) -- (R);
-
-\draw [->] (Ca) -- (Cb);
-\draw [->] (Cb) -- (C);
-
-\draw [->] (I.east) -- (R);
-\draw [->] (E) -- (A);
-
-\draw [->, bendy] (S) to (T1);
-\draw [->, bendy] (T1) to (S);
-
-\draw [->, bendy] (E) to (T2);
-\draw [->, bendy] (T2) to (E);
-
-\draw [->, bendy] (A) to (T3);
-\draw [->, bendy] (T3) to (A);
-
-\end{tikzpicture}
-}
 \end{center}
@@
 
+\begin{frame}{Log-likelihood values for six models}
+
+  \resizebox{12cm}{!}{
+\begin{tabular}{c|rr|l}
+ & loglik & df & description
+\\
+\hline
+$M_1$ &
+  $-\infty$ & 10 &
+  SEAIR model, parameter values and mobility of Li et al (2020)
+\\
+$M_2$ &  -14240.5 & 10 &
+  Adjusted mobility and measurement in $M_1$
+\\
+$M_3$ &
+ -11257.9  & 374 &
+  Independent identically distributed negative binomial
+\\
+$M_4$ &
+ -10825.3 & 375 &
+  Autoregressive negative binomial
+\\
+$M_5$ &
+  -9088.2&
+  12 &
+  Adding overdispersed dynamics to $M_2$ and refitting
+\\
+$M_6$ &
+  -9116.5&
+  10 &
+  Latent and infectious periods unchanged by lockdown in $M_5$
+\\
+
+\hline
+\end{tabular}
+}
+
+  \vspace{10mm}
+
+Model comparisons by log-likelihood. The degrees of freedom (df) is the number of estimated parameters.
+
+\end{frame}
+
+\begin{frame}{Data analysis progression}
+  \bi
+\item Some data errors meant that the original data was incompatible with the model.
+\item Tracked down by assessing conditional log-likelihood for each unit and each time.
+\item Even then, the likelihood was poor compared to non-mechanistic benchmarks. Including over-dispersion fixed that.
+\item Some weak identifiability remained. extra assumptions can lead to plausible parameter estimates without much loss of model fit.
+\item Mild outliers remain, e.g., Wuhan on day 18.  
+  \ei
+
+\end{frame}
+
 \begin{frame}
@@
 \begin{frame}
- \includegraphics[width=12cm]{covid/filter_tests10.png}
+ \includegraphics[width=10cm]{covid/filter_tests10.png}
 
@@
 
-  \includegraphics[trim={0 0 0 10mm},clip,width=9cm]{IBPF_workflow.pdf}
+%  \includegraphics[trim={0 0 0 10mm},clip,width=9cm]{IBPF_workflow.pdf}
+  \includegraphics[width=11cm]{ibpf.png}
 
@@
 
-\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23,li23}
+\nocite{asfaw24,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides24,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler24,li24}
 
+\nocite{asfaw24tutorial}
 \begin{frame}[allowframebreaks]

### 2025-06-15 talk/cmo25.tex
cmo25 talk revised

@@
 \item Bonus question: How do we know if our model is statistically adequate, or needs more work?
-Hint: we need {\bf performance benchmarks}, i.e., comparison with non-mechanistic machine learning methods.
-If there is performance shortfall, we need diagnostics to figure out where and how.
 
+  \begin{itemize}
+  \item Hint: we need {\bf performance benchmarks}, i.e., comparison with non-mechanistic machine learning methods. If there is performance shortfall, we need diagnostics to figure out where and how.
+  \end{itemize}
     \end{enumerate}
@@
 
-\begin{frame}{Inference challenges in population dynamics}
+\begin{frame}{Inference challenges in population dynamics: A brief history}
 
@@
   1--5 are largely solved, from a methodological perspective.\\
-  6 is our immediate topic.\\
+  \myemph{6 is our immediate topic.}\\
   7 is exciting but not the focus of this talk.
@@
     \begin{itemize}
-    \item  Iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference for nonlinear, non-Gaussian, partially observed dynamic systems:
+    \item  \myemph{Iterated particle filtering} via \texttt{mif2} in the R package \texttt{pomp} \citep{king16} enables plug-and-play likelihood-based inference for nonlinear, non-Gaussian, partially observed Markov process (POMP) models.
+
+      \item Allows consideration of arbitrary POMP models for low-dimensional time series data.
+
+      \item Accessible for Masters-level statisticians:
 
@@
 
-  Asfaw et al ({\it J. Open Source Software}, 2024)
+  Asfaw et al (2024, {\it J. Open Source Software})
   \url{https://github.com/spatPomp-org}
@@
   \bi
-\item Permits specification of arbitrary non-linear non-Gaussian spatiotemporal POMP models.
-\item Adds spatial structure to \texttt{pomp} \citep{king16}.
-\item Monte Carlo algorithms designed to address the \myemph{curse of dimensionality}.
+\item Permits specification of arbitrary non-linear non-Gaussian spatiotemporal POMP (SpatPOMP) models.
+\item \texttt{spatPomp} spatial structure to \texttt{pomp}.
+\item Includes Monte Carlo algorithms designed to address the \myemph{curse of dimensionality}.
   \item Algorithms compile via C to assist performance.
@@
   \bi
+  \item $U=373$ cities (spatial units) labeled $u=1,\dots,U$.
 \item 
@@
   12 &
-  Adding overdispersed dynamics to $M_2$ and refitting
+  Adding overdispersed dynamics to $M_2$
 \\
@@
 \item Some weak identifiability remained. extra assumptions can lead to plausible parameter estimates without much loss of model fit.
-\item Mild outliers remain, e.g., Wuhan on day 18.  
+\item Less severe outliers remain, e.g., Wuhan on day 18.  
   \ei
@@
   \begin{itemize}
-\item BPF is the best available filter for this COVID model.
+\item Block particle filter (BPF) beats alternatives for this COVID model.
 \item Similar results hold for other models (Ionides et al, {\it JASA}, 2023).
@@
 
-    \item Two recent case studies (Wheeler et al, 2023; Li et al, 2023) demonstrate data analysis using IBPF. Code and data are provided via R packages extending \code{spatPomp}.
+    \item Two recent case studies (Wheeler et al, 2024; Li et al, 2024) demonstrate data analysis using IBPF. Code and data are provided via R packages extending \code{spatPomp}.
 
@@
 
-\nocite{asfaw24,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides24,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler24,li24}
+%\nocite{asfaw24,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides23,ionides24,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler24,li24}
+
+\nocite{asfaw24,bjornstad01,grenfell04,ionides23,ionides24,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler24,li24}
 
